### 2024-04-26 notes/UQ Verification.tex
Write-up of UQ study for LM

\documentclass{article}

\usepackage{amsmath, amssymb}
\usepackage{xcolor}
\usepackage{natbib}
\usepackage{caption, subcaption}
\usepackage{graphicx}

\newcommand{\bG}{\mathbb{G}}
\newcommand{\bF}{\mathbb{F}}
\newcommand{\bP}{\mathbb{P}}
\newcommand{\bE}{\mathbb{E}}
\newcommand{\bV}{\mathbb{V}}

\title{Verification of SE Formulas for Mediation Analysis}

\begin{document}

\maketitle

I'm doing some Monte Carlo to verify the new SE formulas. Recall that we're doing mediation analysis, so we've got a response, $Y$, an exposure, $X$, and a mediation, $M$. We also have some number of confounders, which will be grouped together in the matrix $W$. Broadly speaking, we fit two regression models, one to predict $M$ using $X$ and $W$, the other to predict $Y$ using $M$, $X$ and $W$. We then compute the mediation effect (specifically, the total effect of $X$ on $Y$) as a function of the coefficients from these two regression models. An asymptotic SE for our mediation effect estimator can then be obtained from the asymptotic standard errors of our fitted regression coefficients using the $\delta$-method.

So far, so simple. There are a few places that things start to get more complicated. First, each of the two regression models can be either linear of logistic depending on whether the corresponding response variable is continuous or binary\footnote{In principle, we could have $Y$ and/or $M$ follow any distribution with a suitable GLM formulation. I don't think I've ever seen count data (i.e. Poisson regression) used here, much less anything more exotic.}. Furthermore, we can add random effects to our regression models. In the trust study, we have random effects for the intercept, $X$ and $M$ (naturally, the latter only applies when predicting $Y$).

We will address each of these extra layers of complexity in turn. First though, we start with the simplest version of the problem.

\section{Continuous Response, Continuous Mediation, Fixed-Effects}

I set the sample size, $n$, to 100. Each regression coefficient is 1 and the residual standard deviation in both regression models is $0.2$. I use a single confounder and generate both it, $W$, and the exposure, $X$, as iid $\mathrm{N}(1,1)$. I generate $1000$ datasets, each with different values for $X$ and $W$.

On each dataset, we fit the two regression models, then extract coefficients and standard errors. Next, we compute our estimate of the mediation effect and its $\delta$-method standard error (see Overleaf for details). After repeating this process $1000$ times, we compute the empirical standard error (SD of our estimates), as well as the mean and median estimated standard error. Values are given in Table \ref{tab:SEs_cont_fix}. As you can see, our $\delta$-method formula works very well.

\begin{table}
    \centering
    \begin{tabular}{|c|c|c|}
        \hline
        Empirical & Mean Analytical & Median Analytical\\
        \hline
        0.0289 & 0.0288 & 0.0286\\
        \hline
    \end{tabular}
    \caption{Standard errors for estimated mediation effect with continuous response and mediation, fixed-effects.}
    \label{tab:SEs_cont_fix}
\end{table}

\end{document}

### 2024-04-26 notes/UQ Verification.tex
Multiple confounders in LM

@@
 In the trust study, we have random effects for the intercept, $X$ and $M$ (naturally, the latter only applies when predicting $Y$).
+The problem is slightly simpler with a single confounder (mostly, the bookkeeping is a bit easier), although I don't want to stick to this for long.
 
@@
 
+The results are very similar if we use multiple confounders (specifically, I use 3). Henceforth, I will use 3 confounders in my analysis.
+
+\section{Categorical Response and Mediation, Fixed-Effects}
+
 \end{document}

### 2024-04-26 scripts/Cont-Resp, Cont-Med, Fixed.jl

@@ -19,7 +19,6 @@ Random.seed!(1)
 
 num_reps = 1000    # Number of datasets to generate
 
-
 n = 100
 
 a_0 = 1
@@ -121,6 +120,137 @@ end
 
 
 par_list_name = @savename num_reps n 
+@save datadir("Cont-Resp, Cont-Med, Fixed-$par_list_name.jld2") all_a_hats all_a_SEs all_b_hats all_b_SEs all_med_hats all_med_SEs
+println("Work complete!")
+
+
+
+
+
+
+#* Multiple confounders
+
+
+Random.seed!(1)
+
+num_reps = 1000    # Number of datasets to generate
+
+n = 100
+p_conf = 3      # Number of confounders
+
+a_0 = 1
+a_1 = 1
+A_2 = repeat([1], p_conf)
+sigma_d = 0.2   # Residual SD for M
+
+b_0 = 1
+b_1 = 1
+b_2 = 1
+B_3 = repeat([1], p_conf)
+sigma_e = 0.2   # Residual SD for Y
+
+
+
+
+
+# X_dist = Bernoulli(0.5)
+# W_dist = Bernoulli(0.5)
+
+X_dist = Normal(1, 1)
+W_dist = Normal(1, 1)
+
+
+
+all_a_hats = []
+all_b_hats = []
+
+all_a_SEs = []
+all_b_SEs = []
+
+all_med_hats = []
+all_med_SEs = []
+
+# @showprogress Threads.@threads for _ in 1:M
+@showprogress for _ in 1:num_reps
+
+    
+    # Make X and W
+    X = rand(X_dist, n)
+    W = rand(W_dist, n, p_conf)
+    
+    # Make data
+    ## M
+    delta = rand(Normal(0, sigma_d), n)
+    M = a_0 .+ a_1 .* X .+ W * A_2 .+ delta
+
+    ## Y
+    epsilon = rand(Normal(0, sigma_e), n)
+    Y = b_0 .+ b_1 .* M .+ b_2 .* X .+ W * B_3 .+ epsilon
+
+
+
+
+    ## Fit M
+    M_data = DataFrame(M=M, X=X)
+    for p in 1:p_conf
+        M_data[!,Symbol("W$p")] = W[:, p]
+    end
+
+    M_formula = term(:M) ~ sum(term.(names(M_data, Not(:M))))
+
+    M_model = lm(M_formula, M_data)
+    a_hat = coef(M_model)
[…]

### 2024-04-30 notes/UQ Verification.tex
Continuous response, binary mediator, fixed effects

@@
 \section{Continuous Response, Continuous Mediation, Fixed-Effects}
+\label{sec:cont_cont_fix}
 
@@
 
-\section{Categorical Response and Mediation, Fixed-Effects}
+\section{Continuous Response, Binary Mediation, Fixed-Effects}
+
+Here things start getting messier.
+I will return to documenting the analysis, but first I want to discuss the proposed estimand.
+Note that the total effect given in Case 2 of Section 1.1, $\gamma(a, \alpha, \beta_1, \beta_2)$ depends on our choice of levels for $X$ and $W$.
+If $X$ is binary then its choice is easy since we're computing the effect of a one-unit increase in $X$.
+For $W$ however, I don't see any natural choice of reference value.
+One option is to evaluate $\gamma$ at each $W$ in the observed dataset and average to ``marginalize out'' $W$.
+I have a few references which discuss doing something along these lines, but it might be worth discussing on Thursday.
+For now, I'm just going to choose a value of $W$ and study ``fixed-confounder'' mediation effects.
+
+I repeated the analysis in Section \ref{tab:SEs_cont_fix}, this time with a binary mediator.
+Datasets are generated with $n=100$ observations, 3 confounders.
+Regression slopes are set to 1, and the intercept is set so that the expected linear predictor is approximately zero (this helps keep the number of 0s and 1s for $M$ approximately balanced).
+The residual variance for $Y$ is set to $0.2^2$.
+
+Following the analysis outlined by Bruno, I use a logistic regression model for the mediator and linear regression for the response.
+Formulas for the $\delta$-method are messier, but conceptually this case is nearly identical to the continuous response setting.
+One difference here is that the total mediation effect depends on levels of the covariates, specifically $X$ and $W$.
+I report results for $X=0$ and $W = [1,1,1]^T$.
+This value of $X$ is very natural when the exposure is binary.
+I have no such reason for choosing this value of $W$.
+See the opening paragraph of this section for more on this.
+
+I generated $1000$ datasets, fit regression models for $M$ (logistic) and $Y$ (linear), then computed our estimate for the mediation effect.
+I also evaluated our SE formula on each dataset.
+Table \ref{tab:SEs_cont_bin_fix} gives the empirical standard error of our total effect estimator, as well as the mean and median analytical SE.
+
+\begin{table}
+    \centering
+    \begin{tabular}{|c|c|c|}
+        \hline
+        Empirical & Mean Analytical & Median Analytical\\
+        \hline
+        0.0540 & 0.0584 & 0.0577\\
+        \hline
+    \end{tabular}
+    \caption{Standard errors for estimated mediation effect with continuous response and binary mediation, fixed-effects.}
+    \label{tab:SEs_cont_bin_fix}
+\end{table}
 

### 2024-04-30 scripts/Bin-Resp, Cont-Med, Fixed.jl

@@ -1,0 +1,183 @@
+#* For now, I'm going to just focus on parameter estimation. This is an easier problem than random effects prediction
+#* Later, I can introduce prediction of random effects
+
+using Distributions, Random, Statistics
+using LogExpFunctions   # For the logit and expit functions
+using DataFrames
+using MixedModels
+using ProgressMeter, DrWatson
+using JLD2
+using GLM   # For linear and logistic regression with fixed effects
+
+include("../src/Helpers - General.jl")
+include("../src/Helpers - Bin-Resp, Cont-Med, Fixed.jl")
+
+
+
+
+
+Random.seed!(1)
+
+num_reps = 1000    # Number of datasets to generate
+
+n = 100
+p_conf = 3      # Number of confounders
+
+a_1 = 1
+A_2 = repeat([1], p_conf)
+a_0 = - (a_1 + sum(A_2))    # Offset mean effect of other coefficients in linear predictor for M
+sigma_d = 0.2   # Residual SD for M
+
+b_1 = 1
+b_2 = 1
+B_3 = repeat([1], p_conf)
+b_0 = - (b_1 + 0.5*b_2 + sum(B_3))    # Approximately offset mean effect of other coefficients in linear predictor for Y
+sigma_e = 0.2   # Residual SD for Y
+
+
+
+
+
+# X_dist = Bernoulli(0.5)
+# W_dist = Bernoulli(0.5)
+
+X_dist = Normal(1, 1)
+W_dist = Normal(1, 1)
+
+
+
+all_a_hats = []
+all_b_hats = []
+
+all_a_SEs = []
+all_b_SEs = []
+
+all_med_hats = []
+all_med_SEs = []
+
+# @showprogress Threads.@threads for _ in 1:M
+@showprogress for _ in 1:num_reps
+
+    
+    # Make X and W
+    X = rand(X_dist, n)
+    W = rand(W_dist, n, p_conf)
+    
+    # Make data
+    ## M
+    eta_M_vec = a_0 .+ a_1 .* X .+ W * A_2
+    p_M_vec = expit.(eta_M_vec)
+    M = rand.(Bernoulli.(p_M_vec))
+
+
+    ## Y
+    epsilon = rand(Normal(0, sigma_e), n)
+    Y = b_0 .+ b_1 .* M .+ b_2 .* X .+ W * B_3 .+ epsilon
+
+
+
+
[…]

### 2024-04-30 scripts/Empirical vs Analytical SEs.jl

@@ -13,9 +13,10 @@ using GLM   # For linear and logistic regression with fixed effects
 
 num_reps = 1000
 n = 100
+p_conf = 3
 
-par_list_name = @savename num_reps n 
-@load datadir("Cont-Resp, Cont-Med, Fixed-$par_list_name.jld2") all_a_hats all_a_SEs all_b_hats all_b_SEs all_med_hats all_med_SEs
+par_list_name = @savename num_reps n p_conf
+@load datadir("Bin-Resp, Cont-Med, Fixed-$par_list_name.jld2") all_a_hats all_a_SEs all_b_hats all_b_SEs all_med_hats all_med_SEs
 
 
 

### 2024-04-30 src/Helpers - Bin-Resp, Cont-Med, Fixed.jl

@@ -1,0 +1,100 @@
+
+# Compute mediation effect
+
+## Increment in conditional expectation of M for a unit increase in X
+function get_delta(eta, a_x)
+    Q1 = exp(-eta)
+    Q2 = (1 - exp(-a_x))
+    Q3 = (1 + exp(-eta - a_x))
+    Q4 = (1 + exp(-eta))
+
+    return Q1 * Q2 / (Q3 * Q4)
+end
+
+## Increment in conditional expectation of Y for a unit increase in X
+## I.e. Mean over M of increment in conditional expectation for Y given M and X
+function get_gamma(delta, b_m, b_x)
+    return b_m * delta + b_x
+end
+
+
+# Compute SE of estimated mediation effect
+# We use the delta method, so many derivatives are required
+
+
+## Helpful intermedeate quantities
+function d_delta_d_eta(eta, a_x)
+    delta = get_delta(eta, a_x)
+
+    Q1 = 1 - exp(-2*eta - a_x)
+    Q2 = 1 + exp(-eta)
+    Q3 = 1 + exp(-eta - a_x)
+
+    return - delta * Q1 / (Q2 * Q3)
+end
+
+function d_delta_d_a_x(eta, a_x)
+    Q1 = exp(-2*eta - a_x)
+    Q2 = 1 + exp(-eta)
+    Q3 = 1 + exp(-eta - a_x)
+
+    return Q1 / (Q2 * Q3^2)
+end
+
+
+## Derivative of gamma wrt each parameter
+function d_gamma_d_a_0(dd_de, b_x)
+    return dd_de * b_x
+end
+
+function d_gamma_d_a_1(dd_de, dd_da_x, x, b_x)
+    return b_x * x * dd_de + b_x * dd_da_x
+end
+
+function d_gamma_d_A_2(dd_de, W, b_x)
+    return dd_de * b_x .* W
+end
+
+
+function d_gamma_d_b_0()
+    return 0
+end
+
+function d_gamma_d_b_1(eta)
+    return eta
+end
+
+function d_gamma_d_b_2()
+    return 1
+end
+
+function d_gamma_d_B_3(W)
+    return 0 .* W
+end
+
+
+
+## Gradient of gamma wrt all parameters
[…]

### 2024-04-30 src/Helpers - General.jl

@@ -1,0 +1,8 @@
+
+function expit(x)
+    return 1 / (1 + exp(-x))
+end
+
+function logit(p)
+    return log(p / (1 - p))
+end

### 2024-05-01 notes/UQ Verification.tex
Fix mistake in a derivative for const, bin, fixed.

@@
 
-I set the sample size, $n$, to 100.
+I set the sample size, $n$, to 1000.
 Each regression coefficient is 1 and the residual standard deviation in both regression models is $0.2$.
@@
         \hline
-        0.0289 & 0.0288 & 0.0286\\
+        0.00833 & 0.00896 & 0.00895\\
         \hline
@@
 \section{Continuous Response, Binary Mediation, Fixed-Effects}
+\label{sec:cont_bin_fix}
 
@@
 I repeated the analysis in Section \ref{tab:SEs_cont_fix}, this time with a binary mediator.
-Datasets are generated with $n=100$ observations, 3 confounders.
+Datasets are generated with $n=1000$ observations, 3 confounders.
 Regression slopes are set to 1, and the intercept is set so that the expected linear predictor is approximately zero (this helps keep the number of 0s and 1s for $M$ approximately balanced).
@@
         \hline
-        0.0540 & 0.0584 & 0.0577\\
+        0.0200 & 0.0203 & 0.0203\\
         \hline
@@
 
+\section{Binary Response, Binary Mediator, Fixed-Effects}
+
+For my own future reference, I'm going to set out some notation here.
+$a$ and $b$ are the vectors of regression coefficients in a model for $M$ and $Y$ respectively, whether linear or logistic.
+We use $\eta$ and $\zeta$ for the linear predictors of $M$ and $Y$ respectively.
+In my code for Section \ref{sec:cont_bin_fix}, I used $\eta$, but didn't really need $\zeta$.
+Here I will.
+I use $\gamma$ for the total mediation effect.
+In the previous section (albeit only in my code), I used $\delta$ for $\bE(M | X=x+1) - \bE(M | X=x)$, although here I don't think I will need this quantity.
+
+\section{Open Questions}
+
+How does the standard error decompose between multiple responses with the same fixed covariates (i.e.
+$X$ and $W$), versus across replicated covariates?
+I can imagine how to do this, at least naively: do a bunch of analyses on each of a bunch of sets of covariates.
+Compute the average ``within-covariates'' variance and the average total variance.
+Compute the fraction of total variance due to response replication and, interestingly, one minus this fraction as a naive representation of the variance fraction due to covariate variability.
+
 \end{document}

### 2024-05-01 scripts/Bin-Resp, Bin-Med, Fixed.jl

@@ -1,0 +1,189 @@
+#* For now, I'm going to just focus on parameter estimation. This is an easier problem than random effects prediction
+#* Later, I can introduce prediction of random effects
+
+using Distributions, Random, Statistics
+using LogExpFunctions   # For the logit and expit functions
+using DataFrames
+using MixedModels
+using ProgressMeter, DrWatson
+using JLD2
+using GLM   # For linear and logistic regression with fixed effects
+
+include("../src/Helpers - General.jl")
+include("../src/Helpers - Bin-Resp, Cont-Med, Fixed.jl")
+
+
+
+
+
+Random.seed!(1)
+
+num_reps = 1000    # Number of datasets to generate
+
+n = 100
+p_conf = 3      # Number of confounders
+
+a_1 = 1
+A_2 = repeat([1], p_conf)
+a_0 = - (a_1 + sum(A_2))    # Offset mean effect of other coefficients in linear predictor for M
+sigma_d = 0.2   # Residual SD for M
+
+b_1 = 1
+b_2 = 1
+B_3 = repeat([1], p_conf)
+b_0 = - (b_1 + 0.5*b_2 + sum(B_3))    # Approximately offset mean effect of other coefficients in linear predictor for Y
+sigma_e = 0.2   # Residual SD for Y
+
+
+# Reference values at which we compute the total mediation effect
+x_pred = 0
+W_pred = repeat([1], p_conf)    # W=[1,1,1]
+
+
+# X_dist = Bernoulli(0.5)
+# W_dist = Bernoulli(0.5)
+
+X_dist = Normal(1, 1)
+W_dist = Normal(1, 1)
+
+
+
+all_a_hats = []
+all_b_hats = []
+
+all_a_SEs = []
+all_b_SEs = []
+
+all_med_hats = []
+all_med_SEs = []
+
+# @showprogress Threads.@threads for _ in 1:M
+@showprogress for _ in 1:num_reps
+
+    
+    # Make X and W
+    X = rand(X_dist, n)
+    W = rand(W_dist, n, p_conf)
+    
+    # Make data
+    ## M
+    eta_vec = a_0 .+ a_1 .* X .+ W * A_2
+    p_M_vec = expit.(eta_vec)
+    M = rand.(Bernoulli.(p_M_vec))
+
+
+    ## Y
[…]

### 2024-05-01 scripts/Cont-Resp, Bin-Med, Fixed.jl

@@ -9,9 +9,11 @@ using ProgressMeter, DrWatson
 using JLD2
 using GLM   # For linear and logistic regression with fixed effects
 
-include("../src/Helpers - General.jl")
-include("../src/Helpers - Bin-Resp, Cont-Med, Fixed.jl")
+# include("src/Helpers - General.jl")
+# include("src/Helpers - Cont-Resp, Bin-Med, Fixed.jl")
 
+include("../src/Helpers - General.jl")
+include("../src/Helpers - Cont-Resp, Bin-Med, Fixed.jl")
 
 
 
@@ -20,7 +22,7 @@ Random.seed!(1)
 
 num_reps = 1000    # Number of datasets to generate
 
-n = 100
+n = 1000
 p_conf = 3      # Number of confounders
 
 a_1 = 1
@@ -36,6 +38,9 @@ sigma_e = 0.2   # Residual SD for Y
 
 
 
+# Reference values at which we compute the total mediation effect
+x_pred = 0
+W_pred = repeat([1], p_conf)    # W=[1,1,1]
 
 
 # X_dist = Bernoulli(0.5)
@@ -55,7 +60,7 @@ all_b_SEs = []
 all_med_hats = []
 all_med_SEs = []
 
-# @showprogress Threads.@threads for _ in 1:M
+# @showprogress Threads.@threads for _ in 1:num_reps
 @showprogress for _ in 1:num_reps
 
     
@@ -116,16 +121,8 @@ all_med_SEs = []
 
 
 
-    ## Extract linear predictor for M at X=0, W=[1,1,1]
-    x_pred = 0
-    W_pred = repeat([1], p_conf)    # W=[1,1,1]
-
-    data_pred = DataFrame(X=x_pred)
-    for p in 1:p_conf
-        data_pred[!,Symbol("W$p")] = W_pred[p:p]    # Can only fill a data frame with vectors, not scalars
-    end
-
-    eta_hat = expit(predict(M_model, data_pred)[1])
+    ## Compute linear predictor for M at reference levels of X and W    
+    eta_hat = a_hat[1] + a_hat[2] * x_pred + sum(a_hat[3:end] .* W_pred)
 
 
 
@@ -154,7 +151,7 @@ all_med_SEs = []
     reg_covs[1:a_size, 1:a_size] = a_cov
     reg_covs[a_size+1:end, a_size+1:end] = b_cov
 
[…]

### 2024-05-01 scripts/Empirical vs Analytical SEs.jl

@@ -16,7 +16,7 @@ n = 100
 p_conf = 3
 
 par_list_name = @savename num_reps n p_conf
-@load datadir("Bin-Resp, Cont-Med, Fixed-$par_list_name.jld2") all_a_hats all_a_SEs all_b_hats all_b_SEs all_med_hats all_med_SEs
+@load datadir("Cont-Resp, Bin-Med, Fixed-$par_list_name.jld2") all_a_hats all_a_SEs all_b_hats all_b_SEs all_med_hats all_med_SEs
 
 
 
@@ -28,3 +28,12 @@ emp_SE = std(all_med_hats)
 # Analytical SEs
 mean_SE = mean(all_med_SEs)
 median_SE = median(all_med_SEs)
+
+std(all_med_SEs)
+
+deleteat!(all_med_SEs, 892)
+
+for i in eachindex(all_med_SEs)
+    println(i)
+    all_med_SEs[i]
+end

### 2024-05-01 src/Helpers - Cont-Resp, Bin-Med, Fixed.jl

@@ -1,0 +1,122 @@
+
+# Compute mediation effect
+
+## Increment in conditional expectation of M for a unit increase in X
+function get_delta(eta, a_x)
+    Q1 = exp(-eta)
+    Q2 = (1 - exp(-a_x))
+    Q3 = (1 + exp(-eta - a_x))
+    Q4 = (1 + exp(-eta))
+
+    return Q1 * Q2 / (Q3 * Q4)
+end
+
+## Increment in conditional expectation of Y for a unit increase in X
+## I.e. Mean over M of increment in conditional expectation for Y given M and X
+function get_gamma(delta, b_m, b_x)
+    return b_m * delta + b_x
+end
+
+
+# Compute SE of estimated mediation effect
+# We use the delta method, so many derivatives are required
+
+
+## Helpful intermedeate quantities
+function d_delta_d_eta(eta, a_x)
+    delta = get_delta(eta, a_x)
+
+    Q1 = 1 - exp(-2*eta - a_x)
+    Q2 = 1 + exp(-eta)
+    Q3 = 1 + exp(-eta - a_x)
+
+    return (-1) * delta * Q1 / (Q2 * Q3)
+end
+
+# Old, incorrect version
+# function d_delta_d_a_x(eta, a_x)
+#     Q1 = exp(-2*eta - a_x)
+#     Q2 = 1 + exp(-eta)
+#     Q3 = 1 + exp(-eta - a_x)
+
+#     return Q1 / (Q2 * Q3^2)
+# end
+
+# Alternative expression
+# function d_delta_d_a_x(eta, a_x)
+#     delta = get_delta(eta, a_x)
+
+#     Q1 = exp(-a_x)
+#     Q2 = 1 + exp(-eta)
+#     Q3 = 1 + exp(-eta - a_x)
+#     Q4 = 1 - exp(-a_x)
+
+#     return delta * Q1 * Q2 / (Q3 * Q4)
+# end
+
+function d_delta_d_a_x(eta, a_x)
+    Q1 = exp(-2*eta - a_x)
+    Q2 = exp(-eta - a_x)
+    Q3 = 1 + exp(-eta)
+    Q4 = 1 + exp(-eta - a_x)
+
+    return (Q1 + Q2)/(Q3 * Q4^2)
+end
+
+
[…]

### 2024-05-01 scripts/Cont-Resp, Cont-Med, Fixed.jl
Increase n to 1000 in const, const, fixed

@@ -19,7 +19,7 @@ Random.seed!(1)
 
 num_reps = 1000    # Number of datasets to generate
 
-n = 100
+n = 1000
 
 a_0 = 1
 a_1 = 1
@@ -135,7 +135,7 @@ Random.seed!(1)
 
 num_reps = 1000    # Number of datasets to generate
 
-n = 100
+n = 1000
 p_conf = 3      # Number of confounders
 
 a_0 = 1

### 2024-05-01 notes/UQ Verification.tex
Binary response, binary mediator, fixed effects

@@
 
+I had to update Bruno's formulas.
+It looks like he forgot to include the $\bP(Y=1 | X=x, \mathbf{W}=\mathbf{w})$ term in the odds, and thus in the odds ratio.
+This has been changed now in the Overleaf document.
+I also worked out all the partial derivatives using Maple, which can produce \textit{Julia} (or \textit{R}, \textit{Python}) code to evaluate these horrendous expressions.
+All partials were validated against a simple finite difference approximation.
+
+Table \ref{tab:SEs_bin_bin_fix} gives empirical and analytical SEs when $n=1000$.
+
+\begin{table}
+    \centering
+    \begin{tabular}{|c|c|c|}
+        \hline
+        Empirical & Mean Analytical & Median Analytical\\
+        \hline
+        0.1060 & 0.1064 & 0.1062\\
+        \hline
+    \end{tabular}
+    \caption{Standard errors for estimated mediation effect with binary response and mediation, fixed-effects.}
+    \label{tab:SEs_bin_bin_fix}
+\end{table}
+
 \section{Open Questions}
@@
 Compute the fraction of total variance due to response replication and, interestingly, one minus this fraction as a naive representation of the variance fraction due to covariate variability.
+\textbf{Every time I've checked this, the difference has been very small.
+This isn't a priority for me right now.}
 

### 2024-05-01 scripts/Bin-Resp, Bin-Med, Fixed.jl

@@ -10,17 +10,18 @@ using JLD2
 using GLM   # For linear and logistic regression with fixed effects
 
 include("../src/Helpers - General.jl")
-include("../src/Helpers - Bin-Resp, Cont-Med, Fixed.jl")
+include("../src/Helpers - Bin-Resp, Bin-Med, Fixed.jl")
+
+# include("src/Helpers - General.jl")
+# include("src/Helpers - Bin-Resp, Bin-Med, Fixed.jl")
 
 
 
-
-
-Random.seed!(1)
+Random.seed!(11111)
 
 num_reps = 1000    # Number of datasets to generate
 
-n = 100
+n = 10000
 p_conf = 3      # Number of confounders
 
 a_1 = 1
@@ -134,23 +135,17 @@ all_med_SEs = []
 
     ## Compute linear predictor for M and Y at reference levels of X and W    
     eta_hat = a_hat[1] + a_hat[2] * x_pred + sum(a_hat[3:end] .* W_pred)    # Lin pred for M
-
-
     zeta_hat = b_hat[1] + b_hat[3] * x_pred + sum(b_hat[4:end] .* W_pred)   # Lin pred for Y (without M contribution)
 
 
-    #! START HERE
-
-
-
-
     ## Compute mediation effect
-    delta = get_delta(eta_hat, a_x)
-    med_hat = get_gamma(delta, b_m, b_x)
-
+    med_hat = get_odds_ratio(eta_hat, a_x, zeta_hat, b_x, b_m)
     push!(all_med_hats, med_hat)
 
-    ## Standard error
+
+
+
+    ## Analytical standard error
 
     ### Regression coefficients' asymptotic covariance matrix
     a_size = size(a_cov, 1)
@@ -160,14 +155,14 @@ all_med_SEs = []
     reg_covs[1:a_size, 1:a_size] = a_cov
     reg_covs[a_size+1:end, a_size+1:end] = b_cov
 
-    reg_asymp_cov = n .* reg_covs_raw
+    reg_asymp_cov = n .* reg_covs
 
 
     ### Derivatives of gamma wrt each parameter
-    dg_dt = d_gamma_d_theta(eta_hat, x_pred, W_pred, a_hat, b_hat)
+    dOR_dt = d_OR_d_theta(eta_hat, a_x, zeta_hat, b_x, b_m, x_pred, W_pred)
 
     ### Compute asymp var of gamma_hat
[…]

### 2024-05-01 scripts/Empirical vs Analytical SEs.jl

@@ -30,10 +30,3 @@ mean_SE = mean(all_med_SEs)
 median_SE = median(all_med_SEs)
 
 std(all_med_SEs)
-
-deleteat!(all_med_SEs, 892)
-
-for i in eachindex(all_med_SEs)
-    println(i)
-    all_med_SEs[i]
-end

### 2024-05-01 src/Helpers - Bin-Resp, Bin-Med, Fixed.jl

@@ -98,3 +98,132 @@ function d_gamma_d_theta(eta, x, W, a, b)
     return [dg_da_0; dg_da_1; dg_dA_2; dg_db_0; dg_db_1; dg_db_2; dg_dB_3]
 end
 
+
+
+
+
+
+
+
+
+
+
+
+function PY1(eta, zeta, beta)
+    Q1 = 1 + exp(-zeta)
+    Q2 = 1 + exp(-eta)
+    Q3 = 1 + exp(-zeta - beta)
+
+    num = Q1 + (Q2 - 1) * Q3
+    den = Q1 * Q2 * Q3
+
+    return num / den
+end
+
+function PY0(eta, zeta, beta)
+    Q1 = exp(-zeta)
+    Q2 = exp(-eta)
+    Q3 = exp(-beta)
+
+    num = Q1 * (Q2 + Q3 + Q1 * Q3 + Q1 * Q2 * Q3)
+    den = (1 + Q1) * (1 + Q2) * (1 + Q1*Q3)
+
+    return num / den
+end
+
+
+function get_odds_ref(eta, zeta, beta)
+    num = PY1(eta, zeta, beta)
+    den = PY0(eta, zeta, beta)
+
+    return num / den
+
+end
+
+function get_odds(eta, zeta, beta)
+    Q1 = exp(-zeta)
+    Q1_inv = 1/Q1
+    Q2 = exp(-eta)
+    Q3 = exp(-beta)
+
+    num = 1 + Q1_inv + Q2 * (Q3 + Q1_inv)
+    den = Q2 + Q3 + Q1 * Q3 + Q1 * Q2 * Q3
+
+    return num / den
+end
+
+function get_odds_ratio(eta, a_x, zeta, b_x, b_m)
+    odds_1 = get_odds(eta + a_x, zeta + b_x, b_m)
+    odds_2 = get_odds(eta, zeta, b_m)
+    OR_hat = odds_1 / odds_2
+end
+
+
+
+# Gradients of OR, obtained from Maple
+## I verified all of these against a simple finite difference calculation
+
+function d_OR_d_eta(eta, a_x, zeta, b_x, b_m)
[…]

### 2024-05-03 notes/UQ Verification.tex
Some progress toward random effects

@@
 \usepackage{graphicx}
+\usepackage{bm}             % For bold math symbols
 
@@
 \newcommand{\bV}{\mathbb{V}}
+
+\newcommand{\zdens}{\frac{e^{-z^2/2}}{\sqrt{2\pi}}}
 
@@
 I have no such reason for choosing this value of $W$.
-See the opening paragraph of this section for more on this.
+See the opening paragraph of this section for more details.
 
@@
 Here I will.
-I use $\gamma$ for the total mediation effect.
-In the previous section (albeit only in my code), I used $\delta$ for $\bE(M | X=x+1) - \bE(M | X=x)$, although here I don't think I will need this quantity.
+In the previous section, I used $\gamma$ for the total mediation effect and $\delta$ for $\bE(M | X=x+1) - \bE(M | X=x)$, although the latter I only used in my code.
+Here, I can just talk about odds and odds ratios.
 
@@
 
+\section{Binary Response, Binary Mediator, Random-Effects}
+
+Before doing any computing, a few notes about Bruno's derivation.
+The calculation for $n^{1/2}(\hat{\sigma}_U(x) - \sigma_U(x))$ contains a few steps which weren't immediately obvious to me.
+I'm going to clarify some of them here.
+First, what is the point of the first step?
+Well, note that $\hat{\sigma}_U(x) + \sigma_U(x) = 2 \sigma_U(x) + o_p(1)$, so the denominator is well-controlled.
+The numerator is now structured so we can use our formula for $\sigma^2_U(x)$ in terms of the components of $\Sigma_U$ from the previous line.
+While not strictly necessary, this saves us from the first step of every derivative calculation being applying the chain rule to deal with a square root.
+Being a little bit more formal, what we're doing here is applying the $\delta$-method to the numerator (with the $n^{1/2}$), then using Slutsky's Theorem/The Continuous Mapping Theorem to handle the ratio.
+No problems.
+
+The second point that tripped me up briefly is how the formula is organized.
+This is actually chosen to match our expression for $\sigma_U^2(x)$ in terms of the components of $\Sigma_U$ from the previous line.
+He then re-arranges to collect like terms in the parameter estimates.
+Third, there are a lot of $\sigma_U(x)$s in the denominators, but no 2s.
+This is because every term in our expression for $\sigma_U^2(x)$ is either a quadratic or multiplied by 2, so every term in the gradient of $\sigma_U^2(x)$ contains a 2.
+
+Fourth, I point out that the definition of $\phi(\mu, \sigma)$ is pretty clever.
+It reduces the problem from one with multiple random effects of a univariate problem with only a single Gaussian.
+This simplification arises from the observation that the random effects only enter into $\bP(Y=1)$ through a weighted sum.
+Since weighted sums of Gaussians are Gaussian, we can replace the whole sum with a new single Gaussian, setting the mean and variance of this new RV appropriately.
+
+Now onto some computation.
+
+\subsection{UQ for GLMM Parameters}
+
+Immediately I've encountered a problem.
+The standard software reports standard errors for the fixed effects' estimators, but not for the random effects' variance parameters\footnote{I'm going to be a bit sloppy here and refer to the standard error of a parameter.
+By this, I mean the standard error of our estimator for that parameter.
+It's just getting to be too much typing to always write the latter.}.
+I guess I'm going to have to work this out myself.
+
+To this end, I need the Hessian of the observed data (log) likelihood.
+There are tricks from the EM algorithm literature for doing this, but I think our problem is simple enough that I can just reduce it to a small-ish number of one-dimensional integrals.
+To start, recall the formula Bruno gives for $\bP(Y=1 | X=x, \mathbf{W}=\mathbf{w}) =: \bP_1$.
+Specifically, he expresses this quantity in terms of the $\phi$ function evaluated at four different arguments.
+Thus, if I have the Hessian of $\phi$, I can compute the Hessian of $\bP_1$ (and $\bP_0$), which in-turn gives me the Hessian of the log-likelihood based on a single point.
+
+This actually isn't quite what I need to do.
+In order to incorporate grouping structure, I need a generalization of the $\phi$ function, whose integrand contains a product of terms with the form $(1 + \exp(-\mu_i - \sigma_i z))^{-1}$, with $i$ ranging over all points in a single group.
+More formally, write
+%
+\begin{equation}
+    \Phi(\bm{\mu}, \bm{\sigma}) := \int \left[ \prod_i \frac{1}{\left( 1 + e^{-\mu_i - \sigma_i z} \right)} \right] \frac{e^{-z^2/2}}{\sqrt{2\pi}} dz
+\end{equation}
+%
+for the generalization of $\phi$ to multiple $\mu$, $\sigma$ from the same group.
+Write $P(z, \bm{\mu}, \bm{\sigma})$ for the product term in the integrand or, when no confusion can arise, $P(z)$ or even $P$.
+The likelihood of a dataset can now be obtained by first evaluating $\Phi$ on each group, then multiplying the results.
+
+In order to compute the Hessian of the log-likelihood, and by extension the asymptotic variance of our MLEs, we will require the gradient and Hessian of $\Phi$ with respect to $\bm{\mu}$ and $\bm{\sigma}$.
+To this end, observe that $\nabla_\mu P = P \bm{g}$, where $\bm{g}$ is a vector with $i$th component $- \nabla_{\mu_i} \log[1 + \exp(-\mu_i - \sigma_i z)] = [1 + \exp(\mu_i + \sigma_i z)]^{-1}$
+%
+\begin{align}
+    \nabla_{\mu} \Phi(\bm{\mu}, \bm{\sigma}) &= \int \nabla_\mu P \frac{e^{-z^2/2}}{\sqrt{2\pi}} dz\\
+    &=  \int P \bm{g} \zdens dz\\
+    &= \int \frac{P}{1 + e^{\bm{\mu} + \bm{\sigma}z}} \zdens dz
+\end{align}
+%
+where arithmetic operations on vectors are interpreted componentwise.
+Similarly, $\nabla_\sigma P = z P \bm{g}$ and
+%
+\begin{equation}
+    \nabla_{\sigma} \Phi(\bm{\mu}, \bm{\sigma}) = \int z P \bm{g} \zdens dz = \int z \frac{P}{1 + e^{\bm{\mu} + \bm{\sigma}z}} \zdens dz
+\end{equation}
+%
+Next, we need the second-derivatives. These are obtained simply using Maple or Wolfram Alpha. \textcolor{red}{To be continued\ldots}
+
+Before working harder on the derivatives of $\Phi$, I want to say exactly what I plan to do with this function.
+First, observe that the likelihood for the observed data, $Y$ and $M$, trivially factors as $\mathcal{L}(Y, M | X, W) = \mathcal{L}(Y|M, X, W) \cdot \mathcal{L}(M| X, W)$.
+Thus, we can treat our regression models for $M$ and $Y|M$ separately.
+Next observe that, within a single group,
+%
+\begin{align}
+    \mathcal{L}(M| X, W) &= \bP(M| X, W)\\
+    &= \bE_V \bP(M | V, X, W)\\
+    &= \bE_V \left( \left[ \prod_{m_i = 1} \bP(M=1 |V, X=x_i, W=w_i) \right]\right. \nonumber \\ 
+    &\left.\left[ \prod_{m_i = 0} \bP(M=0 |V, X=x_i, W=w_i)\right]\right)\\
+    &= 
+\end{align}
+
 \section{Open Questions}

### 2024-05-03 scripts/Bin-Resp, Bin-Med, Random.jl

@@ -1,0 +1,232 @@
+#* For now, I'm going to just focus on parameter estimation. This is an easier problem than random effects prediction
+#* Later, I can introduce prediction of random effects
+
+using Distributions, Random, Statistics
+using LogExpFunctions   # For the logit and expit functions
+using DataFrames
+using MixedModels
+using ProgressMeter, DrWatson
+using JLD2
+using GLM   # For linear and logistic regression with fixed effects
+
+include("../src/Helpers - General.jl")
+include("../src/Helpers - Bin-Resp, Bin-Med, Fixed.jl")
+
+# include("src/Helpers - General.jl")
+# include("src/Helpers - Bin-Resp, Bin-Med, Fixed.jl")
+
+
+
+Random.seed!(11111)
+
+num_reps = 1000    # Number of datasets to generate
+
+n = 100         # Sample size for each group
+K = 5           # Number of groups
+
+p_conf = 3      # Number of confounders
+
+
+# M model
+## Fixed effects
+a_1 = 1
+A_2 = repeat([1], p_conf)
+a_0 = - (a_1 + sum(A_2))    # Offset mean effect of other coefficients in linear predictor for M
+
+## Random effects
+sigma_a_0 = 0.2 * abs(a_0)
+sigma_a_1 = 0.2 * abs(a_1)
+
+
+
+# Y model
+## Fixed effects
+b_1 = 1
+b_2 = 1
+B_3 = repeat([1], p_conf)
+b_0 = - (b_1 + 0.5*b_2 + sum(B_3))    # Approximately offset mean effect of other coefficients in linear predictor for Y
+
+## Random effects
+sigma_b_0 = 0.2 * abs(b_0)
+# sigma_b_1 = 0.2 * abs(b_1)
+sigma_b_2 = 0.2 * abs(b_2)
+
+
+
+# Reference values at which we compute the total mediation effect
+x_pred = 0
+W_pred = repeat([1], p_conf)    # W=[1,1,1]
+
+
+# X_dist = Bernoulli(0.5)
+# W_dist = Bernoulli(0.5)
+
+X_dist = Normal(1, 1)
+W_dist = Normal(1, 1)
+
+
+
+all_a_hats = []
+all_b_hats = []
+
+all_a_SEs = []
+all_b_SEs = []
+
[…]

### 2024-05-03 notes/For the Group.Rmd
Convert code from Julia to RMarkdown. Finished cont, cont, fix and cont, bin, fix

---
title: "Exact Asymptotics for Causal Mediation Analysis"
author: "William Ruth"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
# Force text wrapping for long lines
library(formatR)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)

library(kableExtra)

knitr::opts_chunk$set(echo = TRUE)

```

I'm doing some Monte Carlo to verify the new SE formulas. Recall that we're doing mediation analysis, so we've got a response, $Y$, an exposure, $X$, and a mediation, $M$. We also have some number of confounders, which will be grouped together in the matrix $W$. Broadly speaking, we fit two regression models, one to predict $M$ using $X$ and $W$, the other to predict $Y$ using $M$, $X$ and $W$. We then compute the mediation effect (specifically, the total effect of $X$ on $Y$) as a function of the coefficients from these two regression models. An asymptotic SE for our mediation effect estimator can then be obtained from the asymptotic standard errors of our fitted regression coefficients using the $\delta$-method.

So far, so simple. There are a few places that things start to get more complicated. First, each of the two regression models can be either linear of logistic depending on whether the corresponding response variable is continuous or binary\footnote{In principle, we could have $Y$ and/or $M$ follow any distribution with a suitable GLM formulation. I don't think I've ever seen count data (i.e. Poisson regression) used here, much less anything more exotic.}. Furthermore, we can add random effects to our regression models. In the trust study, we have random effects for the intercept, $X$ and $M$ (naturally, the latter only applies when predicting $Y$).

In each setting, I generate $X$ and three confounders, $W = [W_1, W_2, W_3]$, each iid $\mathrm{N}(0,1)$. In both models, I set regression slopes to 1, and choose the intercept so that the mean of the linear predictor is approximately zero\footnote{In the fully continuous case, the mean linear predictor is exactly zero. When the mediator is binary, I choose the $Y$-intecept as though $\mathbb{E}M=\mathrm{expit}(0)=0.5$}. I generate some datasets (typically 1000, can be adjusted as necessary) with each of $n=100, 1000$ and $10000$ observations. On each dataset, we estimate the mediation effect and its SE (using the $\delta$-method). We then compare the empirical SE (i.e. SD over the $M$ Monte Carlo replicates) to the mean and median of our $M$ estimated SEs.

In summary, we can setup each model with the same boilerplate code as follows:

```{r, boilerplate, eval=FALSE}
# Note: This code chunk is not run here, but it is called and run at the start of every subsequent section. Any changes made here will be duplicated throughout the rest of the document.

set.seed(1)

num_reps = 1000

all_Ns = c(100, 200, 500, 1000, 2000, 5000, 10000)
p_conf = 3  # Number of confounders

# Regression coefficients
## Intercepts are set separately for each model so that the mean of the linear predictor is approximately zero

## M model
a_1 = 1                       # Coefficient for X
A_2 = rep(1, times = p_conf)  # Coefficients for confounders

# Y model
b_1 = 1                       # Coefficient for M
b_2 = 1                       # Coefficient for X
B_3 = rep(1, times = p_conf)  # Coefficients for confounders

# Containers for output
all_a_hats = list()
all_b_hats = list()

all_a_SEs = list()
all_b_SEs = list()

all_med_hats = list()
all_med_SEs = list()

```

In the rest of this document, we carry our Monte Carlo studies for different model configurations. We start with the simplest case: continuous response, continuous mediator, and fixed-effects. We then move to a binary mediator, followed by binary response. Finally, we introduce random effects.

# Continuous Response, Continuous Mediator, Fixed-Effects

```{r, include=FALSE}
# Run the boilerplate setup
<<boilerplate>>
```

## Generate data

Continuous variables are modelled as linear predictor plus residual. We therefore need to set the residual variance.

```{r}
sigma_M = 0.2
sigma_Y = 0.2
```

We also need to set the intercepts for the two models.
```{r}
a_0 = 0
b_0 = 0
```

For pedagogical purposes, I will demonstrate the analysis on a single simulated dataset, then run the full MC study invisibly and only show the results. First, we choose a sample size, $n$, and generate $X$ and $W$, then use them to generate $M$ and $Y$.

```{r}
n = all_Ns[1]

X = rnorm(n, mean=0, sd=1)
W = matrix(rnorm(n*p_conf, mean=0, sd=1), nrow = n, ncol = p_conf)

# Generate M
e_M = rnorm(n, 0, sigma_M)
M = a_0 + a_1*X + W%*%A_2 + e_M

# Generate Y
e_Y = rnorm(n, 0, sigma_Y)
Y = b_0 + b_1*M + b_2*X + W%*%B_3 + e_Y
```

## Estimate Mediation Effect

Next, we fit regression models for $M$ and $Y$ and extract relevant output (in the loop version, these are stored at each iteration)

```{r}
M_data = data.frame(M, X, W1 = W[,1], W2 = W[,2], W3 = W[,3])
M_model = lm(M ~ X + W1 + W2 + W3, data = M_data)

a_hat = summary(M_model)$coefficients[,1]
a_SE = summary(M_model)$coefficients[,2]
a_cov = vcov(M_model)

Y_data = data.frame(Y, M, X, W1 = W[,1], W2 = W[,2], W3 = W[,3])
Y_model = lm(Y ~ M + X + W1 + W2 + W3, data = Y_data)

b_hat = summary(Y_model)$coefficients[,1]
b_SE = summary(Y_model)$coefficients[,2]
b_cov = vcov(Y_model)
```

Now we can extract relevant coefficients and estimate the mediation effect

```{r}
a_x = a_hat[2]
b_x = b_hat[3]
b_m = b_hat[2]

med_hat = a_x*b_m + b_x
```

## Estimate Standard Error

Finally, we can estimate the SE of the mediation effect using the $\delta$-method. The way I do this is excessive here since the mediation effect has such a simple formula. Later though, the extra structure will make our lives easier.

We start by constructing the joint covariance matrix of the regression coefficients from the two models. This matrix is block-diagonal, with the blocks corresponding to the covariance matrices of the coefficients from the two models and off-diagonal entries zero.

```{r}
a_length = nrow(a_cov)
b_length = nrow(b_cov)

joint_cov = matrix(0, nrow = a_length + b_length, ncol = a_length + b_length)
joint_cov[1:a_length, 1:a_length] = a_cov
joint_cov[(a_length+1):(a_length+b_length), (a_length+1):(a_length+b_length)] = b_cov
```

Next, we compute the gradient of the mediation effect with respect to each regression coefficient.

```{r}
grad_a_0 = 0
grad_a_1 = b_m
grad_A_2 = rep(0, times = p_conf)

grad_b_0 = 0
grad_b_1 = a_x
grad_b_2 = 1
grad_B_3 = rep(0, times = p_conf)

grad_med = c(grad_a_0, grad_a_1, grad_A_2, grad_b_0, grad_b_1, grad_b_2, grad_B_3)
```

Finally, we can use the delta method to estimate the SE of the mediation effect. Note that the $\delta$-method works on asymptotic covariances, so we need to make sure to multiply/divide by $n$ or $\sqrt{n}$ where appropriate.

```{r}
asymp_reg_cov = n * joint_cov

med_asymp_var = grad_med %*% asymp_reg_cov %*% grad_med
med_asymp_SE = sqrt(med_asymp_var)

med_SE = med_asymp_SE / sqrt(n)
```

Putting everything together, we have an estimated mediation effect of `r med_hat` with an estimated SE of `r med_SE`.

## Monte Carlo Study

We now repeat the above analysis `r num_reps` times for each of various values of $n$. We also re-run the boilerplate code (not shown)

```{r, include=FALSE}
<<boilerplate>>
```

```{r, cache=TRUE, dependson=c("boilerplate")}

for(j in seq_along(all_Ns)){
  n = all_Ns[j]

  # Containers for output with this value of n
  some_a_hats = list()
  some_b_hats = list()

  some_a_SEs = list()
  some_b_SEs = list()

  some_med_hats = list()
  some_med_SEs = list()

  for(i in 1:num_reps){
    # Generate data
    X = rnorm(n, mean=0, sd=1)
    W = matrix(rnorm(n*p_conf, mean=0, sd=1), nrow = n, ncol = p_conf)

    ## Generate M
    e_M = rnorm(n, 0, sigma_M)
    M = a_0 + a_1*X + W%*%A_2 + e_M

    ## Generate Y
    e_Y = rnorm(n, 0, sigma_Y)
    Y = b_0 + b_1*M + b_2*X + W%*%B_3 + e_Y

    # Fit models

    ## M
    M_data = data.frame(M, X, W1 = W[,1], W2 = W[,2], W3 = W[,3])
    M_model = lm(M ~ X + W1 + W2 + W3, data = M_data)

    a_hat = summary(M_model)$coefficients[,1]
    a_SE = summary(M_model)$coefficients[,2]
    a_cov = vcov(M_model)

    some_a_hats[[i]] = a_hat
    some_a_SEs[[i]] = a_SE

    ## Y
    Y_data = data.frame(Y, M, X, W1 = W[,1], W2 = W[,2], W3 = W[,3])
    Y_model = lm(Y ~ M + X + W1 + W2 + W3, data = Y_data)

    b_hat = summary(Y_model)$coefficients[,1]
    b_SE = summary(Y_model)$coefficients[,2]
    b_cov = vcov(Y_model)

    some_b_hats[[i]] = b_hat
    some_b_SEs[[i]] = b_SE

    # Estimate mediation effect

    ## Extract coefficients
    a_x = a_hat[2]
    b_x = b_hat[3]
    b_m = b_hat[2]

    ## Compute estimate
    med_hat = a_x*b_m + b_x
    some_med_hats[[i]] = med_hat

    # Estimate SE

    ## Build joint covariance matrix
    a_length = nrow(a_cov)
    b_length = nrow(b_cov)

    joint_cov = matrix(0, nrow = a_length + b_length, ncol = a_length + b_length)
    joint_cov[1:a_length, 1:a_length] = a_cov
    joint_cov[(a_length+1):(a_length+b_length), (a_length+1):(a_length+b_length)] = b_cov

    ## Compute gradient of mediation effect
    grad_a_0 = 0
    grad_a_1 = b_m
    grad_A_2 = rep(0, times = p_conf)

    grad_b_0 = 0
    grad_b_1 = a_x
    grad_b_2 = 1
    grad_B_3 = rep(0, times = p_conf)

    grad_med = c(grad_a_0, grad_a_1, grad_A_2, grad_b_0, grad_b_1, grad_b_2, grad_B_3)

    ## Apply delta-method
    asymp_reg_cov = n * joint_cov

    med_asymp_var = grad_med %*% asymp_reg_cov %*% grad_med
    med_asymp_SE = sqrt(med_asymp_var)

    med_SE = med_asymp_SE / sqrt(n)
    some_med_SEs[[i]] = med_SE
  }

  # Store output for current value of n
  all_a_hats[[j]] = some_a_hats
  all_b_hats[[j]] = some_b_hats

  all_a_SEs[[j]] = some_a_SEs
  all_b_SEs[[j]] = some_b_SEs

  all_med_hats[[j]] = some_med_hats
  all_med_SEs[[j]] = some_med_SEs
}

```

We now process the output from our simulation and summarize the results in a table. We give the mean and median of the estimated standard errors, as well as their relative errors (in $\%$) as estimates of the empirical standard error. See Table \ref{tab:results_cont_cont_fix}. Note that the relative error decreased and became more consistent when I increased the number of Monte Carlo replicates from 100 to 1000.

```{r}
# Re-format output to be more useful. 
data_med_hats = lapply(all_med_hats, unlist)
data_med_SEs = lapply(all_med_SEs, unlist)

# SD of estimates across Monte Carlo samples
emp_SEs = sapply(data_med_hats, sd)

# Summaries of estimated SEs (mean and median)
mean_delta_SEs = sapply(data_med_SEs, mean)
median_delta_SEs = sapply(data_med_SEs, median)

# Combine into a table
results = data.frame(n = all_Ns, Empirical = emp_SEs, 
                     Mean = mean_delta_SEs, Median = median_delta_SEs,
                     "Percent_Err_Mean" = 100*(mean_delta_SEs - emp_SEs)/emp_SEs,
                     "Percent_Err_Median" = 100*(median_delta_SEs - emp_SEs)/emp_SEs)

```

```{r results="asis", echo=FALSE}
## Print table
kbl(results, format = "latex", booktabs = T, caption = "Summary of SE estimates for mediation effect under continuous response, continuous mediator, fixed-effects model \\label{tab:results_cont_cont_fix}")
```

# Continuous Response, Binary Mediator, Fixed-Effects

```{r, include=FALSE}
# Run the boilerplate setup
<<boilerplate>>
```

We now generate $M$ as a binary variable using a logistic regression model. We continue to use a continuous model for $Y|M$. This analysis is sufficiently complicated that I wrote some helper functions in another script. See `Helpers.R`.

```{r}
source("../src/Helpers.R")
```

## Generate data

The continuous variable is modelled as linear predictor plus residual. We therefore need to set the residual variance.

```{r}
sigma_Y = 0.2
```

We also need to set the intercepts for the two models. Recall that we want the mean of the linear predictor to be zero in both models, and that we approximate the mean of $M$ as the logit of the mean of its linear predictor (i.e. $\mathbb{E}M \approx \mathrm{expit}(0) = 0.5$).
```{r}
a_0 = 0
b_0 = -0.5
```

As in the previous section, I demonstrate the analysis on a single simulated dataset, then run the full MC study invisibly and only show the results. First, we choose a sample size, $n$, and generate $X$ and $W$, then use them to generate $M$ and $Y$.

```{r}
n = all_Ns[1]

X = rnorm(n, mean=0, sd=1)
W = matrix(rnorm(n*p_conf, mean=0, sd=1), nrow = n, ncol = p_conf)

# Generate M
eta_vec = a_0 + a_1*X + W%*%A_2
p_M_vec = expit(eta_vec)
M = rbinom(n, size = 1, prob = p_M_vec)

# Generate Y
e_Y = rnorm(n, 0, sigma_Y)
Y = b_0 + b_1*M + b_2*X + W%*%B_3 + e_Y
```

## Estimate Mediation Effect

Next, we fit regression models for $M$ and $Y$ and extract relevant output (in the loop version, these are stored at each iteration)

```{r}
M_data = data.frame(M, X, W1 = W[,1], W2 = W[,2], W3 = W[,3])
M_model = glm(M ~ X + W1 + W2 + W3, data = M_data, family = binomial(link = "logit"))

a_hat = summary(M_model)$coefficients[,1]
a_SE = summary(M_model)$coefficients[,2]
a_cov = vcov(M_model)

Y_data = data.frame(Y, M, X, W1 = W[,1], W2 = W[,2], W3 = W[,3])
Y_model = lm(Y ~ M + X + W1 + W2 + W3, data = Y_data)

b_hat = summary(Y_model)$coefficients[,1]
b_SE = summary(Y_model)$coefficients[,2]
b_cov = vcov(Y_model)
```

Now we can extract relevant coefficients and estimate the mediation effect. Note that this is more involved than the fully continuous case. In particular, the total effect of $X$ on $Y$ depends on the levels of $X$ and the confounders, $W$. I evaluate the effect at $X=0$ and $W = [1,1,1]$. The former represents the effect of a binary exposure, but the latter is chosen completely arbitrarily.

```{r}
x_pred = 0
W_pred = c(1,1,1)

a_0_hat = a_hat[1]
a_x_hat = a_hat[2]
A_2_hat = a_hat[3:5]

b_0_hat = b_hat[1]
b_m_hat = b_hat[2]
b_x_hat = b_hat[3]
B_3_hat = b_hat[4:6]

# Linear predictor for M
eta_hat = a_0_hat + a_x_hat*x_pred + W_pred%*%A_2_hat

# Increment in the conditional expectation of M
delta_hat = get_delta(eta_hat, a_x_hat)

# Increment in the conditional expectation of Y
med_hat = get_gamma(delta_hat, b_m_hat, b_x_hat)
```

## Estimate Standard Error

Finally, we can estimate the SE of the mediation effect using the $\delta$-method. Now that we have a more complicated expression for the total effect, it makes more sense to use the general formula for the $\delta$-method standard error.

We start by constructing the joint covariance matrix of the regression coefficients from the two models. This matrix is block-diagonal, with the blocks corresponding to the covariance matrices of the coefficients from the two models and off-diagonal entries zero.

```{r}
a_length = nrow(a_cov)
b_length = nrow(b_cov)

joint_cov = matrix(0, nrow = a_length + b_length, ncol = a_length + b_length)
joint_cov[1:a_length, 1:a_length] = a_cov
joint_cov[(a_length+1):(a_length+b_length), (a_length+1):(a_length+b_length)] = b_cov
```

Next, we compute the gradient of the mediation effect with respect to each regression coefficient.

```{r}
d_gamma_d_theta(eta_hat, x_pred, W_pred, a_hat, b_hat)
```

Finally, we can use the delta method to estimate the SE of the mediation effect. Note that the $\delta$-method works on asymptotic covariances, so we need to make sure to multiply/divide by $n$ or $\sqrt{n}$ where appropriate.

```{r}
asymp_reg_cov = n * joint_cov

med_asymp_var = grad_med %*% asymp_reg_cov %*% grad_med
med_asymp_SE = sqrt(med_asymp_var)

med_SE = med_asymp_SE / sqrt(n)
```

Putting everything together, we have an estimated mediation effect of `r med_hat` with an estimated SE of `r med_SE`.

## Monte Carlo Study

We now repeat the above analysis `r num_reps` times for each of various values of $n$. Values of $X$ and $W$ for which we compute the total effect are set at the beginning. We also re-run the boilerplate code (not shown)

```{r, include=FALSE}
<<boilerplate>>
```

```{r}
# Values of X and W for which we compute the total effect
x_pred = 0
W_pred = c(1,1,1)
```

```{r, cache=TRUE, dependson=c("boilerplate")}

for(j in seq_along(all_Ns)){
  n = all_Ns[j]

  # Containers for output with this value of n
  some_a_hats = list()
  some_b_hats = list()

  some_a_SEs = list()
  some_b_SEs = list()

  some_med_hats = list()
  some_med_SEs = list()

  for(i in 1:num_reps){
    # Generate data
    X = rnorm(n, mean=0, sd=1)
    W = matrix(rnorm(n*p_conf, mean=0, sd=1), nrow = n, ncol = p_conf)

    ## Generate M
    eta_vec = a_0 + a_1*X + W%*%A_2
    p_M_vec = expit(eta_vec)
    M = rbinom(n, size = 1, prob = p_M_vec)

    ## Generate Y
    e_Y = rnorm(n, 0, sigma_Y)
    Y = b_0 + b_1*M + b_2*X + W%*%B_3 + e_Y

    # Fit models

    ## M
    M_data = data.frame(M, X, W1 = W[,1], W2 = W[,2], W3 = W[,3])
    M_model = glm(M ~ X + W1 + W2 + W3, data = M_data, family = binomial(link = "logit"))

    a_hat = summary(M_model)$coefficients[,1]
    a_SE = summary(M_model)$coefficients[,2]
    a_cov = vcov(M_model)

    some_a_hats[[i]] = a_hat
    some_a_SEs[[i]] = a_SE

    ## Y
    Y_data = data.frame(Y, M, X, W1 = W[,1], W2 = W[,2], W3 = W[,3])
    Y_model = lm(Y ~ M + X + W1 + W2 + W3, data = Y_data)

    b_hat = summary(Y_model)$coefficients[,1]
    b_SE = summary(Y_model)$coefficients[,2]
    b_cov = vcov(Y_model)

    some_b_hats[[i]] = b_hat
    some_b_SEs[[i]] = b_SE

    # Estimate mediation effect

    ## Extract coefficients
    a_0_hat = a_hat[1]
    a_x_hat = a_hat[2]
    A_2_hat = a_hat[3:length(a_hat)]

    b_0_hat = b_hat[1]
    b_m_hat = b_hat[2]
    b_x_hat = b_hat[3]
    B_3_hat = b_hat[4:length(b_hat)]

    ## Linear predictor for M
    eta_hat = a_0_hat + a_x_hat*x_pred + W_pred%*%A_2_hat

    ## Increment in the conditional expectation of M
    delta_hat = get_delta(eta_hat, a_x_hat)

    ## Increment in the conditional expectation of Y (i.e. mediation effect)
    med_hat = get_gamma(delta_hat, b_m_hat, b_x_hat)
    some_med_hats[[i]] = med_hat

    # Estimate SE

    ## Build joint covariance matrix
    a_length = nrow(a_cov)
    b_length = nrow(b_cov)

    joint_cov = matrix(0, nrow = a_length + b_length, ncol = a_length + b_length)
    joint_cov[1:a_length, 1:a_length] = a_cov
    joint_cov[(a_length+1):(a_length+b_length), (a_length+1):(a_length+b_length)] = b_cov

    ## Compute gradient of mediation effect
    grad_med = d_gamma_d_theta(eta_hat, x_pred, W_pred, a_hat, b_hat)

    ## Apply delta-method
    asymp_reg_cov = n * joint_cov

    med_asymp_var = grad_med %*% asymp_reg_cov %*% grad_med
    med_asymp_SE = sqrt(med_asymp_var)

    med_SE = med_asymp_SE / sqrt(n)
    some_med_SEs[[i]] = med_SE
  }

  # Store output for current value of n
  all_a_hats[[j]] = some_a_hats
  all_b_hats[[j]] = some_b_hats

  all_a_SEs[[j]] = some_a_SEs
  all_b_SEs[[j]] = some_b_SEs

  all_med_hats[[j]] = some_med_hats
  all_med_SEs[[j]] = some_med_SEs
}

```

We now process the output from our simulation and summarize the results in a table. We give the mean and median of the estimated standard errors, as well as their relative errors (in $\%$) as estimates of the empirical standard error. See Table \ref{tab:results_cont_bin_fix}. Note that the relative error decreased and became more consistent when I increased the number of Monte Carlo replicates from 100 to 1000.

```{r}
# Re-format output to be more useful. 
data_med_hats = lapply(all_med_hats, unlist)
data_med_SEs = lapply(all_med_SEs, unlist)

# SD of estimates across Monte Carlo samples
emp_SEs = sapply(data_med_hats, sd)

# Summaries of estimated SEs (mean and median)
mean_delta_SEs = sapply(data_med_SEs, mean)
median_delta_SEs = sapply(data_med_SEs, median)

# Combine into a table
results = data.frame(n = all_Ns, Empirical = emp_SEs, 
                     Mean = mean_delta_SEs, Median = median_delta_SEs,
                     "Percent_Err_Mean" = 100*(mean_delta_SEs - emp_SEs)/emp_SEs,
                     "Percent_Err_Median" = 100*(median_delta_SEs - emp_SEs)/emp_SEs)

```

```{r results="asis", echo=FALSE}
## Print table
kbl(results, format = "latex", booktabs = T, caption = "Summary of SE estimates for mediation effect under continuous response, continuous mediator, fixed-effects model \\label{tab:results_cont_bin_fix}")
```

### 2024-05-03 scripts/Cont-Resp, Bin-Med, Fixed.jl

@@ -28,7 +28,6 @@ p_conf = 3      # Number of confounders
 a_1 = 1
 A_2 = repeat([1], p_conf)
 a_0 = - (a_1 + sum(A_2))    # Offset mean effect of other coefficients in linear predictor for M
-sigma_d = 0.2   # Residual SD for M
 
 b_1 = 1
 b_2 = 1
@@ -48,6 +47,7 @@ W_pred = repeat([1], p_conf)    # W=[1,1,1]
 
 X_dist = Normal(1, 1)
 W_dist = Normal(1, 1)
+
 
 
 
@@ -126,10 +126,7 @@ all_med_SEs = []
 
 
 
-    ## Get coefficients from models for M and Y
-    a_hat = coef(M_model)
-    b_hat = coef(Y_model)
-
+    ## Get relevant coefficients from models for M and Y
     a_x = a_hat[2]
     b_m = b_hat[2]
     b_x = b_hat[3]

### 2024-05-03 src/Helpers.R

@@ -1,0 +1,111 @@
+# General ####
+
+logit = function(x) log(x/(1-x))
+
+expit = function(x) exp(x)/(1+exp(x))
+
+
+
+# Cont Resp, Bin Med, Fixed-Effs ####
+## Some notation:
+  ## eta: a_0 + a_1 x + A_2 W = a_0 + a_x * x + A_2' w. Linear predictor for M
+  ## zeta: b_0 + b_1 * m + b_2 * x + B_3' w = b_0 + b_m * m + b_x * x + B_3' w. Linear predictor for Y
+
+  ## delta: E(M | X=x+1) - E(M | X=x)
+  ## gamma: E(Y | X=x+1) - E(Y | X=x). Total effect of X on Y
+
+
+get_delta = function(eta, a_x){
+  Q1 = exp(-eta)
+  Q2 = (1 - exp(-a_x))
+  Q3 = (1 + exp(-eta - a_x))
+  Q4 = (1 + exp(-eta))
+
+  return(Q1 * Q2 / (Q3 * Q4))
+}
+
+
+get_gamma = function(delta, b_m, b_x){
+  return(delta * b_m + b_x)
+}
+
+
+## Gradient of total effect wrt the a's and b's
+
+### First, some intermediate quantities
+d_delta_d_eta = function(eta, a_x){
+  delta = get_delta(eta, a_x)
+
+  Q1 = 1 - exp(-2*eta - a_x)
+  Q2 = 1 + exp(-eta)
+  Q3 = 1 + exp(-eta - a_x)
+
+  return( (-1) * delta * Q1 / (Q2 * Q3))
+}
+
+d_delta_d_a_x = function(eta, a_x){
+  Q1 = exp(-2*eta - a_x)
+  Q2 = exp(-eta - a_x)
+  Q3 = 1 + exp(-eta)
+  Q4 = 1 + exp(-eta - a_x)
+
+  return((Q1 + Q2)/(Q3 * Q4^2))
+}
+
+
+### Now, the gradient
+d_gamma_d_a_0 <- function(dd_de, b_m) {
+  return(dd_de * b_m)
+}
+
+d_gamma_d_a_1 <- function(dd_de, dd_da_x, x, b_m) {
+  return(b_m * x * dd_de + b_m * dd_da_x)
+}
+
[…]

### 2024-05-06 notes/Literature Review/Literature Review.tex
Expand literature review

\documentclass{article}

\usepackage{amsmath, amssymb}
\usepackage{xcolor}
\usepackage{caption, subcaption}
\usepackage{graphicx}
\usepackage{bm}             % For bold math symbols

\usepackage{natbib}
\bibliographystyle{plainnat}

\newcommand{\bG}{\mathbb{G}}
\newcommand{\bF}{\mathbb{F}}
\newcommand{\bP}{\mathbb{P}}
\newcommand{\bE}{\mathbb{E}}
\newcommand{\bV}{\mathbb{V}}

\newcommand{\zdens}{\frac{e^{-z^2/2}}{\sqrt{2\pi}}}

\title{Verification of SE Formulas for Mediation Analysis}

\begin{document}

\maketitle

There is a short exchange in the American Journal of Epidemiology

\section{The \citeauthor{Sam23} group}

This group does essentially what B\&B are proposing with exact formulas and $\delta$-method standard errors. In particular, they avoid any requirement for rare events. They also do direct and indirect effects. However, I haven't seen any mention of random-effects/multilevel-models.

\subsection{\citet{Sam18}}

Gives exact formulas for direct and indirect mediation effects, as well as $\delta$-method standard errors. Binary outcome, binary mediator. No analytical SE when exposure-mediator interaction is present.

\subsection{\citet{Sam21}}

Extends analytical results of \citet{Sam18} to handle exposure-mediator interaction term in model for outcome.

\subsection{\citet{Sam23}}

As \citet{Sam18}, but with continuous mediator and binary response. No mixed-effects. Very nice simulation study.

\subsection{\citet{Cau24}}

Extends the work of \citet{Sam23} to case-control data.

\section{Derivative-Based}

Defines mediation effects in terms of derivatives. E.g. They define the indirect effect of $X$ on $Y$ as $(\partial \bE Y/ \partial \bE M) \cdot (\partial \bE M / \partial X)$, although that's not quite how they state it in the paper. This approach has had a small, but consistent following.

\subsection{\citet{Mar24}}

Incorporates random-effects in the derivative-based approach. Mostly does Bayesian inference. Mentions bootstrap.

\section{GLMM UQ}

\subsection{\citet{Zhe21}}

Does joint uncertainty quantification for the parameters and random effects. Based on maximizing the Laplace Approximation to the marginal likelihood. Gives $\delta$-method SEs with gradients from automatic differentiation

\section{Applications}

\subsection{Not Relevant}

\citet{Bro22} talks about mediation analysis and includes models with mixed-effects, but the latter are only used for imputation.

\bibliography{Lit_Review_Bib}

\end{document}

### 2024-05-06 notes/William Analysis.Rmd
Re-name write-up
Implement binary, binary, fixed in R. SEs are currently wrong

--- 2024-05-03 notes/For the Group.Rmd
@@
 
-```{r, boilerplate, eval=FALSE}
+```{r, boilerplate, include=FALSE, cached=TRUE}
 # Note: This code chunk is not run here, but it is called and run at the start of every subsequent section. Any changes made here will be duplicated throughout the rest of the document.
@@
 ```
+
+# Binary Response, Binary Mediator, Fixed-Effects
+
+For our last fixed-effect model, we now take both the response and mediator to be binary.
+I wrote some helper functions to compute relevant derivatives, all of which validated against a simple finite-difference approximation.
+See `Helpers.R`.
+
+```{r}
+source("../src/Helpers.R")
+```
+
+## Generate Data
+
+As in the previous case, we set intercepts so that the means of both linear predictors are at least approximately zero. Here however, no residuals are required.
+
+```{r}
+a_0 = 0
+b_0 = -0.5
+```
+
+We first demonstrate our analysis on a single dataset, then put this analysis inside a `for` loop.
+The first step is to set the sample size and generate some data.
+We use $\eta$ to denote the linear predictor for $M$, and $\zeta$ for the linear predictor of $Y$ without the $M$ term.
+That is, $\zeta = b_0 + b_2 X + W B_3$.
+
+```{r}
+n = all_Ns[1]
+
+X = rnorm(n, mean=0, sd=1)
+W = matrix(rnorm(n*p_conf, mean=0, sd=1), nrow = n, ncol = p_conf)
+
+# Generate M
+eta_vec = a_0 + a_1*X + W%*%A_2
+p_M_vec = expit(eta_vec)
+M = rbinom(n, size = 1, prob = p_M_vec)
+
+# Generate Y
+zeta_vec = b_0 + b_1*M + b_2*X + W%*%B_3
+p_Y_vec = expit(zeta_vec)
+Y = rbinom(n, size = 1, prob = p_Y_vec)
+```
+
+## Estimate Mediation Effect
+
+We now fit logistic regression models to predict $M$ and $Y$, and extract output that we will need later.
+
+```{r}
+M_data = data.frame(M, X, W1 = W[, 1], W2 = W[, 2], W3 = W[,3])
+M_model = glm(M ~ X + W1 + W2 + W3, data = M_data, family = binomial(link = "logit"))
+
+a_hat = summary(M_model)$coefficients[, 1]
+a_SE = summary(M_model)$coefficients[, 2]
+a_cov = vcov(M_model)
+
+Y_data = data.frame(Y, M, X, W1 = W[, 1], W2 = W[, 2], W3 = W[,3])
+Y_model = glm(Y ~ M + X + W1 + W2 + W3, data = Y_data, family = binomial(link = "logit"))
+
+b_hat = summary(Y_model)$coefficients[, 1]
+b_SE = summary(Y_model)$coefficients[, 2]
+b_cov = vcov(Y_model)
+```
+
+Next, we estimate the mediation effect.
+Note that, as in the contiuous outcome, binary mediator setting, our mediation effect depends on $X$ and $W$.
+We again use $X=0$ and $W=[1,1,1]$, although Samoilenko and Lefebvre (2023) use the sample means of these covariates.
+
+```{r}
+x_pred = 0
+W_pred = c(1,1,1)
+
+a_0_hat = a_hat[1]
+a_x_hat = a_hat[2]
+A_2_hat = a_hat[3:5]
+
+b_0_hat = b_hat[1]
+b_m_hat = b_hat[2]
+b_x_hat = b_hat[3]
+B_3_hat = b_hat[4:6]
+
+# Linear predictors
+eta_hat = a_0_hat + a_x_hat*x_pred + W_pred%*%A_2_hat
+zeta_hat = b_0_hat + b_x_hat * x_pred + W_pred %*% B_3_hat
+
+# Mediation effect
+## See Helpers.R for the function get_odds_ratio
+med_hat = get_odds_ratio(eta_hat, a_x, zeta_hat, b_x, b_m)
+```
+
+We now compute the gradient of the odds ratio and use the $\delta$-method to get the standard error.
+
+```{r}
+# Gradient of OR wrt regression coefficients
+grad_med = d_OR_d_theta(eta_hat, a_x_hat, zeta_hat, b_x_hat, b_m_hat, x_pred, W_pred)
+
+# Get asymptotic SE using delta method
+
+## Build joint covariance matrix of regression coefficients
+a_length = nrow(a_cov)
+b_length = nrow(b_cov)
+joint_cov = matrix(0, nrow = a_length + b_length, ncol = a_length +
+b_length)
+joint_cov[1:a_length, 1:a_length] = a_cov
+joint_cov[(a_length + 1):(a_length + b_length), (a_length + 1):(a_length +
+b_length)] = b_cov
+
+## Convert to asymptotic covariance matrix
+asymp_reg_cov = n * joint_cov
+
+## Pre- and post-multiply asymptotic covariance by gradient of transformation
+med_asymp_var = grad_med %*% asymp_reg_cov %*% grad_med
+
+## Get small-sample standard error
+med_asymp_SE = sqrt(med_asymp_var)
+med_SE = med_asymp_SE / sqrt(n)
+```
+
+Putting everything together, we have an estimated mediation effect of `r med_hat`, with an estimated SE of `r med_SE`.
+
+## Monte Carlo Study
+
+We now repeat the above analysis `r num_reps` times for each of various values of $n$.
+Values of $X$ and $W$ for which we compute the total effect are set at the beginning.
+We also re-run the boilerplate code (not shown).
+
+```{r, include=FALSE}
+<<boilerplate>>
+```
+
+```{r}
+# Values of X and W for which we compute the total effect
+x_pred = 0
+W_pred = c(1,1,1)
+
+# Values of a_0 and b_0
+a_0 = 0
+b_0 = -0.5
+```
+
+```{r, cache=TRUE, dependson=c("boilerplate")}
+
+for(j in seq_along(all_Ns)){
+  n = all_Ns[j]
+
+  # Containers for output with this value of n
+  some_a_hats = list()
+  some_b_hats = list()
+
+  some_a_SEs = list()
+  some_b_SEs = list()
+
+  some_med_hats = list()
+  some_med_SEs = list()
+
+  for(i in 1:num_reps){
+    # Generate data
+    X = rnorm(n, mean=0, sd=1)
+    W = matrix(rnorm(n*p_conf, mean=0, sd=1), nrow = n, ncol = p_conf)
+
+    ## Generate M
+    eta_vec = a_0 + a_1*X + W%*%A_2
+    p_M_vec = expit(eta_vec)
+    M = rbinom(n, size = 1, prob = p_M_vec)
+
+    ## Generate Y
+    zeta_vec = b_0 + b_1*M + b_2*X + W%*%B_3
+    p_Y_vec = expit(zeta_vec)
+    Y = rbinom(n, size = 1, prob = p_Y_vec)
+
+    # Fit models
+
+    ## M
+    M_data = data.frame(M, X, W1 = W[,1], W2 = W[,2], W3 = W[,3])
+    M_model = glm(M ~ X + W1 + W2 + W3, data = M_data, family = binomial(link = "logit"))
+
+    a_hat = summary(M_model)$coefficients[,1]
+    a_SE = summary(M_model)$coefficients[,2]
+    a_cov = vcov(M_model)
+
+    some_a_hats[[i]] = a_hat
+    some_a_SEs[[i]] = a_SE
+
+    ## Y
+    Y_data = data.frame(Y, M, X, W1 = W[,1], W2 = W[,2], W3 = W[,3])
+    Y_model = glm(Y ~ M + X + W1 + W2 + W3, data = Y_data, family = binomial(link = "logit"))
+
+    b_hat = summary(Y_model)$coefficients[,1]
+    b_SE = summary(Y_model)$coefficients[,2]
+    b_cov = vcov(Y_model)
+
+    some_b_hats[[i]] = b_hat
+    some_b_SEs[[i]] = b_SE
+
+    # Estimate mediation effect
+
+    ## Extract coefficients
+    a_0_hat = a_hat[1]
+    a_x_hat = a_hat[2]
+    A_2_hat = a_hat[3:length(a_hat)]
+
+    b_0_hat = b_hat[1]
+    b_m_hat = b_hat[2]
+    b_x_hat = b_hat[3]
+    B_3_hat = b_hat[4:length(b_hat)]
+
+    ## Linear predictor for M
+    eta_hat = a_0_hat + a_x_hat*x_pred + W_pred%*%A_2_hat
+    zeta_hat = b_0_hat + b_x_hat * x_pred + W_pred %*% B_3_hat
+
+    # Mediation effect
+    ## See Helpers.R for the function get_odds_ratio
+    med_hat = get_odds_ratio(eta_hat, a_x, zeta_hat, b_x, b_m)
+    some_med_hats[[i]] = med_hat
+
+    # Estimate SE
+
+    ## Build joint covariance matrix
+    a_length = nrow(a_cov)
+    b_length = nrow(b_cov)
+
+    joint_cov = matrix(0, nrow = a_length + b_length, ncol = a_length + b_length)
+    joint_cov[1:a_length, 1:a_length] = a_cov
+    joint_cov[(a_length+1):(a_length+b_length), (a_length+1):(a_length+b_length)] = b_cov
+
+    ## Compute gradient of mediation effect
+    grad_med = d_OR_d_theta(eta_hat, a_x_hat, zeta_hat, b_x_hat, b_m_hat, x_pred, W_pred)
+
+    ## Apply delta-method
+    asymp_reg_cov = n * joint_cov
+
+    med_asymp_var = grad_med %*% asymp_reg_cov %*% grad_med
+    med_asymp_SE = sqrt(med_asymp_var)
+
+    med_SE = med_asymp_SE / sqrt(n)
+    some_med_SEs[[i]] = med_SE
+  }
+
+  # Store output for current value of n
+  all_a_hats[[j]] = some_a_hats
+  all_b_hats[[j]] = some_b_hats
+
+  all_a_SEs[[j]] = some_a_SEs
+  all_b_SEs[[j]] = some_b_SEs
+
+  all_med_hats[[j]] = some_med_hats
+  all_med_SEs[[j]] = some_med_SEs
+}
+
+```
+
+We now process the output from our simulation and summarize the results in a table.
+We give the mean and median of the estimated standard errors, as well as their relative errors (in $\%$) as estimates of the empirical standard error.
+See Table \ref{tab:results_bin_bin_fix}.
+
+```{r}
+# Re-format output to be more useful. 
+data_med_hats = lapply(all_med_hats, unlist)
+data_med_SEs = lapply(all_med_SEs, unlist)
+
+# SD of estimates across Monte Carlo samples
+emp_SEs = sapply(data_med_hats, sd)
+
+# Summaries of estimated SEs (mean and median)
+mean_delta_SEs = sapply(data_med_SEs, mean)
+median_delta_SEs = sapply(data_med_SEs, median)
+
+# Combine into a table
+results = data.frame(n = all_Ns, Empirical = emp_SEs, 
+                     Mean = mean_delta_SEs, Median = median_delta_SEs,
+                     "Percent_Err_Mean" = 100*(mean_delta_SEs - emp_SEs)/emp_SEs,
+                     "Percent_Err_Median" = 100*(median_delta_SEs - emp_SEs)/emp_SEs)
+
+```
+
+```{r results="asis", echo=FALSE}
+## Print table
+kbl(results, format = "latex", booktabs = T, caption = "Summary of SE estimates for mediation effect under continuous response, continuous mediator, fixed-effects model \\label{tab:results_bin_bin_fix}")
+```

### 2024-05-06 src/Helpers - Bin-Resp, Bin-Med, Fixed.jl

@@ -1,114 +1,7 @@
 
 # Compute mediation effect
 
-## Increment in conditional expectation of M for a unit increase in X
-function get_delta(eta, a_x)
-    Q1 = exp(-eta)
-    Q2 = (1 - exp(-a_x))
-    Q3 = (1 + exp(-eta - a_x))
-    Q4 = (1 + exp(-eta))
-
-    return Q1 * Q2 / (Q3 * Q4)
-end
-
-## Increment in conditional expectation of Y for a unit increase in X
-## I.e. Mean over M of increment in conditional expectation for Y given M and X
-function get_gamma(delta, b_m, b_x)
-    return b_m * delta + b_x
-end
-
-
-# Compute SE of estimated mediation effect
-# We use the delta method, so many derivatives are required
-
-
-## Helpful intermedeate quantities
-function d_delta_d_eta(eta, a_x)
-    delta = get_delta(eta, a_x)
-
-    Q1 = 1 - exp(-2*eta - a_x)
-    Q2 = 1 + exp(-eta)
-    Q3 = 1 + exp(-eta - a_x)
-
-    return - delta * Q1 / (Q2 * Q3)
-end
-
-function d_delta_d_a_x(eta, a_x)
-    Q1 = exp(-2*eta - a_x)
-    Q2 = 1 + exp(-eta)
-    Q3 = 1 + exp(-eta - a_x)
-
-    return Q1 / (Q2 * Q3^2)
-end
-
-
-## Derivative of gamma wrt each parameter
-function d_gamma_d_a_0(dd_de, b_x)
-    return dd_de * b_x
-end
-
-function d_gamma_d_a_1(dd_de, dd_da_x, x, b_x)
-    return b_x * x * dd_de + b_x * dd_da_x
-end
-
-function d_gamma_d_A_2(dd_de, W, b_x)
-    return dd_de * b_x .* W
-end
-
-
-function d_gamma_d_b_0()
-    return 0
-end
-
-function d_gamma_d_b_1(eta)
-    return eta
-end
-
-function d_gamma_d_b_2()
-    return 1
-end
-
-function d_gamma_d_B_3(W)
-    return 0 .* W
-end
-
-
-
-## Gradient of gamma wrt all parameters
[…]

### 2024-05-06 src/Helpers.R

@@ -109,3 +109,146 @@ d_gamma_d_theta <- function(eta, x, W, a, b) {
 
   return(c(dg_da_0, dg_da_1, dg_dA_2, dg_db_0, dg_db_1, dg_db_2, dg_dB_3))
 }
+
+
+
+
+
+
+
+######################################################
+#### Binary Response, Binary Mediator, Fixed-Effs ####
+######################################################
+
+
+# Compute mediation effect
+
+## Probability of Y=1 given inputs
+PY1 <- function(eta, zeta, beta) {
+  Q1 <- 1 + exp(-zeta)
+  Q2 <- 1 + exp(-eta)
+  Q3 <- 1 + exp(-zeta - beta)
+
+  num <- Q1 + (Q2 - 1) * Q3
+  den <- Q1 * Q2 * Q3
+
+  return(num / den)
+}
+
+## Probability of Y=0 given inputs
+PY0 <- function(eta, zeta, beta) {
+  Q1 <- exp(-zeta)
+  Q2 <- exp(-eta)
+  Q3 <- exp(-beta)
+
+  num <- Q1 * (Q2 + Q3 + Q1 * Q3 + Q1 * Q2 * Q3)
+  den <- (1 + Q1) * (1 + Q2) * (1 + Q1 * Q3)
+
+  return(num / den)
+}
+
+
+get_odds_ref <- function(eta, zeta, beta) {
+  num <- PY1(eta, zeta, beta)
+  den <- PY0(eta, zeta, beta)
+
+  return(num / den)
+}
+
+get_odds <- function(eta, zeta, beta) {
+  Q1 <- exp(-zeta)
+  Q1_inv <- 1 / Q1
+  Q2 <- exp(-eta)
+  Q3 <- exp(-beta)
+
+  num <- 1 + Q1_inv + Q2 * (Q3 + Q1_inv)
+  den <- Q2 + Q3 + Q1 * Q3 + Q1 * Q2 * Q3
+
+  return(num / den)
+}
+
+get_odds_ratio <- function(eta, a_x, zeta, b_x, b_m) {
+  odds_1 <- get_odds(eta + a_x, zeta + b_x, b_m)
+  odds_2 <- get_odds(eta, zeta, b_m)
+  OR_hat <- odds_1 / odds_2
+  return(OR_hat)
[…]

### 2024-05-08 notes/William Analysis.Rmd
Fix SEs for bin-bib-fix

@@
 
-```{r, boilerplate, include=FALSE, cached=TRUE}
+```{r, boilerplate}
+# , cached=TRUE}
 # Note: This code chunk is not run here, but it is called and run at the start of every subsequent section. Any changes made here will be duplicated throughout the rest of the document.
@@
 
-```{r, cache=TRUE, dependson=c("boilerplate")}
+```{r MC-cont-cont-fix, include=FALSE, cache=TRUE}
+# , dependson=c("boilerplate")}
 
@@
 
-```{r, cache=TRUE, dependson=c("boilerplate")}
+```{r MC-cont-bin-fix, include=FALSE, cache=TRUE}
+# , dependson=c("boilerplate")}
 
@@
 
-```{r, cache=TRUE, dependson=c("boilerplate")}
+```{r MC-bin-bin-fix, include=FALSE, cache=TRUE}
+# , dependson=c("boilerplate")}
 
@@
     eta_hat = a_0_hat + a_x_hat*x_pred + W_pred%*%A_2_hat
-    zeta_hat = b_0_hat + b_x_hat * x_pred + W_pred %*% B_3_hat
+    zeta_hat = b_0_hat + b_x_hat*x_pred + W_pred %*% B_3_hat
 
@@
     ## See Helpers.R for the function get_odds_ratio
-    med_hat = get_odds_ratio(eta_hat, a_x, zeta_hat, b_x, b_m)
+    med_hat = get_odds_ratio(eta_hat, a_x_hat, zeta_hat, b_x_hat, b_m_hat)
     some_med_hats[[i]] = med_hat

### 2024-05-08 scripts/Bin-Resp, Bin-Med, Fixed.jl

@@ -47,6 +47,9 @@ W_pred = repeat([1], p_conf)    # W=[1,1,1]
 X_dist = Normal(1, 1)
 W_dist = Normal(1, 1)
 
+# X_dist = Normal(0, 1)
+# W_dist = Normal(0, 1)
+
 
 
 all_a_hats = []
@@ -160,6 +163,8 @@ all_med_SEs = []
 
     ### Derivatives of gamma wrt each parameter
     dOR_dt = d_OR_d_theta(eta_hat, a_x, zeta_hat, b_x, b_m, x_pred, W_pred)
+
+    norm(dOR_dt)
 
     ### Compute asymp var of gamma_hat
     asymp_var = dOR_dt' * reg_asymp_cov * dOR_dt

### 2024-05-08 notes/Literature Review/Literature Review.tex
Lit review: SEM, derivatives

@@
 
-\title{Verification of SE Formulas for Mediation Analysis}
+\title{Review of the Literature on Exact Formulas for Mediation Effects}
 
@@
 
-There is a short exchange in the American Journal of Epidemiology
+\section{Summary and Future Work}
+
+We have a nice expression for the total effect of $X$ on $Y$.
+It would be nice to also do direct and indirect effects.
+These are given by the so-called ``mediation formula'' \citep{Pea12}.
+There is the whole business of linking our analysis to causal inference, but I'm happy to leave that for later.
+For now, we can just use the expressions in terms of conditional expectations which result when the counterfactual-based definition is identifiable\footnote{\citet{Pea12} points out that, in the absence of random-effects, his mediation effects can be estimated directly from a contingency table, with no parametric model.
+If there are confounders (as in our setting), this approach may break down due to the large number of possible values.
+\textcolor{red}{Uncertainty quantification?
+Extension to multilevel models?}}.
+
+\textcolor{red}{Revisit after I review the relevant paper some more.} There is an approach to UQ for GLMMs based on automatic differentiation.
+See Section \ref{sec:GLMM_UQ}.
+I'm still working to understand \citet{Zhe21}, but it would be helpful if one of you could look over this paper.
+For our current purposes, we could just use the automatic differentiation-based asymptotic standard errors for the GLMM parameters.
+Later, when we want to do uncertainty quantification for the predicted random effects, I think that this paper will be useful.
 
@@
 However, I haven't seen any mention of random-effects/multilevel-models.
+
+These methods are implemented in the \texttt{R} package \texttt{ExactMed} \citep{Cau23}. This package does not address mixed-effects models.
 
@@
 Defines mediation effects in terms of derivatives.
-E.g.
-They define the indirect effect of $X$ on $Y$ as $(\partial \bE Y/ \partial \bE M) \cdot (\partial \bE M / \partial X)$, although that's not quite how they state it in the paper.
-This approach has had a small, but consistent following.
+Originally proposed by \citet{Yu14}, and later extended to multilevel models by \citet{Yu20}.
+They use the name ``third-variable effect analysis''.
+Documented extensively in a book by \citet{Yu22}.
+Implemented in the \texttt{R} packages \texttt{mma} and \texttt{mlma} for single level and multilevel models respectively.
+The latter uses the bootstrap for UQ.
 
@@
 
+\subsection{\citet{Gel18}}
+
+Defines mediation effects as partial derivatives.
+Does UQ by bootstrap or by a very rough method called MCCI.
+This MCCI simulates parameter values from a normal distribution parameterized by the estimates, then evaluates the mediation effects on these simulated parameters.
+This gives a distribution, but it seems to me that they would be better off just using the $\delta$-method.
+
+\subsection{\citet{Dor22}}
+
+Gives exact formulas for mediation effects and $\delta$-method SEs. Decomposes total mediation effect more finely than just direct and indirect (i.e. path-specific effects). No mixed-effects.
+
+\section{Structural Equation Modelling}
+
+There are a bunch of papers which use SEM to do mediation analysis.
+Many of them include multilevel/mixed-effects models.
+I'm not very familiar with SEM methodology, but it does seem to be connected with our work.
+See \citet{Val13} for a review and comparison with the causal-inference framework.
+Generally limited to linear models (i.e.
+continuous response and mediator), where the mediation effects' formulas are very simple.
+They seem to only focus on indirect effects\footnote{There is a short paragraph in \citet{Zit21} near the top of column 1 on page 532 which says ``the direct effect is only of little interest''.
+They don't give a reference for this.}.
+
+\subsection{\citet{Zig19}}
+
+Review of multilevel mediation analysis using SEM. Does a Monte Carlo study comparing different approaches. UQ is based on \citet{Bau06}.
+
+\subsection{\citet{Bau06}}
+
+Does mixed-effects mediation analysis based on SEMs.
+Only handles continuous outcome and mediator.
+Mediation effects are simple enough that variances are computed directly (in terms of asymptotic covariances of the regression parameter estimates).
+Fits the two regression models (for $M$ and $Y$) simultaneously by concatenating the response vectors and multiplying covariates by indicators.
+
+\subsection{\citet{Pre10}}
+
+Does mediation analysis using structural equation models. Does UQ by bootstrap or based on the exact distribution of the product of two dependent normals \citep{Mac07}.
+
 \section{GLMM UQ}
+\label{sec:GLMM_UQ}
 
@@
 Based on maximizing the Laplace Approximation to the marginal likelihood.
-Gives $\delta$-method SEs with gradients from automatic differentiation
+Gives $\delta$-method SEs with gradients from automatic differentiation.
 
-\section{Applications}
+Based on the \texttt{R} package \texttt{TMB}, which does automatic differentiation \citep{Kri16}.
+There is a related package, \texttt{glmmTMB}, which is specifically designed to apply automatic differentiation methods to GLMMs \citep{Bro17} (their paper focuses on zero-inflated count GLMMs, but their methodology applies more generally).
+\citet{Zhe21} raises some concerns about the two \texttt{TMB} packages, and presents a solution.
 
-\subsection{Not Relevant}
+\section{Others}
 
-\citet{Bro22} talks about mediation analysis and includes models with mixed-effects, but the latter are only used for imputation.
+\subsection{\citet{Ken03}}
+
+Gives a simplistic definition of mediation effects for continuous models.
+Does UQ based on exact distribution of mediation effects obtained as a transformation of Gaussians.
+A nice approach, but only really works because of the simple setting.
+
+There is a book by \citet{Mac17} on mediation analysis, which has a whole chapter called ``Multilevel Mediation Models''. However, it is mostly based on the \citet{Ken03} and small extensions.
 

### 2024-05-08 notes/Literature Review/Literature Review.tex
Expand literature review. Add summary and next steps

@@
 
-We have a nice expression for the total effect of $X$ on $Y$.
-It would be nice to also do direct and indirect effects.
-These are given by the so-called ``mediation formula'' \citep{Pea12}.
-There is the whole business of linking our analysis to causal inference, but I'm happy to leave that for later.
-For now, we can just use the expressions in terms of conditional expectations which result when the counterfactual-based definition is identifiable\footnote{\citet{Pea12} points out that, in the absence of random-effects, his mediation effects can be estimated directly from a contingency table, with no parametric model.
-If there are confounders (as in our setting), this approach may break down due to the large number of possible values.
-\textcolor{red}{Uncertainty quantification?
-Extension to multilevel models?}}.
+As far as I can tell, nobody has done exact asymptotics for mixed-effects mediation analysis.
+I do want to do some more searching, but the recent literature (e.g., since 2020) presents exact formulas for mediation effects with $\delta$-method uncertainty quantification as novel.
 
-\textcolor{red}{Revisit after I review the relevant paper some more.} There is an approach to UQ for GLMMs based on automatic differentiation.
-See Section \ref{sec:GLMM_UQ}.
-I'm still working to understand \citet{Zhe21}, but it would be helpful if one of you could look over this paper.
-For our current purposes, we could just use the automatic differentiation-based asymptotic standard errors for the GLMM parameters.
-Later, when we want to do uncertainty quantification for the predicted random effects, I think that this paper will be useful.
+Below, I have summarized a bunch of work on mediation analysis, roughly divided into sections.
+Section \ref{sec:Sam} covers work by Samoilenko, Lefebvre and others.
+Their work looks pretty close to what we're doing.
+They give exact formulas and $\delta$-method UQ, but no random effects.
+Section \ref{sec:deriv} defines mediation effects in terms of partial derivatives of the mean of $Y$ given $X$.
+They have done a bit with random effects, but only do Bayesian or bootstrap-based uncertainty quantification.
+Section \ref{sec:SEM} uses structural equation modelling (SEM) to do mediation analysis.
+I'm not very familiar with this methodology, but I tried to summarize what I could.
+Some of these papers do include random effects, but only seem to do continuous response and mediator.
+They make heavy use of the simple mediation formulas in the fully continuous setting, so uncertainty quantification is pretty basic.
+Finally, Section \ref{sec:other} is an assortment of papers.
+Highlights include \citet{Che21}, which gives definitions of direct and indirect effects which we can use, and \citet{Zhe21}, which uses automatic differentiation to get the full Hessian for GLMMs.
+\citeauthor{Zhe21} also do UQ for the predicted random effects, although I'm still working on understanding exactly how they do this.
+
+\subsection{What to do Next}
+
+Next week, my plan is to continue the literature search, focusing on SEMs, the UQ methodology proposed by \citet{Zhe21}, and also looking more widely to see if I've missed any other lines of work.
+It would be helpful for me if you could look at \citet{Zhe21} to see if you think it has merit.
+I'm also pretty shaky on SEMs, so it would be nice if you could look at \citet{Zig19} to see how it connects with our methodology.
+
+In terms of advancing our own work, I have finished translating my analysis into \texttt{R}, up to the binary response and mediator.
+My next steps in this direction for next week are to work out as much as I can for the binary/binary setting with mixed effects.
+This is going to involve computing the gradient of the mediation effect, and looking for whether \texttt{lme4} gives standard errors for the random effects' covariance parameters.
+If not, then I will look around for other sources.
+
+Once the mixed-effects case is done, I would like to extend our analysis to the direct and indirect effects.
+Formulas for these are given in \citet{Sam18} or \citet{Che21}.
+We should also consider interaction terms between the exposure and mediator in our model for the outcome.
+I understand that this interaction term arises from causal inference-based concerns, but I'm still fuzzy on the details.
+\citet{Sam21} extend their previous work on fixed-effects models to include interactions.
+We also need to extend all of our methods to handle predicted group-specific effects, although this makes the uncertainty quantification more challenging, both conceptually and computationally.
+See, e.g., \citet{Boo98,Flo19,Zig19}.
 
 \section{The \citeauthor{Sam23} group}
+\label{sec:Sam}
 
@@
 \section{Derivative-Based}
+\label{sec:deriv}
 
@@
 \section{Structural Equation Modelling}
+\label{sec:SEM}
 
@@
 They don't give a reference for this.}.
+
+There does seem to be some work on applying mixed-effects methodology to SEMs. I'm not sure yet if this has made its way to the SEM-based mediation analysis literature. I'll look into this next week.
 
@@
 
-\section{GLMM UQ}
-\label{sec:GLMM_UQ}
+\section{Others}
+\label{sec:other}
+
+\subsection{\citet{Ken03}}
+
+Gives a simplistic definition of mediation effects for continuous models.
+Does UQ based on exact distribution of mediation effects obtained as a transformation of Gaussians.
+A nice approach, but only really works because of the simple setting.
+
+There is a book by \citet{Mac17} on mediation analysis, which has a whole chapter called ``Multilevel Mediation Models''. However, it is mostly based on the \citet{Ken03} and small extensions.
+
+\subsection{\citet{Rit04}}
+
+Points-out that, in mixed-effects models, the regression structure for $Y$ only holds conditional on the random effects.
+Thus, it's hard to give a marginal interpretation of the regression coefficients.
+They give some conditions under which the conditional mean structure is preserved (in a particular sense) after marginalizing away the random effects.
+See, e.g., \citet{Neu91} for more discussion of the problem being addressed.
+
+I think I'm missing something here, because it seems obvious to me that there will be a difference between the parameters in an assumed regression model for the marginal mean of $Y$ and the conditional mean of $Y$ given the random effects.
+Regardless, I don't think this one is particularly relevant to us.
+We have a conditional regression model and don't make any effort to model the marginal mean of $Y$ using regression.
+
+\subsection{\citet{Van10}}
+
+Standard reference for mediation analysis with binary outcome.
+Focuses on the causal inference aspects of the problem.
+Makes heavy use of the ``rare outcome'' assumption.
+Resulting model is sufficiently simple that UQ is easy.
+
+\subsection{\citet{Pea12}}
+
+Presents the so-called ``mediation formula''.
+A set of identities for the total, direct and indirect mediation effects based on counterfactuals, but expressed in terms of conditional expectations.
+Effects are on risk-difference scale.
+
+\subsection{\citet{Nev17}}
+
+Does inference for mediation effects defined using the ``difference method''.
+Briefly, fit models to predict $Y$ using $X$ with or without $M$, then check how much the coefficient on $X$ changes.
+They apply their analysis to continuous, binary (log- and logit-link), and survival data (Cox model).
+
+Inference is done with the $\delta$-method.
+They need the covariance between coefficient estimates from different models (i.e.
+with and without $M$).
+Since this doesn't come directly from a standard analysis, they propose a ``data duplication algorithm''.
+
+No mixed effects. Not obviously relevant.
+
+\subsection{\citet{Che21}}
+
+Defines mediation effects (indirect and direct) as differences between transformed conditional expectations.
+This facilitates use with GLM link functions.
+Does continuous and binary outcome and mediator using GEEs.
+Does UQ with $\delta$-method.
+
+Includes a Monte Carlo study comparing $\delta$-method with bootstrap.
+Finds that pretty large samples are required for $\delta$-method UQ to be accurate, although bootstrap works with somewhat smaller samples.
+
+No mixed-effects.
+
+Does $\delta$-method UQ for GLMs (more precisely, GEEs) with exact formulas for mediation effects.
+These mediation effects (indirect and direct effects) are defined as differences between transformed conditional expectations.
 
@@
 \citet{Zhe21} raises some concerns about the two \texttt{TMB} packages, and presents a solution.
-
-\section{Others}
-
-\subsection{\citet{Ken03}}
-
-Gives a simplistic definition of mediation effects for continuous models.
-Does UQ based on exact distribution of mediation effects obtained as a transformation of Gaussians.
-A nice approach, but only really works because of the simple setting.
-
-There is a book by \citet{Mac17} on mediation analysis, which has a whole chapter called ``Multilevel Mediation Models''. However, it is mostly based on the \citet{Ken03} and small extensions.
+See \citet{Ska06} for more discussion of automatic differentiation in GLMM fitting.
 

### 2024-05-16 notes/Literature Review/Literature Review.tex
Expand lit review. Add Imai group and assorted miscellany

@@
 \bibliographystyle{plainnat}
+
+\usepackage{hyperref}   % For the \url{} command
 
@@
 \section{Summary and Future Work}
+
+\subsection{May 16}
+
+This week, I added work by Imai and colleagues, as well as a few general references.
+See Section \ref{sec:Imai} for the Imai group.
+There is also a method proposed by \citet{Lan12} based on inverse probability weighting to estimate unobserved counterfactuals.
+Both of these methods are included in the extensive Monte Carlo study of \citet{Sam23} and found to be comparable to inference for mediation effects based on the $\delta$-method.
+However, the method of \citet{Ima10I} is computationally intensive, and that of \citet{Lan12} is quite different from our approach of combining estimates from two regression models.
+
+I also found an \texttt{R} package, \texttt{merDeriv} \citep{Wan18} which computes the full asymptotic covariance matrix for a GLMM fit using \texttt{lme4}.
+This will save me a lot of time in implementing the $\delta$-method uncertainty quantification for mixed-effects models.
+
+\subsubsection{What to do next}
+
+I'm still not totally confident about the SEM literature, but otherwise I don't think anyone has done exact asymptotic theory for multilevel mediation analysis.
+I feel ready to start working on the mixed-effects version of our analysis.
+I will continue reading, but I think it makes sense to start spending some time writing code.
+
+There are some extensions to the methodology outlined by B \& B that I think we should consider.
+First, we will want to include direct and indirect effects in addition to our total effects.
+We can follow \citet{Sam23} for this, although we will also need to give some though to how we incorporate random effects into the definitions.
+It would also be nice to explicitly estimate the ``mediation proportion''.
+For continuous response, this is the indirect effect divided by the total effect.
+It's less clear to me what definition to use for multiplicative effects (e.g.
+effects on odds-ratio scale).
+
+Second, it is common to include an interaction effect between the exposure and mediator in our model for the outcome\footnote{There is no interaction effect in our trust study because a preliminary analysis found that it didn't have a significant effect on the outcome.}.
+This isn't conceptually any harder, it just adds another component to our formulas.
+
+I think that we could address just the above two points and, with some nice Monte Carlo studies, call it a paper.
+This would resemble the work by \citeauthor{Sam23} (e.g.
+\citealp{Sam23}).
+After that, there are some more dramatic extensions we can consider.
+
+One extension that we really should look at is predicted cluster-specific effects.
+The uncertainty quantification here is subtle; see, e.g., \citet{Skr09,Boo98}.
+The Imai group does give cluster-specific mediation effects, but only under their restricted framework.
+Otherwise, this seems to still be an open problem
+
+Finally (for now), I think we should eventually look at sensitivity analysis.
+There are a bunch of assumptions required for us to give a causal interpretation to our results.
+My understanding is that sensitivity analysis lets us put a bound on the extent to which these assumptions can be violated without invalidating our conclusions.
+I'm still looking into how this works.
+
+\subsection{May 2}
 
@@
 
-\subsection{What to do Next}
+\subsubsection{What to do Next}
 
@@
 
-\section{The \citeauthor{Sam23} group}
+\section{The Samoilenko and Lefebvre group}
 \label{sec:Sam}
@@
 As \citet{Sam18}, but with continuous mediator and binary response. No mixed-effects. Very nice simulation study.
+
+Compares their approach to those of the Imai group (see Section \ref{sec:Imai}), as well as \citet{Lan12}.
+Found that the $\delta$-method based approach proposed by the \citeauthor{Sam23} group gives comparable performance with its computationally intensive alternatives.
 
@@
 This MCCI simulates parameter values from a normal distribution parameterized by the estimates, then evaluates the mediation effects on these simulated parameters.
-This gives a distribution, but it seems to me that they would be better off just using the $\delta$-method.
+This is the same UQ method used by the Imai group (see Section \ref{sec:Imai}).
 
@@
 
-There does seem to be some work on applying mixed-effects methodology to SEMs. I'm not sure yet if this has made its way to the SEM-based mediation analysis literature. I'll look into this next week.
+There does seem to be some work on applying mixed-effects methodology to SEMs. See, e.g., \citet{Rab04} for early work in this direction.
+
+I'm not sure yet if this has made its way to the SEM-based mediation analysis literature. I'll look into this next week.
 
@@
 
+\section{The Imai Group}
+\label{sec:Imai}
+
+This group focuses mainly on the assumptions required to link regression-based mediation analysis with causal inference.
+They give definitions of mediation effects which closely resemble the ``mediation formula'' \citep{Pea12}, but only for differences of conditional expectation (i.e.
+the ``risk-difference scale''; see also \citealp{Sam23}).
+Their approach to UQ is very computational.
+They recommend either the bootstrap or a simulation-based approach outlined in \citet{Kin00}.
+Briefly, simulate parameters from their fitted sampling distributions, then evaluate the quantity of interest on each sample and observe the resulting distribution.
+I think this is connected to empirical Bayes methodology, but I haven't looked into it extensively.
+
+Their methods are implemented in the \texttt{R} package \texttt{mediation} \citep{Tin14}. They can handle mixed-effects models, including prediction of group-level effects.
+
+\subsection{\citet{Ima10I}}
+
+This paper presents an UQ algorithm for estimating mediation effects defined as differences of expected values.
+The algorithm closely resembles the bootstrap, but is slightly different.
+Specifically, parameters are repeatedly sampled from the fitted sampling distributions of their estimators, then mediation effects are estimated using the sampled parameter estimates \citep[see, e.g.,][]{Kin00}.
+They also give an alternative method which just uses the bootstrap.
+
+Refers to the causal inference theory from \citet{Ima10II} to justify identification of counterfactuals with functions of conditional expectations.
+
+\subsection{\citet{Ima10II}}
+
+Focuses on estimating the indirect effect for a continuous response.
+Defines this effect as a difference of counterfactuals and identifies this difference with a conditional expectation in the same manner as the mediation formula given by \citet{Pea12}.
+
+Also gives a ``non-parametric estimator'' for categorical mediator based on a contingency table based approach.
+The idea is to partition the data based on the mediator, estimate the mediation effect within each group with a difference of averages, then re-combine by averaging.
+
+UQ is given either by the $\delta$-method or by the bootstrap. Some discussion of sensitivity analysis.
+
+\subsection{\citet{Ima11}}
+
+Very similar to \citet{Ima10I} and \citet{Ima10II}. Focuses more on causal inference interpretation and applications to political science. Doesn't appear to advance the methodology.
+
+\subsection{\citet{Tin14}}
+
+This paper overviews the \texttt{mediation} package, which implements the methods developed by this group.
+Can handle GLMMs, but only operates on differences in the expected value of the outcome (i.e.
+risk-difference scale for binary outcome).
+Does UQ for parameters \textbf{and predicted random effects}, but doesn't go into detail on how.
+I'm guessing that it just follows the Monte Carlo method given by \citet{Ima10I}.
+
+Package last updated 2019-10-08
+
+\section{The \texttt{SPSS} Macro \texttt{MLmed}}
+\label{sec:MLmed}
+
+This macro appears to be the standard way to do multilevel mediation analysis in \texttt{SPSS}.
+As such, it is likely to be used extensively by applied researchers, and we should understand it.
+The macro can only handle continuous outcomes and mediators.
+
+\subsection{\citet{Roc19}}
+
+User manual for the \texttt{MLmed} macro. Can only handle continuous outcomes and mediators. UQ is either by the bootstrap or the ``Monte Carlo method''. See \citet{Pre12} or \citet{Wil08}.
+
+\subsection{\citet{Hay20}}
+
+A recommended reference for the methodology used in the macro.
+
+\subsection{\citet{Roc17}}
+
+A conference poster presenting the macro. Popular reference for people looking to cite the macro.
+
+\section{Applications}
+
+It doesn't hurt for us to point to what people are doing in the applied literature. Here is some of what shows up when I search ``multilevel mediation analysis'' on Google Scholar.
+
+\subsection{\citet{Vea19}}
+
+Does multilevel mediation analysis with continuous outcome and mediator.
+UQ is done with ``Monte Carlo confidence intervals''.
+Uses an \texttt{SPSS} macro called \texttt{MLmed}.
+See Section \ref{sec:MLmed}.
+
+\subsection{\citet{Ara21}}
+
+It looks like their variables are all ordinal (1-6 or 1-8 Likert scale).
+I'm not sure how they fit this into the mediation analysis framework.
+Moving on, they use \texttt{Mplus} to do their analysis via ``maximum likelihood''.
+Not much detail given.
+
+\subsection{\citet{Hwa12}}
+
+Follows an earlier edition of \citet{Mac17}, as well as the very early work of \citet{Bar86}.
+As in \citet{Ara21}, variables are ordinal/Likert.
+Analysis is done in \texttt{SPSS}, although few details are given.
+Note that this work was published before the \texttt{SPSS} macro of \citet{Roc19}.
+
 \section{Others}
 \label{sec:other}
+
+\subsection{\citet{Kru01}}
+
+Throughout some parts of the literature on multilevel mediation analysis, you will see people talk about, e.g., the 2-1-1 model.
+This means that the exposure is measured at ``level 2'' (group-level), while the mediator and outcome are measured at ``level 1'' (individual-level).
+That is, in this model we would include a random effect for the mediator, but not the exposure.
+Other examples of such models include 2-2-1 and 1-1-1.
+As far as I can tell, this notation was proposed by \citet{Kru01}.
+
+This paper also does $\delta$-method uncertainty quantification for the very simple mediation effects that people were using at that time (i.e.
+a product of coefficients).
+In fact, this approach for testing the presence/absence of mediation effects dates back to \citet{Sob82}.
 
@@
 Effects are on risk-difference scale.
+
+\subsection{\citet{Lan12}}
+
+Proposes estimating mediation effects using inverse probability weighting.
+Duplicates the observed data and uses weighting to estimate unobserved counterfactual outcomes.
+Referred to as the ``weighting method''.
+Does give analytical (asymptotic) standard errors based on the sandwich variance estimator that arises from the theory of generalized estimating equations.
+See their ``Web Appendix 3'' for more details.
+They also suggest using the bootstrap.
+
+Implemented in the \texttt{R} package \texttt{medflex} \citep{Ste17}.
 
@@
 
+\subsection{\citet{Wan18}}
+
+Standard reference for the \texttt{merDeriv} package in \texttt{R}.
+This is a small package that computes the Hessian of the observed data likelihood with respect to all parameters (including RE covariance parameters) for LMMs and GLMMs fit using \texttt{lme4}.
+In particular, use the \texttt{vcov} function with \texttt{full=TRUE} to get the asymptotic covariance matrix of the parameters (i.e.
+the inverse of the negative Hessian of the log-likelihood).
+You can also choose your parameterization for the RE parameters from variances, SDs or terms in the Cholesky decomposition.
+The last one is referred to as \texttt{theta} in \texttt{lme4} and is used extensively therein.
+
 \bibliography{Lit_Review_Bib}

### 2024-05-17 notes/UQ Verification.tex
Gradient of Phi function for bin-bin-mixed case in R

@@
 
-\subsection{UQ for GLMM Parameters}
+The approximate covariance matrix of the GLMM parameter estimates can be obtained directly from a fitted \texttt{lme4} object using the \texttt{merDeriv} package in \texttt{R}.
+This saves all the work I was starting in the following subsection.
+Please disregard.
+
+\subsection{UQ for GLMM Parameters (Old)}
 
@@
 There are tricks from the EM algorithm literature for doing this, but I think our problem is simple enough that I can just reduce it to a small-ish number of one-dimensional integrals.
-To start, recall the formula Bruno gives for $\bP(Y=1 | X=x, \mathbf{W}=\mathbf{w}) =: \bP_1$.
+To start, recall the formula Bruno gives for $\bP(Y=1 | X=x, \mathbf{W}=\mathbf{w}) =: P_1$ (similar for $P_0$).
 Specifically, he expresses this quantity in terms of the $\phi$ function evaluated at four different arguments.
-Thus, if I have the Hessian of $\phi$, I can compute the Hessian of $\bP_1$ (and $\bP_0$), which in-turn gives me the Hessian of the log-likelihood based on a single point.
+Thus, if I have the Hessian of $\phi$, I can compute the Hessian of $P_1$ (and $P_0$), which in-turn gives me the Hessian of the log-likelihood based on a single point.
 
@@
 
+\subsection{UQ for Mediation Effects}
+
+Following Section 2.2 of B \& B's document, we have an expression for the mediation effect, $\Psi$, in terms of a small number of integrals, $\phi$.
+Each of these integrals is univariate, and can thus be computed accurately using quadrature.
+As for uncertainty quantification, we still need to compute the partial derivative of $\Psi$ wrt each of its arguments.
+This shouldn't be too hard to do in Maple (hopefully).
+I will probably need to leave $\phi$ undefined, and get Maple to return an answer in terms of the partial derivatives of $\phi$.
+
+Note that $\Psi$ depends explicitly on a combination of model parameters and transformations thereof.
+In particular, the functions $\sigma_U$ and $\sigma_V$ depend implicitly on the mixed effects' covariance matrices.
+Conveniently, B \& B have worked out the gradients of $\sigma_U$ and $\sigma_V$ wrt the model parameters.
+They're messy, but ultimately code-able.
+
+\textcolor{red}{As of Friday, May 17, I have coded the gradient of $\Phi$. Next I need to do UQ for each argument of $\Phi$ (e.g., $\sigma_V(x+1)$), then I can apply the $\delta$-method.}
+
 \section{Open Questions}

### 2024-05-17 notes/William Analysis.Rmd
Gradient of Phi function for bin-bin-mixed case in R

@@
 The first step is to set the sample size and generate some data.
-We use $\eta$ to denote the linear predictor for $M$, and $\zeta$ for the linear predictor of $Y$ without the $M$ term.
+We use $\eta$ to denote the linear predictor for $M$, and $\zeta$ for the linear predictor of $Y$.
+<!-- (This is a comment) without the $M$ term.
 That is, $\zeta = b_0 + b_2 X + W B_3$.
+-->
 
@@
 ```
+
+# Binary Response, Binary Mediator, Mixed-Effects
+
+We now introduce random effects to our model.
+This changes many of the details, but the overall structure of the analysis remains the same.
+I still need to write some helper functions to compute relevant derivatives, all of which validated against a simple finite-difference approximation.
+See `Helpers.R`.
+
+```{r}
+source("../src/Helpers.R")
+
+library(lme4)     # For fitting GLMMs using glmer()
+library(MASS)     # For simulating multivariate normals using mvrnorm()
+library(merDeriv) # For computing the full information matrix from models fit using lme4
+```
+
+## Generate Data
+
+As in the previous case, we set intercepts so that the means of both linear predictors are at least approximately zero.
+We also need to set covariance parameters for the random effects.
+For this, we use either $20\%$ of the true value or $0.2$ if the true value is zero.
+Correlations between all random effects are set to $0.5$.
+Note that we include random effects for the intercept and the exposure, but not for the mediator.
+
+Another difference from the previous settings is that we must now construct our dataset in clusters.
+We will use $K=5$ such clusters, with $n$ observations in each cluster.
+We will set $n=100$ for the first iteration of the loop.
+
+```{r setup_mixed}
+# Fixed effects for intercepts
+a_0 = 0
+b_0 = -0.5
+
+# Random effects
+sigma_a_0 = 0.2
+sigma_a_1 = 0.2 * abs(a_1)
+cor_a0_a1 = 0.2
+cov_a0_a1 = sigma_a_0 * sigma_a_1 * cor_a0_a1
+Sigma_a = matrix(c(sigma_a_0^2, cov_a0_a1, cov_a0_a1, sigma_a_1^2), nrow = 2, ncol = 2)
+
+sigma_b_0 = 0.2 * abs(b_0)
+sigma_b_1 = 0.2 * abs(b_1)
+cor_b0_b1 = 0.2
+cov_b0_b1 = sigma_b_0 * sigma_b_1 * cor_b0_b1
+Sigma_b = matrix(c(sigma_b_0^2, cov_b0_b1, cov_b0_b1, sigma_b_1^2), nrow = 2, ncol = 2)
+
+# Number of clusters
+K=5
+```
+
+We first demonstrate our analysis on a single dataset, then put this analysis inside a `for` loop.
+The first step is to set the sample size and generate some data.
+We use $\eta$ to denote the linear predictor for $M$, and $\zeta$ for the linear predictor of $Y$.
+<!-- (This is also a comment) without the $M$ term.
+That is, $\zeta = b_0 + b_2 X + W B_3$.
+-->
+
+```{r generate_data_mixed}
+n = all_Ns[2]
+# n = all_Ns[1]
+
+all_Xs = list()
+all_Ws = list()
+
+for(k in 1:K){
+  X = rnorm(n, mean=0, sd=1)
+  W = matrix(rnorm(n*p_conf, mean=0, sd=1), nrow = n, ncol = p_conf)
+
+  all_Xs[[k]] = X
+  all_Ws[[k]] = W
+}
+
+# Generate M
+all_Ms = list()
+for(k in 1:K){
+  eta_vec_fixed = a_0 + a_1*all_Xs[[k]] + all_Ws[[k]]%*%A_2
+
+  ## Add random effects
+  a_ran = mvrnorm(1, mu = rep(0, 2), Sigma = Sigma_a)
+  eta_vec = eta_vec_fixed + a_ran[1] + a_ran[2]*all_Xs[[k]]
+
+  ## Generate M
+  p_M_vec = expit(eta_vec)
+  M = rbinom(n, size = 1, prob = p_M_vec)
+  all_Ms[[k]] = M
+}
+
+# Generate Y
+all_Ys = list()
+for(k in 1:K){
+  zeta_vec_fixed = b_0 + b_1*all_Ms[[k]] + b_2 * all_Xs[[k]] + all_Ws[[k]]%*%B_3
+
+  ## Add random effects
+  b_ran = mvrnorm(1, mu = rep(0, 2), Sigma = Sigma_b)
+  zeta_vec = zeta_vec_fixed + b_ran[1] + b_ran[2]*all_Xs[[k]]
+
+  ## Generate Y
+  p_Y_vec = expit(zeta_vec)
+  Y = rbinom(n, size = 1, prob = p_Y_vec)
+  all_Ys[[k]] = Y
+}
+
+# Consolidate groups
+X = do.call(c, all_Xs)
+W = do.call(rbind, all_Ws)
+M = do.call(c, all_Ms)
+Y = do.call(c, all_Ys)
+group = rep(1:K, each = n)
+```
+
+## Estimate Mediation Effect
+
+We now fit logistic regression models to predict $M$ and $Y$, and extract output that we will need later.
+
+```{r fit_models_mixed}
+M_data = data.frame(M, X, W1 = W[, 1], W2 = W[, 2], W3 = W[,3], group = group)
+M_model = glmer(M ~ X + W1 + W2 + W3 + (1 + X | group), data = M_data, family = binomial(link = "logit"))
+M_model_info = attributes(VarCorr(M_model)$group)
+
+a_hat = fixef(M_model)
+a_RE_sds = M_model_info$stddev
+a_RE_cor = M_model_info$correlation[2,1]
+a_cov = vcov(M_model, full=TRUE, ranpar="sd")
+
+# a_hat = summary(M_model)$coefficients[, 1]
+# a_SE = summary(M_model)$coefficients[, 2]
+# a_cov = vcov(M_model)
+
+Y_data = data.frame(Y, M, X, W1 = W[, 1], W2 = W[, 2], W3 = W[,3], group = group)
+Y_model = glmer(Y ~ M + X + W1 + W2 + W3 + (1 + X | group), data = Y_data, family = binomial(link = "logit"))
+Y_model_info = attributes(VarCorr(Y_model)$group)
+
+b_hat = fixef(Y_model)
+b_RE_sds = Y_model_info$stddev
+b_RE_cor = Y_model_info$correlation[2,1]
+b_cov = vcov(Y_model, full=TRUE, ranpar="sd")
+
+# b_hat = summary(Y_model)$coefficients[, 1]
+# b_SE = summary(Y_model)$coefficients[, 2]
+# b_cov = vcov(Y_model)
+```
+
+Next, we estimate the mediation effect.
+Note that, as in the contiuous outcome, binary mediator setting, our mediation effect depends on $X$ and $W$.
+We again use $X=0$ and $W=[1,1,1]$, although Samoilenko and Lefebvre (2023) use the sample means of these covariates.
+
+```{r}
+x_pred = 0
+W_pred = c(1,1,1)
+
+a_0_hat = a_hat[1]
+a_x_hat = a_hat[2]
+A_2_hat = a_hat[3:5]
+
+b_0_hat = b_hat[1]
+b_m_hat = b_hat[2]
+b_x_hat = b_hat[3]
+B_3_hat = b_hat[4:6]
+
+# Linear predictors
+eta_hat = a_0_hat + a_x_hat*x_pred + W_pred%*%A_2_hat
+zeta_hat = b_0_hat + b_x_hat * x_pred + W_pred %*% B_3_hat
+
+# Mediation effect
+## See Helpers.R for the function get_odds_ratio
+med_hat = get_odds_ratio(eta_hat, a_x, zeta_hat, b_x, b_m)
+```
+
+We now compute the gradient of the odds ratio and use the $\delta$-method to get the standard error.
+
+```{r}
+# Gradient of OR wrt regression coefficients
+grad_med = d_OR_d_theta(eta_hat, a_x_hat, zeta_hat, b_x_hat, b_m_hat, x_pred, W_pred)
+
+# Get asymptotic SE using delta method
+
+## Build joint covariance matrix of regression coefficients
+a_length = nrow(a_cov)
+b_length = nrow(b_cov)
+joint_cov = matrix(0, nrow = a_length + b_length, ncol = a_length +
+b_length)
+joint_cov[1:a_length, 1:a_length] = a_cov
+joint_cov[(a_length + 1):(a_length + b_length), (a_length + 1):(a_length +
+b_length)] = b_cov
+
+## Convert to asymptotic covariance matrix
+asymp_reg_cov = n * joint_cov
+
+## Pre- and post-multiply asymptotic covariance by gradient of transformation
+med_asymp_var = grad_med %*% asymp_reg_cov %*% grad_med
+
+## Get small-sample standard error
+med_asymp_SE = sqrt(med_asymp_var)
+med_SE = med_asymp_SE / sqrt(n)
+```
+
+Putting everything together, we have an estimated mediation effect of `r med_hat`, with an estimated SE of `r med_SE`.

### 2024-05-17 scripts/Bin-Resp, Bin-Med, Random.jl

@@ -101,7 +101,7 @@ all_med_SEs = []
     all_zeta_vecs_fixed = [b_0 .+ b_1 .* M .+ b_2 .* X .+ W * B_3 for (M, X, W) in zip(all_Ms, all_Xs, all_Ws)]
 
     ### Add random effects
-    all_zeta_vecs = [zeta_vec_fixed .+ rand(Normal(0, sigma_b_0), n) .+ rand(Normal(0, sigma_b_2), n) .* X for (zeta_vec_fixed, X) in zip(all_zeta_vecs_fixed, X)]
+    all_zeta_vecs = [zeta_vec_fixed .+ rand(Normal(0, sigma_b_0), n) .+ rand(Normal(0, sigma_b_2), n) .* X for (zeta_vec_fixed, X) in zip(all_zeta_vecs_fixed, all_Xs)]
     
     ### Get probabilities
     all_p_Y_vecs = [expit.(zeta_vec) for zeta_vec in all_zeta_vecs]
@@ -126,7 +126,7 @@ all_med_SEs = []
         M_data[!,Symbol("W$p")] = W[:, p]
     end
 
-    ### This code tries to automate construction of the formula. I ran into issues specifying the random effects
+    ### This commented code tries to automate construction of the formula. I ran into issues specifying the random effects
     # M_formula_fix = term(:M) ~ sum(term.(names(M_data, Not(:M, :group))))
     # RE_terms = term(1) + term(:X)
     # group_term = term(:group)
@@ -136,7 +136,7 @@ all_med_SEs = []
 
     M_model = fit(MixedModel, M_formula, M_data, Bernoulli(), verbose=false, progress=false)
 
-    M_model.InterceptTerm
+    # M_model.InterceptTerm
 
     a_hat = coef(M_model)
     a_SE = stderror(M_model)

### 2024-05-17 src/Helpers.R

@@ -252,3 +252,165 @@ d_OR_d_theta <- function(eta, a_x, zeta, b_x, b_m, x_ref, W_ref) {
            d_OR_d_B3(eta, a_x, zeta, b_x, b_m, W_ref)))
 }
 
+
+
+
+
+
+
+
+
+######################################################
+#### Binary Response, Binary Mediator, Mixed-Effs ####
+######################################################
+
+
+phi = function(mu, sigma){
+  integrand = function(x){
+    A = 1 + exp(-mu - sigma*x)
+    B = dnorm(x)
+
+    return(B/A)
+  }
+
+  integral = integrate(integrand, -Inf, Inf)
+  # print(integral$message)
+  # print(integral$abs.error)
+
+  return(integral$value)
+}
+
+# Partial derivatives of phi
+# I validated these against a simple finite difference approximation
+
+## d phi / d mu
+d1_phi = function(mu, sigma){
+  integrand = function(x){
+    A = 1 + exp(-mu - sigma*x)
+    B = dnorm(x)
+    C = exp(-mu - sigma*x)
+
+    return(C*B/A^2)
+  }
+
+  return(integrate(integrand, -Inf, Inf)$value)
+}
+
+## d phi / d sigma
+d2_phi = function(mu, sigma){
+  integrand = function(x){
+    A = 1 + exp(-mu - sigma*x)
+    B = dnorm(x)
+    C = x * exp(-mu - sigma*x)
+
+    return(C*B/A^2)
+  }
+
+  return(integrate(integrand, -Inf, Inf)$value)
+}
+
+
+
+# Total mediation effect (on odds-ratio scale)
+Phi = function(a, b, a1, b1, b2, s1, s2, s3, s4){
[…]

### 2024-05-20 notes/Literature Review/Literature Review.tex
Add summary and future steps to lit review
Document Imai group UQ for predicted group-wise effects

@@
 
-\subsection{May 2}
+\subsection{May 9}
 
@@
 I think this is connected to empirical Bayes methodology, but I haven't looked into it extensively.
+They use the name ``quasi-Bayesian confidence intervals'' in their \texttt{R} package.
 
@@
 Package last updated 2019-10-08
+
+\subsubsection{UQ for Group-Level Effects}
+
+The \citet{Tin14} paper discusses inference for predicted group-level effects, but doesn't go into detail.
+As far as I can tell, this is also the only paper from the Imai group which discusses multilevel mediation analysis.
+Worse still, they give very few references.
+
+After digging into the source code for the \texttt{mediation} package, I found what they do.
+First, they get predicted group-wise random effects from \texttt{lme4}.
+Second, it turns out that \texttt{lme4} can give a conditional covariance matrix for the random effects ``conditional on the estimates of the model parameters and on the data''.
+For now, just note that the full covariance matrix of the random effects is block diagonal, with one block per group.
+The \texttt{mediation} package uses these group-wise conditional covariance matrices, together with the predicted group-wise means, and simulates random effects for each group (independently, from normal distributions with the corresponding moments).
+The package then proceeds with its ``quasi-Bayesian'' inference.
+
+Just a brief note on the conditional covariances given by \texttt{lme4}.
+Initially, when I was reading this portion of the documentation, I got excited that we might not need the \texttt{merDeriv} package to get standard errors for the random effects' covariance parameters.
+That's definitely not what I found.
+These are standard errors for the predicted random effects (conditional on the observed data), not standard errors for the estimators of any model parameters.
+In particular, the former can be thought of as $K$ separate covariance matrices, one per group, whereas the latter is a single matrix representing estimation uncertainty based on the entire dataset.
+
+Unfortunately, no references are given in the \texttt{lme4} documentation, but it is fairly clear what is being done.
+The predicted random effect within a group is the mode of the conditional distribution of the random effects given the observed data from that group (alternatively, the mean; although \texttt{lme4} uses the mode).
+The prediction standard error given by \texttt{lme4} is the covariance of this same conditional distribution.
 
@@
 Note that this work was published before the \texttt{SPSS} macro of \citet{Roc19}.
+
+\subsection{\citet{Rij21}}
+
+Review of applied papers using mediation analysis.
 

### 2024-05-20 scripts/Imai Investigation.R

@@ -1,0 +1,29 @@
+
+library(mediation)
+library(lme4)
+library(dplyr)
+
+set.seed(1)
+
+data("student")
+
+student
+student_small = filter(student, SCH_ID < 100)
+
+med.fit <- glmer(attachment ~ catholic + gender + income + pared + (1 + catholic | SCH_ID), family = binomial(link = "logit"), data = student_small)
+
+out.fit <- glmer(fight ~ catholic * attachment + gender + income + pared + (1 + attachment | SCH_ID), family = binomial(link = "logit"), data = student_small)
+
+
+med.out <- mediate(med.fit, out.fit, treat = "catholic", mediator = "attachment", sims = 500)
+summary(med.out)
+summary(med.out, output="bygroup")
+
+
+
+ranef(med.fit, condVar = TRUE)
+ranef(med.fit, condVar = FALSE)
+
+attr(ranef(med.fit, condVar = TRUE)[[1]], "postVar")
+
+

### 2024-05-20 notes/Literature Review/Literature Review.tex
Add imputation method and some general notes

@@
 \section{Summary and Future Work}
+
+\subsection{May 23}
+
+A thought on how to frame the definition of mediation effects on different scales: the mediation formula of \citet{Pea12} identifies causal direct/indirect effects with specific functions of conditional expectations of the outcome and mediator.
+This makes it natural to work on the risk-difference scale, as done by the Imai group (see Section \ref{sec:Imai}).
+However, if we prefer the risk-ratio or odds-ratio, we can just express our target quantities in terms of the functions identified by \citet{Pea12}.
+There may or may not be simplification that leads to interpretable formulas, but that's not the point.
+You can define mediation effects on whatever scale you like, then we will give you estimates and standard errors, irrespective of how messy the formulas may be (they will all be hidden in the code anyway).
 
@@
 
-Review of applied papers using mediation analysis.
+Review of applied papers using mediation analysis. \textcolor{red}{Read and document this!}
 
@@
 
-Implemented in the \texttt{R} package \texttt{medflex} \citep{Ste17}.
+Implemented in the \texttt{R} package \texttt{medflex} \citep{Ste17} alongside the imputation method of \citet{Van12}.
+
+\subsection{\citet{Van12}}
+
+Proposes estimating mediation effects using regression-based imputation.
+Recall that the indirect and direct mediation effects depend on unobservable nested counterfactuals of the form $\phi(x, x') := Y(x, M(x'))$, where $x \neq x'$.
+\citet{Van12} proposes to estimate these counterfactuals directly by first fitting a regression model for the dependence of $\phi$ on $x$ and $x'$ using configurations of these covariates (i.e.
+the exposure) observed in the data.
+We then use this regression model to predict the value of $\phi$ at unobserved configurations.
+
+Caution: This whole method is based on extrapolating from a regression model into a region where we, by definition, have no information.
+The usual warnings apply, both philosophical and practical (e.g.
+don't include high-order polynomial terms in your regression model).
+
+Implemented in the \texttt{R} package \texttt{medflex} \citep{Ste17} alongside the weighting method of \citet{Lan12}.
 

### 2024-05-24 notes/Literature_Review/Literature Review.tex
Re-organize directory

--- 2024-05-20 notes/Literature Review/Literature Review.tex
@@
 
+I'm just finishing-up my code for the binary response, binary mediator, mixed effects case. My analysis is running, but the output doesn't look right so I need to do some debugging.
+
+I now better understand the uncertainty quantification method used by the Imai group.
+See Section \ref{sec:Imai_UQ} for details.
+Briefly, they do a Monte Carlo analog of the $\delta$-method.
+If the target of inference is a function of some model parameters, they start by estimating the sampling distribution of our estimator for those model parameters.
+They then simulate from this estimated sampling distribution and evaluate the function of interest on each simulated observation.
+This produces a sample from the pushforward measure induced on the target of inference by our estimated sampling distribution for the model parameters.
+It's not completely clear to me how they generate confidence intervals from the Monte Carlo samples, but they call their method ``quasi-Bayesian'', so I suspect it's just percentile intervals.
+
+Incidentally, if we could do exact (i.e.
+non-asymptotic) distribution theory for the model parameter (e.g.
+ordinary linear regression), then Imai's method is equivalent to the parametric bootstrap.
+They just skip the generation of datasets and work directly with the bootstrap distribution of the estimator.
+
 A thought on how to frame the definition of mediation effects on different scales: the mediation formula of \citet{Pea12} identifies causal direct/indirect effects with specific functions of conditional expectations of the outcome and mediator.
@@
 You can define mediation effects on whatever scale you like, then we will give you estimates and standard errors, irrespective of how messy the formulas may be (they will all be hidden in the code anyway).
+
+\subsubsection{What to do Next}
+
+I think that at this week's meeting, we should plan out what we want to put in the paper.
+I propose that we do total, direct and indirect effects on risk difference, risk ratio and odds ratio scales, but stick to fixed-effects (i.e.
+avoid predicting group-level effects).
+Based on other work in the literature.
+I think that's enough for a paper, provided that we do a nice Monte Carlo study comparing our UQ with other methods, such as Imai's quasi-Bayesian method, the weighting method of \citet{Lan12} and the imputation method of \citet{Van12}.
+All of these methods are implemented to varying extents in \texttt{R} packages, \texttt{mediation} for the former and \texttt{medflex} for the latter two.
+
+Prediction of group-level effects barely comes up in the literature.
+The \texttt{mediation} package from Imai's group does it, but with very little discussion.
+They give a few sentences in the package vignette saying they can do group-level effect prediction, but don't give any further discussion or references.
+I had to read their source code to know what method they're using.
+Prediction of group-level mediation effects seems like an untapped area that we are well-positioned to develop.
 
@@
 \subsubsection{UQ for Group-Level Effects}
+\ref{sec:Imai_UQ}
 

### 2024-05-24 notes/William Analysis.Rmd
Run analysis for bin-bin-mix case

@@
 
-## Monte Carlo Study
-
-We now repeat the above analysis `r num_reps` times for each of various values of $n$.
-Values of $X$ and $W$ for which we compute the total effect are set at the beginning.
-We also re-run the boilerplate code (not shown).
-
-```{r, include=FALSE}
-<<boilerplate>>
-```
-
-```{r}
-# Values of X and W for which we compute the total effect
-x_pred = 0
-W_pred = c(1,1,1)
-
-# Values of a_0 and b_0
-a_0 = 0
-b_0 = -0.5
-```
-
-```{r MC-bin-bin-fix, include=FALSE, cache=TRUE}
-# , dependson=c("boilerplate")}
-
-for(j in seq_along(all_Ns)){
-  n = all_Ns[j]
-
-  # Containers for output with this value of n
-  some_a_hats = list()
-  some_b_hats = list()
-
-  some_a_SEs = list()
-  some_b_SEs = list()
-
-  some_med_hats = list()
-  some_med_SEs = list()
-
-  for(i in 1:num_reps){
-    # Generate data
-    X = rnorm(n, mean=0, sd=1)
-    W = matrix(rnorm(n*p_conf, mean=0, sd=1), nrow = n, ncol = p_conf)
-
-    ## Generate M
-    eta_vec = a_0 + a_1*X + W%*%A_2
-    p_M_vec = expit(eta_vec)
-    M = rbinom(n, size = 1, prob = p_M_vec)
-
-    ## Generate Y
-    zeta_vec = b_0 + b_1*M + b_2*X + W%*%B_3
-    p_Y_vec = expit(zeta_vec)
-    Y = rbinom(n, size = 1, prob = p_Y_vec)
-
-    # Fit models
-
-    ## M
-    M_data = data.frame(M, X, W1 = W[,1], W2 = W[,2], W3 = W[,3])
-    M_model = glm(M ~ X + W1 + W2 + W3, data = M_data, family = binomial(link = "logit"))
-
-    a_hat = summary(M_model)$coefficients[,1]
-    a_SE = summary(M_model)$coefficients[,2]
-    a_cov = vcov(M_model)
-
-    some_a_hats[[i]] = a_hat
-    some_a_SEs[[i]] = a_SE
-
-    ## Y
-    Y_data = data.frame(Y, M, X, W1 = W[,1], W2 = W[,2], W3 = W[,3])
-    Y_model = glm(Y ~ M + X + W1 + W2 + W3, data = Y_data, family = binomial(link = "logit"))
-
-    b_hat = summary(Y_model)$coefficients[,1]
-    b_SE = summary(Y_model)$coefficients[,2]
-    b_cov = vcov(Y_model)
-
-    some_b_hats[[i]] = b_hat
-    some_b_SEs[[i]] = b_SE
-
-    # Estimate mediation effect
-
-    ## Extract coefficients
-    a_0_hat = a_hat[1]
-    a_x_hat = a_hat[2]
-    A_2_hat = a_hat[3:length(a_hat)]
-
-    b_0_hat = b_hat[1]
-    b_m_hat = b_hat[2]
-    b_x_hat = b_hat[3]
-    B_3_hat = b_hat[4:length(b_hat)]
-
-    ## Linear predictor for M
-    eta_hat = a_0_hat + a_x_hat*x_pred + W_pred%*%A_2_hat
-    zeta_hat = b_0_hat + b_x_hat*x_pred + W_pred %*% B_3_hat
-
-    # Mediation effect
-    ## See Helpers.R for the function get_odds_ratio
-    med_hat = get_odds_ratio(eta_hat, a_x_hat, zeta_hat, b_x_hat, b_m_hat)
-    some_med_hats[[i]] = med_hat
-
-    # Estimate SE
-
-    ## Build joint covariance matrix
-    a_length = nrow(a_cov)
-    b_length = nrow(b_cov)
-
-    joint_cov = matrix(0, nrow = a_length + b_length, ncol = a_length + b_length)
-    joint_cov[1:a_length, 1:a_length] = a_cov
-    joint_cov[(a_length+1):(a_length+b_length), (a_length+1):(a_length+b_length)] = b_cov
-
-    ## Compute gradient of mediation effect
-    grad_med = d_OR_d_theta(eta_hat, a_x_hat, zeta_hat, b_x_hat, b_m_hat, x_pred, W_pred)
-
-    ## Apply delta-method
-    asymp_reg_cov = n * joint_cov
-
-    med_asymp_var = grad_med %*% asymp_reg_cov %*% grad_med
-    med_asymp_SE = sqrt(med_asymp_var)
-
-    med_SE = med_asymp_SE / sqrt(n)
-    some_med_SEs[[i]] = med_SE
-  }
-
-  # Store output for current value of n
-  all_a_hats[[j]] = some_a_hats
-  all_b_hats[[j]] = some_b_hats
-
-  all_a_SEs[[j]] = some_a_SEs
-  all_b_SEs[[j]] = some_b_SEs
-
-  all_med_hats[[j]] = some_med_hats
-  all_med_SEs[[j]] = some_med_SEs
-}
-
-```
-
-We now process the output from our simulation and summarize the results in a table.
-We give the mean and median of the estimated standard errors, as well as their relative errors (in $\%$) as estimates of the empirical standard error.
-See Table \ref{tab:results_bin_bin_fix}.
-
-```{r}
-# Re-format output to be more useful. 
-data_med_hats = lapply(all_med_hats, unlist)
-data_med_SEs = lapply(all_med_SEs, unlist)
-
-# SD of estimates across Monte Carlo samples
-emp_SEs = sapply(data_med_hats, sd)
-
-# Summaries of estimated SEs (mean and median)
-mean_delta_SEs = sapply(data_med_SEs, mean)
-median_delta_SEs = sapply(data_med_SEs, median)
-
-# Combine into a table
-results = data.frame(n = all_Ns, Empirical = emp_SEs, 
-                     Mean = mean_delta_SEs, Median = median_delta_SEs,
-                     "Percent_Err_Mean" = 100*(mean_delta_SEs - emp_SEs)/emp_SEs,
-                     "Percent_Err_Median" = 100*(median_delta_SEs - emp_SEs)/emp_SEs)
-
-```
-
-```{r results="asis", echo=FALSE}
-## Print table
-kbl(results, format = "latex", booktabs = T, caption = "Summary of SE estimates for mediation effect under continuous response, continuous mediator, fixed-effects model \\label{tab:results_bin_bin_fix}")
-```
-
 # Binary Response, Binary Mediator, Mixed-Effects
@@
 a_RE_cor = M_model_info$correlation[2,1]
-a_cov = vcov(M_model, full=TRUE, ranpar="sd")
+theta_hat = c(a_RE_sds, a_RE_cor)
+M_cov = vcov(M_model, full=TRUE, ranpar="sd")
 
@@
 b_RE_cor = Y_model_info$correlation[2,1]
-b_cov = vcov(Y_model, full=TRUE, ranpar="sd")
+gamma_hat = c(b_RE_sds, b_RE_cor)
+Y_cov = vcov(Y_model, full=TRUE, ranpar="sd")
 
@@
 ```{r}
+
+# Fixed-effects
 x_pred = 0
@@
 # Linear predictors
-eta_hat = a_0_hat + a_x_hat*x_pred + W_pred%*%A_2_hat
-zeta_hat = b_0_hat + b_x_hat * x_pred + W_pred %*% B_3_hat
+eta_hat = as.numeric(a_0_hat + a_x_hat*x_pred + W_pred%*%A_2_hat)
+zeta_hat = as.numeric(b_0_hat + b_x_hat * x_pred + W_pred %*% B_3_hat)
+
+# Random effects covariances
+s_M_0 = a_RE_sds[1]
+s_M_x = a_RE_sds[2]
+rho_M = a_RE_cor
+
+s_Y_0 = b_RE_sds[1]
+s_Y_x = b_RE_sds[2]
+rho_Y = b_RE_cor
+
+# Sigma functions
+sigma_M1 = sigma_fun(x_pred, s_M_0, s_M_x, rho_M)
+sigma_M2 = sigma_fun(x_pred + 1, s_M_x, s_M_0, rho_M)
+
+sigma_Y1 = sigma_fun(x_pred, s_Y_0, s_Y_x, rho_Y)
+sigma_Y2 = sigma_fun(x_pred + 1, s_Y_x, s_Y_0, rho_Y)
 
 # Mediation effect
-## See Helpers.R for the function get_odds_ratio
-med_hat = get_odds_ratio(eta_hat, a_x, zeta_hat, b_x, b_m)
-```
-
+## See Helpers.R for the function Phi, which computes the mediation effect on odds-ratio scale
+med_hat = Phi(eta_hat, zeta_hat, a_x_hat, b_m_hat, b_x_hat, sigma_M2, sigma_Y2, sigma_M1, sigma_Y1)
+```
+
+Applying the $\delta$-method here is a bit more complicated than in the other settings considered so far.
+Following the notes written by B & B, we take a two-step approach.
+We first transform model parameters from the two GLMMs into the vector of arguments to $\Phi$; call such a function $\xi$.
+We then compute the gradient of $\Phi$ with respect to these arguments, and use the $\delta$-method to get the standard error.
+
+In order to perform the first step, we need the gradient of $\xi$.
+Fortunately, much of this has already been done for us.
+Specifically, B & B worked out the gradients of each of the four $\sigma$ functions.
+All that remains is to repeat this process for the other arguments of $\Phi$, and stack the results in a matrix.
+See the `Helpers.R` file for the relevant functions.
+
+Starting first with 
 We now compute the gradient of the odds ratio and use the $\delta$-method to get the standard error.
@@
 # Gradient of OR wrt regression coefficients
-grad_med = d_OR_d_theta(eta_hat, a_x_hat, zeta_hat, b_x_hat, b_m_hat, x_pred, W_pred)
+grad_Phi_obs = grad_Phi(eta_hat, zeta_hat, a_x_hat, b_m_hat, b_x_hat, sigma_M2, sigma_Y2, sigma_M1, sigma_Y1)
+grad_xi_obs = grad_xi(a_hat, theta_hat, b_hat, gamma_hat, x_pred, W_pred)
+d_Phi_d_GLMM_pars = grad_Phi_obs %*% grad_xi_obs
 
@@
 ## Build joint covariance matrix of regression coefficients
-a_length = nrow(a_cov)
-b_length = nrow(b_cov)
-joint_cov = matrix(0, nrow = a_length + b_length, ncol = a_length +
-b_length)
-joint_cov[1:a_length, 1:a_length] = a_cov
-joint_cov[(a_length + 1):(a_length + b_length), (a_length + 1):(a_length +
-b_length)] = b_cov
+M_length = nrow(M_cov)
+Y_length = nrow(Y_cov)
+joint_cov = matrix(0, nrow = M_length + Y_length, ncol = M_length +
+Y_length)
+joint_cov[1:M_length, 1:M_length] = M_cov
+joint_cov[(M_length + 1):(M_length + Y_length), (M_length + 1):(M_length +
+Y_length)] = Y_cov
 
@@
 
-## Pre- and post-multiply asymptotic covariance by gradient of transformation
-med_asymp_var = grad_med %*% asymp_reg_cov %*% grad_med
+## Pre- and post-multiply asymptotic covariance by gradient of Phi wrt GLMM parameters
+med_asymp_var = d_Phi_d_GLMM_pars %*% asymp_reg_cov %*% t(d_Phi_d_GLMM_pars)
 
@@
 Putting everything together, we have an estimated mediation effect of `r med_hat`, with an estimated SE of `r med_SE`.
+
+## Monte Carlo Study
+
+We now repeat the above analysis `r num_reps` times for each of various values of $n$.
+Values of $X$ and $W$ for which we compute the total effect are set at the beginning.
+We also re-run the boilerplate code (not shown).
+
+```{r, include=FALSE}
+<<boilerplate>>
+
+```
+
+```{r}
+source("../src/Helpers.R")
+
+library(lme4)     # For fitting GLMMs using glmer()
+library(MASS)     # For simulating multivariate normals using mvrnorm()
+library(merDeriv) # For computing the full information matrix from models fit using lme4
+
+# Fewer large sample sizes
+all_Ns = c(100, 200, 500, 1000, 2000)
+
+# Values of X and W for which we compute the total effect
+x_pred = 0
+W_pred = c(1,1,1)
+
+# Fixed effects for intercepts
+a_0 = 0
+a = c(a_0, a_1, A_2)
+b_0 = -0.5
+b = c(b_0, b_1, b_2, B_3)
+
+# Random effects
+sigma_a_0 = 0.2
+sigma_a_1 = 0.2 * abs(a_1)
+cor_a0_a1 = 0.2
+cov_a0_a1 = sigma_a_0 * sigma_a_1 * cor_a0_a1
+Sigma_a = matrix(c(sigma_a_0^2, cov_a0_a1, cov_a0_a1, sigma_a_1^2), nrow = 2, ncol = 2)
+
+sigma_b_0 = 0.2 * abs(b_0)
+sigma_b_1 = 0.2 * abs(b_1)
+cor_b0_b1 = 0.2
+cov_b0_b1 = sigma_b_0 * sigma_b_1 * cor_b0_b1
+Sigma_b = matrix(c(sigma_b_0^2, cov_b0_b1, cov_b0_b1, sigma_b_1^2), nrow = 2, ncol = 2)
+
+# Number of clusters
+K=5
+
+# Number of replicates for each sample size
+# We use fewer here because model fitting takes longer
+num_reps = 200
+```
+
+```{r MC-bin-bin-mix, include=FALSE, cache=TRUE}
+# , dependson=c("boilerplate")}
+
+tic()
+
+# Containers for output
+all_a_hats = list()
+all_theta_hats = list()
+all_b_hats = list()
+all_gamma_hats = list()
+
+all_M_covs = list()
+all_Y_covs = list()
+
+all_med_hats = list()
+all_med_SEs = list()
+
+for(j in seq_along(all_Ns)){
+  n = all_Ns[j]
+
+  # Containers for output with this value of n
+  some_a_hats = list()
+  some_theta_hats = list()
+  some_b_hats = list()
+  some_gamma_hats = list()
+
+  some_M_covs = list()
+  some_Y_covs = list()
+
+  some_med_hats = list()
+  some_med_SEs = list()
+
+  for(i in 1:num_reps){
+    print(paste0("n = ", n, ", i = ", i, " of ", num_reps))
+
+    # Use tryCatch to omit any datasets leading to convergence issues
+    tryCatch({
+      all_Xs = list()
+      all_Ws = list()
+
+      for(k in 1:K){
+        X = rnorm(n, mean=0, sd=1)
+        W = matrix(rnorm(n*p_conf, mean=0, sd=1), nrow = n, ncol = p_conf)
+
+        all_Xs[[k]] = X
+        all_Ws[[k]] = W
+      }
+
+      # Generate M
+      all_Ms = list()
+      for(k in 1:K){
+        eta_vec_fixed = a_0 + a_1*all_Xs[[k]] + all_Ws[[k]]%*%A_2
+
+        ## Add random effects
+        a_ran = mvrnorm(1, mu = rep(0, 2), Sigma = Sigma_a)
+        eta_vec = eta_vec_fixed + a_ran[1] + a_ran[2]*all_Xs[[k]]
+
+        ## Generate M
+        p_M_vec = expit(eta_vec)
+        M = rbinom(n, size = 1, prob = p_M_vec)
+        all_Ms[[k]] = M
+      }
+
+      # Generate Y
+      all_Ys = list()
+      for(k in 1:K){
+        zeta_vec_fixed = b_0 + b_1*all_Ms[[k]] + b_2 * all_Xs[[k]] + all_Ws[[k]]%*%B_3
+
+        ## Add random effects
+        b_ran = mvrnorm(1, mu = rep(0, 2), Sigma = Sigma_b)
+        zeta_vec = zeta_vec_fixed + b_ran[1] + b_ran[2]*all_Xs[[k]]
+
+        ## Generate Y
+        p_Y_vec = expit(zeta_vec)
+        Y = rbinom(n, size = 1, prob = p_Y_vec)
+        all_Ys[[k]] = Y
+      }
+
+      # Consolidate groups
+      X = do.call(c, all_Xs)
+      W = do.call(rbind, all_Ws)
+      M = do.call(c, all_Ms)
+      Y = do.call(c, all_Ys)
+      group = rep(1:K, each = n)
+
+      # Fit models
+
+      ## M
+      M_data = data.frame(M, X, W1 = W[, 1], W2 = W[, 2], W3 = W[,3], group = group)
+      M_model = glmer(M ~ X + W1 + W2 + W3 + (1 + X | group), data = M_data, family = binomial(link = "logit"))
+      M_model_info = attributes(VarCorr(M_model)$group)
+
+      ### Fitted parameters
+      a_hat = fixef(M_model)
+      a_RE_sds = M_model_info$stddev
+      a_RE_cor = M_model_info$correlation[2,1]
+      theta_hat = c(a_RE_sds, a_RE_cor)
+      if(any(is.nan(theta_hat))) stop("NaNs in theta_hat")  # Skip rest of current analysis if correlation is 0/0
+
+      ### Estimated SE covariance matrix
+      M_cov = vcov(M_model, full=TRUE, ranpar="sd")
+
+      some_a_hats[[i]] = a_hat
+      some_theta_hats[[i]] = theta_hat
+      some_M_covs[[i]] = M_cov
+
+      ## Y
+      Y_data = data.frame(Y, M, X, W1 = W[, 1], W2 = W[, 2], W3 = W[,3], group = group)
+      Y_model = glmer(Y ~ M + X + W1 + W2 + W3 + (1 + X | group), data = Y_data, family = binomial(link = "logit"))
+      Y_model_info = attributes(VarCorr(Y_model)$group)
+
+      ### Fitted parameters
+      b_hat = fixef(Y_model)
+      b_RE_sds = Y_model_info$stddev
+      b_RE_cor = Y_model_info$correlation[2,1]
+      gamma_hat = c(b_RE_sds, b_RE_cor)
+      if(any(is.nan(gamma_hat))) stop("NaNs in gamma_hat")  # Skip rest of current analysis if correlation is 0/0
+
+      ### Estimated SE covariance matrix
+      Y_cov = vcov(Y_model, full=TRUE, ranpar="sd")
+
+      some_b_hats[[i]] = b_hat
+      some_gamma_hats[[i]] = gamma_hat
+      some_Y_covs[[i]] = Y_cov
+
+      # Estimate mediation effect
+
+      ## Fixed-effects
+      x_pred = 0
+      W_pred = c(1,1,1)
+
+      a_0_hat = a_hat[1]
+      a_x_hat = a_hat[2]
+      A_2_hat = a_hat[3:5]
+
+      b_0_hat = b_hat[1]
+      b_m_hat = b_hat[2]
+      b_x_hat = b_hat[3]
+      B_3_hat = b_hat[4:6]
+
+      ## Linear predictors
+      eta_hat = as.numeric(a_0_hat + a_x_hat*x_pred + W_pred%*%A_2_hat)
+      zeta_hat = as.numeric(b_0_hat + b_x_hat * x_pred + W_pred %*% B_3_hat)
+
+      ## Random effects covariances
+      s_M_0 = a_RE_sds[1]
+      s_M_x = a_RE_sds[2]
+      rho_M = a_RE_cor
+
+      s_Y_0 = b_RE_sds[1]
+      s_Y_x = b_RE_sds[2]
+      rho_Y = b_RE_cor
+
+      ## Sigma functions
+      sigma_M1 = sigma_fun(x_pred, s_M_0, s_M_x, rho_M)
+      sigma_M2 = sigma_fun(x_pred + 1, s_M_x, s_M_0, rho_M)
+
+      sigma_Y1 = sigma_fun(x_pred, s_Y_0, s_Y_x, rho_Y)
+      sigma_Y2 = sigma_fun(x_pred + 1, s_Y_x, s_Y_0, rho_Y)
+
+      ## Mediation effect
+      ### See Helpers.R for the function Phi, which computes the mediation effect on odds-ratio scale
+      med_hat = Phi(eta_hat, zeta_hat, a_x_hat, b_m_hat, b_x_hat, sigma_M2, sigma_Y2, sigma_M1, sigma_Y1)
+      some_med_hats[[i]] = med_hat
+
+      # Estimate SE
+
+      ## Gradient of OR wrt regression coefficients
+      grad_Phi_obs = grad_Phi(eta_hat, zeta_hat, a_x_hat, b_m_hat, b_x_hat, sigma_M2, sigma_Y2, sigma_M1, sigma_Y1)
+      grad_xi_obs = grad_xi(a_hat, theta_hat, b_hat, gamma_hat, x_pred, W_pred)
+      d_Phi_d_GLMM_pars = grad_Phi_obs %*% grad_xi_obs
+
+      ## Get asymptotic SE using delta method
+
+      ### Build joint covariance matrix of regression coefficients
+      M_length = nrow(M_cov)
+      Y_length = nrow(Y_cov)
+      joint_cov = matrix(0, nrow = M_length + Y_length, ncol = M_length +
+      Y_length)
+      joint_cov[1:M_length, 1:M_length] = M_cov
+      joint_cov[(M_length + 1):(M_length + Y_length), (M_length + 1):(M_length +
+      Y_length)] = Y_cov
+
+      ### Convert to asymptotic covariance matrix
+      asymp_reg_cov = n * joint_cov
+
+      ### Pre- and post-multiply asymptotic covariance by gradient of Phi wrt GLMM parameters
+      med_asymp_var = d_Phi_d_GLMM_pars %*% asymp_reg_cov %*% t(d_Phi_d_GLMM_pars)
+
+      ### Get small-sample standard error
+      med_asymp_SE = sqrt(med_asymp_var)
+      med_SE = med_asymp_SE / sqrt(n)
+
+      some_med_SEs[[i]] = med_SE
+    }, error = function(e){})
+  }
+
+  # Store output for current value of n
+  all_a_hats[[j]] = some_a_hats
+  all_theta_hats[[j]] = some_theta_hats
+  all_b_hats[[j]] = some_b_hats
+  all_gamma_hats[[j]] = some_gamma_hats
+
+  all_M_covs[[j]] = some_M_covs
+  all_Y_covs[[j]] = some_Y_covs
+
+  all_med_hats[[j]] = some_med_hats
+  all_med_SEs[[j]] = some_med_SEs
+}
+toc()
+
+```
+
+We now process the output from our simulation and summarize the results in a table.
+We give the mean and median of the estimated standard errors, as well as their relative errors (in $\%$) as estimates of the empirical standard error.
+See Table \ref{tab:results_bin_bin_fix}.
+
+```{r}
+# Re-format output to be more useful. 
+data_med_hats = lapply(all_med_hats, unlist)
+data_med_SEs = all_med_SEs %>% lapply(unlist) %>% lapply(na.omit)
+
+# SD of estimates across Monte Carlo samples
+emp_SEs = sapply(data_med_hats, sd)
+
+# Summaries of estimated SEs (mean and median)
+mean_delta_SEs = sapply(data_med_SEs, mean)
+median_delta_SEs = sapply(data_med_SEs, median)
+
+# Combine into a table
+results = data.frame(Empirical = emp_SEs, 
+                     Mean = mean_delta_SEs, Median = median_delta_SEs,
+                     "Percent_Err_Mean" = 100*(mean_delta_SEs - emp_SEs)/emp_SEs,
+                     "Percent_Err_Median" = 100*(median_delta_SEs - emp_SEs)/emp_SEs)
+# results = data.frame(n = all_Ns, Empirical = emp_SEs, 
+#                      Mean = mean_delta_SEs, Median = median_delta_SEs,
+#                      "Percent_Err_Mean" = 100*(mean_delta_SEs - emp_SEs)/emp_SEs,
+#                      "Percent_Err_Median" = 100*(median_delta_SEs - emp_SEs)/emp_SEs)
+
+```
+
+```{r results="asis", echo=FALSE}
+## Print table
+kbl(results, format = "latex", booktabs = T, caption = "Summary of SE estimates for mediation effect under continuous response, continuous mediator, fixed-effects model \\label{tab:results_bin_bin_fix}")
+```
+
+I don't really understand the results I'm getting here. It should help if I can compare my estimates of the mediation effect with the true value.
+
+```{r}
+
+```

### 2024-05-24 src/Helpers.R

@@ -312,6 +312,9 @@ d2_phi = function(mu, sigma){
 
 
 # Total mediation effect (on odds-ratio scale)
+## In my notation: Phi(eta, zeta, a_x, b_m, b_x, sigma_U(x+1), sigma_V(x+1), sigma_U(x), sigma_V(x))
+## Note: eta, a and U come from the M-model, while zeta, b and V come from the Y-model
+## eta and zeta are linear predictors
 Phi = function(a, b, a1, b1, b2, s1, s2, s3, s4){
   (phi(b + b1 + b2, s1) * phi(a + a1, s2) + phi(b + b2, s1) * (1 - phi(a + a1, s2))) * (1 - phi(b, s3) + phi(a, s4) * (phi(b, s3) - phi(b + b1, s3))) / (1 - phi(b + b2, s1) + phi(a + a1, s2) * (phi(b + b2, s1) - phi(b + b1 + b2, s2))) / (phi(b + b1, s3) * phi(a, s3) + phi(b, s3) * (1 - phi(a, s4)))
 
@@ -396,21 +399,425 @@ grad_Phi = function(a, b, a1, b1, b2, s1, s2, s3, s4){
 }
 
 
-# Finite difference verification
-a = 1
-b = 1
-a1 = 1
-b1 = 1
-b2 = 1
-s1 = 0.5
-s2 = 0.5
-s3 = 0.5
-s4 = 0.5
+# # Finite difference verification
+# a = 1
+# b = 1
+# a1 = 1
+# b1 = 1
+# b2 = 1
+# s1 = 0.5
+# s2 = 0.5
+# s3 = 0.5
+# s4 = 0.5
+#
+# e = 0.000001
+#
+# (Phi(a, b, a1, b1, b2, s1, s2, s3, s4 + e) - Phi(a, b, a1, b1, b2, s1, s2, s3, s4)) / e
+# dPhi_ds4(a, b, a1, b1, b2, s1, s2, s3, s4)
+#
+#
+# grad_Phi(a, b, a1, b1, b2, s1, s2, s3, s4)
 
-e = 0.000001
-
-(Phi(a, b, a1, b1, b2, s1, s2, s3, s4 + e) - Phi(a, b, a1, b1, b2, s1, s2, s3, s4)) / e
-dPhi_ds4(a, b, a1, b1, b2, s1, s2, s3, s4)
 
 
-grad_Phi(a, b, a1, b1, b2, s1, s2, s3, s4)
+# Transform from GLMM parameters to arguments of Phi
[…]

### 2024-05-29 notes/Analysis Write-Up/Direct and Indirect Effects.tex
Define mediation effects in write-up

\documentclass{article}

\usepackage{amsmath, amssymb}
\usepackage{xcolor}
\usepackage{caption, subcaption}
\usepackage{graphicx}
\usepackage{bm}             % For bold math symbols
\usepackage{mathrsfs}       % For the \mathscr{} font

\usepackage{natbib}
\bibliographystyle{plainnat}

\usepackage{hyperref}   % For the \url{} command

\usepackage{tabu,multirow}

\newcommand{\bG}{\mathbb{G}}
\newcommand{\bF}{\mathbb{F}}
\newcommand{\bP}{\mathbb{P}}
\newcommand{\bE}{\mathbb{E}}
\newcommand{\bV}{\mathbb{V}}

\newcommand{\sY}{\mathcal{Y}}

\newcommand{\zdens}{\frac{e^{-z^2/2}}{\sqrt{2\pi}}}

\title{Analysis of Direct and Indirect Mediation Effects in Causal Mediation Analysis}

\begin{document}

\maketitle

In this document, we develop definitions and formulas for the direct and indirect mediation effects to complement the total effect given by B \& B. We follow the work of \citet{Pea12} and the Samoilenko and Lefebvre group (e.g., \citealp{Sam23}).

\section{Expected Nested Counterfactuals}
\label{sec:nest_CFs}

To start, we define the nested counterfactual $Y(x, M(x'))$ as the value that $Y$ would assume when $X$ is set to $x$ and $M$ is set to whatever value it would have assumed if we had set $X$ to $x'$. \citet{Pea12} identifies the expected value of this nested counterfactual with the following expression based on conditional expectations:
%
\begin{align}
    \bE Y(x, M(x')) = \int \bE(Y | M=m, X=x) \bP(M = dm | X = x') \label{eq:med_gen}
\end{align}
%
Focusing now on the case with binary response and binary mediator, \ref{eq:med_gen} becomes
%
\begin{multline}
    \bE Y(x, M(x')) = \bP(Y=1 | M=1, X=x) \bP(M=1 | X=x') + \\\bP(Y=1 | M=0, X=x) \bP(M=0 | X=x') \label{eq:med_bin}
\end{multline}
%
Consider using  logistic regression models for $Y$ and $M$. Write $\mathrm{logit}(\bE(Y | M=m, X=x)) = \beta_0 + \beta_m m + \beta_x x$ and $\mathrm{logit}(\bE(M | X=x)) = \alpha_0 + \alpha_x x$. Then (\ref{eq:med_bin}) becomes
%
\begin{multline}
    \bE Y(x, M(x')) = \frac{1}{1 + \exp(-\beta_0 - \beta_m - \beta_x x)} \frac{1}{1 + \exp(-\alpha_0 - \alpha_x x')} +\\
    \frac{1}{1 + \exp(-\beta_0 - \beta_x x)} \frac{1}{1 + \exp(\alpha_0 + \alpha_x x')} \label{eq:med_bin_fix}
\end{multline}
%
Equation (\ref{eq:med_bin_fix}) holds for logistic regression with fixed-effects only. If we instead use  mixed-effects logistic regressions for $Y$ and $M$, then (\ref{eq:med_gen}) and (\ref{eq:med_bin}) still hold. For the mixed-effects models, first, write $V = (V_0, V_m, V_x) \sim \mathrm{N}(0, \Sigma_V)$ and $U = (U_0, U_x) \sim \mathrm{N}(0, \Sigma_U)$ for the random-effects in our models for $Y$ and $M$ respectively. Next, write $\mathrm{logit}(\bE(Y | V, M=m, X=x)) = \beta_0 + V_0 + (\beta_m + V_m) m + (\beta_x + V_x) x$ and $\mathrm{logit}(\bE(M |U, X=x)) = (\alpha_0 + U_0) + (\alpha_x + U_x) x$. Returning now to identification of the expected counterfactual for $Y$, we get
%
\begin{align}
    \bE Y(x, M(x')) =& \left[\int \bP(Y=1 |V=v, M=1, X=x) \bP(V = dv)\right] \cdot \\
    & \left[\int \bP(M=1 |U=u, X=x') \bP(U=du)\right] +\\
     & \left[\int \bP(Y=1 |V=v, M=0, X=x) \bP(V = dv)\right] \cdot \\
     & \left[\int \bP(M=0 |U=u, X=x') \bP(U=du)\right]
\end{align}
%
and, in the logistic regression context,
%
\begin{align}
    \bE Y(x, M(x')) =& \left[ \int \frac{\phi(v; 0, \Sigma_V)}{1 + \exp(-(\beta_0 + v_0) - (\beta_m + v_m) - (\beta_x + v_x) x)}  dv \right] \cdot \label{eq:med_bin_ran}\\
    &\left[ \int \frac{\phi(u; 0, \Sigma_U)}{1 + \exp(-(\alpha_0 + u_0) - (\alpha_x + u_x) x')}  du \right] + \nonumber\\
    & \left[\int \frac{\phi(v; 0, \Sigma_V)}{1 + \exp(-(\beta_0 + v_0) - (\beta_x + v_x) x)} dv \right] \cdot \nonumber\\
    & \left[\int \frac{\phi(u; 0, \Sigma_U)}{1 + \exp((\alpha_0 + u_0) + (\alpha_x + u_x) x')} du \right], \nonumber
\end{align}
%
where $\phi(.; \mu, \Sigma)$ is the multivariate normal density with mean $\mu$ and covariance matrix $\Sigma$.

Note that the four integrals in (\ref{eq:med_bin_ran}) are all multivariate, but can be transformed to univariate integrals by suitable changes of variables. To this end, write $\eta = \alpha_0 + \alpha_x x$ and $\zeta = \beta_0 + \beta_x x$ for two linear predictors (note that $\zeta$ does not contain $\beta_m$), and $\gamma^2_\Sigma(c_1, \ldots, c_r) = (c_1, \ldots, c_r) \Sigma (c_1, \ldots, c_r)^T$, where $\Sigma$ is an $r$-by-$r$ covariance matrix. We will generally set $\Sigma = \Sigma_V$ or $\Sigma = \Sigma_U$, in which case we write $\gamma^2_V$ or $\gamma^2_U$ respectively. We now define the function $\psi$ as a template for the four integrals in (\ref{eq:med_bin_ran}). 
%
$$
\psi(\mu, \sigma^2) := \int \frac{\phi(z; 0, 1)}{1 + \exp(-\mu - \sigma z)} dz.
$$
%
Note that $\psi$ is a univariate integral, so we can expect it to be well-approximated by routine numerical quadrature techniques. We now re-write (\ref{eq:med_bin_ran}) in terms of $\psi$ as follows,
%
\begin{align}
    \bE Y(x, M(x')) =& \psi(\zeta + \beta_m, \gamma^2_V(1, 1, x)) \cdot \psi(\eta, \gamma^2_U(1, x)) +\\
    & \psi(\zeta, \gamma^2_V(1, 0, x)) \cdot \psi(-\eta, \gamma^2_U(1, x)). \nonumber 
\end{align}

\section{Mediation Effects}
Denote the expected nested counterfactual defined in (\ref{eq:med_gen}) by $\mathscr{Y}(x, x') = \bE Y(x, M(x'))$. We can define the various mediation effects in terms of expected counterfactuals. Note that mediation effects for a binary outcome are commonly defined on three different scales: risk difference, risk ratio and odds ratio. Table \ref{tab:med_eff_defs} gives all such definitions explicitly. Combined with the formulas given throughout Section \ref{sec:nest_CFs}, we now have everything we need to do point estimation for the various mediation effects. It still remains however, to address uncertainty quantification.
%
\begin{table}[ht]
    \centering
    \caption{Definitions of various mediation effects; $x$ and $x'$ denote different values of the exposure.}
    \label{tab:med_eff_defs}
    \begin{tabu}{|[1pt]c|c|c|[1pt]}
        \tabucline[1pt]{-}
        \multirow{3}{*}{Risk Difference} & Total Effect & $\sY(x, x) - \sY(x', x')$ \\
        \cline{2-3}
        & Direct Effect & $\sY(x, x') - \sY(x', x')$\\
        \cline{2-3}
        & Indirect Effect & $\sY(x, x) - \sY(x, x')$ \\
        \tabucline[1pt]{-}
        \multirow{3}{*}{Risk Ratio} & Total Effect & $\sY(x, x) / \sY(x', x')$ \\
        \cline{2-3}
        & Direct Effect & $\sY(x, x') / \sY(x', x')$\\
        \cline{2-3}
        & Indirect Effect & $\sY(x, x) / \sY(x, x')$ \\
        \tabucline[1pt]{-}
        \multirow{3}{*}{Odds Ratio} & Total Effect & $\left. \frac{\sY(x, x)}{1 - \sY(x, x)} \right/ \frac{\sY(x', x')}{1 - \sY(x', x')} $\\
        \cline{2-3}
        & Direct Effect & $\left. \frac{\sY(x, x')}{1 - \sY(x, x')} \right/ \frac{\sY(x', x')}{1 - \sY(x', x')} $\\
        \cline{2-3}
        & Indirect Effect & $\left. \frac{\sY(x, x)}{1 - \sY(x, x)} \right/ \frac{\sY(x, x')}{1 - \sY(x, x')} $\\
        \tabucline[1pt]{-}
    \end{tabu}
\end{table}

\subsection{Uncertainty Quantification}

Note that all mediation effects are defined in terms of expected nested counterfactuals $\sY$. Thus, if we can produce a formula for the asymptotic covariance matrix of two (or more) values of $\sY$, we can use a simple application of the $\delta$-method to obtain the variance of any mediation effect. In fact, we can obtain the covariance matrix for all three mediation effects defined on a particular scale (e.g., risk ratio), or, indeed, between all nine mediation effects given in Table \ref{tab:med_eff_defs}.

Toward a variance formula for $\sY$, we first write $\theta$ for all parameters upon which $\sY$ depends. That is, $\theta$ contains both sets of regression coefficients, $\beta_0, \beta_m, \beta_x$ and $\alpha_0, \alpha_x$, as well as both sets of covariance parameters. For consistency with B \& B, we parameterize these as $\tau_0, \tau_m, \tau_x$ for the standard deviations of $V_0, V_m, V_x$, and $\tau_{0,m}, \tau_{0,x}, \tau_{m,x}$ for the corresponding correlations\footnote{While our notation doesn't match that given by B \& B, parameterizing in terms of the standard deviations and correlations does. Alternative choices include the variances and covariances, or the unique components of the Cholesky factorizations of $\Sigma_V$ and $\Sigma_U$.}. Similarly, we use $\sigma_0, \sigma_x$ for the standard deviations of $U_0, U_x$, and $\sigma_{0,x}$ for their correlation. We are now equipped to write-out $\theta$ in full. The order of parameters is chosen to match my code (and to avoid refactoring thereof).
$$
\theta = (\alpha_0, \alpha_x, \sigma_0, \sigma_x, \sigma_{0,x}, \beta_0, \beta_m, \beta_x, \tau_0, \tau_m, \tau_x, \tau_{0,m}, \tau_{0,x}, \tau_{m,x})
$$
%
There are two problems left to solve for the quantification of uncertainty in $\sY$. First, we need the (asymptotic) covariance matrix of $\theta$. Second, we need the $\theta$-gradient of the function $\sY$. Note that $\sY$ is also a function of $x$ and $x'$, but we treat these values as fixed for our analysis.

\section{To Do}
\begin{itemize}
    \item Incorporate an interaction term between $X$ and $M$ in the model for $Y$.
\end{itemize}

\bibliography{../Literature_Review/Lit_Review_Bib.bib}
% \bibliography{Lit_Review_Bib}
% \bibliography{notes/Literature Review/Lit_Review_Bib}

\end{document}

### 2024-05-29 notes/William Analysis.Rmd
Define mediation effects in write-up

@@
 
-set.seed(1)
+# set.seed(1)
+set.seed(12345)
 
@@
 library(merDeriv) # For computing the full information matrix from models fit using lme4
+library(tictoc)
 
 # Fewer large sample sizes
-all_Ns = c(100, 200, 500, 1000, 2000)
+# all_Ns = c(100, 200, 500, 1000, 2000)
+# all_Ns = c(200, 500, 1000, 2000)
+all_Ns = c(200, 500, 1000)
 
@@
 cov_a0_a1 = sigma_a_0 * sigma_a_1 * cor_a0_a1
+theta = c(sigma_a_0, sigma_a_1, cor_a0_a1)
 Sigma_a = matrix(c(sigma_a_0^2, cov_a0_a1, cov_a0_a1, sigma_a_1^2), nrow = 2, ncol = 2)
@@
 cov_b0_b1 = sigma_b_0 * sigma_b_1 * cor_b0_b1
+gamma = c(sigma_b_0, sigma_b_1, cor_b0_b1)
 Sigma_b = matrix(c(sigma_b_0^2, cov_b0_b1, cov_b0_b1, sigma_b_1^2), nrow = 2, ncol = 2)
@@
 tic()
+
+set.seed(2222)
 
@@
       ## Fixed-effects
-      x_pred = 0
-      W_pred = c(1,1,1)
 
@@
 
+# Save output
+# save(all_a_hats, all_theta_hats, all_b_hats, all_gamma_hats, all_M_covs, all_Y_covs, all_med_hats, all_med_SEs, file = "William-Analysis_cache/MC-bin-bin-mix.RData")
+# load("William-Analysis_cache/MC-bin-bin-mix.RData")
 ```
@@
 # Combine into a table
-results = data.frame(Empirical = emp_SEs, 
+# results = data.frame(Empirical = emp_SEs,
+#                      Mean = mean_delta_SEs, Median = median_delta_SEs,
+#                      "Percent_Err_Mean" = 100*(mean_delta_SEs - emp_SEs)/emp_SEs,
+#                      "Percent_Err_Median" = 100*(median_delta_SEs - emp_SEs)/emp_SEs)
+results = data.frame(n = all_Ns, Empirical = emp_SEs,
                      Mean = mean_delta_SEs, Median = median_delta_SEs,
@@
                      "Percent_Err_Median" = 100*(median_delta_SEs - emp_SEs)/emp_SEs)
-# results = data.frame(n = all_Ns, Empirical = emp_SEs, 
-#                      Mean = mean_delta_SEs, Median = median_delta_SEs,
-#                      "Percent_Err_Mean" = 100*(mean_delta_SEs - emp_SEs)/emp_SEs,
-#                      "Percent_Err_Median" = 100*(median_delta_SEs - emp_SEs)/emp_SEs)
 
@@
 ```{r}
-
-```
+load("../data/Exact_Asymptotics/Bin-Bin-Mix_Results.RData")
+
+true_med = Phi_of_xi(a, theta, b, gamma, x_pred, W_pred)
+
+load("William-Analysis_cache/latex/unnamed-chunk-37_44bd8d1152289e689fbff9185d17042c.RData", ex <- new.env())
+ls.str(ex)
+```
+
+Investigate normality of estimators
+```{r}
+library(ggplot2)
+library(magrittr)
+
+med_hats_to_plot = unlist(data_med_hats)
+Ns_to_plot = rep(all_Ns, times = sapply(data_med_hats, length))
+data_med_hat_to_plot = data.frame(med_hat = med_hats_to_plot, N = Ns_to_plot)
+
+data_med_hat_to_plot_OLD = data_med_hat_to_plot
+data_med_hat_to_plot %<>% dplyr::filter(med_hat < 4)
+
+ggplot(data_med_hat_to_plot, aes(x = med_hat)) +
+  # geom_histogram(bins = 30) +
+  geom_density() +
+  facet_wrap(~N, scales = "free") +
+  theme_minimal()
+
+# QQ plot for each sample size
+ggplot(data_med_hat_to_plot, aes(sample = med_hat)) +
+  stat_qq() +
+  stat_qq_line() +
+  facet_wrap(~N, scales = "free") +
+  theme_minimal()
+
+# Test normality in each sample size
+all_norm_p_vals = c()
+for(n in all_Ns){
+  p_val = shapiro.test(data_med_hat_to_plot$med_hat[data_med_hat_to_plot$N == n])$p.value
+  all_norm_p_vals = c(all_norm_p_vals, p_val)
+}
+all_norm_p_vals
+```

### 2024-05-29 src/Helpers.R

@@ -286,24 +286,28 @@ phi = function(mu, sigma){
 ## d phi / d mu
 d1_phi = function(mu, sigma){
   integrand = function(x){
-    A = 1 + exp(-mu - sigma*x)
-    B = dnorm(x)
-    C = exp(-mu - sigma*x)
+    A = dnorm(x)
 
-    return(C*B/A^2)
+    r = mu + sigma*x
+    B = exp(-r/2) + exp(r/2)
+
+    # return(C*B/A^2)
+    return(A / (B^2))
   }
 
   return(integrate(integrand, -Inf, Inf)$value)
 }
 
+
 ## d phi / d sigma
 d2_phi = function(mu, sigma){
   integrand = function(x){
-    A = 1 + exp(-mu - sigma*x)
-    B = dnorm(x)
-    C = x * exp(-mu - sigma*x)
+    A = dnorm(x)
 
-    return(C*B/A^2)
+    r = mu + sigma*x
+    B = exp(-r/2) + exp(r/2)
+
+    return(x * A / B^2)
   }
 
   return(integrate(integrand, -Inf, Inf)$value)
@@ -315,11 +319,17 @@ d2_phi = function(mu, sigma){
 ## In my notation: Phi(eta, zeta, a_x, b_m, b_x, sigma_U(x+1), sigma_V(x+1), sigma_U(x), sigma_V(x))
 ## Note: eta, a and U come from the M-model, while zeta, b and V come from the Y-model
 ## eta and zeta are linear predictors
-Phi = function(a, b, a1, b1, b2, s1, s2, s3, s4){
-  (phi(b + b1 + b2, s1) * phi(a + a1, s2) + phi(b + b2, s1) * (1 - phi(a + a1, s2))) * (1 - phi(b, s3) + phi(a, s4) * (phi(b, s3) - phi(b + b1, s3))) / (1 - phi(b + b2, s1) + phi(a + a1, s2) * (phi(b + b2, s1) - phi(b + b1 + b2, s2))) / (phi(b + b1, s3) * phi(a, s3) + phi(b, s3) * (1 - phi(a, s4)))
+Phi = function(eta, zeta, a1, b1, b2, s1, s2, s3, s4){
[…]

### 2024-05-30 notes/William Analysis.Rmd
Run analysis for binary-binary-mixed case

@@
 # all_Ns = c(200, 500, 1000, 2000)
+# all_Ns = c(200, 500, 1000)
 all_Ns = c(200, 500, 1000)
@@
 # Number of clusters
-K=5
+all_Ks = c(5, 20, 50)
+# K=5
+
+N_K_combs = expand.grid(N = all_Ns, K = all_Ks)
 
@@
 
-for(j in seq_along(all_Ns)){
-  n = all_Ns[j]
+for(j in seq_len(nrow(N_K_combs))){
+  n = N_K_combs[j, "N"]
+  K = N_K_combs[j, "K"]
 
@@
   for(i in 1:num_reps){
-    print(paste0("n = ", n, ", i = ", i, " of ", num_reps))
+    print(paste0("n = ", n, ", K = ", K, ", i = ", i, " of ", num_reps))
 
@@
 # Save output
-# save(all_a_hats, all_theta_hats, all_b_hats, all_gamma_hats, all_M_covs, all_Y_covs, all_med_hats, all_med_SEs, file = "William-Analysis_cache/MC-bin-bin-mix.RData")
+# save(all_a_hats, all_theta_hats, all_b_hats, all_gamma_hats, all_M_covs, all_Y_covs, all_med_hats, all_med_SEs, file = "William-Analysis_cache/MC-bin-bin-mix-n=200,500,1000.RData")
 # load("William-Analysis_cache/MC-bin-bin-mix.RData")
@@
 #                      "Percent_Err_Median" = 100*(median_delta_SEs - emp_SEs)/emp_SEs)
-results = data.frame(n = all_Ns, Empirical = emp_SEs,
+# results = data.frame(n = all_Ns, Empirical = emp_SEs,
+#                      Mean = mean_delta_SEs, Median = median_delta_SEs,
+#                      "Percent_Err_Mean" = 100*(mean_delta_SEs - emp_SEs)/emp_SEs,
+#                      "Percent_Err_Median" = 100*(median_delta_SEs - emp_SEs)/emp_SEs)
+results = data.frame(n = N_K_combs$N, K = N_K_combs$K, Empirical = emp_SEs,
                      Mean = mean_delta_SEs, Median = median_delta_SEs,
@@
 med_hats_to_plot = unlist(data_med_hats)
-Ns_to_plot = rep(all_Ns, times = sapply(data_med_hats, length))
-data_med_hat_to_plot = data.frame(med_hat = med_hats_to_plot, N = Ns_to_plot)
+Ns_to_plot = rep(N_K_combs$N, times = sapply(data_med_hats, length))
+Ks_to_plot = rep(N_K_combs$K, times = sapply(data_med_hats, length))
+data_med_hat_to_plot = data.frame(med_hat = med_hats_to_plot, N = Ns_to_plot, K = Ks_to_plot)
 
@@
   # geom_histogram(bins = 30) +
-  geom_density() +
-  facet_wrap(~N, scales = "free") +
-  theme_minimal()
+  geom_density() +   theme_minimal() +
+  facet_grid(rows = vars(N), cols = vars(K), scales = "free")
 
@@
   stat_qq_line() +
-  facet_wrap(~N, scales = "free") +
+  facet_grid(rows = vars(N), cols = vars(K), scales = "free")
   theme_minimal()
@@
 all_norm_p_vals = c()
-for(n in all_Ns){
-  p_val = shapiro.test(data_med_hat_to_plot$med_hat[data_med_hat_to_plot$N == n])$p.value
+for(i in 1:nrow(N_K_combs)){
+  p_val = shapiro.test(data_med_hats[[i]])$p.value
   all_norm_p_vals = c(all_norm_p_vals, p_val)
 }
-all_norm_p_vals
-```
+
+results$Shapiro_p = all_norm_p_vals
+results
+```

### 2024-05-30 notes/Analysis Write-Up/Direct and Indirect Effects.tex
Write-up definitions of direct and indirect effects in terms of nested counterfactuals

@@
 
-\usepackage{hyperref}   % For the \url{} command
+\usepackage[breaklinks,hidelinks]{hyperref}   % For the \url{} command
 
@@
 Equation (\ref{eq:med_bin_fix}) holds for logistic regression with fixed-effects only.
-If we instead use  mixed-effects logistic regressions for $Y$ and $M$, then (\ref{eq:med_gen}) and (\ref{eq:med_bin}) still hold.
-For the mixed-effects models, first, write $V = (V_0, V_m, V_x) \sim \mathrm{N}(0, \Sigma_V)$ and $U = (U_0, U_x) \sim \mathrm{N}(0, \Sigma_U)$ for the random-effects in our models for $Y$ and $M$ respectively.
+If we instead use  mixed-effects logistic regressions for $Y$ and $M$, then (\ref{eq:med_gen}) and (\ref{eq:med_bin}) still hold, but (\ref{eq:med_bin_fix}) must be modified.
+For the mixed-effects models, first write $V = (V_0, V_m, V_x) \sim \mathrm{N}(0, \Sigma_V)$ and $U = (U_0, U_x) \sim \mathrm{N}(0, \Sigma_U)$ for the random-effects in our models for $Y$ and $M$ respectively.
 Next, write $\mathrm{logit}(\bE(Y | V, M=m, X=x)) = \beta_0 + V_0 + (\beta_m + V_m) m + (\beta_x + V_x) x$ and $\mathrm{logit}(\bE(M |U, X=x)) = (\alpha_0 + U_0) + (\alpha_x + U_x) x$.
@@
 Table \ref{tab:med_eff_defs} gives all such definitions explicitly.
-Combined with the formulas given throughout Section \ref{sec:nest_CFs}, we now have everything we need to do point estimation for the various mediation effects.
-It still remains however, to address uncertainty quantification.
+In the rest of this section, we outline the processes of estimation and uncertainty quantification for the mediation effects defined in Table \ref{tab:med_eff_defs}.
+
 %
@@
 
-\subsection{Uncertainty Quantification}
+\subsection{Estimation}
 
-Note that all mediation effects are defined in terms of expected nested counterfactuals $\sY$.
-Thus, if we can produce a formula for the asymptotic covariance matrix of two (or more) values of $\sY$, we can use a simple application of the $\delta$-method to obtain the variance of any mediation effect.
-In fact, we can obtain the covariance matrix for all three mediation effects defined on a particular scale (e.g., risk ratio), or, indeed, between all nine mediation effects given in Table \ref{tab:med_eff_defs}.
-
-Toward a variance formula for $\sY$, we first write $\theta$ for all parameters upon which $\sY$ depends.
+Mediation effects are all defined in terms of the expected nested counterfactuals, $\sY(x, x')$.
+As such, the estimation of any mediation effect given in Table \ref{tab:med_eff_defs} centers around estimation of $\sY$.
+To this end, write $\theta$ for all parameters upon which $\sY$ depends.
 That is, $\theta$ contains both sets of regression coefficients, $\beta_0, \beta_m, \beta_x$ and $\alpha_0, \alpha_x$, as well as both sets of covariance parameters.
@@
 %
-There are two problems left to solve for the quantification of uncertainty in $\sY$.
-First, we need the (asymptotic) covariance matrix of $\theta$.
-Second, we need the $\theta$-gradient of the function $\sY$.
-Note that $\sY$ is also a function of $x$ and $x'$, but we treat these values as fixed for our analysis.
+In order to estimate $\theta$, we fit two mixed-effects logistic regression models.
+The first model predicts the mediator, $M$, using the exposure, $X$, while the second predicts the outcome, $Y$, using $M$ and $X$.
+Both models also contain intercepts.
+We include random effects for all regression coefficients in both models.
+\textcolor{red}{Either or both models may also contain one or more confounders, $\mathbf{C}$, although we assign these confounders only fixed effects.} Following the algorithm used by the \texttt{lme4} package in \texttt{R} \citep{Bat15, Wal23}, we estimate the parameters in both regression models by maximizing an approximation to the marginal likelihood of the response ($M$ or $Y$).
+This approximation is obtained using Laplace's Method, maximizing the joint likelihood with the random effects using a penalized iteratively re-weighted least squares algorithm.
+Applying this procedure to both of our regression models yields estimates for all components of $\theta$.
+
+Combining our estimate of $\theta$ with Equation \ref{eq:med_bin_ran} from Section \ref{sec:nest_CFs} and the formulas given in Table \ref{tab:med_eff_defs}, we are now able to estimate all mediation effects on the three scales we consider.
+It remains however, to address uncertainty quantification.
+
+\subsection{Uncertainty Quantification}
+
+As mentioned above, all mediation effects are defined in terms of expected nested counterfactuals $\sY$, which themselves depend on the parameter vector $\theta$.
+Quantification of uncertainty for our estimator of a mediation effect thus proceeds in three stages.
+First, we obtain a covariance matrix for our estimator of $\theta$.
+Second, we apply the $\delta$-method to get a joint covariance matrix for all values of $\sY$ required to evaluate the mediation effect of interest.
+Finally, we use the $\delta$-method again to translate uncertainty in the $\sY$s to uncertainty in the mediation effect.
+In fact, we can obtain the covariance matrix for all three mediation effects defined on a particular scale (e.g., risk ratio), or, indeed, between all nine mediation effects given in Table \ref{tab:med_eff_defs}.
+Note that the variances and covariances discussed above are asymptotic.
+
+We now proceed with the three steps discussed above, starting with an asymptotic covariance matrix for $\theta$.
+\textcolor{red}{For now, we treat the two regression models as independent.
+That is, we assume that the covariance between all parameter estimators in our $M$ model and all parameter estimators in our $Y$ model are zero.
+This is likely not true; call it a working assumption.
+See \citet{Bau06} for a method to model this inter-model dependence.} This step is fairly straightforward, since both regression models are fit using (approximate) maximum likelihood.
+We simply compute the negative Hessian of the marginal log-likelihood for the observed data in both models, and stack the results into a block-diagonal matrix.
+Call the resulting covariance matrix $\hat{\Sigma}_\theta$.
+This computation is performed by the \texttt{merDeriv} package in \texttt{R} \citep{Wan18}.
+
+To quantify uncertainty in $\sY$ based on our estimator for $\theta$, we use the $\delta$-method \citep[see, e.g., Chapter 3 of][]{vdV98}.
+Briefly, given the asymptotic covariance of our estimator for $\theta$, $\sqrt{n}(\hat{\theta} - \theta) \rightsquigarrow \mathrm{N}(0, \Sigma)$, the asymptotic covariance of some function of our estimator can be obtained by differentiating that function and multiplying by the limiting distribution.
+That is, $\sqrt{n}[f(\hat{\theta}) - f(\theta)] \rightsquigarrow \nabla f(\theta)\mathrm{N}(0, \Sigma) \overset{d}{=} \mathrm{N}(0, \nabla f(\theta) \Sigma \nabla f(\theta)^T)$.
+We estimate this limiting covariance by $\nabla f(\hat{\theta}) \hat{\Sigma} \nabla f(\hat{\theta})^T$, where typically $\Sigma = \Sigma(\theta)$ and $\hat{\Sigma} = \Sigma(\hat{\theta})$.
+
+Returning now to our problem, in order to apply the $\delta$-method we need the gradient of $\sY$ with respect to $\theta$.
+While tedious, this calculation is easily performed with the help of symbolic differentiation software like Maple \citep{Map20}.
+We then evaluate our gradient formula at $\hat{\theta}$, $\left.
+\nabla \sY \right|_{\theta = \hat{\theta}}$, then pre- and post-multiply $\hat{\Sigma}_\theta$ by our estimated gradient.
+If we require the covariance matrix for multiple values of $\sY$ (e.g., $\sY(x,x)$ and $\sY(x,x')$ for some $x \neq x'$), we frame the problem as a vector-valued transformation, and compute the Jacobian by stacking the gradient for each $\sY$.
+More precisely, to get a covariance matrix for $\sY_1, \ldots, \sY_r$, we first construct the $r$-by-$|\theta|$ Jacobian matrix $\mathcal{J} := [\nabla \sY_1, \ldots, \nabla \sY_r]$.
+We then evaluate $\mathcal{J}$ at $\theta = \hat{\theta}$, call the result $\hat{\mathcal{J}}$, and pre- and post-multiply $\hat{\Sigma}_\theta$ by $\hat{\mathcal{J}}$, $\hat{\mathcal{J}} \hat{\Sigma}_\theta \hat{\mathcal{J}}^T$.
+
+Finally, to get the asymptotic variance of one or more estimated mediation effects, we just apply the $\delta$-method to the corresponding formulas from Table \ref{tab:med_eff_defs}.
+Since the mediation effects are all simple functions of the $\sY$, any gradients required for this step are easily obtained.
 
@@
     \item Incorporate an interaction term between $X$ and $M$ in the model for $Y$.
+    \item Explore dependence between the models for $M$ and $Y$ using the `stacking' technique described in \citet{Bau06}.
+This will give non-zero covariance between parameter estimators from the two models.
 \end{itemize}

### 2024-05-30 scripts/Bin-Resp, Bin-Med, Random.jl
Explore convergence rate for GLMM MLEs

@@ -2,7 +2,7 @@
 #* Later, I can introduce prediction of random effects
 
 using Distributions, Random, Statistics
-using LogExpFunctions   # For the logit and expit functions
+using LogExpFunctions  
 using DataFrames
 using MixedModels
 using ProgressMeter, DrWatson

### 2024-05-30 scripts/GLMM Asymptotics Exploration/GLMM Asymptotics Exploration - GLMM.jl

@@ -1,0 +1,144 @@
+
+using Random, Distributions
+using LinearAlgebra   # For the identity matrix, I()
+using DataFrames, DataFramesMeta        # The latter works like dplyr in R with different verb names
+using MixedModels, GLM
+using ProgressMeter
+
+
+# Set parameters
+
+# N = 1000        # Total sample size
+# K = 5           # Number of groups
+# n = Int(N / K)       # Number of observations per group
+p = 2           # Number of (mixed-effects) covariates
+
+all_Ns = [500, 1000, 2000, 5000, 10000]
+all_Ks = [5, 10, 25, 50, 100]
+
+beta = [1.0, 2.0, 3.0]           # Fixed effects
+Gamma = 0.2 * I(1 + p)      # Covariance matrix of random-effects
+
+sigma = 1.0     # Standard deviation of the residuals
+
+
+# Vector of all GLMM parameters
+theta = [beta; sqrt.(diag(Gamma)); fill(0, 1 + p)]
+
+
+
+Random.seed!(123)    
+
+
+M = 100        # Number of Monte Carlo simulations
+
+all_theta_hats = []
+
+all_parameter_pairs = vec(collect(Iterators.product(all_Ns, all_Ks)))
+
+@showprogress for (N, K) in all_parameter_pairs
+
+    n = Int(N / K)       # Number of observations per group
+
+    some_theta_hats = []
+
+    for _ in 1:M
+
+        # Generate data
+        all_Xs = [randn(n, p) for _ in 1:K]
+
+        ## Fixed-effects
+        all_eta_vecs_fixed = [beta[1] .+ X * beta[2:end] for X in all_Xs]
+
+        ## Add random effects
+        all_bs = [rand(MvNormal(zeros(1 + p), Gamma)) for _ in 1:K]
+        all_eta_vecs = [eta_vec_fixed .+ fill(b[1], n) .+ X * b[2:end] for (eta_vec_fixed, X, b) in zip(all_eta_vecs_fixed, all_Xs, all_bs)]
+
+        ## Construct Y
+        all_p_vecs = [1 ./ (1 .+ exp.(-eta_vec)) for eta_vec in all_eta_vecs]
+        all_Ys = [rand.(Bernoulli.(p_vec)) for p_vec in all_p_vecs]
+
+        ## Combine data from all groups
+        X = vcat(all_Xs...)
+        Y = vcat(all_Ys...)
+        group = repeat(1:K, inner=n)
+
[…]

### 2024-05-30 scripts/GLMM Asymptotics Exploration/GLMM Asymptotics Exploration - LMM.jl

@@ -1,0 +1,140 @@
+
+using Random, Distributions
+using LinearAlgebra   # For the identity matrix, I()
+using DataFrames, DataFramesMeta        # The latter works like dplyr in R with different verb names
+using MixedModels, GLM
+using ProgressMeter
+
+
+# Set parameters
+
+# N = 1000        # Total sample size
+# K = 5           # Number of groups
+# n = Int(N / K)       # Number of observations per group
+p = 2           # Number of (mixed-effects) covariates
+
+all_Ns = [500, 1000, 2000, 5000, 10000]
+all_Ks = [5, 10, 25, 50, 100]
+
+beta = [1.0, 2.0, 3.0]           # Fixed effects
+Gamma = 0.2 * I(1 + p)      # Covariance matrix of random-effects
+
+sigma = 1.0     # Standard deviation of the residuals
+
+
+# Vector of all GLMM parameters
+theta = [beta; sqrt.(diag(Gamma)); fill(0, 1 + p)]
+
+
+
+Random.seed!(123)    
+
+
+M = 100        # Number of Monte Carlo simulations
+
+all_theta_hats = []
+
+all_parameter_pairs = vec(collect(Iterators.product(all_Ns, all_Ks)))
+
+@showprogress for (N, K) in all_parameter_pairs
+
+    n = Int(N / K)       # Number of observations per group
+
+    some_theta_hats = []
+
+    for _ in 1:M
+
+        # Generate data
+        all_Xs = [randn(n, p) for _ in 1:K]
+
+        ## Fixed-effects
+        all_eta_vecs_fixed = [beta[1] .+ X * beta[2:end] for X in all_Xs]
+
+        ## Add random effects
+        all_bs = [rand(MvNormal(zeros(1 + p), Gamma)) for _ in 1:K]
+        all_eta_vecs = [eta_vec_fixed .+ fill(b[1], n) .+ X * b[2:end] for (eta_vec_fixed, X, b) in zip(all_eta_vecs_fixed, all_Xs, all_bs)]
+
+        ## Construct Y
+        all_Ys = [eta_vec .+ rand(Normal(0, sigma), n) for eta_vec in all_eta_vecs] 
+
+        ## Combine data from all groups
+        X = vcat(all_Xs...)
+        Y = vcat(all_Ys...)
+        group = repeat(1:K, inner=n)
+
+        data = DataFrame(Y = Y, X1 = X[:,1], X2 = X[:,2], group = group)
+
+
+
[…]

### 2024-05-30 presentations/SSC 2024/SSC 2024.tex
Presentation for SSC 2024

\documentclass[14pt]{beamer}

\usetheme{Madrid}

% Add frame numbers
\setbeamertemplate{page number in head/foot}[framenumber]

\usepackage{amsmath, amssymb}
\usepackage{tikz}
\usepackage{bm}
\usepackage{outlines}   % For multilevel lists using the outline environment

\newcommand{\CMA}{Causal Mediation Analysis}
\newcommand{\bE}{\mathbb{E}}
\newcommand{\GLMMs}{Mixed-Effects Models}

\title[]{Statistical Considerations in Multilevel Mediation Analysis}
\author{William Ruth}
\institute[]{Collaborators: Rado Ramasy, Rowin Alfaro, Ariel Mundo, Bruno Remillard, Bouchra Nasri}
\date{\vspace{-3cm}}
\titlegraphic{\includegraphics[width=2cm]{../Logos/CANSSI_Logo.png} \hspace{2cm} \includegraphics[width=4cm]{../Logos/Logo_UdeM-CMJN.jpg}}

\begin{document}

\begin{frame}
    \titlepage
\end{frame}

\begin{frame}{Outline}
    \begin{itemize}
        \setlength{\itemsep}{0.75em}
        \item[1)] The Problem
        \item[2)] Mediation Analysis
        \item[3)] Causal Inference
        \item[4)] Mixed-Effects Models
        \item[234)] Mixed-Effects Models in Causal Mediation Analysis
    \end{itemize}
\end{frame}

\begin{frame}{Example}
    \begin{itemize}
        \item Goal: Understand adherence to restrictive measures
        \begin{itemize}
            \item E.g. Lockdowns
            \item Both past and future \newline
        \end{itemize}
        \item Influence of news source
        \begin{itemize}
            \item How trustworthy? \newline
        \end{itemize}
        \item Disentangle influence on future from influence on past
    \end{itemize}

\end{frame}

\begin{frame}{Example}
        \begin{figure}[H]
            \begin{tikzpicture}
                % Circles with labels
                \node at (0,0) [circle, draw, minimum size=2.5cm] (news) {News Source};
                \node at (4,0) [circle, draw, minimum size=2.5cm] (past) {\shortstack{Past\\Compliance}};
                \node at (8,0) [circle, draw, minimum size=2.5cm] (future) {\shortstack{Future\\Compliance}};

                % Arrows
                \draw[->, line width=1.5pt] (news.east) -- (past.west);
                \draw[->, line width=1.5pt] (past.east) -- (future.west);
                \draw[->, line width=1.5pt, bend left] (news.north) to [out=45,in=135] (future.north);

                \end{tikzpicture}
    \end{figure}
\end{frame}

\begin{frame}{Example}
    Terminology
    \begin{columns}
        \column{0.6\textwidth}
        \begin{itemize}
            \item Top path: {Direct effect}
            \item Center path: {Indirect effect}
            \item Combined: {Total effect}
        \end{itemize}
        \column{0.4\textwidth}
        \begin{itemize}
            \item Exposure: $X$
            \item Outcome: $Y$
            \item Mediator: $M$
        \end{itemize}
    \end{columns}

\end{frame}

\begin{frame}{Mediation Analysis}
    \begin{figure}[H]
        \begin{tikzpicture}
            % Circles with labels
            \node at (0,0) [circle, draw, minimum size=2.5cm] (X) {\shortstack{Exposure\\($X$)}};
            \node at (4,0) [circle, draw, minimum size=2.5cm] (M) {\shortstack{Mediator\\($M$)}};
            \node at (8,0) [circle, draw, minimum size=2.5cm] (Y) {\shortstack{Outcome\\($Y$)}};

            % Arrows
            \draw[->, line width=1.5pt] (X.east) -- (M.west);
            \draw[->, line width=1.5pt] (M.east) -- (Y.west);
            \draw[->, line width=1.5pt, bend left] (X.north) to [out=45,in=135] (Y.north);

            \end{tikzpicture}
    \end{figure}
\end{frame}

\begin{frame}{Mediation Analysis}
    Separate \textbf{Total Effect} of $X$ on $Y$ into
    \begin{itemize}
        \item \textbf{Direct Effect}
        \item \textbf{Indirect Effect}\newline
    \end{itemize}

    Traditionally, use regression
\end{frame}

\begin{frame}{Mediation Analysis}
    Continuous outcome and mediator:
    \begin{itemize}
        \item $Y = \alpha_0 + \alpha_1 M + \alpha_2 X + \varepsilon_Y$
        \item $M = \beta_0 + \beta_1 X + \varepsilon_M$ \newline
    \end{itemize}

    \textbf{Direct Effect}: $\alpha_2$
    \begin{itemize}
        \item ``$X$ in $Y$''
    \end{itemize}
    \textbf{Indirect Effect}: $\alpha_1 \cdot \beta_1$
    \begin{itemize}
        \item ``$M$ in $Y$'' $\cdot$ ``$X$ in $M$''
    \end{itemize}
    \textbf{Total Effect}: $\alpha_2 + \alpha_1 \cdot \beta_1$

\end{frame}

\begin{frame}{Mediation Analysis}
    Popular approach
    \begin{itemize}
        \item A bit outdated\ldots \newline
    \end{itemize}

    More popular: Causal mediation analysis

\end{frame}

\begin{frame}{Causal Inference}
    Assume that $X$ \textit{causes} $Y$\newline

    Counterfactuals:
    \begin{itemize}
        \item What value would $Y$ take if $X$ were set to a particular level?
        \item Write $Y_x$ for the value of $Y$ when $X=x$
        \item If $X\neq x$ then $Y_x$ is literally a ``counterfactual'' 
    \end{itemize}
\end{frame}

\begin{frame}{Causal Inference}
    Example:
    \begin{itemize}
        \item Alice only reads scientific publications and will follow all lockdown mandates
        \item What if she instead only read Facebook?\newline
        \item $Y_{Science}(Alice) = \mathrm{follow}$
        \item $Y_{Facebook}(Alice) = \mathrm{follow}$
    \end{itemize}
\end{frame}

\begin{frame}{Causal Inference}
    Example:
    \begin{itemize}
        \item Bob also only reads scientific publications and will follow all lockdown mandates, but is more susceptible to being influenced \newline
        \item $Y_{Science}(Bob) = \mathrm{follow}$
        \item $Y_{Facebook}(Bob) = \mathrm{not\ follow}$
    \end{itemize}
\end{frame}

\begin{frame}{Causal Inference}
    \begin{itemize}
        \item We only observe one outcome per individual\newline

        \item Explore population-level effects by averaging\newline

        \item Define mediation effects in terms of expected counterfactuals
    \end{itemize}
\end{frame}

\begin{frame}{Causal Inference}
    \textbf{Total Effect}: $\mathbb{E}(Y_{x'} - Y_{x})$
    \begin{itemize}
        \item Effect on outcome when we change exposure from $X=x$ to $X=x'$ \newline
    \end{itemize}

    Other effects involve dependence on a mediator:
    \begin{itemize}
        \item $Y_{xm}$: Value of outcome when
        \begin{itemize}
            \item Exposure ($X$) is set to $x$
            \item Mediator ($M$) is set to $m$
        \end{itemize}
        \item $M_x$: Value of mediator when
        \begin{itemize}
            \item Exposure ($X$) is set to $x$
        \end{itemize}
        \item ``Nested Counterfactuals'': $Y_{x M_x}$ or $Y_{x M_{x'}}$
    \end{itemize}

\end{frame}

\begin{frame}{Causal Mediation Analysis}
    \textbf{Controlled Direct Effect}: $\mathbb{E}(Y_{x'm} - Y_{xm})$
    \begin{itemize}
        \item Effect of changing exposure with mediator held fixed \newline
    \end{itemize}

    \textbf{Natural Direct Effect}: $\mathbb{E}(Y_{x'M_{x}} - Y_{xM_{x}})$
    \begin{itemize}
        \item Effect of changing exposure when we don't interfere with the mediator  \newline
    \end{itemize}

    \textbf{Natural Indirect Effect}: $\mathbb{E}(Y_{xM_{x'}} - Y_{xM_{x}})$
    \begin{itemize}
        \item Effect of changing which exposure value is seen by the mediator while holding fixed which exposure value is seen by the outcome
    \end{itemize}
\end{frame}

\begin{frame}{\CMA}
    In our example
    \begin{itemize}
        \item Controlled Direct Effect: Effect of increasing news trustworthiness if the whole population followed guidelines in the past
        \item Natural Direct Effect: Effect of increasing news trustworthiness independent of any induced change in past compliance
        \item Natural Indirect Effect: Effect of changing past compliance if everyone only got news from Facebook
    \end{itemize}
\end{frame}

\begin{frame}{\CMA}
    We can't measure all required counterfactuals
    \begin{itemize}
        \item E.g., $Y_x$ or $Y_{x'}$, not both \newline
    \end{itemize}

    Expected counterfactuals related to conditional expectations

    \begin{itemize}
        \item     Under strong assumptions, $\bE Y_x = \bE(Y | X=x)$ \newline
    \end{itemize}

    Nested counterfactuals more complicated
    \begin{itemize}
        \item More on this later
    \end{itemize}
\end{frame}

\begin{frame}{\CMA}
    How does causality change our analysis? \newline

    Still fit regression models, but include interaction terms between exposure and mediator
    \begin{itemize}
        \item $Y = \alpha_0 + \alpha_1 M + \alpha_2 X + \alpha_3 M \cdot X + \varepsilon_Y$
        \item $M = \beta_0 + \beta_1 X + \varepsilon_M$ \newline
    \end{itemize}

    Direct and indirect effects now depend on the levels of the exposure

\end{frame}

\begin{frame}{\CMA\ -- Extensions}
    Discussion so far has involved continuous mediator and outcome
    \begin{itemize}
        \item What about binary or categorical? \newline
    \end{itemize}

    Individuals might also be clustered
    \begin{itemize}
        \item E.g. Within countries
    \end{itemize}

\end{frame}

\begin{frame}{\CMA\ -- Extensions}
    Handling binary variables is pretty straightforward
    \begin{outline}
        \1 Instead of linear regression, use logistic regression
        \1 Re-define mediation effects based on expected counterfactuals
            \2 I.e. Counterfactual probabilities
        \1 New formulas for relating mediation effects to regression coefficients \newline
    \end{outline}

    Extend to more than 2 categories using binary indicators
\end{frame}

\begin{frame}{\CMA\ -- Extensions}
    Clustered data more complicated \newline

    Standard approach is multi-level modelling
    \begin{itemize}
        \item I.e. Add random effects which vary across clusters \newline
    \end{itemize}

    Combined with categorical variables: 
    \begin{itemize}
        \item Generalized linear mixed models (GLMMs)
    \end{itemize}
\end{frame}

\begin{frame}{\GLMMs}
    The core idea is to augment our set of covariates
    \begin{itemize}
        \item Coefficients of these new covariates are random variables which vary across groups/clusters \newline
    \end{itemize}

    In the linear setting:
    \begin{itemize}
        \item Old model: $Y = \alpha_0 + \alpha_1 X_1 + \ldots + \alpha_p X_p + \varepsilon$
        \item New model: $Y = \alpha_0 + \alpha_1 X_1 + \ldots + \alpha_p X_p + \mathbf{u_1 Z_1 + \ldots + u_q Z_q} + \varepsilon$ \newline
    \end{itemize}

\end{frame}

\begin{frame}{\GLMMs}
    The $Z$'s are fixed, known covariates
    The $u$'s are random variables
    \begin{itemize}
        \item I.e. Random effects \newline
    \end{itemize}

    It's possible for the $X$'s and $Z$'s to overlap
    \begin{itemize}
        \item The coefficient on such a covariate has the form $\alpha_j + u_{k}$
        \item I.e. Mixed effect
    \end{itemize}

\end{frame}

\begin{frame}{\GLMMs}
    Extend to generalized linear models in the usual way \newline

    Choose response distribution and link function as for ordinary GLMs \newline

    Linear predictor now has a random effects component \newline
\end{frame}

\begin{frame}{\GLMMs}
    Why bother?
    \begin{itemize}
        \item E.g. Measured some but not all levels of a categorical variable
        \item Estimate covariance matrix of random effects
        \item Test for non-zero variance of each random effect \newline
    \end{itemize}

    ``Predict'' level of random effects for each group
    \begin{itemize}
        \item Conditional mean or conditional mode of random effects given response
    \end{itemize}
\end{frame}

\begin{frame}{\GLMMs}
    In our example:
    \begin{itemize}
        \item Data collected from 11 different countries
        \item Explicitly model inter-country variability
        \item Predict country-specific random effects
        \item Use country-specific coefficients in formulas for mediation effects
        \item Test for significant mediation effects within each country \newline
    \end{itemize}
\end{frame}

\begin{frame}{Multilevel Mediation Analysis}
    Uncertainty quantification for mixed-effects models can be challenging  \newline

    Strategies include:
    \begin{itemize}
        \item Bootstrap
        \item Quasi-Bayesian Monte Carlo
        \item $\delta$-method
    \end{itemize}
\end{frame}

\begin{frame}{Multilevel Mediation Analysis}
    Uncertainty quantification for mixed-effects models can be challenging  \newline

    Strategies include:
    \begin{itemize}
        \item \textbf{Bootstrap}
        \item Quasi-Bayesian Monte Carlo
        \item $\delta$-method
    \end{itemize}
\end{frame}

\begin{frame}{Multilevel Mediation Analysis}
    Uncertainty quantification for mixed-effects models can be challenging  \newline

    Strategies include:
    \begin{itemize}
        \item Bootstrap
        \item Quasi-Bayesian Monte Carlo
        \item $\bm{\delta}$\textbf{-method}
    \end{itemize}
\end{frame}

\begin{frame}{Multilevel Mediation Analysis}
    Recall: Mediation effects defined using nested counterfactuals -- $ Y_{xM_{x'}}$
    \begin{itemize}
        \item Value of $Y$ when $X$ is set to $x$, and
        \item $M$ is set to whatever value it would have if $X$ were set to $x'$ \newline
    \end{itemize}

    Under strong assumptions, use the ``Mediation Formula''

\end{frame}

\begin{frame}{Multilevel Mediation Analysis}
    Mediation Formula:
    \begin{align*}
        \action<+->{\bE Y_{xM_{x'}} &= \bE_{M | X=x'} \bE (Y | X=x, M) \\}
    &\action<+->{= \sum_{m=0}^1 \mathbb{P}(Y = 1 | X=x, M=m) \mathbb{P}(M = m | X = x')}
    \end{align*}   
    \action<+->{
    Logistic regression model makes this (relatively) simple:
    \begin{align*}
        & \mathbb{P}(Y = 1 | X=x, M=m) = \mathrm{logit}^{-1}(\mathrm{linear\ predictor})\\
        & \mathbb{P}(M = m | X = x') = \mathrm{logit}^{-1}(\mathrm{different\ linear\ predictor})
    \end{align*}}

\end{frame}

\begin{frame}{Multilevel Mediation Analysis}
    Estimating $\bE Y_{xM_{x'}}$ messy, but not \textit{hard}\newline

    Uncertainty quantification based on asymptotic covariance of regression parameters
    \begin{itemize}
        \item Use $\delta$-method
        \item Need derivatives of $\bE Y_{xM_{x'}}$ wrt regression parameters
        \item Very messy, not particularly hard
    \end{itemize}

\end{frame}

\begin{frame}{Multilevel Mediation Analysis}
    Uncertainty quantification for mediation effects similar
    \begin{itemize}
        \item More $\delta$-method \newline
    \end{itemize}

    Very similar for GLMs and GLMMs
    \begin{itemize}
        \item Latter has more regression parameters
        \item Use \texttt{merDeriv} package to supplement \texttt{lme4}
    \end{itemize}

\end{frame}

\begin{frame}{Putting it All Together}
    \begin{itemize}
        \item Define direct, indirect and total effects using counterfactuals \newline

        \item Estimate these effects across countries using generalized linear mixed models \newline

        \item Compute standard errors for estimated effects using the $\delta$-method

    \end{itemize}

\end{frame}

\begin{frame}{Acknowledgements}
    Collaborators:
    \begin{itemize}
        \item Rado Ramasy
        \item Rowin Alfaro
        \item Ariel Mundo
        \item Bruno Remillard
        \item Bouchra Nasri \newline
    \end{itemize}

    Funding:
    \begin{itemize}
        \item Canadian Statistical Sciences Institute
    \end{itemize}
\end{frame}

\begin{frame}
    \centering
    \Huge Thank You
\end{frame}

\end{document}

### 2024-06-14 notes/Analysis Write-Up/Direct and Indirect Effects.tex
Consolidate notes into one document

@@
 \begin{multline}
-    \bE Y(x, M(x')) = \frac{1}{1 + \exp(-\beta_0 - \beta_m - \beta_x x)} \frac{1}{1 + \exp(-\alpha_0 - \alpha_x x')} +\\
-    \frac{1}{1 + \exp(-\beta_0 - \beta_x x)} \frac{1}{1 + \exp(\alpha_0 + \alpha_x x')} \label{eq:med_bin_fix}
+    \bE Y(x, M(x')) = \frac{1}{1 + \exp(-\beta_0 - \beta_m - \beta_x x)} \cdot \frac{1}{1 + \exp(-\alpha_0 - \alpha_x x')} +\\
+    \frac{1}{1 + \exp(-\beta_0 - \beta_x x)} \cdot \frac{1}{1 + \exp(\alpha_0 + \alpha_x x')} \label{eq:med_bin_fix}
 \end{multline}
@@
 For the mixed-effects models, first write $V = (V_0, V_m, V_x) \sim \mathrm{N}(0, \Sigma_V)$ and $U = (U_0, U_x) \sim \mathrm{N}(0, \Sigma_U)$ for the random-effects in our models for $Y$ and $M$ respectively.
-Next, write $\mathrm{logit}(\bE(Y | V, M=m, X=x)) = \beta_0 + V_0 + (\beta_m + V_m) m + (\beta_x + V_x) x$ and $\mathrm{logit}(\bE(M |U, X=x)) = (\alpha_0 + U_0) + (\alpha_x + U_x) x$.
+Next, write $\mathrm{logit}(\bE(Y | V, M=m, X=x)) = (\beta_0 + V_0) + (\beta_m + V_m) m + (\beta_x + V_x) x$ and $\mathrm{logit}(\bE(M |U, X=x)) = (\alpha_0 + U_0) + (\alpha_x + U_x) x$.
 Returning now to identification of the expected counterfactual for $Y$, we get
@@
 \section{Mediation Effects}
-Denote the expected nested counterfactual defined in (\ref{eq:med_gen}) by $\mathscr{Y}(x, x') = \bE Y(x, M(x'))$.
+Denote the expected nested counterfactual defined in (\ref{eq:med_gen}) by $\sY(x, x') = \bE Y(x, M(x'))$.
 We can define the various mediation effects in terms of expected counterfactuals.
@@
     \label{tab:med_eff_defs}
-    \begin{tabu}{|[1pt]c|c|c|[1pt]}
+    \begin{tabu}{|[1pt]c|[1pt]c|c|[1pt]}
         \tabucline[1pt]{-}
@@
 For consistency with B \& B, we parameterize these as $\tau_0, \tau_m, \tau_x$ for the standard deviations of $V_0, V_m, V_x$, and $\tau_{0,m}, \tau_{0,x}, \tau_{m,x}$ for the corresponding correlations\footnote{While our notation doesn't match that given by B \& B, parameterizing in terms of the standard deviations and correlations does.
-Alternative choices include the variances and covariances, or the unique components of the Cholesky factorizations of $\Sigma_V$ and $\Sigma_U$.}.
+Alternative choices include the variances and covariances, or the unique components of the Cholesky factorizations of $\Sigma_V$ and $\Sigma_U$ \citep{Wan18}.}.
 Similarly, we use $\sigma_0, \sigma_x$ for the standard deviations of $U_0, U_x$, and $\sigma_{0,x}$ for their correlation.
@@
 \textcolor{red}{Either or both models may also contain one or more confounders, $\mathbf{C}$, although we assign these confounders only fixed effects.} Following the algorithm used by the \texttt{lme4} package in \texttt{R} \citep{Bat15, Wal23}, we estimate the parameters in both regression models by maximizing an approximation to the marginal likelihood of the response ($M$ or $Y$).
-This approximation is obtained using Laplace's Method, maximizing the joint likelihood with the random effects using a penalized iteratively re-weighted least squares algorithm.
+This approximation is obtained using Laplace's Method, with a penalized iteratively re-weighted least squares algorithm maximizing to maximize the joint likelihood of the observed data and the random effects.
 Applying this procedure to both of our regression models yields estimates for all components of $\theta$.
@@
 While tedious, this calculation is easily performed with the help of symbolic differentiation software like Maple \citep{Map20}.
-We then evaluate our gradient formula at $\hat{\theta}$, $\left.
-\nabla \sY \right|_{\theta = \hat{\theta}}$, then pre- and post-multiply $\hat{\Sigma}_\theta$ by our estimated gradient.
-If we require the covariance matrix for multiple values of $\sY$ (e.g., $\sY(x,x)$ and $\sY(x,x')$ for some $x \neq x'$), we frame the problem as a vector-valued transformation, and compute the Jacobian by stacking the gradient for each $\sY$.
+Next, we evaluate our gradient formula at $\hat{\theta}$, $\left.
+\nabla \sY \right|_{\theta = \hat{\theta}}$, then pre- and post-multiply $\hat{\Sigma}_\theta$ by this estimated gradient.
+If we require the covariance matrix for multiple values of $\sY$ (e.g., $\sY(x,x)$ and $\sY(x,x')$ for some $x \neq x'$), we frame the problem as a vector-valued transformation, and compute the Jacobian by stacking the gradient for each individual $\sY$.
 More precisely, to get a covariance matrix for $\sY_1, \ldots, \sY_r$, we first construct the $r$-by-$|\theta|$ Jacobian matrix $\mathcal{J} := [\nabla \sY_1, \ldots, \nabla \sY_r]$.
-We then evaluate $\mathcal{J}$ at $\theta = \hat{\theta}$, call the result $\hat{\mathcal{J}}$, and pre- and post-multiply $\hat{\Sigma}_\theta$ by $\hat{\mathcal{J}}$, $\hat{\mathcal{J}} \hat{\Sigma}_\theta \hat{\mathcal{J}}^T$.
+We then evaluate $\mathcal{J}$ at $\theta = \hat{\theta}$, call the result $\hat{\mathcal{J}}$, and pre- and post-multiply $\hat{\Sigma}_\theta$ by $\hat{\mathcal{J}}$, giving $\hat{\mathcal{J}} \hat{\Sigma}_\theta \hat{\mathcal{J}}^T$.
 
@@
 This will give non-zero covariance between parameter estimators from the two models.
+    \item Incorporate covariates/confounders
 \end{itemize}

### 2024-06-14 notes/Consolidated Notes/Consolidated Notes.tex
Consolidate notes into one document

--- 2024-05-24 notes/Literature_Review/Literature Review.tex
@@
-\documentclass{article}
+\documentclass{report}
 
@@
 
+\usepackage{tabu,multirow}
+
 \newcommand{\bG}{\mathbb{G}}
@@
 
-\title{Review of the Literature on Exact Formulas for Mediation Effects}
+\newcommand{\sY}{\mathcal{Y}}
+
+\title{An Update on William's Work}
 
@@
 
+\chapter*{Preface}
+\addcontentsline{toc}{chapter}{Preface}
+
+I have been working on three main areas: a literature review of estimation and UQ for multilevel mediation, implementing the UQ developed by B \& B for the estimated total effect, and expanding their development to address direct and indirect effects.
+In this document, I attempt to consolidate this disparate work in one place.
+Unfortunately, since my implementation was written in RMarkdown, I only overview that work here and refer to an external PDF.
+
+In the interest of minimizing the amount of re-writing necessary, I have organized this document into chapters.
+
+\tableofcontents
+
+\chapter{Literature Review}
+
 \section{Summary and Future Work}
@@
 
-I'm just finishing-up my code for the binary response, binary mediator, mixed effects case. My analysis is running, but the output doesn't look right so I need to do some debugging.
+I'm just finishing-up my code for the binary response, binary mediator, mixed effects case. My analysis is running, but the output doesn't look right, so I need to do some debugging.
 
@@
 \subsubsection{UQ for Group-Level Effects}
-\ref{sec:Imai_UQ}
+\label{sec:Imai_UQ}
 
@@
 
-\bibliography{Lit_Review_Bib}
+\chapter{Implementation}
+
+I don't have a whole lot to say here.
+I implemented the $\delta$-method based uncertainty quantification developed by B \& B for the total mediation effect.
+Please see the RMarkdown document and accompanying PDF called ``William Analysis.Rmd'' and ``William-Analysis.pdf'' for discussion and results.
+
+My findings for the fixed-effects models look pretty reasonable. However, the mixed-effects setting is much less nice. I'm still exploring why this is happening. I hope to have more to say here soon.
+
+\chapter{Extension to Other Mediation Effects}
+
+In this chapter, we develop definitions and formulas for the direct and indirect mediation effects to complement the total effect given by B \& B.
+We follow the work of \citet{Pea12} and the Samoilenko and Lefebvre group (e.g., \citealp{Sam23}).
+
+\section{Expected Nested Counterfactuals}
+\label{sec:nest_CFs}
+
+To start, we define the nested counterfactual $Y(x, M(x'))$ as the value that $Y$ would assume when $X$ is set to $x$ and $M$ is set to whatever value it would have assumed if we had set $X$ to $x'$.
+\citet{Pea12} identifies the expected value of this nested counterfactual with the following expression based on conditional expectations:
+%
+\begin{align}
+    \bE Y(x, M(x')) = \int \bE(Y | M=m, X=x) \bP(M = dm | X = x') \label{eq:med_gen}
+\end{align}
+%
+Focusing now on the case with binary response and binary mediator, \ref{eq:med_gen} becomes
+%
+\begin{multline}
+    \bE Y(x, M(x')) = \bP(Y=1 | M=1, X=x) \bP(M=1 | X=x') + \\\bP(Y=1 | M=0, X=x) \bP(M=0 | X=x') \label{eq:med_bin}
+\end{multline}
+%
+Consider using  logistic regression models for $Y$ and $M$.
+Write $\mathrm{logit}(\bE(Y | M=m, X=x)) = \beta_0 + \beta_m m + \beta_x x$ and $\mathrm{logit}(\bE(M | X=x)) = \alpha_0 + \alpha_x x$.
+Then (\ref{eq:med_bin}) becomes
+%
+\begin{multline}
+    \bE Y(x, M(x')) = \frac{1}{1 + \exp(-\beta_0 - \beta_m - \beta_x x)} \cdot \frac{1}{1 + \exp(-\alpha_0 - \alpha_x x')} +\\
+    \frac{1}{1 + \exp(-\beta_0 - \beta_x x)} \cdot \frac{1}{1 + \exp(\alpha_0 + \alpha_x x')} \label{eq:med_bin_fix}
+\end{multline}
+%
+Equation (\ref{eq:med_bin_fix}) holds for logistic regression with fixed-effects only.
+If we instead use  mixed-effects logistic regressions for $Y$ and $M$, then (\ref{eq:med_gen}) and (\ref{eq:med_bin}) still hold, but (\ref{eq:med_bin_fix}) must be modified.
+For the mixed-effects models, first write $V = (V_0, V_m, V_x) \sim \mathrm{N}(0, \Sigma_V)$ and $U = (U_0, U_x) \sim \mathrm{N}(0, \Sigma_U)$ for the random-effects in our models for $Y$ and $M$ respectively.
+Next, write $\mathrm{logit}(\bE(Y | V, M=m, X=x)) = (\beta_0 + V_0) + (\beta_m + V_m) m + (\beta_x + V_x) x$ and $\mathrm{logit}(\bE(M |U, X=x)) = (\alpha_0 + U_0) + (\alpha_x + U_x) x$.
+Returning now to identification of the expected counterfactual for $Y$, we get
+%
+\begin{align}
+    \bE Y(x, M(x')) =& \left[\int \bP(Y=1 |V=v, M=1, X=x) \bP(V = dv)\right] \cdot \\
+    & \left[\int \bP(M=1 |U=u, X=x') \bP(U=du)\right] +\\
+     & \left[\int \bP(Y=1 |V=v, M=0, X=x) \bP(V = dv)\right] \cdot \\
+     & \left[\int \bP(M=0 |U=u, X=x') \bP(U=du)\right]
+\end{align}
+%
+and, in the logistic regression context,
+%
+\begin{align}
+    \bE Y(x, M(x')) =& \left[ \int \frac{\phi(v; 0, \Sigma_V)}{1 + \exp(-(\beta_0 + v_0) - (\beta_m + v_m) - (\beta_x + v_x) x)}  dv \right] \cdot \label{eq:med_bin_ran}\\
+    &\left[ \int \frac{\phi(u; 0, \Sigma_U)}{1 + \exp(-(\alpha_0 + u_0) - (\alpha_x + u_x) x')}  du \right] + \nonumber\\
+    & \left[\int \frac{\phi(v; 0, \Sigma_V)}{1 + \exp(-(\beta_0 + v_0) - (\beta_x + v_x) x)} dv \right] \cdot \nonumber\\
+    & \left[\int \frac{\phi(u; 0, \Sigma_U)}{1 + \exp((\alpha_0 + u_0) + (\alpha_x + u_x) x')} du \right], \nonumber
+\end{align}
+%
+where $\phi(.; \mu, \Sigma)$ is the multivariate normal density with mean $\mu$ and covariance matrix $\Sigma$.
+
+Note that the four integrals in (\ref{eq:med_bin_ran}) are all multivariate, but can be transformed to univariate integrals by suitable changes of variables.
+To this end, write $\eta = \alpha_0 + \alpha_x x$ and $\zeta = \beta_0 + \beta_x x$ for two linear predictors (note that $\zeta$ does not contain $\beta_m$), and $\gamma^2_\Sigma(c_1, \ldots, c_r) = (c_1, \ldots, c_r) \Sigma (c_1, \ldots, c_r)^T$, where $\Sigma$ is an $r$-by-$r$ covariance matrix.
+We will generally set $\Sigma = \Sigma_V$ or $\Sigma = \Sigma_U$, in which case we write $\gamma^2_V$ or $\gamma^2_U$ respectively.
+We now define the function $\psi$ as a template for the four integrals in (\ref{eq:med_bin_ran}).
+%
+$$
+\psi(\mu, \sigma^2) := \int \frac{\phi(z; 0, 1)}{1 + \exp(-\mu - \sigma z)} dz.
+$$
+%
+Note that $\psi$ is a univariate integral, so we can expect it to be well-approximated by routine numerical quadrature techniques. We now re-write (\ref{eq:med_bin_ran}) in terms of $\psi$ as follows,
+%
+\begin{align}
+    \bE Y(x, M(x')) =& \psi(\zeta + \beta_m, \gamma^2_V(1, 1, x)) \cdot \psi(\eta, \gamma^2_U(1, x)) +\\
+    & \psi(\zeta, \gamma^2_V(1, 0, x)) \cdot \psi(-\eta, \gamma^2_U(1, x)). \nonumber 
+\end{align}
+
+\section{Mediation Effects}
+Denote the expected nested counterfactual defined in (\ref{eq:med_gen}) by $\sY(x, x') = \bE Y(x, M(x'))$.
+We can define the various mediation effects in terms of expected counterfactuals.
+Note that mediation effects for a binary outcome are commonly defined on three different scales: risk difference, risk ratio and odds ratio.
+Table \ref{tab:med_eff_defs} gives all such definitions explicitly.
+In the rest of this section, we outline the processes of estimation and uncertainty quantification for the mediation effects defined in Table \ref{tab:med_eff_defs}.
+
+%
+\begin{table}[ht]
+    \centering
+    \caption{Definitions of various mediation effects; $x$ and $x'$ denote different values of the exposure.}
+    \label{tab:med_eff_defs}
+    \begin{tabu}{|[1pt]c|[1pt]c|c|[1pt]}
+        \tabucline[1pt]{-}
+        \multirow{3}{*}{Risk Difference} & Total Effect & $\sY(x, x) - \sY(x', x')$ \\
+        \cline{2-3}
+        & Direct Effect & $\sY(x, x') - \sY(x', x')$\\
+        \cline{2-3}
+        & Indirect Effect & $\sY(x, x) - \sY(x, x')$ \\
+        \tabucline[1pt]{-}
+        \multirow{3}{*}{Risk Ratio} & Total Effect & $\sY(x, x) / \sY(x', x')$ \\
+        \cline{2-3}
+        & Direct Effect & $\sY(x, x') / \sY(x', x')$\\
+        \cline{2-3}
+        & Indirect Effect & $\sY(x, x) / \sY(x, x')$ \\
+        \tabucline[1pt]{-}
+        \multirow{3}{*}{Odds Ratio} & Total Effect & $\left. \frac{\sY(x, x)}{1 - \sY(x, x)} \right/ \frac{\sY(x', x')}{1 - \sY(x', x')} $\\
+        \cline{2-3}
+        & Direct Effect & $\left. \frac{\sY(x, x')}{1 - \sY(x, x')} \right/ \frac{\sY(x', x')}{1 - \sY(x', x')} $\\
+        \cline{2-3}
+        & Indirect Effect & $\left. \frac{\sY(x, x)}{1 - \sY(x, x)} \right/ \frac{\sY(x, x')}{1 - \sY(x, x')} $\\
+        \tabucline[1pt]{-}
+    \end{tabu}
+\end{table}
+
+\subsection{Estimation}
+
+Mediation effects are all defined in terms of the expected nested counterfactuals, $\sY(x, x')$.
+As such, the estimation of any mediation effect given in Table \ref{tab:med_eff_defs} centers around estimation of $\sY$.
+To this end, write $\theta$ for all parameters upon which $\sY$ depends.
+That is, $\theta$ contains both sets of regression coefficients, $\beta_0, \beta_m, \beta_x$ and $\alpha_0, \alpha_x$, as well as both sets of covariance parameters.
+For consistency with B \& B, we parameterize these as $\tau_0, \tau_m, \tau_x$ for the standard deviations of $V_0, V_m, V_x$, and $\tau_{0,m}, \tau_{0,x}, \tau_{m,x}$ for the corresponding correlations\footnote{While our notation doesn't match that given by B \& B, parameterizing in terms of the standard deviations and correlations does.
+Alternative choices include the variances and covariances, or the unique components of the Cholesky factorizations of $\Sigma_V$ and $\Sigma_U$ \citep{Wan18}.}.
+Similarly, we use $\sigma_0, \sigma_x$ for the standard deviations of $U_0, U_x$, and $\sigma_{0,x}$ for their correlation.
+We are now equipped to write-out $\theta$ in full.
+The order of parameters is chosen to match my code (and to avoid refactoring thereof).
+$$
+\theta = (\alpha_0, \alpha_x, \sigma_0, \sigma_x, \sigma_{0,x}, \beta_0, \beta_m, \beta_x, \tau_0, \tau_m, \tau_x, \tau_{0,m}, \tau_{0,x}, \tau_{m,x})
+$$
+%
+In order to estimate $\theta$, we fit two mixed-effects logistic regression models.
+The first model predicts the mediator, $M$, using the exposure, $X$, while the second predicts the outcome, $Y$, using $M$ and $X$.
+Both models also contain intercepts.
+We include random effects for all regression coefficients in both models.
+\textcolor{red}{Either or both models may also contain one or more confounders, $\mathbf{C}$, although we assign these confounders only fixed effects.} Following the algorithm used by the \texttt{lme4} package in \texttt{R} \citep{Bat15, Wal23}, we estimate the parameters in both regression models by maximizing an approximation to the marginal likelihood of the response ($M$ or $Y$).
+This approximation is obtained using Laplace's Method, with a penalized iteratively re-weighted least squares algorithm maximizing to maximize the joint likelihood of the observed data and the random effects.
+Applying this procedure to both of our regression models yields estimates for all components of $\theta$.
+
+Combining our estimate of $\theta$ with Equation \ref{eq:med_bin_ran} from Section \ref{sec:nest_CFs} and the formulas given in Table \ref{tab:med_eff_defs}, we are now able to estimate all mediation effects on the three scales we consider.
+It remains however, to address uncertainty quantification.
+
+\subsection{Uncertainty Quantification}
+
+As mentioned above, all mediation effects are defined in terms of expected nested counterfactuals $\sY$, which themselves depend on the parameter vector $\theta$.
+Quantification of uncertainty for our estimator of a mediation effect thus proceeds in three stages.
+First, we obtain a covariance matrix for our estimator of $\theta$.
+Second, we apply the $\delta$-method to get a joint covariance matrix for all values of $\sY$ required to evaluate the mediation effect of interest.
+Finally, we use the $\delta$-method again to translate uncertainty in the $\sY$s to uncertainty in the mediation effect.
+In fact, we can obtain the covariance matrix for all three mediation effects defined on a particular scale (e.g., risk ratio), or, indeed, between all nine mediation effects given in Table \ref{tab:med_eff_defs}.
+Note that the variances and covariances discussed above are asymptotic.
+
+We now proceed with the three steps discussed above, starting with an asymptotic covariance matrix for $\theta$.
+\textcolor{red}{For now, we treat the two regression models as independent.
+That is, we assume that the covariance between all parameter estimators in our $M$ model and all parameter estimators in our $Y$ model are zero.
+This is likely not true; call it a working assumption.
+See \citet{Bau06} for a method to model this inter-model dependence.} This step is fairly straightforward, since both regression models are fit using (approximate) maximum likelihood.
+We simply compute the negative Hessian of the marginal log-likelihood for the observed data in both models, and stack the results into a block-diagonal matrix.
+Call the resulting covariance matrix $\hat{\Sigma}_\theta$.
+This computation is performed by the \texttt{merDeriv} package in \texttt{R} \citep{Wan18}.
+
+To quantify uncertainty in $\sY$ based on our estimator for $\theta$, we use the $\delta$-method \citep[see, e.g., Chapter 3 of][]{vdV98}.
+Briefly, given the asymptotic covariance of our estimator for $\theta$, $\sqrt{n}(\hat{\theta} - \theta) \rightsquigarrow \mathrm{N}(0, \Sigma)$, the asymptotic covariance of some function of our estimator can be obtained by differentiating that function and multiplying by the limiting distribution.
+That is, $\sqrt{n}[f(\hat{\theta}) - f(\theta)] \rightsquigarrow \nabla f(\theta)\mathrm{N}(0, \Sigma) \overset{d}{=} \mathrm{N}(0, \nabla f(\theta) \Sigma \nabla f(\theta)^T)$.
+We estimate this limiting covariance by $\nabla f(\hat{\theta}) \hat{\Sigma} \nabla f(\hat{\theta})^T$, where typically $\Sigma = \Sigma(\theta)$ and $\hat{\Sigma} = \Sigma(\hat{\theta})$.
+
+Returning now to our problem, in order to apply the $\delta$-method we need the gradient of $\sY$ with respect to $\theta$.
+While tedious, this calculation is easily performed with the help of symbolic differentiation software like Maple \citep{Map20}.
+Next, we evaluate our gradient formula at $\hat{\theta}$, $\left.
+\nabla \sY \right|_{\theta = \hat{\theta}}$, then pre- and post-multiply $\hat{\Sigma}_\theta$ by this estimated gradient.
+If we require the covariance matrix for multiple values of $\sY$ (e.g., $\sY(x,x)$ and $\sY(x,x')$ for some $x \neq x'$), we frame the problem as a vector-valued transformation, and compute the Jacobian by stacking the gradient for each individual $\sY$.
+More precisely, to get a covariance matrix for $\sY_1, \ldots, \sY_r$, we first construct the $r$-by-$|\theta|$ Jacobian matrix $\mathcal{J} := [\nabla \sY_1, \ldots, \nabla \sY_r]$.
+We then evaluate $\mathcal{J}$ at $\theta = \hat{\theta}$, call the result $\hat{\mathcal{J}}$, and pre- and post-multiply $\hat{\Sigma}_\theta$ by $\hat{\mathcal{J}}$, giving $\hat{\mathcal{J}} \hat{\Sigma}_\theta \hat{\mathcal{J}}^T$.
+
+Finally, to get the asymptotic variance of one or more estimated mediation effects, we just apply the $\delta$-method to the corresponding formulas from Table \ref{tab:med_eff_defs}.
+Since the mediation effects are all simple functions of the $\sY$, any gradients required for this step are easily obtained.
+
+\section{To Do}
+\begin{itemize}
+    \item Incorporate an interaction term between $X$ and $M$ in the model for $Y$.
+    \item Explore dependence between the models for $M$ and $Y$ using the `stacking' technique described in \citet{Bau06}.
+This will give non-zero covariance between parameter estimators from the two models.
+    \item Incorporate covariates/confounders
+\end{itemize}
+
+\bibliography{../Literature_Review/Lit_Review_Bib}
 

### 2024-06-14 notes/William Analysis.Rmd
Consolidate notes into one document

@@
 # Number of clusters
-all_Ks = c(5, 20, 50)
+all_Ks = c(5, 20, 50, 100)
 # K=5
@@
 
+library(doParallel)
+library(pbapply)
+
 tic()
 
-set.seed(2222)
+set.seed(1)
 
@@
 
+# Initialize cluster on my machine ----
+# n_cores = 2
+# n_cores = 10
+n_cores = parallel::detectCores() - 1
+my_cluster = makeCluster(n_cores, type = "PSOCK")
+registerDoParallel(my_cluster)
+clusterEvalQ(my_cluster,{
+  library(MASS)
+  library(lme4)
+  library(merDeriv)
+  source("../src/Helpers.R")
+})
+clusterExport(my_cluster, c("N_K_combs", "num_reps", "a_0", "a_1", "A_2", "a", "b_0", "b_1", "b_2", "B_3", "b", "Sigma_a", "theta", "Sigma_b", "gamma", "p_conf", "x_pred", "W_pred"))
+clusterSetRNGStream(my_cluster, iseed = 1)
+
+# output_bin_bin_ran = pblapply(seq_len(nrow(N_K_combs)), function(j){
+# output_bin_bin_ran = pblapply(nrow(N_K_combs), function(j){
 for(j in seq_len(nrow(N_K_combs))){
@@
 
-  # Containers for output with this value of n
-  some_a_hats = list()
-  some_theta_hats = list()
-  some_b_hats = list()
-  some_gamma_hats = list()
-
-  some_M_covs = list()
-  some_Y_covs = list()
-
-  some_med_hats = list()
-  some_med_SEs = list()
-
-  for(i in 1:num_reps){
-    print(paste0("n = ", n, ", K = ", K, ", i = ", i, " of ", num_reps))
+  clusterExport(my_cluster, c("n", "K"))
+
+  print(paste0("j = ", j, " of ", nrow(N_K_combs),  " (n=", n, ", K=", K, ")"))
+
+  # # Containers for output with this value of n
+  # some_a_hats = list()
+  # some_theta_hats = list()
+  # some_b_hats = list()
+  # some_gamma_hats = list()
+  # 
+  # some_M_covs = list()
+  # some_Y_covs = list()
+  # 
+  # some_med_hats = list()
+  # some_med_SEs = list()
+
+  # for(i in 1:num_reps){
+  some_output = pblapply(seq_len(num_reps), function(i){
+    # print(paste0("n = ", n, ", K = ", K, ", i = ", i, " of ", num_reps))
+
+    this_output = NULL
 
     # Use tryCatch to omit any datasets leading to convergence issues
-    tryCatch({
+    tryCatch({                                                                    ############## Un-quote
       all_Xs = list()
@@
 
-      some_a_hats[[i]] = a_hat
-      some_theta_hats[[i]] = theta_hat
-      some_M_covs[[i]] = M_cov
+      # some_a_hats[[i]] = a_hat
+      # some_theta_hats[[i]] = theta_hat
+      # some_M_covs[[i]] = M_cov
 
@@
 
-      some_b_hats[[i]] = b_hat
-      some_gamma_hats[[i]] = gamma_hat
-      some_Y_covs[[i]] = Y_cov
+      # some_b_hats[[i]] = b_hat
+      # some_gamma_hats[[i]] = gamma_hat
+      # some_Y_covs[[i]] = Y_cov
 
@@
       med_hat = Phi(eta_hat, zeta_hat, a_x_hat, b_m_hat, b_x_hat, sigma_M2, sigma_Y2, sigma_M1, sigma_Y1)
-      some_med_hats[[i]] = med_hat
+      # some_med_hats[[i]] = med_hat
 
@@
 
-      some_med_SEs[[i]] = med_SE
-    }, error = function(e){})
+      # some_med_SEs[[i]] = med_SE
+
+      this_output = list(a_hat = a_hat, theta_hat = theta_hat, b_hat = b_hat, gamma_hat = gamma_hat, M_cov = M_cov, Y_cov = Y_cov, med_hat = med_hat, med_SE = med_SE)
+    # }, error = function(e){grad_Phi_obs <<- e}) # For troubleshooting parallel errors
+    }, error = function(e){})                                                       ############ Un-quote
+
+    return(this_output)
+
+  }, cl = my_cluster)
+  # })
+
+  # Store output for current values of n and K
+  all_a_hats[[j]] = lapply(some_output, function(x) x$a_hat)
+  all_theta_hats[[j]] = lapply(some_output, function(x) x$theta_hat)
+  all_b_hats[[j]] = lapply(some_output, function(x) x$b_hat)
+  all_gamma_hats[[j]] = lapply(some_output, function(x) x$gamma_hat)
+  all_M_covs[[j]] = lapply(some_output, function(x) x$M_cov)
+  all_Y_covs[[j]] = lapply(some_output, function(x) x$Y_cov)
+  all_med_hats[[j]] = lapply(some_output, function(x) x$med_hat)
+  all_med_SEs[[j]] = lapply(some_output, function(x) x$med_SE)
 }
 
-  # Store output for current value of n
-  all_a_hats[[j]] = some_a_hats
-  all_theta_hats[[j]] = some_theta_hats
-  all_b_hats[[j]] = some_b_hats
-  all_gamma_hats[[j]] = some_gamma_hats
-
-  all_M_covs[[j]] = some_M_covs
-  all_Y_covs[[j]] = some_Y_covs
-
-  all_med_hats[[j]] = some_med_hats
-  all_med_SEs[[j]] = some_med_SEs
-}
-toc()
+runtime = toc()
+
+parallel::stopCluster(my_cluster)
 
 # Save output
-# save(all_a_hats, all_theta_hats, all_b_hats, all_gamma_hats, all_M_covs, all_Y_covs, all_med_hats, all_med_SEs, file = "William-Analysis_cache/MC-bin-bin-mix-n=200,500,1000.RData")
+# save(all_a_hats, all_theta_hats, all_b_hats, all_gamma_hats, all_M_covs, all_Y_covs, all_med_hats, all_med_SEs, runtime, file = "William-Analysis_cache/MC-bin-bin-mix.RData")
 # load("William-Analysis_cache/MC-bin-bin-mix.RData")
@@
 results
-```
+
+```

### 2024-06-18 notes/William Analysis.Rmd
Update consolidated notes

@@
 library(tictoc)
+library(ggplot2)
+library(magrittr)
+library(dplyr)
 
@@
 num_reps = 200
-```
-
-```{r MC-bin-bin-mix, include=FALSE, cache=TRUE}
+
+# Easier version of problem
+all_Ns = c(200, 500)
+all_Ks = c(5, 20)
+
+sigma_a_0 = 0.5
+sigma_a_1 = 0.5 * abs(a_1)
+cor_a0_a1 = 0.5
+cov_a0_a1 = sigma_a_0 * sigma_a_1 * cor_a0_a1
+theta = c(sigma_a_0, sigma_a_1, cor_a0_a1)
+Sigma_a = matrix(c(sigma_a_0^2, cov_a0_a1, cov_a0_a1, sigma_a_1^2), nrow = 2, ncol = 2)
+
+sigma_b_0 = 0.5 * abs(b_0)
+sigma_b_1 = 0.5 * abs(b_1)
+cor_b0_b1 = 0.5
+cov_b0_b1 = sigma_b_0 * sigma_b_1 * cor_b0_b1
+gamma = c(sigma_b_0, sigma_b_1, cor_b0_b1)
+Sigma_b = matrix(c(sigma_b_0^2, cov_b0_b1, cov_b0_b1, sigma_b_1^2), nrow = 2, ncol = 2)
+```
+
+This simulation takes a while. As such, I have saved the results and don't re-run things every time I compile this document. To re-simulate, un-comment the following code chunk.
+
+```{r MC-bin-bin-mix, include=FALSE}
 # , dependson=c("boilerplate")}
@@
   # Store output for current values of n and K
+
   all_a_hats[[j]] = lapply(some_output, function(x) x$a_hat)
@@
 # Save output
-# save(all_a_hats, all_theta_hats, all_b_hats, all_gamma_hats, all_M_covs, all_Y_covs, all_med_hats, all_med_SEs, runtime, file = "William-Analysis_cache/MC-bin-bin-mix.RData")
-# load("William-Analysis_cache/MC-bin-bin-mix.RData")
+# save(N_K_combs, all_a_hats, all_theta_hats, all_b_hats, all_gamma_hats, all_M_covs, all_Y_covs, all_med_hats, all_med_SEs, runtime, file = "William-Analysis_cache/MC-bin-bin-mix.RData")
+# load("D:/OneDrive/OneDrive - Simon Fraser University (1sfu)/Research/Postdoc/Exact_Asymptotics/notes/William-Analysis_cache/MC-bin-bin-mix.RData", verbose = TRUE)
 ```
@@
 We give the mean and median of the estimated standard errors, as well as their relative errors (in $\%$) as estimates of the empirical standard error.
-See Table \ref{tab:results_bin_bin_fix}.
+See Table \ref{tab:results_bin_bin_mix}.
 
@@
 # Combine into a table
-# results = data.frame(Empirical = emp_SEs,
-#                      Mean = mean_delta_SEs, Median = median_delta_SEs,
-#                      "Percent_Err_Mean" = 100*(mean_delta_SEs - emp_SEs)/emp_SEs,
-#                      "Percent_Err_Median" = 100*(median_delta_SEs - emp_SEs)/emp_SEs)
-# results = data.frame(n = all_Ns, Empirical = emp_SEs,
-#                      Mean = mean_delta_SEs, Median = median_delta_SEs,
-#                      "Percent_Err_Mean" = 100*(mean_delta_SEs - emp_SEs)/emp_SEs,
-#                      "Percent_Err_Median" = 100*(median_delta_SEs - emp_SEs)/emp_SEs)
 results = data.frame(n = N_K_combs$N, K = N_K_combs$K, Empirical = emp_SEs,
@@
 
-```
+# Add extra values of n and K to list
+# all_N_combs = c(N_K_combs$N)
+# all_K_combs = c(N_K_combs$K)
+# results = data.frame(n = all_N_combs, K = all_K_combs, Empirical = emp_SEs,
+#                      Mean = mean_delta_SEs, Median = median_delta_SEs,
+#                      "Percent_Err_Mean" = 100*(mean_delta_SEs - emp_SEs)/emp_SEs,
+#                      "Percent_Err_Median" = 100*(median_delta_SEs - emp_SEs)/emp_SEs)
+
+```
+
+<!-- I'm now trying to figure out what's causing an error which only occurs on knit. Trying to isolate it in the following code chunk: -->
+
+<!-- ```{r} -->
+<!-- load("D:/OneDrive/OneDrive - Simon Fraser University (1sfu)/Research/Postdoc/Exact_Asymptotics/notes/William-Analysis_cache/MC-bin-bin-mix.RData", verbose = TRUE) -->
+
+<!-- # print(all_med_hats) -->
+<!-- ``` -->
 
@@
 ## Print table
-kbl(results, format = "latex", booktabs = T, caption = "Summary of SE estimates for mediation effect under continuous response, continuous mediator, fixed-effects model \\label{tab:results_bin_bin_fix}")
-```
-
-I don't really understand the results I'm getting here. It should help if I can compare my estimates of the mediation effect with the true value.
-
-```{r}
-load("../data/Exact_Asymptotics/Bin-Bin-Mix_Results.RData")
-
-true_med = Phi_of_xi(a, theta, b, gamma, x_pred, W_pred)
-
-load("William-Analysis_cache/latex/unnamed-chunk-37_44bd8d1152289e689fbff9185d17042c.RData", ex <- new.env())
-ls.str(ex)
-```
-
-Investigate normality of estimators
-```{r}
-library(ggplot2)
-library(magrittr)
-
+kbl(results, format = "latex", booktabs = T, caption = "Summary of SE estimates for mediation effect under continuous response, continuous mediator, fixed-effects model \\label{tab:results_bin_bin_mix}")
+```
+
+<!-- I don't really understand the results I'm getting here. It should help if I can compare my estimates of the mediation effect with the true value. -->
+
+<!-- ```{r} -->
+<!-- load("../data/Exact_Asymptotics/Bin-Bin-Mix_Results.RData") -->
+
+<!-- true_med = Phi_of_xi(a, theta, b, gamma, x_pred, W_pred) -->
+
+<!-- load("William-Analysis_cache/latex/unnamed-chunk-37_44bd8d1152289e689fbff9185d17042c.RData", ex <- new.env()) -->
+<!-- ``` -->
+
+There's lots that I don't understand about the above results.
+One think I decided to explore is the normality of the estimators.
+To this end, I make some density plots, and also perform Shapiro-Wilk tests for normality.
+See Table \ref{tab:results_bin_bin_mix-Shapiro} for the test results.
+
+```{r}
 med_hats_to_plot = unlist(data_med_hats)
@@
 data_med_hat_to_plot %<>% dplyr::filter(med_hat < 4)
-
+```
+
+```{r}
 ggplot(data_med_hat_to_plot, aes(x = med_hat)) +
-  # geom_histogram(bins = 30) +
-  geom_density() +   theme_minimal() +
-  facet_grid(rows = vars(N), cols = vars(K), scales = "free")
-
+  geom_density() +
+  facet_grid(rows = vars(N), cols = vars(K), scales = "free") +
+  ggtitle("Density of mediation effect estimates")
+  # theme_minimal()
+```
+
+```{r}
 # QQ plot for each sample size
@@
   stat_qq_line() +
-  facet_grid(rows = vars(N), cols = vars(K), scales = "free")
-  theme_minimal()
-
+  facet_grid(rows = vars(N), cols = vars(K), scales = "free") +
+  ggtitle("QQ plot of mediation effect estimates")
+  # theme_minimal()
+```
+
+```{r}
 # Test normality in each sample size
@@
 results$Shapiro_p = all_norm_p_vals
-results
-
-```
+# results
+
+# # Check how p-value varies with N and K
+# p_vals_by_N = results %>% group_by(n) %>% summarize(mean_p = mean(Shapiro_p), sd_p = sd(Shapiro_p))
+# p_vals_by_K = results %>% group_by(K) %>% summarize(mean_p = mean(Shapiro_p), sd_p = sd(Shapiro_p))
+
+```
+
+```{r results="asis", echo=FALSE}
+## Print table
+kbl(results, format = "latex", booktabs = T, caption = "Summary of SE estimates for mediation effect under continuous response, continuous mediator, fixed-effects model \\label{tab:results_bin_bin_mix-Shapiro}")
+```

### 2024-06-19 notes/William Analysis.Rmd
Start forming notes into a manuscript

@@
 all_Ks = c(5, 20)
+N_K_combs = expand.grid(N = all_Ns, K = all_Ks)
 
@@
 # n_cores = 2
-# n_cores = 10
-n_cores = parallel::detectCores() - 1
+n_cores = 10
+# n_cores = parallel::detectCores() - 1
 my_cluster = makeCluster(n_cores, type = "PSOCK")

### 2024-06-19 papers/Manuscript - Exact Asymptotics.tex
Start forming notes into a manuscript

--- 2024-05-24 notes/Literature_Review/Literature Review.tex
@@
 \usepackage{bm}             % For bold math symbols
+\usepackage{soul}
 
@@
 
-\title{Review of the Literature on Exact Formulas for Mediation Effects}
+\title{Analytical Uncertainty Quantification for Multilevel Mediation Analysis}
 
@@
 \maketitle
+
+\section{Introduction}
+
+Mediation analysis is a central problem in modern causal inference.
+Many scientific and public health questions have the form of separating the direct effect of some exposure on an outcome from the indirect effect of that exposure via a mediator.
+Many authors have developed methods to address this problem; gradually increasing in complexity.
+Early work by \citet{Bar86} laid the foundation for future developments, but did not use the machinery of counterfactual outcomes.
+
+There have been several approaches to the analysis of causal mediation.
+One group established a non-parametric identification result \citep{Ima10II}, and used this result to estimate mediation effects in various contexts \citep{Ima10I, Ima11}.
+Their methodology is implemented in the \texttt{R} package \texttt{mediation} \citep{Tin14}.
+Note that this group only estimates mediation effects as expected differences in counterfactuals, so if the outcome is binary then mediation effects are only available on the risk-difference scale.
+
+Another approach began with the work of \citet{Van09} on continuous outcomes, and was later extended to handle various modifications to the basic model \citep{Van10, Van13, Van14}.
+Of particular interest to us is the modification to handle binary outcomes \citep{Van10}.
+Effects are defined on the odds-ratio scale, and the outcome is assumed to be rare.
+Later work by \citet{Sam21} removes the rare-outcome assumption and extends the work of \citet{Van09} to handle effects on the risk-difference, risk-ratio and odds-ratio scales.
+See also \citet{Sam18,Sam23} for more details.
+
+Briefly, causal mediation analysis is based on the counterfactual, or potential outcome framework.
+Let $Y$ be an outcome of interest and $X$ be an exposure.
+We write $Y(x)$ for the value $Y$ would have attained if, possibly counter to fact, $X$ had been set to the value $x$.
+Introducing a mediator, $M$, we write $M(x)$ for the value $M$ would have attained if $X=x$, and $Y(x, m)$ for the value of $Y$ when $X=x$ and $M=m$.
+Note that every individual in the population has values for each of the above quantities, $Y(x)$, $M(x)$, $Y(x,m)$ at every possible value of $x$ and $m$.
+Unfortunately, in practice we only observe $Y$ and $M$ for the values of $x$ and $m$ which actually occurred.
+This is known as the ``fundamental problem of causal inference'' \citep{Din18, Hol86}.
+
+The standard approach to solving this fundamental problem is to avoid estimating individual-level counterfactuals and instead estimate population averages thereof.
+Under standard assumptions, such as consistency and no unmeasured confounders \citep[see, e.g.,][]{Pea09}, we can estimate expected conterfactuals as functions of conditional expectations.
+From this point, mediation analysis reduces to a problem of classical statistics; one which can be solved using traditional regression methodology.
+The difference between continuous and binary outcomes (or mediators) is essentially addressed by choosing between linear and logistic regression.
+Other data types (e.g., count, survival), or more flexible relationships (e.g., splines), can also be incorporated by selecting the appropriate regression methodology.
+
+One extension which is of particular interest is to dependent data via multilevel, or mixed-effects, models.
+Mixed-effects regression models involve the introduction of random, unobserved coefficients to an existing regression \hl{(awk)}.
+A common setting in which such a model arises is clustered data, where the random coefficients differ across clusters, but are constant within each cluster.
+See, e.g., \citet{Dem04} for an overview of mixed-effects methodology.
+When applied to mediation analysis, mixed-effects methods allow for modelling group-specific mediation effects, and the effect of this heterogeneity on the estimation of global effects.
+
+In this paper, we present a general framework for multilevel mediation analysis based on the estimation of nested counterfactuals.
+In particular, our method can be applied to estimate mediation effects on whatever scale is of interest (e.g., risk difference, risk ration, odds ratio).
+
+\section{Multilevel Mediation Analysis}
+
+In this section, we define the total, direct and indirect mediation effects in terms of nested counterfactuals.
+We begin by applying the Mediation Formula of \citet{Pea12} to identify nested counterfactuals with simple functions of conditional expectations.
+
+We begin by observing that mediation effects are often defined in terms of nested counterfactuals (regardless of the scale on which these effects are reported).
+\hl{There is some causal inference theory to be done here.
+For now, I'm just going to write what I expect to be true, then later I will go back and add the necessary assumptions.} As such, we begin by identifying a general nested counterfactual with estimable quantities.
+To this end, write $Y(x, m)$ for the counterfactual value of $Y$ when $X$ and $M$ are set to $x$ and $m$ respectively.
+Similarly, write $M(x)$ for the counterfactual value of $M$ when $X=x$.
+We write $Y(x, M(x'))$ for the value of $Y$ when $X$ is set to $x$ and $M$ is set to what it would have been if $X$ were $x'$.
+We refer to $Y(x, M(x'))$ as a nested counterfactual.
+Note that, when $x \neq x'$, the nested counterfactual is necessarily unobservable.
+Nevertheless, given regularity conditions, we can write
+%
+\begin{equation}
+    \bE Y(x, M(x')) = \int \bE(Y | M=m, X=x) \bP(M = dm | X = x') \label{eq:id_nest}
+\end{equation}
+%
+A similar expression holds conditional on pre-treatment covariates,
+\begin{equation}
+    \bE Y_c(x, M(x')) = \int \bE(Y | M=m, X=x, C=c) \bP(M = dm | X = x', C=c) \label{eq:id_nest_cov}
+\end{equation}
+%
+Having identified expected nested counterfactuals with integrals of conditional quantities, we can use regression to estimate the RHS of Equations \ref{eq:id_nest} and \ref{eq:id_nest_cov}.
+This identification with regression-based quantities is especially simple if $M$ is binary.
+Here, we get $\bP(M=m | \cdot) = \bE(M=m | \cdot)$, where the latter quantity is popularly modelled with logistic regression.
+For simplicity, we hereafter assume that the mediator, $M$, is binary.
+\hl{If not, we need to do something different.
+One option is to extract the conditional density of $(M | \cdot)$, then perform the integral.
+Another is to do regression with response $\bE(Y | M=m, X=x, C=c)$, viewed as a function of $m$.
+There is likely an equivalence between these two methods; I can come back to this later.}
+
+The above development allows only fixed-effects in the regression models.
+If we want to also incorporate random effects, we simply view Equations \ref{eq:id_nest} and \ref{eq:id_nest_cov} as having had these random effects marginalized out.
+That is, if we write $U$ and $V$ for the random effects in our models for $Y$ and $M$ respectively, and $G_V$, $G_V$ for the random effects' distributions, then we can re-write Equations \ref{eq:id_nest} and \ref{eq:id_nest_cov} as
+%
+\begin{align}
+    \bE Y(x, M(x')) &= \int \bE(Y | U=u, M=m, X=x) \bP(M = dm |V=v, X = x') G_U(du) G_V(dv) \label{eq:ran_id_nest} \\
+    \bE Y_c(x, M(x')) &= \int \bE(Y | U=u, M=m, X=x, C=c) \bP(M = dm | V=v, X = x', C=c) G_U(du) G_V(dv) \label{eq:ran_id_nest_cov}
+\end{align}
+
+\subsection{Ideas for CI Theory}
+Applying the Counterfactual Unnesting Theorem of \citet{Cor21}, and provided that the conditions specified therein are satisfied, we have
+
+Alternatively, see Theorem 1 of \citet{Ima10I}.
+This is specifically the result I want, not the corresponding identification results of \citet{Ima10II}, which focuses more narrowly on identification of mediation effects.
+Note that, in \citet{Ima10I}, the function $f$ is a density (with respect to some unspecified measure, possibly counting measure).
+In order to get the above formulas for expected values of $Y$, we do need to assume that $\bE Y$ is finite so we can apply Fubini's Theorem (the one for finite integrals).
+
+\section{Estimation and Inference}
 
@@
 
-\bibliography{Lit_Review_Bib}
+\bibliography{../Bib-Exact_Asymptotics}
 

### 2024-06-25 notes/Consolidated Notes/Consolidated Notes.tex
Update consolidated notes
Send updated analysis to B & B

@@
 
-I'm just finishing-up my code for the binary response, binary mediator, mixed effects case. My analysis is running, but the output doesn't look right, so I need to do some debugging.
+I'm just finishing-up my code for the binary response, binary mediator, mixed effects case. My analysis is running, but the output doesn't look right so I need to do some debugging.
 
@@
 
+\section{\citeauthor{Van10}}
+
+In a sequence of papers over the early-to-mid 2010s, \citeauthor{Van10} developed much of the theory for causal-based mediation analysis.
+Unfortunately, they make extensive use of the ``rare event'' assumption.
+Nevertheless, many of their papers have been influential, and they are certainly worth cataloging here.
+
+\subsection{\citet{Van09}}
+
+This paper lays out the basic theoretical assumptions required for causal mediation analysis.
+These include consistency, composition and no unmeasured confounders.
+They also define the various direct and indirect effects which are typically of interest, and discuss how to estimate these effects using regression (provided, of course, that the necessary assumptions are satisfied).
+The focus is entirely on continuous outcome and mediator, although they do address conditional mediation effects given covariates/confounders, and averaging over the confounder(s) to recover marginal effects.
+Next, they give several pages' discussion of how to think about the various assumptions.
+Finally, an appendix gives proofs of the asserted regression-based formulas, alongside some simple uncertainty quantification.
+
+This is really a foundational paper, and is worth citing in pretty much any work on causal mediation analysis.
+
+\subsection{\citet{Van10}}
+
+This paper modifies the work of \citet{Van09} to handle a binary outcome, although the mediator is still taken to be continuous.
+Results are presented on odds-ratio scale.
+They set-up their analysis exactly, but then present their results under the rare-outcome assumption.
+As far as I can tell, this is done to simplify formulas.
+The estimators they end up with are sufficiently simple that uncertainty quantification is easy.
+
+This paper is a standard reference for mediation analysis with a binary outcome.
+
+\subsection{\citet{Val13}}
+
+Extends the work of \citet{Van10} to include a binary mediator. Still uses the rare outcome assumption and only reports results on odds-ratio scale.
+
+\subsection{\citet{Van13}}
+
+This paper extends the work of \citet{Van09} to handle multiple mediators.
+
+\subsection{\citet{Van14}}
+
+This paper extends the work of \citet{Van09} by giving an alternative decomposition of the total effect.
+The goal of this decomposition is to separate mediation from interaction between the exposure and mediator.
+
+\subsection{\citet{Van15}}
+
+A book, which consolidates much of the work of this group from the previous decade.
+
 \section{The \texttt{SPSS} Macro \texttt{MLmed}}
@@
 We have a conditional regression model and don't make any effort to model the marginal mean of $Y$ using regression.
-
-\subsection{\citet{Van10}}
-
-Standard reference for mediation analysis with binary outcome.
-Focuses on the causal inference aspects of the problem.
-Makes heavy use of the ``rare outcome'' assumption.
-Resulting model is sufficiently simple that UQ is easy.
 
@@
 I implemented the $\delta$-method based uncertainty quantification developed by B \& B for the total mediation effect.
-Please see the RMarkdown document and accompanying PDF called ``William Analysis.Rmd'' and ``William-Analysis.pdf'' for discussion and results.
+Please see the PDF called ``William-Analysis.pdf'' for discussion and results.
 
@@
 
-\bibliography{../Literature_Review/Lit_Review_Bib}
+% \bibliography{../Literature_Review/Lit_Review_Bib}
+\bibliography{../../Bib-Exact_Asymptotics}
 

### 2024-06-25 notes/William Analysis.Rmd
Send updated analysis to B & B

--- 2024-06-18 notes/William Analysis.Rmd
@@
 
-# Fewer large sample sizes
-# all_Ns = c(100, 200, 500, 1000, 2000)
-# all_Ns = c(200, 500, 1000, 2000)
-# all_Ns = c(200, 500, 1000)
-all_Ns = c(200, 500, 1000)
+### Easier version of problem ###
+
+all_Ns = c(50, 200, 500, 1000)  # Sample size for each cluster
+all_Ks = c(5, 20, 50)           # Number of clusters
+N_K_combs = expand.grid(N = all_Ns, K = all_Ks)
+
+# Number of replicates for each sample size
+num_reps = 200
 
@@
 # Random effects
-sigma_a_0 = 0.2
-sigma_a_1 = 0.2 * abs(a_1)
-cor_a0_a1 = 0.2
-cov_a0_a1 = sigma_a_0 * sigma_a_1 * cor_a0_a1
-theta = c(sigma_a_0, sigma_a_1, cor_a0_a1)
-Sigma_a = matrix(c(sigma_a_0^2, cov_a0_a1, cov_a0_a1, sigma_a_1^2), nrow = 2, ncol = 2)
-
-sigma_b_0 = 0.2 * abs(b_0)
-sigma_b_1 = 0.2 * abs(b_1)
-cor_b0_b1 = 0.2
-cov_b0_b1 = sigma_b_0 * sigma_b_1 * cor_b0_b1
-gamma = c(sigma_b_0, sigma_b_1, cor_b0_b1)
-Sigma_b = matrix(c(sigma_b_0^2, cov_b0_b1, cov_b0_b1, sigma_b_1^2), nrow = 2, ncol = 2)
-
-# Number of clusters
-all_Ks = c(5, 20, 50, 100)
-# K=5
-
-N_K_combs = expand.grid(N = all_Ns, K = all_Ks)
-
-# Number of replicates for each sample size
-# We use fewer here because model fitting takes longer
-num_reps = 200
-
-# Easier version of problem
-all_Ns = c(200, 500)
-all_Ks = c(5, 20)
-
 sigma_a_0 = 0.5
@@
 Sigma_b = matrix(c(sigma_b_0^2, cov_b0_b1, cov_b0_b1, sigma_b_1^2), nrow = 2, ncol = 2)
+
+### Harder version ###
+
+# # Fewer large sample sizes
+# # all_Ns = c(100, 200, 500, 1000, 2000)
+# # all_Ns = c(200, 500, 1000, 2000)
+# # all_Ns = c(200, 500, 1000)
+# # all_Ns = c(200, 500, 1000)
+# 
+# Values of X and W for which we compute the total effect
+# x_pred = 0
+# W_pred = c(1,1,1)
+# 
+# # Fixed effects for intercepts
+# a_0 = 0
+# a = c(a_0, a_1, A_2)
+# b_0 = -0.5
+# b = c(b_0, b_1, b_2, B_3)
+# 
+# # # Random effects
+# # sigma_a_0 = 0.2
+# # sigma_a_1 = 0.2 * abs(a_1)
+# # cor_a0_a1 = 0.2
+# # cov_a0_a1 = sigma_a_0 * sigma_a_1 * cor_a0_a1
+# # theta = c(sigma_a_0, sigma_a_1, cor_a0_a1)
+# # Sigma_a = matrix(c(sigma_a_0^2, cov_a0_a1, cov_a0_a1, sigma_a_1^2), nrow = 2, ncol = 2)
+# # 
+# # sigma_b_0 = 0.2 * abs(b_0)
+# # sigma_b_1 = 0.2 * abs(b_1)
+# # cor_b0_b1 = 0.2
+# # cov_b0_b1 = sigma_b_0 * sigma_b_1 * cor_b0_b1
+# # gamma = c(sigma_b_0, sigma_b_1, cor_b0_b1)
+# # Sigma_b = matrix(c(sigma_b_0^2, cov_b0_b1, cov_b0_b1, sigma_b_1^2), nrow = 2, ncol = 2)
+# 
+# # Number of clusters
+# # all_Ks = c(5, 20, 50, 100)
+# # K=5
+# 
+# # N_K_combs = expand.grid(N = all_Ns, K = all_Ks)
+# 
+# # Number of replicates for each sample size
+# # We use fewer here because model fitting takes longer
+# num_reps = 200
+
 ```
@@
 
+# Commented out to avoid lengthy runtime
+
 library(doParallel)
@@
 
-tic()
-
-set.seed(1)
-
-# Containers for output
-all_a_hats = list()
-all_theta_hats = list()
-all_b_hats = list()
-all_gamma_hats = list()
-
-all_M_covs = list()
-all_Y_covs = list()
-
-all_med_hats = list()
-all_med_SEs = list()
-
-# Initialize cluster on my machine ----
-# n_cores = 2
+# tic()
+# 
+# set.seed(1)
+# 
+# # Containers for output
+# all_a_hats = list()
+# all_theta_hats = list()
+# all_b_hats = list()
+# all_gamma_hats = list()
+# 
+# all_M_covs = list()
+# all_Y_covs = list()
+# 
+# all_med_hats = list()
+# all_med_SEs = list()
+# 
+# 
+# # Initialize cluster on my machine ----
+# # n_cores = 2
 # n_cores = 10
-n_cores = parallel::detectCores() - 1
-my_cluster = makeCluster(n_cores, type = "PSOCK")
-registerDoParallel(my_cluster)
-clusterEvalQ(my_cluster,{
-  library(MASS)
-  library(lme4)
-  library(merDeriv)
-  source("../src/Helpers.R")
-})
-clusterExport(my_cluster, c("N_K_combs", "num_reps", "a_0", "a_1", "A_2", "a", "b_0", "b_1", "b_2", "B_3", "b", "Sigma_a", "theta", "Sigma_b", "gamma", "p_conf", "x_pred", "W_pred"))
-clusterSetRNGStream(my_cluster, iseed = 1)
-
-# output_bin_bin_ran = pblapply(seq_len(nrow(N_K_combs)), function(j){
-# output_bin_bin_ran = pblapply(nrow(N_K_combs), function(j){
-for(j in seq_len(nrow(N_K_combs))){
-  n = N_K_combs[j, "N"]
-  K = N_K_combs[j, "K"]
-
-  clusterExport(my_cluster, c("n", "K"))
-
-  print(paste0("j = ", j, " of ", nrow(N_K_combs),  " (n=", n, ", K=", K, ")"))
-
-  # # Containers for output with this value of n
-  # some_a_hats = list()
-  # some_theta_hats = list()
-  # some_b_hats = list()
-  # some_gamma_hats = list()
-  #
-  # some_M_covs = list()
-  # some_Y_covs = list()
-  #
-  # some_med_hats = list()
-  # some_med_SEs = list()
-
-  # for(i in 1:num_reps){
-  some_output = pblapply(seq_len(num_reps), function(i){
-    # print(paste0("n = ", n, ", K = ", K, ", i = ", i, " of ", num_reps))
-
-    this_output = NULL
-
-    # Use tryCatch to omit any datasets leading to convergence issues
-    tryCatch({                                                                    ############## Un-quote
-      all_Xs = list()
-      all_Ws = list()
-
-      for(k in 1:K){
-        X = rnorm(n, mean=0, sd=1)
-        W = matrix(rnorm(n*p_conf, mean=0, sd=1), nrow = n, ncol = p_conf)
-
-        all_Xs[[k]] = X
-        all_Ws[[k]] = W
-      }
-
-      # Generate M
-      all_Ms = list()
-      for(k in 1:K){
-        eta_vec_fixed = a_0 + a_1*all_Xs[[k]] + all_Ws[[k]]%*%A_2
-
-        ## Add random effects
-        a_ran = mvrnorm(1, mu = rep(0, 2), Sigma = Sigma_a)
-        eta_vec = eta_vec_fixed + a_ran[1] + a_ran[2]*all_Xs[[k]]
-
-        ## Generate M
-        p_M_vec = expit(eta_vec)
-        M = rbinom(n, size = 1, prob = p_M_vec)
-        all_Ms[[k]] = M
-      }
-
-      # Generate Y
-      all_Ys = list()
-      for(k in 1:K){
-        zeta_vec_fixed = b_0 + b_1*all_Ms[[k]] + b_2 * all_Xs[[k]] + all_Ws[[k]]%*%B_3
-
-        ## Add random effects
-        b_ran = mvrnorm(1, mu = rep(0, 2), Sigma = Sigma_b)
-        zeta_vec = zeta_vec_fixed + b_ran[1] + b_ran[2]*all_Xs[[k]]
-
-        ## Generate Y
-        p_Y_vec = expit(zeta_vec)
-        Y = rbinom(n, size = 1, prob = p_Y_vec)
-        all_Ys[[k]] = Y
-      }
-
-      # Consolidate groups
-      X = do.call(c, all_Xs)
-      W = do.call(rbind, all_Ws)
-      M = do.call(c, all_Ms)
-      Y = do.call(c, all_Ys)
-      group = rep(1:K, each = n)
-
-      # Fit models
-
-      ## M
-      M_data = data.frame(M, X, W1 = W[, 1], W2 = W[, 2], W3 = W[,3], group = group)
-      M_model = glmer(M ~ X + W1 + W2 + W3 + (1 + X | group), data = M_data, family = binomial(link = "logit"))
-      M_model_info = attributes(VarCorr(M_model)$group)
-
-      ### Fitted parameters
-      a_hat = fixef(M_model)
-      a_RE_sds = M_model_info$stddev
-      a_RE_cor = M_model_info$correlation[2,1]
-      theta_hat = c(a_RE_sds, a_RE_cor)
-      if(any(is.nan(theta_hat))) stop("NaNs in theta_hat")  # Skip rest of current analysis if correlation is 0/0
-
-      ### Estimated SE covariance matrix
-      M_cov = vcov(M_model, full=TRUE, ranpar="sd")
-
-      # some_a_hats[[i]] = a_hat
-      # some_theta_hats[[i]] = theta_hat
-      # some_M_covs[[i]] = M_cov
-
-      ## Y
-      Y_data = data.frame(Y, M, X, W1 = W[, 1], W2 = W[, 2], W3 = W[,3], group = group)
-      Y_model = glmer(Y ~ M + X + W1 + W2 + W3 + (1 + X | group), data = Y_data, family = binomial(link = "logit"))
-      Y_model_info = attributes(VarCorr(Y_model)$group)
-
-      ### Fitted parameters
-      b_hat = fixef(Y_model)
-      b_RE_sds = Y_model_info$stddev
-      b_RE_cor = Y_model_info$correlation[2,1]
-      gamma_hat = c(b_RE_sds, b_RE_cor)
-      if(any(is.nan(gamma_hat))) stop("NaNs in gamma_hat")  # Skip rest of current analysis if correlation is 0/0
-
-      ### Estimated SE covariance matrix
-      Y_cov = vcov(Y_model, full=TRUE, ranpar="sd")
-
-      # some_b_hats[[i]] = b_hat
-      # some_gamma_hats[[i]] = gamma_hat
-      # some_Y_covs[[i]] = Y_cov
-
-      # Estimate mediation effect
-
-      ## Fixed-effects
-
-      a_0_hat = a_hat[1]
-      a_x_hat = a_hat[2]
-      A_2_hat = a_hat[3:5]
-
-      b_0_hat = b_hat[1]
-      b_m_hat = b_hat[2]
-      b_x_hat = b_hat[3]
-      B_3_hat = b_hat[4:6]
-
-      ## Linear predictors
-      eta_hat = as.numeric(a_0_hat + a_x_hat*x_pred + W_pred%*%A_2_hat)
-      zeta_hat = as.numeric(b_0_hat + b_x_hat * x_pred + W_pred %*% B_3_hat)
-
-      ## Random effects covariances
-      s_M_0 = a_RE_sds[1]
-      s_M_x = a_RE_sds[2]
-      rho_M = a_RE_cor
-
-      s_Y_0 = b_RE_sds[1]
-      s_Y_x = b_RE_sds[2]
-      rho_Y = b_RE_cor
-
-      ## Sigma functions
-      sigma_M1 = sigma_fun(x_pred, s_M_0, s_M_x, rho_M)
-      sigma_M2 = sigma_fun(x_pred + 1, s_M_x, s_M_0, rho_M)
-
-      sigma_Y1 = sigma_fun(x_pred, s_Y_0, s_Y_x, rho_Y)
-      sigma_Y2 = sigma_fun(x_pred + 1, s_Y_x, s_Y_0, rho_Y)
-
-      ## Mediation effect
-      ### See Helpers.R for the function Phi, which computes the mediation effect on odds-ratio scale
-      med_hat = Phi(eta_hat, zeta_hat, a_x_hat, b_m_hat, b_x_hat, sigma_M2, sigma_Y2, sigma_M1, sigma_Y1)
-      # some_med_hats[[i]] = med_hat
-
-      # Estimate SE
-
-      ## Gradient of OR wrt regression coefficients
-      grad_Phi_obs = grad_Phi(eta_hat, zeta_hat, a_x_hat, b_m_hat, b_x_hat, sigma_M2, sigma_Y2, sigma_M1, sigma_Y1)
-      grad_xi_obs = grad_xi(a_hat, theta_hat, b_hat, gamma_hat, x_pred, W_pred)
-      d_Phi_d_GLMM_pars = grad_Phi_obs %*% grad_xi_obs
-
-      ## Get asymptotic SE using delta method
-
-      ### Build joint covariance matrix of regression coefficients
-      M_length = nrow(M_cov)
-      Y_length = nrow(Y_cov)
-      joint_cov = matrix(0, nrow = M_length + Y_length, ncol = M_length +
-      Y_length)
-      joint_cov[1:M_length, 1:M_length] = M_cov
-      joint_cov[(M_length + 1):(M_length + Y_length), (M_length + 1):(M_length +
-      Y_length)] = Y_cov
-
-      ### Convert to asymptotic covariance matrix
-      asymp_reg_cov = n * joint_cov
-
-      ### Pre- and post-multiply asymptotic covariance by gradient of Phi wrt GLMM parameters
-      med_asymp_var = d_Phi_d_GLMM_pars %*% asymp_reg_cov %*% t(d_Phi_d_GLMM_pars)
-
-      ### Get small-sample standard error
-      med_asymp_SE = sqrt(med_asymp_var)
-      med_SE = med_asymp_SE / sqrt(n)
-
-      # some_med_SEs[[i]] = med_SE
-
-      this_output = list(a_hat = a_hat, theta_hat = theta_hat, b_hat = b_hat, gamma_hat = gamma_hat, M_cov = M_cov, Y_cov = Y_cov, med_hat = med_hat, med_SE = med_SE)
-    # }, error = function(e){grad_Phi_obs <<- e}) # For troubleshooting parallel errors
-    }, error = function(e){})                                                       ############ Un-quote
-
-    return(this_output)
-
-  }, cl = my_cluster)
+# # n_cores = parallel::detectCores() - 1
+# my_cluster = makeCluster(n_cores, type = "PSOCK")
+# registerDoParallel(my_cluster)
+# clusterEvalQ(my_cluster,{
+#   library(MASS)
+#   library(lme4)
+#   library(merDeriv)
+#   source("../src/Helpers.R")
 # })
-
-  # Store output for current values of n and K
-
-  all_a_hats[[j]] = lapply(some_output, function(x) x$a_hat)
-  all_theta_hats[[j]] = lapply(some_output, function(x) x$theta_hat)
-  all_b_hats[[j]] = lapply(some_output, function(x) x$b_hat)
-  all_gamma_hats[[j]] = lapply(some_output, function(x) x$gamma_hat)
-  all_M_covs[[j]] = lapply(some_output, function(x) x$M_cov)
-  all_Y_covs[[j]] = lapply(some_output, function(x) x$Y_cov)
-  all_med_hats[[j]] = lapply(some_output, function(x) x$med_hat)
-  all_med_SEs[[j]] = lapply(some_output, function(x) x$med_SE)
-}
-
-runtime = toc()
-
-parallel::stopCluster(my_cluster)
+# clusterExport(my_cluster, c("N_K_combs", "num_reps", "a_0", "a_1", "A_2", "a", "b_0", "b_1", "b_2", "B_3", "b", "Sigma_a", "theta", "Sigma_b", "gamma", "p_conf", "x_pred", "W_pred"))
+# clusterSetRNGStream(my_cluster, iseed = 1)
+# 
+# # output_bin_bin_ran = pblapply(seq_len(nrow(N_K_combs)), function(j){
+# # output_bin_bin_ran = pblapply(nrow(N_K_combs), function(j){
+# for(j in seq_len(nrow(N_K_combs))){
+#   n = N_K_combs[j, "N"]
+#   K = N_K_combs[j, "K"]
+# 
+#   clusterExport(my_cluster, c("n", "K"))
+# 
+#   print(paste0("j = ", j, " of ", nrow(N_K_combs),  " (n=", n, ", K=", K, ")"))
+# 
+# 
+#   # # Containers for output with this value of n
+#   # some_a_hats = list()
+#   # some_theta_hats = list()
+#   # some_b_hats = list()
+#   # some_gamma_hats = list()
+#   #
+#   # some_M_covs = list()
+#   # some_Y_covs = list()
+#   #
+#   # some_med_hats = list()
+#   # some_med_SEs = list()
+# 
+#   # for(i in 1:num_reps){
+#   some_output = pblapply(seq_len(num_reps), function(i){
+#     # print(paste0("n = ", n, ", K = ", K, ", i = ", i, " of ", num_reps))
+# 
+#     this_output = NULL
+# 
+#     # Use tryCatch to omit any datasets leading to convergence issues
+#     tryCatch({                                                                    ############## Un-quote
+#       all_Xs = list()
+#       all_Ws = list()
+# 
+#       for(k in 1:K){
+#         X = rnorm(n, mean=0, sd=1)
+#         W = matrix(rnorm(n*p_conf, mean=0, sd=1), nrow = n, ncol = p_conf)
+# 
+#         all_Xs[[k]] = X
+#         all_Ws[[k]] = W
+#       }
+# 
+#       # Generate M
+#       all_Ms = list()
+#       for(k in 1:K){
+#         eta_vec_fixed = a_0 + a_1*all_Xs[[k]] + all_Ws[[k]]%*%A_2
+# 
+#         ## Add random effects
+#         a_ran = mvrnorm(1, mu = rep(0, 2), Sigma = Sigma_a)
+#         eta_vec = eta_vec_fixed + a_ran[1] + a_ran[2]*all_Xs[[k]]
+# 
+#         ## Generate M
+#         p_M_vec = expit(eta_vec)
+#         M = rbinom(n, size = 1, prob = p_M_vec)
+#         all_Ms[[k]] = M
+#       }
+# 
+#       # Generate Y
+#       all_Ys = list()
+#       for(k in 1:K){
+#         zeta_vec_fixed = b_0 + b_1*all_Ms[[k]] + b_2 * all_Xs[[k]] + all_Ws[[k]]%*%B_3
+# 
+#         ## Add random effects
+#         b_ran = mvrnorm(1, mu = rep(0, 2), Sigma = Sigma_b)
+#         zeta_vec = zeta_vec_fixed + b_ran[1] + b_ran[2]*all_Xs[[k]]
+# 
+#         ## Generate Y
+#         p_Y_vec = expit(zeta_vec)
+#         Y = rbinom(n, size = 1, prob = p_Y_vec)
+#         all_Ys[[k]] = Y
+#       }
+# 
+# 
+#       # Consolidate groups
+#       X = do.call(c, all_Xs)
+#       W = do.call(rbind, all_Ws)
+#       M = do.call(c, all_Ms)
+#       Y = do.call(c, all_Ys)
+#       group = rep(1:K, each = n)
+# 
+# 
+#       # Fit models
+# 
+#       ## M
+#       M_data = data.frame(M, X, W1 = W[, 1], W2 = W[, 2], W3 = W[,3], group = group)
+#       M_model = glmer(M ~ X + W1 + W2 + W3 + (1 + X | group), data = M_data, family = binomial(link = "logit"))
+#       M_model_info = attributes(VarCorr(M_model)$group)
+# 
+#       ### Fitted parameters
+#       a_hat = fixef(M_model)
+#       a_RE_sds = M_model_info$stddev
+#       a_RE_cor = M_model_info$correlation[2,1]
+#       theta_hat = c(a_RE_sds, a_RE_cor)
+#       if(any(is.nan(theta_hat))) stop("NaNs in theta_hat")  # Skip rest of current analysis if correlation is 0/0
+# 
+#       ### Estimated SE covariance matrix
+#       M_cov = vcov(M_model, full=TRUE, ranpar="sd")
+# 
+#       # some_a_hats[[i]] = a_hat
+#       # some_theta_hats[[i]] = theta_hat
+#       # some_M_covs[[i]] = M_cov
+# 
+# 
+#       ## Y
+#       Y_data = data.frame(Y, M, X, W1 = W[, 1], W2 = W[, 2], W3 = W[,3], group = group)
+#       Y_model = glmer(Y ~ M + X + W1 + W2 + W3 + (1 + X | group), data = Y_data, family = binomial(link = "logit"))
+#       Y_model_info = attributes(VarCorr(Y_model)$group)
+# 
+#       ### Fitted parameters
+#       b_hat = fixef(Y_model)
+#       b_RE_sds = Y_model_info$stddev
+#       b_RE_cor = Y_model_info$correlation[2,1]
+#       gamma_hat = c(b_RE_sds, b_RE_cor)
+#       if(any(is.nan(gamma_hat))) stop("NaNs in gamma_hat")  # Skip rest of current analysis if correlation is 0/0
+# 
+#       ### Estimated SE covariance matrix
+#       Y_cov = vcov(Y_model, full=TRUE, ranpar="sd")
+# 
+#       # some_b_hats[[i]] = b_hat
+#       # some_gamma_hats[[i]] = gamma_hat
+#       # some_Y_covs[[i]] = Y_cov
+# 
+# 
+# 
+#       # Estimate mediation effect
+# 
+#       ## Fixed-effects
+# 
+#       a_0_hat = a_hat[1]
+#       a_x_hat = a_hat[2]
+#       A_2_hat = a_hat[3:5]
+# 
+#       b_0_hat = b_hat[1]
+#       b_m_hat = b_hat[2]
+#       b_x_hat = b_hat[3]
+#       B_3_hat = b_hat[4:6]
+# 
+# 
+#       ## Linear predictors
+#       eta_hat = as.numeric(a_0_hat + a_x_hat*x_pred + W_pred%*%A_2_hat)
+#       zeta_hat = as.numeric(b_0_hat + b_x_hat * x_pred + W_pred %*% B_3_hat)
+# 
+# 
+#       ## Random effects covariances
+#       s_M_0 = a_RE_sds[1]
+#       s_M_x = a_RE_sds[2]
+#       rho_M = a_RE_cor
+# 
+#       s_Y_0 = b_RE_sds[1]
+#       s_Y_x = b_RE_sds[2]
+#       rho_Y = b_RE_cor
+# 
+# 
+#       ## Sigma functions
+#       sigma_M1 = sigma_fun(x_pred, s_M_0, s_M_x, rho_M)
+#       sigma_M2 = sigma_fun(x_pred + 1, s_M_x, s_M_0, rho_M)
+# 
+#       sigma_Y1 = sigma_fun(x_pred, s_Y_0, s_Y_x, rho_Y)
+#       sigma_Y2 = sigma_fun(x_pred + 1, s_Y_x, s_Y_0, rho_Y)
+# 
+#       ## Mediation effect
+#       ### See Helpers.R for the function Phi, which computes the mediation effect on odds-ratio scale
+#       med_hat = Phi(eta_hat, zeta_hat, a_x_hat, b_m_hat, b_x_hat, sigma_M2, sigma_Y2, sigma_M1, sigma_Y1)
+#       # some_med_hats[[i]] = med_hat
+# 
+# 
+# 
+# 
+# 
+#       # Estimate SE
+# 
+#       ## Gradient of OR wrt regression coefficients
+#       grad_Phi_obs = grad_Phi(eta_hat, zeta_hat, a_x_hat, b_m_hat, b_x_hat, sigma_M2, sigma_Y2, sigma_M1, sigma_Y1)
+#       grad_xi_obs = grad_xi(a_hat, theta_hat, b_hat, gamma_hat, x_pred, W_pred)
+#       d_Phi_d_GLMM_pars = grad_Phi_obs %*% grad_xi_obs
+# 
+#       ## Get asymptotic SE using delta method
+# 
+#       ### Build joint covariance matrix of regression coefficients
+#       M_length = nrow(M_cov)
+#       Y_length = nrow(Y_cov)
+#       joint_cov = matrix(0, nrow = M_length + Y_length, ncol = M_length +
+#       Y_length)
+#       joint_cov[1:M_length, 1:M_length] = M_cov
+#       joint_cov[(M_length + 1):(M_length + Y_length), (M_length + 1):(M_length +
+#       Y_length)] = Y_cov
+# 
+#       ### Convert to asymptotic covariance matrix
+#       asymp_reg_cov = n * joint_cov
+# 
+#       ### Pre- and post-multiply asymptotic covariance by gradient of Phi wrt GLMM parameters
+#       med_asymp_var = d_Phi_d_GLMM_pars %*% asymp_reg_cov %*% t(d_Phi_d_GLMM_pars)
+# 
+#       ### Get small-sample standard error
+#       med_asymp_SE = sqrt(med_asymp_var)
+#       med_SE = med_asymp_SE / sqrt(n)
+# 
+#       # some_med_SEs[[i]] = med_SE
+# 
+#       this_output = list(a_hat = a_hat, theta_hat = theta_hat, b_hat = b_hat, gamma_hat = gamma_hat, M_cov = M_cov, Y_cov = Y_cov, med_hat = med_hat, med_SE = med_SE)
+#     # }, error = function(e){grad_Phi_obs <<- e}) # For troubleshooting parallel errors
+#     }, error = function(e){})                                                       ############ Un-quote
+# 
+#     return(this_output)
+# 
+# 
+#   }, cl = my_cluster)
+#   # })
+# 
+#   # Store output for current values of n and K
+# 
+#   all_a_hats[[j]] = lapply(some_output, function(x) x$a_hat)
+#   all_theta_hats[[j]] = lapply(some_output, function(x) x$theta_hat)
+#   all_b_hats[[j]] = lapply(some_output, function(x) x$b_hat)
+#   all_gamma_hats[[j]] = lapply(some_output, function(x) x$gamma_hat)
+#   all_M_covs[[j]] = lapply(some_output, function(x) x$M_cov)
+#   all_Y_covs[[j]] = lapply(some_output, function(x) x$Y_cov)
+#   all_med_hats[[j]] = lapply(some_output, function(x) x$med_hat)
+#   all_med_SEs[[j]] = lapply(some_output, function(x) x$med_SE)
+# }
+# 
+# 
+# runtime = toc()
+# 
+# parallel::stopCluster(my_cluster)
 
@@
 # save(N_K_combs, all_a_hats, all_theta_hats, all_b_hats, all_gamma_hats, all_M_covs, all_Y_covs, all_med_hats, all_med_SEs, runtime, file = "William-Analysis_cache/MC-bin-bin-mix.RData")
-# load("D:/OneDrive/OneDrive - Simon Fraser University (1sfu)/Research/Postdoc/Exact_Asymptotics/notes/William-Analysis_cache/MC-bin-bin-mix.RData", verbose = TRUE)
+load("William-Analysis_cache/MC-bin-bin-mix.RData", verbose = TRUE)
 ```
@@
 
-<!-- I'm now trying to figure out what's causing an error which only occurs on knit. Trying to isolate it in the following code chunk: -->
-
-<!-- ```{r} -->
-<!-- load("D:/OneDrive/OneDrive - Simon Fraser University (1sfu)/Research/Postdoc/Exact_Asymptotics/notes/William-Analysis_cache/MC-bin-bin-mix.RData", verbose = TRUE) -->
-
-<!-- # print(all_med_hats) -->
-<!-- ``` -->
-
 ```{r results="asis", echo=FALSE}
@@
 ```
-
-<!-- I don't really understand the results I'm getting here. It should help if I can compare my estimates of the mediation effect with the true value. -->
-
-<!-- ```{r} -->
-<!-- load("../data/Exact_Asymptotics/Bin-Bin-Mix_Results.RData") -->
-
-<!-- true_med = Phi_of_xi(a, theta, b, gamma, x_pred, W_pred) -->
-
-<!-- load("William-Analysis_cache/latex/unnamed-chunk-37_44bd8d1152289e689fbff9185d17042c.RData", ex <- new.env()) -->
-<!-- ``` -->
 

### 2024-06-25 notes/For Bruno/Helpers.R

@@ -1,0 +1,981 @@
+# General ####
+
+logit = function(x) log(x/(1-x))
+
+expit = function(x) exp(x)/(1+exp(x))
+
+
+
+# Cont Resp, Bin Med, Fixed-Effs ####
+## Some notation:
+  ## eta: a_0 + a_1 x + A_2 W = a_0 + a_x * x + A_2' w. Linear predictor for M
+  ## zeta: b_0 + b_1 * m + b_2 * x + B_3' w = b_0 + b_m * m + b_x * x + B_3' w. Linear predictor for Y
+
+  ## delta: E(M | X=x+1) - E(M | X=x)
+  ## gamma: E(Y | X=x+1) - E(Y | X=x). Total effect of X on Y
+
+
+get_delta = function(eta, a_x){
+  Q1 = exp(-eta)
+  Q2 = (1 - exp(-a_x))
+  Q3 = (1 + exp(-eta - a_x))
+  Q4 = (1 + exp(-eta))
+
+  return(Q1 * Q2 / (Q3 * Q4))
+}
+
+
+get_gamma = function(delta, b_m, b_x){
+  return(delta * b_m + b_x)
+}
+
+
+## Gradient of total effect wrt the a's and b's
+
+### First, some intermediate quantities
+d_delta_d_eta = function(eta, a_x){
+  delta = get_delta(eta, a_x)
+
+  Q1 = 1 - exp(-2*eta - a_x)
+  Q2 = 1 + exp(-eta)
+  Q3 = 1 + exp(-eta - a_x)
+
+  return( (-1) * delta * Q1 / (Q2 * Q3))
+}
+
+d_delta_d_a_x = function(eta, a_x){
+  Q1 = exp(-2*eta - a_x)
+  Q2 = exp(-eta - a_x)
+  Q3 = 1 + exp(-eta)
+  Q4 = 1 + exp(-eta - a_x)
+
+  return((Q1 + Q2)/(Q3 * Q4^2))
+}
+
+
+### Now, the gradient
+d_gamma_d_a_0 <- function(dd_de, b_m) {
+  return(dd_de * b_m)
+}
+
+d_gamma_d_a_1 <- function(dd_de, dd_da_x, x, b_m) {
+  return(b_m * x * dd_de + b_m * dd_da_x)
+}
+
+d_gamma_d_A_2 <- function(dd_de, W, b_m) {
[…]

### 2024-07-09 papers/Manuscript - Exact Asymptotics.tex
Some manuscript updates

@@
 \usepackage{soul}
+\usepackage{outlines}   % For multilevel lists using the outline environment
+\usepackage{tikz}
+\usetikzlibrary{arrows.meta, positioning}
 
@@
 \section{Introduction}
+
+\begin{outline}
+    \1 Literature review
+        \2 \citeauthor{Sam23}
+        \2 Imai et al.
+        \2 SEM world
+    \1 Overview of our contribution
+        \2 Analytical UQ for mediation effects
+        \2 Existing work is Monte Carlo-based. E.g. Imai's method, bootstrap
+\end{outline}
 
@@
 \section{Multilevel Mediation Analysis}
+
+\begin{outline}
+    \1 Define counterfactuals (ctfs) and nested counterfactuals 
+    \1 Define mediation effects
+        \2 Continuous outcome
+        \2 Binary outcome on risk difference, risk ratio and odds ratio scales
+    \1 Identification of mediation effects
+        \2 See \citet{Ima10I} for non-parametric identification of nested counterfactuals
+    \1 Regression modelling
+        \2 Restrict to binary outcome?
+        \2 Fixed effects regression (brief)
+        \2 Mixed effects regression
+        \2 Prediction of random effects
+\end{outline}
+
+\subsection{2024-06-25}
+
+Our approach to causal mediation analysis is based on the counterfactual framework of ???.
+Briefly, let $Y$ be an outcome of interest, $X$ be an exposure which is a causal driver of $Y$, and $M$ be a mediator, which influences $Y$ and is influenced by $X$.
+Figure \ref{fig:simp_med} shows a causal diagram representing this relationship.
+We call the top arrow the direct effect of $X$ on $Y$, and the bottom path through $M$ the indirect effect of $Y$.
+Taken together, these two pathways constitute the total effect of $X$ on $Y$.
+
+\begin{figure}
+    \centering
+    \begin{tikzpicture}[>=Stealth, node distance=2cm, on grid, auto]
+
+        % Nodes
+        \node (X) [draw, circle] {X};
+        \node (M) [draw, circle, right=of X] {M};
+        \node (Y) [draw, circle, right=of M] {Y};
+
+        % Arrows
+        \draw[->] (X) -- (M);
+        \draw[->] (M) -- (Y);
+        \draw[->] (X) to[bend left] (Y);
+
+      \end{tikzpicture}
+      \caption{Causal diagram showing $M$ mediating the effect of $X$ on $Y$.}    
+      \label{fig:simp_med}
+\end{figure}
+
+We define the counterfactual $Y_x$ as the value $Y$ would assume when $X=x$.
+Similarly, write $M_x$ for the value $M$ would assume when $X=x$.
+Next, write $Y_{x,m}$ for the value $Y$ would assume if $X$ and $M$ were set to $x$ and $m$ respectively.
+Combining these ideas, we get the ``nested counterfactual'' $Y_{x, M_{x'}}$, which is the value of $Y$ when $X=x$ and $M$ is set to whatever it would have been when $X=x'$.
+It is common to link ordinary and nested conterfactuals by making the consistency assumption, which states $Y_x = Y_{x, M_x}$.
+Note that if $x \neq x'$ then the nested counterfactual is necessarily unobservable.
+It is nevertheless possible, given certain assumptions, to estimate the expected values of ordinary and nested counterfactuals.
+We will return later to the identification of expected counterfactuals with estimable quantities \hl{(I'm not certain whether it's better to do identification here or later).}
+
+We now move on to the definition of various mediation effects which are of interest throughout the literature.
+There are three main types of mediation effect: total, direct and indirect, although the scale on which these are measured can vary.
+In the interest of generality, we write $d(y_1, y_2)$ for an arbitrary measure of the discrepancy between $y_1$ and $y_2$.
+The most natural form is $d(y_1, y_2) = y_1 - y_2$, although others may be more appropriate depending on context.
+Note that we do not require $d$ to be symmetric; in fact, the sign of $d$ is often of central scientific importance.
+
+Proceeding now to the actual definitions, we define the total effect of $X$ on $Y$ to be $TE(x, x') = d(EY_x, EY_{x'})$.
+The direct effect is defined as $DE(x, x') = d(Y_{x, M_{x'}}, Y_{x', M_{x'}})$, and the indirect effect is $IE(x, x') = d(Y_{x, M_{x}}, Y_{x, M_{x'}})$.
+See, e.g., ???
+for motivation and discussion of these definitions \hl{($g$-function?)}.
+
+When the outcome, $Y$, is continuous, it is natural to take the discrepancy function to be $d(y_1, y_2) = y_1 - y_2$ as mentioned above.
+However, when $Y$ is binary, it is less obvious which scale to use.
+In this setting, expected counterfactuals are probabilities, so mediation effects measure the discrepancy between two probabilities.
+One option is to use the same difference preferred for continuous outcome; we refer to this as the risk-difference scale.
+Alternatives include $d(y_1, y_2) = y_1 / y_2$, the risk-ratio scale, and $d(y_1, y_2) = [y_1 / (1 - y_1)] / [y_2 / (1 - y_2)]$, the odds-ratio scale.
+We focus only on the binary outcome setting, although our results are easily extended to handle continuous outcomes.
+While our method can handle any choice of discrepancy function $d$ \hl{(differentiable?
+Whatever is necessary to apply the $\delta$-method later; might need to be continuously differentiable)}, in the interest of accessibility we specialize our results to the three discrepancies just mentioned for binary outcomes.
+
+A fundamental problem of causal inference is that we can only ever observe one counterfactual outcome on a particular individual.
+In mediation analysis, this problem is even worse, since many of our definitions involve the nested counterfactual, $Y_{x, M_{x'}}$, which when $x \neq x'$ cannot be observed on any individual.
+Nevertheless, \citet{Ima10I} give conditions under which the population average of a nested counterfactual can be expressed in terms of conditional expectations, possibly conditional on one or more additional covariates, $W$.
+Specifically, their Theorem 1 states that, under a condition they call ``Sequential Ignorability'', we can write
+%
+\begin{equation}
+    \bE (Y_{x, M_{x'}} | W=w) = \bE_M \left[ \bE_Y \left( Y | X=x, M, W=w \right) | X=x', W=w \right]. \label{eq:ID}
+\end{equation}
+%
+In fact, their Theorem 1 is somewhat more general, giving an expression for the density of the nested counterfactual rather than its expected value.
+
+Using Equation (\ref{eq:ID}), we can estimate expected nested counterfactuals by working with the more tractable conditional expectations of $Y$ and $M$.
+We model the latter using regression, either linear or logistic depending on the forms of $Y$ and $M$.
+If $Y$ and $M$ are binary, then Equation (\ref{eq:ID}) has a particularly simple form:
+%
+\begin{multline}
+    \bE (Y_{x, M_{x'}} | W=w) = \bP \left( Y=1 | X=x, M=1, W=w \right) \bP(M=1 | X = x', W=w) \\ + \bP \left( Y=1 | X=x, M=0, W=w \right) \bP(M=0 | X = x', W=w). \label{eq:ID_bin}
+\end{multline}
+%
+We estimate the conditional probabilities on the right-hand side of Equation (\ref{eq:ID_bin}) using logistic regression in the usual way.
+Note that this requires us to fit two models, one to predict $M$ using $X$ and $W$, while the other predicts $Y$ using $M$, $X$ and $W$.
+
+We extend this regression modelling with the introduction of random effects \citep[see, e.g.][]{Dem04}.
+Specifically, we include random effects for the intercept and $X$ in our model for $M$, and for the intercept, $M$ and $X$ in our model for $Y$.
+We model these random effects as normally distributed, allowing for correlation within a model, but independent across models.
+Let $U \sim N(0, \Gamma_Y)$ and $V \sim N(0, \Gamma_M)$ be the random effects in our models for $Y$ and $M$ respectively.
+We can now re-write Equation (\ref{eq:ID_bin}) as
+%
+\begin{align}
+    \bE (Y_{x, M_{x'}} | W=w) = & \left[\bE_U \bP \left( Y=1 |U, X=x, M=1, W=w \right) \right. \cdot \label{eq:ID_bin_ran}\\
+    &  \left. \bE_V \bP(M=1 |V, X = x', W=w)\right] + \nonumber \\ 
+    & \left[ \bE_U \bP \left( Y=1 |U, X=x, M=0, W=w \right) \right. \cdot \nonumber \\
+    & \left. \bE_V \bP(M=0 |V, X = x', W=w) \right]. \nonumber 
+\end{align}
+%
+
+%! START HERE
+
+\textcolor{red}{The following discussion is awkward.
+I want to present a general expression for $E_U P(Y | ...)$, then specialize to Equation (\ref{eq:ID_bin_ran}).
+I have defined too much notation, but for now I'm just going to focus on getting ideas down on the page.
+Later, I will edit and tighten-up the presentation.}
+
+We show an expansion for one term in Equation (\ref{eq:ID_bin_ran}); the rest follow similarly.
+First, write $\eta = (\beta_0 + \beta_X x + \beta_M m + B_W^T w) + (U_0 + U_X x + U_M m)$ for the linear predictor of $Y$ based on $X$, $M$ and $W$.
+Let $\mu = \beta_0 + \beta_X x + \beta_M m + B_W^T w$ be the fixed-effects component of $\eta$, and $\xi = U_0 + U_X x + U_M m$ be its random-effects component.
+For convenience, we will also write $\gamma^2_Y(c_1, c_2, c_3) = (c_1, c_2, c_3) \Gamma_Y (c_1, c_2, c_3)^T$, so that $\bV \xi = \gamma^2_Y(1, m, x)$.
+
+It is a well-known fact about logistic regression that
+%
+\begin{equation}
+    \bP \left( Y=1 |U, X=x, M=1, W=w \right) = \left[1 + \exp(-\eta)\right]^{-1},
+\end{equation}
+%
+so the first term in Equation (\ref{eq:ID_bin_ran}) can be written as
+%
+\begin{align}
+    \bE_U \bP \left( Y=1 |U, X=x, M=1, W=w \right) & = \int \frac{\phi_3(u; 0, \Gamma_Y)}{1 + \exp(-\eta)} du, \label{eq:pY1_bad}
+\end{align}
+%
+where $\phi_d$ is the $d$-variate normal density. A straightforward change of variables gives us the alternative expression
+%
+\begin{align}
+    \bE_U \bP \left( Y=1 |U, X=x, M=1, W=w \right) & = \int \frac{\phi_1(z; 0, 1)}{1 + \exp(-\mu - \gamma_Y(1, m, x) z)} dz. \label{eq:pY1}
+\end{align}
+%
+Importantly, the integral in Equation (\ref{eq:pY1}) is univariate (and thus amenable to numerical evaluation using quadrature), while the integral in Equation (\ref{eq:pY1_bad}) is multivariate.
+
+The integral in Equation (\ref{eq:pY1}) arises often enough that we give it a name. Let
+%
+\begin{equation}
+    \Psi(\mu, \sigma^2) = \int \frac{\phi(z; 0, 1)}{1 + \exp(-\mu - \sigma z)} dz.
+\end{equation}
+
+\subsection{Old}
 
@@
 
-\section{Summary and Future Work}
+\begin{outline}
+    \1 Estimate models using \texttt{lme4}
+    \1 UQ for model parameters (mention \texttt{merDeriv} package)
+    \1 UQ for nested counterfactuals via $\delta$-method
+    \1 Simultaneous UQ for triples of mediation effects (i.e. total, direct and indirect)
+        \2 Can also do all 9 simultaneously
+    \1 UQ for predicted group-level effects?
+\end{outline}
 
-\subsection{May 23}
+\section{Empirical Investigation}
 
-I'm just finishing-up my code for the binary response, binary mediator, mixed effects case. My analysis is running, but the output doesn't look right so I need to do some debugging.
-
-I now better understand the uncertainty quantification method used by the Imai group.
-See Section \ref{sec:Imai_UQ} for details.
-Briefly, they do a Monte Carlo analog of the $\delta$-method.
-If the target of inference is a function of some model parameters, they start by estimating the sampling distribution of our estimator for those model parameters.
-They then simulate from this estimated sampling distribution and evaluate the function of interest on each simulated observation.
-This produces a sample from the pushforward measure induced on the target of inference by our estimated sampling distribution for the model parameters.
-It's not completely clear to me how they generate confidence intervals from the Monte Carlo samples, but they call their method ``quasi-Bayesian'', so I suspect it's just percentile intervals.
-
-Incidentally, if we could do exact (i.e.
-non-asymptotic) distribution theory for the model parameter (e.g.
-ordinary linear regression), then Imai's method is equivalent to the parametric bootstrap.
-They just skip the generation of datasets and work directly with the bootstrap distribution of the estimator.
-
-A thought on how to frame the definition of mediation effects on different scales: the mediation formula of \citet{Pea12} identifies causal direct/indirect effects with specific functions of conditional expectations of the outcome and mediator.
-This makes it natural to work on the risk-difference scale, as done by the Imai group (see Section \ref{sec:Imai}).
-However, if we prefer the risk-ratio or odds-ratio, we can just express our target quantities in terms of the functions identified by \citet{Pea12}.
-There may or may not be simplification that leads to interpretable formulas, but that's not the point.
-You can define mediation effects on whatever scale you like, then we will give you estimates and standard errors, irrespective of how messy the formulas may be (they will all be hidden in the code anyway).
-
-\subsubsection{What to do Next}
-
-I think that at this week's meeting, we should plan out what we want to put in the paper.
-I propose that we do total, direct and indirect effects on risk difference, risk ratio and odds ratio scales, but stick to fixed-effects (i.e.
-avoid predicting group-level effects).
-Based on other work in the literature.
-I think that's enough for a paper, provided that we do a nice Monte Carlo study comparing our UQ with other methods, such as Imai's quasi-Bayesian method, the weighting method of \citet{Lan12} and the imputation method of \citet{Van12}.
-All of these methods are implemented to varying extents in \texttt{R} packages, \texttt{mediation} for the former and \texttt{medflex} for the latter two.
-
-Prediction of group-level effects barely comes up in the literature.
-The \texttt{mediation} package from Imai's group does it, but with very little discussion.
-They give a few sentences in the package vignette saying they can do group-level effect prediction, but don't give any further discussion or references.
-I had to read their source code to know what method they're using.
-Prediction of group-level mediation effects seems like an untapped area that we are well-positioned to develop.
-
-\subsection{May 16}
-
-This week, I added work by Imai and colleagues, as well as a few general references.
-See Section \ref{sec:Imai} for the Imai group.
-There is also a method proposed by \citet{Lan12} based on inverse probability weighting to estimate unobserved counterfactuals.
-Both of these methods are included in the extensive Monte Carlo study of \citet{Sam23} and found to be comparable to inference for mediation effects based on the $\delta$-method.
-However, the method of \citet{Ima10I} is computationally intensive, and that of \citet{Lan12} is quite different from our approach of combining estimates from two regression models.
-
-I also found an \texttt{R} package, \texttt{merDeriv} \citep{Wan18} which computes the full asymptotic covariance matrix for a GLMM fit using \texttt{lme4}.
-This will save me a lot of time in implementing the $\delta$-method uncertainty quantification for mixed-effects models.
-
-\subsubsection{What to do next}
-
-I'm still not totally confident about the SEM literature, but otherwise I don't think anyone has done exact asymptotic theory for multilevel mediation analysis.
-I feel ready to start working on the mixed-effects version of our analysis.
-I will continue reading, but I think it makes sense to start spending some time writing code.
-
-There are some extensions to the methodology outlined by B \& B that I think we should consider.
-First, we will want to include direct and indirect effects in addition to our total effects.
-We can follow \citet{Sam23} for this, although we will also need to give some though to how we incorporate random effects into the definitions.
-It would also be nice to explicitly estimate the ``mediation proportion''.
-For continuous response, this is the indirect effect divided by the total effect.
-It's less clear to me what definition to use for multiplicative effects (e.g.
-effects on odds-ratio scale).
-
-Second, it is common to include an interaction effect between the exposure and mediator in our model for the outcome\footnote{There is no interaction effect in our trust study because a preliminary analysis found that it didn't have a significant effect on the outcome.}.
-This isn't conceptually any harder, it just adds another component to our formulas.
-
-I think that we could address just the above two points and, with some nice Monte Carlo studies, call it a paper.
-This would resemble the work by \citeauthor{Sam23} (e.g.
-\citealp{Sam23}).
-After that, there are some more dramatic extensions we can consider.
-
-One extension that we really should look at is predicted cluster-specific effects.
-The uncertainty quantification here is subtle; see, e.g., \citet{Skr09,Boo98}.
-The Imai group does give cluster-specific mediation effects, but only under their restricted framework.
-Otherwise, this seems to still be an open problem
-
-Finally (for now), I think we should eventually look at sensitivity analysis.
-There are a bunch of assumptions required for us to give a causal interpretation to our results.
-My understanding is that sensitivity analysis lets us put a bound on the extent to which these assumptions can be violated without invalidating our conclusions.
-I'm still looking into how this works.
-
-\subsection{May 9}
-
-As far as I can tell, nobody has done exact asymptotics for mixed-effects mediation analysis.
-I do want to do some more searching, but the recent literature (e.g., since 2020) presents exact formulas for mediation effects with $\delta$-method uncertainty quantification as novel.
-
-Below, I have summarized a bunch of work on mediation analysis, roughly divided into sections.
-Section \ref{sec:Sam} covers work by Samoilenko, Lefebvre and others.
-Their work looks pretty close to what we're doing.
-They give exact formulas and $\delta$-method UQ, but no random effects.
-Section \ref{sec:deriv} defines mediation effects in terms of partial derivatives of the mean of $Y$ given $X$.
-They have done a bit with random effects, but only do Bayesian or bootstrap-based uncertainty quantification.
-Section \ref{sec:SEM} uses structural equation modelling (SEM) to do mediation analysis.
-I'm not very familiar with this methodology, but I tried to summarize what I could.
-Some of these papers do include random effects, but only seem to do continuous response and mediator.
-They make heavy use of the simple mediation formulas in the fully continuous setting, so uncertainty quantification is pretty basic.
-Finally, Section \ref{sec:other} is an assortment of papers.
-Highlights include \citet{Che21}, which gives definitions of direct and indirect effects which we can use, and \citet{Zhe21}, which uses automatic differentiation to get the full Hessian for GLMMs.
-\citeauthor{Zhe21} also do UQ for the predicted random effects, although I'm still working on understanding exactly how they do this.
-
-\subsubsection{What to do Next}
-
-Next week, my plan is to continue the literature search, focusing on SEMs, the UQ methodology proposed by \citet{Zhe21}, and also looking more widely to see if I've missed any other lines of work.
-It would be helpful for me if you could look at \citet{Zhe21} to see if you think it has merit.
-I'm also pretty shaky on SEMs, so it would be nice if you could look at \citet{Zig19} to see how it connects with our methodology.
-
-In terms of advancing our own work, I have finished translating my analysis into \texttt{R}, up to the binary response and mediator.
-My next steps in this direction for next week are to work out as much as I can for the binary/binary setting with mixed effects.
-This is going to involve computing the gradient of the mediation effect, and looking for whether \texttt{lme4} gives standard errors for the random effects' covariance parameters.
-If not, then I will look around for other sources.
-
-Once the mixed-effects case is done, I would like to extend our analysis to the direct and indirect effects.
-Formulas for these are given in \citet{Sam18} or \citet{Che21}.
-We should also consider interaction terms between the exposure and mediator in our model for the outcome.
-I understand that this interaction term arises from causal inference-based concerns, but I'm still fuzzy on the details.
-\citet{Sam21} extend their previous work on fixed-effects models to include interactions.
-We also need to extend all of our methods to handle predicted group-specific effects, although this makes the uncertainty quantification more challenging, both conceptually and computationally.
-See, e.g., \citet{Boo98,Flo19,Zig19}.
-
-\section{The Samoilenko and Lefebvre group}
-\label{sec:Sam}
-
-This group does essentially what B\&B are proposing with exact formulas and $\delta$-method standard errors.
-In particular, they avoid any requirement for rare events.
-They also do direct and indirect effects.
-However, I haven't seen any mention of random-effects/multilevel-models.
-
-These methods are implemented in the \texttt{R} package \texttt{ExactMed} \citep{Cau23}. This package does not address mixed-effects models.
-
-\subsection{\citet{Sam18}}
-
-Gives exact formulas for direct and indirect mediation effects, as well as $\delta$-method standard errors.
-Binary outcome, binary mediator.
-No analytical SE when exposure-mediator interaction is present.
-
-\subsection{\citet{Sam21}}
-
-Extends analytical results of \citet{Sam18} to handle exposure-mediator interaction term in model for outcome.
-
-\subsection{\citet{Sam23}}
-
-As \citet{Sam18}, but with continuous mediator and binary response. No mixed-effects. Very nice simulation study.
-
-Compares their approach to those of the Imai group (see Section \ref{sec:Imai}), as well as \citet{Lan12}.
-Found that the $\delta$-method based approach proposed by the \citeauthor{Sam23} group gives comparable performance with its computationally intensive alternatives.
-
-\subsection{\citet{Cau24}}
-
-Extends the work of \citet{Sam23} to case-control data.
-
-\section{Derivative-Based}
-\label{sec:deriv}
-
-Defines mediation effects in terms of derivatives.
-Originally proposed by \citet{Yu14}, and later extended to multilevel models by \citet{Yu20}.
-They use the name ``third-variable effect analysis''.
-Documented extensively in a book by \citet{Yu22}.
-Implemented in the \texttt{R} packages \texttt{mma} and \texttt{mlma} for single level and multilevel models respectively.
-The latter uses the bootstrap for UQ.
-
-\subsection{\citet{Mar24}}
-
-Incorporates random-effects in the derivative-based approach. Mostly does Bayesian inference. Mentions bootstrap.
-
-\subsection{\citet{Gel18}}
-
-Defines mediation effects as partial derivatives.
-Does UQ by bootstrap or by a very rough method called MCCI.
-This MCCI simulates parameter values from a normal distribution parameterized by the estimates, then evaluates the mediation effects on these simulated parameters.
-This is the same UQ method used by the Imai group (see Section \ref{sec:Imai}).
-
-\subsection{\citet{Dor22}}
-
-Gives exact formulas for mediation effects and $\delta$-method SEs. Decomposes total mediation effect more finely than just direct and indirect (i.e. path-specific effects). No mixed-effects.
-
-\section{Structural Equation Modelling}
-\label{sec:SEM}
-
-There are a bunch of papers which use SEM to do mediation analysis.
-Many of them include multilevel/mixed-effects models.
-I'm not very familiar with SEM methodology, but it does seem to be connected with our work.
-See \citet{Val13} for a review and comparison with the causal-inference framework.
-Generally limited to linear models (i.e.
-continuous response and mediator), where the mediation effects' formulas are very simple.
-They seem to only focus on indirect effects\footnote{There is a short paragraph in \citet{Zit21} near the top of column 1 on page 532 which says ``the direct effect is only of little interest''.
-They don't give a reference for this.}.
-
-There does seem to be some work on applying mixed-effects methodology to SEMs. See, e.g., \citet{Rab04} for early work in this direction.
-
-I'm not sure yet if this has made its way to the SEM-based mediation analysis literature. I'll look into this next week.
-
-\subsection{\citet{Zig19}}
-
-Review of multilevel mediation analysis using SEM. Does a Monte Carlo study comparing different approaches. UQ is based on \citet{Bau06}.
-
-\subsection{\citet{Bau06}}
-
-Does mixed-effects mediation analysis based on SEMs.
-Only handles continuous outcome and mediator.
-Mediation effects are simple enough that variances are computed directly (in terms of asymptotic covariances of the regression parameter estimates).
-Fits the two regression models (for $M$ and $Y$) simultaneously by concatenating the response vectors and multiplying covariates by indicators.
-
-\subsection{\citet{Pre10}}
-
-Does mediation analysis using structural equation models. Does UQ by bootstrap or based on the exact distribution of the product of two dependent normals \citep{Mac07}.
-
-\section{The Imai Group}
-\label{sec:Imai}
-
-This group focuses mainly on the assumptions required to link regression-based mediation analysis with causal inference.
-They give definitions of mediation effects which closely resemble the ``mediation formula'' \citep{Pea12}, but only for differences of conditional expectation (i.e.
-the ``risk-difference scale''; see also \citealp{Sam23}).
-Their approach to UQ is very computational.
-They recommend either the bootstrap or a simulation-based approach outlined in \citet{Kin00}.
-Briefly, simulate parameters from their fitted sampling distributions, then evaluate the quantity of interest on each sample and observe the resulting distribution.
-I think this is connected to empirical Bayes methodology, but I haven't looked into it extensively.
-They use the name ``quasi-Bayesian confidence intervals'' in their \texttt{R} package.
-
-Their methods are implemented in the \texttt{R} package \texttt{mediation} \citep{Tin14}. They can handle mixed-effects models, including prediction of group-level effects.
-
-\subsection{\citet{Ima10I}}
-
-This paper presents an UQ algorithm for estimating mediation effects defined as differences of expected values.
-The algorithm closely resembles the bootstrap, but is slightly different.
-Specifically, parameters are repeatedly sampled from the fitted sampling distributions of their estimators, then mediation effects are estimated using the sampled parameter estimates \citep[see, e.g.,][]{Kin00}.
-They also give an alternative method which just uses the bootstrap.
-
-Refers to the causal inference theory from \citet{Ima10II} to justify identification of counterfactuals with functions of conditional expectations.
-
-\subsection{\citet{Ima10II}}
-
-Focuses on estimating the indirect effect for a continuous response.
-Defines this effect as a difference of counterfactuals and identifies this difference with a conditional expectation in the same manner as the mediation formula given by \citet{Pea12}.
-
-Also gives a ``non-parametric estimator'' for categorical mediator based on a contingency table based approach.
-The idea is to partition the data based on the mediator, estimate the mediation effect within each group with a difference of averages, then re-combine by averaging.
-
-UQ is given either by the $\delta$-method or by the bootstrap. Some discussion of sensitivity analysis.
-
-\subsection{\citet{Ima11}}
-
-Very similar to \citet{Ima10I} and \citet{Ima10II}. Focuses more on causal inference interpretation and applications to political science. Doesn't appear to advance the methodology.
-
-\subsection{\citet{Tin14}}
-
-This paper overviews the \texttt{mediation} package, which implements the methods developed by this group.
-Can handle GLMMs, but only operates on differences in the expected value of the outcome (i.e.
-risk-difference scale for binary outcome).
-Does UQ for parameters \textbf{and predicted random effects}, but doesn't go into detail on how.
-I'm guessing that it just follows the Monte Carlo method given by \citet{Ima10I}.
-
-Package last updated 2019-10-08
-
-\subsubsection{UQ for Group-Level Effects}
-\label{sec:Imai_UQ}
-
-The \citet{Tin14} paper discusses inference for predicted group-level effects, but doesn't go into detail.
-As far as I can tell, this is also the only paper from the Imai group which discusses multilevel mediation analysis.
-Worse still, they give very few references.
-
-After digging into the source code for the \texttt{mediation} package, I found what they do.
-First, they get predicted group-wise random effects from \texttt{lme4}.
-Second, it turns out that \texttt{lme4} can give a conditional covariance matrix for the random effects ``conditional on the estimates of the model parameters and on the data''.
-For now, just note that the full covariance matrix of the random effects is block diagonal, with one block per group.
-The \texttt{mediation} package uses these group-wise conditional covariance matrices, together with the predicted group-wise means, and simulates random effects for each group (independently, from normal distributions with the corresponding moments).
-The package then proceeds with its ``quasi-Bayesian'' inference.
-
-Just a brief note on the conditional covariances given by \texttt{lme4}.
-Initially, when I was reading this portion of the documentation, I got excited that we might not need the \texttt{merDeriv} package to get standard errors for the random effects' covariance parameters.
-That's definitely not what I found.
-These are standard errors for the predicted random effects (conditional on the observed data), not standard errors for the estimators of any model parameters.
-In particular, the former can be thought of as $K$ separate covariance matrices, one per group, whereas the latter is a single matrix representing estimation uncertainty based on the entire dataset.
-
-Unfortunately, no references are given in the \texttt{lme4} documentation, but it is fairly clear what is being done.
-The predicted random effect within a group is the mode of the conditional distribution of the random effects given the observed data from that group (alternatively, the mean; although \texttt{lme4} uses the mode).
-The prediction standard error given by \texttt{lme4} is the covariance of this same conditional distribution.
-
-\section{The \texttt{SPSS} Macro \texttt{MLmed}}
-\label{sec:MLmed}
-
-This macro appears to be the standard way to do multilevel mediation analysis in \texttt{SPSS}.
-As such, it is likely to be used extensively by applied researchers, and we should understand it.
-The macro can only handle continuous outcomes and mediators.
-
-\subsection{\citet{Roc19}}
-
-User manual for the \texttt{MLmed} macro. Can only handle continuous outcomes and mediators. UQ is either by the bootstrap or the ``Monte Carlo method''. See \citet{Pre12} or \citet{Wil08}.
-
-\subsection{\citet{Hay20}}
-
-A recommended reference for the methodology used in the macro.
-
-\subsection{\citet{Roc17}}
-
-A conference poster presenting the macro. Popular reference for people looking to cite the macro.
-
-\section{Applications}
-
-It doesn't hurt for us to point to what people are doing in the applied literature. Here is some of what shows up when I search ``multilevel mediation analysis'' on Google Scholar.
-
-\subsection{\citet{Vea19}}
-
-Does multilevel mediation analysis with continuous outcome and mediator.
-UQ is done with ``Monte Carlo confidence intervals''.
-Uses an \texttt{SPSS} macro called \texttt{MLmed}.
-See Section \ref{sec:MLmed}.
-
-\subsection{\citet{Ara21}}
-
-It looks like their variables are all ordinal (1-6 or 1-8 Likert scale).
-I'm not sure how they fit this into the mediation analysis framework.
-Moving on, they use \texttt{Mplus} to do their analysis via ``maximum likelihood''.
-Not much detail given.
-
-\subsection{\citet{Hwa12}}
-
-Follows an earlier edition of \citet{Mac17}, as well as the very early work of \citet{Bar86}.
-As in \citet{Ara21}, variables are ordinal/Likert.
-Analysis is done in \texttt{SPSS}, although few details are given.
-Note that this work was published before the \texttt{SPSS} macro of \citet{Roc19}.
-
-\subsection{\citet{Rij21}}
-
-Review of applied papers using mediation analysis. \textcolor{red}{Read and document this!}
-
-\section{Others}
-\label{sec:other}
-
-\subsection{\citet{Kru01}}
-
-Throughout some parts of the literature on multilevel mediation analysis, you will see people talk about, e.g., the 2-1-1 model.
-This means that the exposure is measured at ``level 2'' (group-level), while the mediator and outcome are measured at ``level 1'' (individual-level).
-That is, in this model we would include a random effect for the mediator, but not the exposure.
-Other examples of such models include 2-2-1 and 1-1-1.
-As far as I can tell, this notation was proposed by \citet{Kru01}.
-
-This paper also does $\delta$-method uncertainty quantification for the very simple mediation effects that people were using at that time (i.e.
-a product of coefficients).
-In fact, this approach for testing the presence/absence of mediation effects dates back to \citet{Sob82}.
-
-\subsection{\citet{Ken03}}
-
-Gives a simplistic definition of mediation effects for continuous models.
-Does UQ based on exact distribution of mediation effects obtained as a transformation of Gaussians.
-A nice approach, but only really works because of the simple setting.
-
-There is a book by \citet{Mac17} on mediation analysis, which has a whole chapter called ``Multilevel Mediation Models''. However, it is mostly based on the \citet{Ken03} and small extensions.
-
-\subsection{\citet{Rit04}}
-
-Points-out that, in mixed-effects models, the regression structure for $Y$ only holds conditional on the random effects.
-Thus, it's hard to give a marginal interpretation of the regression coefficients.
-They give some conditions under which the conditional mean structure is preserved (in a particular sense) after marginalizing away the random effects.
-See, e.g., \citet{Neu91} for more discussion of the problem being addressed.
-
-I think I'm missing something here, because it seems obvious to me that there will be a difference between the parameters in an assumed regression model for the marginal mean of $Y$ and the conditional mean of $Y$ given the random effects.
-Regardless, I don't think this one is particularly relevant to us.
-We have a conditional regression model and don't make any effort to model the marginal mean of $Y$ using regression.
-
-\subsection{\citet{Van10}}
-
-Standard reference for mediation analysis with binary outcome.
-Focuses on the causal inference aspects of the problem.
-Makes heavy use of the ``rare outcome'' assumption.
-Resulting model is sufficiently simple that UQ is easy.
-
-\subsection{\citet{Pea12}}
-
-Presents the so-called ``mediation formula''.
-A set of identities for the total, direct and indirect mediation effects based on counterfactuals, but expressed in terms of conditional expectations.
-Effects are on risk-difference scale.
-
-\subsection{\citet{Lan12}}
-
-Proposes estimating mediation effects using inverse probability weighting.
-Duplicates the observed data and uses weighting to estimate unobserved counterfactual outcomes.
-Referred to as the ``weighting method''.
-Does give analytical (asymptotic) standard errors based on the sandwich variance estimator that arises from the theory of generalized estimating equations.
-See their ``Web Appendix 3'' for more details.
-They also suggest using the bootstrap.
-
-Implemented in the \texttt{R} package \texttt{medflex} \citep{Ste17} alongside the imputation method of \citet{Van12}.
-
-\subsection{\citet{Van12}}
-
-Proposes estimating mediation effects using regression-based imputation.
-Recall that the indirect and direct mediation effects depend on unobservable nested counterfactuals of the form $\phi(x, x') := Y(x, M(x'))$, where $x \neq x'$.
-\citet{Van12} proposes to estimate these counterfactuals directly by first fitting a regression model for the dependence of $\phi$ on $x$ and $x'$ using configurations of these covariates (i.e.
-the exposure) observed in the data.
-We then use this regression model to predict the value of $\phi$ at unobserved configurations.
-
-Caution: This whole method is based on extrapolating from a regression model into a region where we, by definition, have no information.
-The usual warnings apply, both philosophical and practical (e.g.
-don't include high-order polynomial terms in your regression model).
-
-Implemented in the \texttt{R} package \texttt{medflex} \citep{Ste17} alongside the weighting method of \citet{Lan12}.
-
-\subsection{\citet{Nev17}}
-
-Does inference for mediation effects defined using the ``difference method''.
-Briefly, fit models to predict $Y$ using $X$ with or without $M$, then check how much the coefficient on $X$ changes.
-They apply their analysis to continuous, binary (log- and logit-link), and survival data (Cox model).
-
-Inference is done with the $\delta$-method.
-They need the covariance between coefficient estimates from different models (i.e.
-with and without $M$).
-Since this doesn't come directly from a standard analysis, they propose a ``data duplication algorithm''.
-
-No mixed effects. Not obviously relevant.
-
-\subsection{\citet{Che21}}
-
-Defines mediation effects (indirect and direct) as differences between transformed conditional expectations.
-This facilitates use with GLM link functions.
-Does continuous and binary outcome and mediator using GEEs.
-Does UQ with $\delta$-method.
-
-Includes a Monte Carlo study comparing $\delta$-method with bootstrap.
-Finds that pretty large samples are required for $\delta$-method UQ to be accurate, although bootstrap works with somewhat smaller samples.
-
-No mixed-effects.
-
-Does $\delta$-method UQ for GLMs (more precisely, GEEs) with exact formulas for mediation effects.
-These mediation effects (indirect and direct effects) are defined as differences between transformed conditional expectations.
-
-\subsection{\citet{Zhe21}}
-
-Does joint uncertainty quantification for the parameters and random effects.
-Based on maximizing the Laplace Approximation to the marginal likelihood.
-Gives $\delta$-method SEs with gradients from automatic differentiation.
-
-Based on the \texttt{R} package \texttt{TMB}, which does automatic differentiation \citep{Kri16}.
-There is a related package, \texttt{glmmTMB}, which is specifically designed to apply automatic differentiation methods to GLMMs \citep{Bro17} (their paper focuses on zero-inflated count GLMMs, but their methodology applies more generally).
-\citet{Zhe21} raises some concerns about the two \texttt{TMB} packages, and presents a solution.
-See \citet{Ska06} for more discussion of automatic differentiation in GLMM fitting.
-
-\subsection{\citet{Wan18}}
-
-Standard reference for the \texttt{merDeriv} package in \texttt{R}.
-This is a small package that computes the Hessian of the observed data likelihood with respect to all parameters (including RE covariance parameters) for LMMs and GLMMs fit using \texttt{lme4}.
-In particular, use the \texttt{vcov} function with \texttt{full=TRUE} to get the asymptotic covariance matrix of the parameters (i.e.
-the inverse of the negative Hessian of the log-likelihood).
-You can also choose your parameterization for the RE parameters from variances, SDs or terms in the Cholesky decomposition.
-The last one is referred to as \texttt{theta} in \texttt{lme4} and is used extensively therein.
+\begin{outline}
+    \1 Monte Carlo study
+        \2 Proof of concept
+        \2 Explore robustness. See \citet{Sam23} for inspiration.
+    \1 Real data
+        \2 Trust study dataset?
+\end{outline}
 

### 2024-07-11 papers/Manuscript - Exact Asymptotics.tex
Add more organization to manuscript
Finish draft presentation of identification for expected nested counterfactual

@@
 \maketitle
+\section{Notes to Self}
+\begin{outline}
+    \1 In the paper, I organize variables as $Y, X, M, W$.
+This is based on defining counterfactuals as $Y_x$, then adding $M$, giving $Y_{x, m}$.
+In my code, it's always $Y, M, X, W$.
+I must be very careful to not mess this ordering up.
+The code is more arbitrary, but it's also harder to change.
+For now, I'm just going to try to juggle this difference manually, but I will need to carefully validate any results when passing from code to paper (this sort of issue has already been the source of one elusive error in my code).
+\end{outline}
 
@@
 
-\subsection{2024-06-25}
+\subsection{Counterfactual-Based Mediation Effects}
 
@@
 It is nevertheless possible, given certain assumptions, to estimate the expected values of ordinary and nested counterfactuals.
-We will return later to the identification of expected counterfactuals with estimable quantities \hl{(I'm not certain whether it's better to do identification here or later).}
+Section \ref{sec:ID} goes into detail on the identification of expected counterfactuals with estimable quantities.
 
-We now move on to the definition of various mediation effects which are of interest throughout the literature.
 There are three main types of mediation effect: total, direct and indirect, although the scale on which these are measured can vary.
-In the interest of generality, we write $d(y_1, y_2)$ for an arbitrary measure of the discrepancy between $y_1$ and $y_2$.
-The most natural form is $d(y_1, y_2) = y_1 - y_2$, although others may be more appropriate depending on context.
-Note that we do not require $d$ to be symmetric; in fact, the sign of $d$ is often of central scientific importance.
+For concreteness, we focus here only on discrepancies defined as differences; i.e., those of the form $Y_{x_1, M_{x_1'}} - Y_{x_2, M_{x_2'}}$.
+Extending our results to effects defined on different scales (e.g.
+ratios) is straightforward.
+We will return briefly to this in Section ???.
 
-Proceeding now to the actual definitions, we define the total effect of $X$ on $Y$ to be $TE(x, x') = d(EY_x, EY_{x'})$.
-The direct effect is defined as $DE(x, x') = d(Y_{x, M_{x'}}, Y_{x', M_{x'}})$, and the indirect effect is $IE(x, x') = d(Y_{x, M_{x}}, Y_{x, M_{x'}})$.
+Proceeding now to the actual definitions, we define the total effect of $X$ on $Y$ to be $TE(x, x') = EY_x - EY_{x'}$.
+The direct effect is defined as $DE(x, x') = Y_{x, M_{x'}} - Y_{x', M_{x'}}$, and the indirect effect is $IE(x, x') = Y_{x, M_{x}} - Y_{x, M_{x'}}$.
 See, e.g., ???
-for motivation and discussion of these definitions \hl{($g$-function?)}.
+for motivation and discussion of these definitions.
 
-When the outcome, $Y$, is continuous, it is natural to take the discrepancy function to be $d(y_1, y_2) = y_1 - y_2$ as mentioned above.
-However, when $Y$ is binary, it is less obvious which scale to use.
-In this setting, expected counterfactuals are probabilities, so mediation effects measure the discrepancy between two probabilities.
-One option is to use the same difference preferred for continuous outcome; we refer to this as the risk-difference scale.
-Alternatives include $d(y_1, y_2) = y_1 / y_2$, the risk-ratio scale, and $d(y_1, y_2) = [y_1 / (1 - y_1)] / [y_2 / (1 - y_2)]$, the odds-ratio scale.
-We focus only on the binary outcome setting, although our results are easily extended to handle continuous outcomes.
-While our method can handle any choice of discrepancy function $d$ \hl{(differentiable?
-Whatever is necessary to apply the $\delta$-method later; might need to be continuously differentiable)}, in the interest of accessibility we specialize our results to the three discrepancies just mentioned for binary outcomes.
+\subsection{Identification and Modelling of Expected Counterfactuals}
+\label{sec:ID}
 
@@
 We model the latter using regression, either linear or logistic depending on the forms of $Y$ and $M$.
-If $Y$ and $M$ are binary, then Equation (\ref{eq:ID}) has a particularly simple form:
+For concreteness, we take $Y$ and $M$ to both be binary.
+The extension of our method to continuous outcome and/or mediator is straightforward \textcolor{red}{(comment on sums $\rightarrow$ integrals and quadrature?)}.
+In this case, Equation (\ref{eq:ID}) has a particularly simple form:
 %
@@
 %
-We estimate the conditional probabilities on the right-hand side of Equation (\ref{eq:ID_bin}) using logistic regression in the usual way.
-Note that this requires us to fit two models, one to predict $M$ using $X$ and $W$, while the other predicts $Y$ using $M$, $X$ and $W$.
+We estimate the conditional probabilities on the right-hand side of Equation (\ref{eq:ID_bin}) using logistic regression.
+Note that we must fit two models: one to predict $M$ using $X$ and $W$, and another to predict $Y$ using $M$, $X$ and $W$.
 
 We extend this regression modelling with the introduction of random effects \citep[see, e.g.][]{Dem04}.
-Specifically, we include random effects for the intercept and $X$ in our model for $M$, and for the intercept, $M$ and $X$ in our model for $Y$.
-We model these random effects as normally distributed, allowing for correlation within a model, but independent across models.
-Let $U \sim N(0, \Gamma_Y)$ and $V \sim N(0, \Gamma_M)$ be the random effects in our models for $Y$ and $M$ respectively.
-We can now re-write Equation (\ref{eq:ID_bin}) as
+Specifically, we include random effects for the intercept and $X$ in our model for $M$, and for the intercept, $X$ and $M$ in our model for $Y$.
+Let $U$ and $V$ be the random effects for our models of $Y$ and $M$ respectively.
+We can re-write Equation (\ref{eq:ID_bin}) as
 %
@@
 %
+Following the usual approach, we model the random effects $U$ and $V$ as normally distributed, allowing for correlation between effects from the same model, but assuming independence between models.
+Let $U \sim N(0, \Gamma_Y)$ and $V \sim N(0, \Gamma_M)$.
 
-%! START HERE
+\subsubsection{Expanding One Term in Equation (\ref{eq:ID_bin_ran})}
 
-\textcolor{red}{The following discussion is awkward.
-I want to present a general expression for $E_U P(Y | ...)$, then specialize to Equation (\ref{eq:ID_bin_ran}).
-I have defined too much notation, but for now I'm just going to focus on getting ideas down on the page.
-Later, I will edit and tighten-up the presentation.}
+We derive an expression for the first term in Equation \eqref{eq:ID_bin_ran}, then report results for the other three terms \textcolor{red}{(awk)}.
 
-We show an expansion for one term in Equation (\ref{eq:ID_bin_ran}); the rest follow similarly.
-First, write $\eta = (\beta_0 + \beta_X x + \beta_M m + B_W^T w) + (U_0 + U_X x + U_M m)$ for the linear predictor of $Y$ based on $X$, $M$ and $W$.
-Let $\mu = \beta_0 + \beta_X x + \beta_M m + B_W^T w$ be the fixed-effects component of $\eta$, and $\xi = U_0 + U_X x + U_M m$ be its random-effects component.
-For convenience, we will also write $\gamma^2_Y(c_1, c_2, c_3) = (c_1, c_2, c_3) \Gamma_Y (c_1, c_2, c_3)^T$, so that $\bV \xi = \gamma^2_Y(1, m, x)$.
+First, write $\eta_Y = (\alpha_0 + \alpha_X x + \alpha_M m + A_W^T w) + (U_0 + U_X x + U_M m)$ for the linear predictor of $Y$ based on $X$, $M$ and $W$.
+Let $\mu_Y = \alpha_0 + \alpha_X x + A_W^T w$, so that the fixed-effects component of $\eta_Y$ is $\mu_Y + \alpha_M m$, and let $\xi_Y = U_0 + U_X x + U_M m$ be the random-effects component of $\eta_Y$.
+For convenience, we will also write $\gamma^2_Y(c_1, c_2, c_3) = (c_1, c_2, c_3) \Gamma_Y (c_1, c_2, c_3)^T$, so that $\bV \xi_Y = \gamma^2_Y(1, x, m)$ (recall that $\Gamma_Y$ is the covariance matrix of $U$).
 
@@
 \begin{equation}
-    \bP \left( Y=1 |U, X=x, M=1, W=w \right) = \left[1 + \exp(-\eta)\right]^{-1},
+    \bP \left( Y=1 |U, X=x, M=1, W=w \right) = \left[1 + \exp(-\eta_Y)\right]^{-1},
 \end{equation}
@@
 \begin{align}
-    \bE_U \bP \left( Y=1 |U, X=x, M=1, W=w \right) & = \int \frac{\phi_3(u; 0, \Gamma_Y)}{1 + \exp(-\eta)} du, \label{eq:pY1_bad}
+    \bE_U \bP \left( Y=1 |U, X=x, M=1, W=w \right) & = \int \frac{\phi_3(u; 0, \Gamma_Y)}{1 + \exp(-\eta_Y)} du, \label{eq:pY1_bad}
 \end{align}
@@
 \begin{align}
-    \bE_U \bP \left( Y=1 |U, X=x, M=1, W=w \right) & = \int \frac{\phi_1(z; 0, 1)}{1 + \exp(-\mu - \gamma_Y(1, m, x) z)} dz. \label{eq:pY1}
+    \bE_U \bP \left( Y=1 |U, X=x, M=1, W=w \right) & = \int \frac{\phi_1(z; 0, 1)}{1 + \exp(-\mu_Y - \alpha_M - \gamma_Y(1, x, 1) z)} dz. \label{eq:pY1}
 \end{align}
 %
-Importantly, the integral in Equation (\ref{eq:pY1}) is univariate (and thus amenable to numerical evaluation using quadrature), while the integral in Equation (\ref{eq:pY1_bad}) is multivariate.
-
-The integral in Equation (\ref{eq:pY1}) arises often enough that we give it a name. Let
+Importantly, the integral in Equation (\ref{eq:pY1}) is univariate (and thus amenable to numerical evaluation using quadrature). This integral arises often enough that we give it a name. Let
 %
 \begin{equation}
-    \Psi(\mu, \sigma^2) = \int \frac{\phi(z; 0, 1)}{1 + \exp(-\mu - \sigma z)} dz.
+    \Psi(a, b) = \int \frac{\phi(z; 0, 1)}{1 + \exp(-a - b z)} dz.
 \end{equation}
+%
+We can now write the first term in \eqref{eq:ID_bin_ran} compactly as
+%
+\begin{equation}
+    \bE_U \bP \left( Y=1 |U, X=x, M=1, W=w \right) = \Psi(\mu_Y + \alpha_M, \gamma_Y(1, x, 1))
+\end{equation}
+%
+Returning now to the other terms in Equation \eqref{eq:ID_bin_ran}, similar expressions hold.
+Write $\eta_M = (\beta_0 + \beta_X x + B_W^T w) + (V_0 + V_X x)$ for the linear predictor of $M$ based on $X$ and $W$.
+Write $\mu_M = \beta_0 + \beta_X x + B_W^T w$ and $\xi_M = V_0 + V_X x$ for the fixed and random components respectively of $\eta_M$.
+Finally, write $\gamma^2_M(c_1, c_2) = (c_1, c_2) \Gamma_M (c_1, c_2)^T$, so that $\bV \xi_M = \gamma^2_M(1, x)$ ($\Gamma_M$ is the covariance matrix of $V$).
+We can now write the expected counterfactual in \eqref{eq:ID_bin_ran} as
+%
+\begin{align}
+    \bE (Y_{x, M_{x'}} | W=w) = & \left[  \Psi(\mu_Y + \alpha_M, \gamma_Y(1, x, 1)) \cdot \Psi(\mu_M , \gamma_Y(1, x')) \right] + \label{eq:Psi_bin_ran}\\
+    & \left[ \Psi(\mu_Y, \gamma_Y(1, x, 0)) \cdot \Psi(- \mu_M , \gamma_Y(1, x')) \right] \nonumber
+\end{align}
+%
+Recall that each term in \eqref{eq:Psi_bin_ran} is a univariate integral, and can thus be accurately evaluated using standard numerical quadrature routines available in most software packages.
+
+\subsection{Alternative Effect Definitions}
+
+\textcolor{red}{This section needs to be completely re-written. The following might contain useable parts, but no guarantees.}
+
+When the outcome, $Y$, is continuous, it is natural to take the discrepancy function to be $d(y_1, y_2) = y_1 - y_2$ as mentioned above.
+However, when $Y$ is binary, it is less obvious which scale to use.
+In this setting, expected counterfactuals are probabilities, so mediation effects measure the discrepancy between two probabilities.
+One option is to use the same difference preferred for continuous outcome; we refer to this as the risk-difference scale.
+Alternatives include $d(y_1, y_2) = y_1 / y_2$, the risk-ratio scale, and $d(y_1, y_2) = [y_1 / (1 - y_1)] / [y_2 / (1 - y_2)]$, the odds-ratio scale.
+We focus only on the binary outcome setting, although our results are easily extended to handle continuous outcomes.
+While our method can handle any choice of discrepancy function $d$ \hl{(differentiable?
+Whatever is necessary to apply the $\delta$-method later; might need to be continuously differentiable)}, in the interest of accessibility we specialize our results to the three discrepancies just mentioned for binary outcomes.
 

### 2024-07-16 papers/Manuscript - Exact Asymptotics.tex
Minor changes to manuscript
Write-up UQ for expected counterfactuals and mediation effects

@@
 For now, I'm just going to try to juggle this difference manually, but I will need to carefully validate any results when passing from code to paper (this sort of issue has already been the source of one elusive error in my code).
+    \1 I need to be very explicit that our analysis of expected counterfactuals is in service to mediation effects.
+Once we have estimates and UQ for expected counterfactuals, the extension to total, direct and indirect effects is straightforward.
 \end{outline}
@@
 
-\subsection{Counterfactual-Based Mediation Effects}
+\subsection{Counterfactual-Based Mediation Analysis}
+\label{sec:MA_defns}
 
-Our approach to causal mediation analysis is based on the counterfactual framework of ???.
+Our approach to causal mediation analysis is based on the counterfactual framework of ????.
 Briefly, let $Y$ be an outcome of interest, $X$ be an exposure which is a causal driver of $Y$, and $M$ be a mediator, which influences $Y$ and is influenced by $X$.
@@
 ratios) is straightforward.
-We will return briefly to this in Section ???.
+We will return briefly to this in Section ????.
 
@@
 The direct effect is defined as $DE(x, x') = Y_{x, M_{x'}} - Y_{x', M_{x'}}$, and the indirect effect is $IE(x, x') = Y_{x, M_{x}} - Y_{x, M_{x'}}$.
-See, e.g., ???
+See, e.g., ????
 for motivation and discussion of these definitions.
@@
 
+\subsection{Estimation and UQ}
+
+Although not obvious from our notation, Equation \eqref{eq:Psi_bin_ran} depends only on the parameters of our two regression models (as well as the levels of our covariates, $x$, $x'$ and $w$).
+Estimation of the expected counterfactual thus proceeds by first fitting these two regression models to our data, then plugging-in our estimates of the two sets of model parameters to Equation \eqref{eq:Psi_bin_ran}.
+More precisely, if we re-write \eqref{eq:Psi_bin_ran} as
+%
+\begin{equation}
+    \bE (Y_{x, M_{x'}} | W=w) = \kappa(x, x', w; \alpha, \Gamma_Y, \beta, \Gamma_Y) \label{eq:kappa},
+\end{equation}
+%
+then our estimate of this expected counterfactual is
+\begin{align}
+    \hat{\bE} (Y_{x, M_{x'}} | W=w) &= \kappa(x, x', w; \hat{\alpha}, \hat{\Gamma}_Y, \hat{\beta}, \hat{\Gamma}_M) \label{eq:kappa_hat}\\
+    &=: \hat{\kappa}(x, x', w)
+\end{align}
+%
+We often omit the parameters from $\kappa$, and $w$ from both $\kappa$ and $\hat{\kappa}$.
+
+Uncertainty quantification for our estimator, $\hat{\kappa}$, proceeds by the $\delta$-method \citep[see, e.g.,][]{vdV98}.
+We first get asymptotic covariance matrices for estimators of our two sets of regression parameters.
+Next, we stack these two covariances into a single $2\times 2$ block-diagonal matrix (\hl{assuming asymptotic independence between models, see} \citealp{Bau06}), then pre- and post-multiply by the Jacobian of $\kappa$.
+From traditional M-estimator theory, the asymptotic covariance of our regression estimators is approximately the inverse Hessian of their respective objective functions at the maximizers.
+These objective functions differ from likelihoods only in that Laplace's method is used to approximate the marginal likelihood of the data from the joint likelihood between the observed data and random effects.
+
+We omit here an explicit formula for the asymptotic covariance of $\hat{\kappa}$, \textcolor{red}{but one can be found in our \texttt{R} package (\texttt{MultMedUQ}?).
+See also the supplemental material for an accompanying derivation.} For concreteness however, write $\sqrt{n} (\hat{\kappa}(x, x') - \kappa(x, x')) \rightsquigarrow N(0, \lambda^2(x, x'))$.
+We will often also need the limiting covariance of $\kappa$ evaluated at two sets of values for $x$ and $x'$.
+In this case we write $\rho(x_1, x'_1; x_2, x'_2)$ for the asymptotic correlation between $\hat{\kappa}(x_1, x'_1)$ and $\hat{\kappa}(x_2, x'_2)$.
+
+\subsection{Mediation Effects}
+
+Recall that we define the three mediation effects, total, direct and indirect, in terms of expected counterfactuals of the form given in Equation \eqref{eq:kappa}.
+Estimation and uncertainty quantification for mediation effects thus follows immediately from the corresponding analysis of expected counterfactuals.
+Recall from Section \ref{sec:MA_defns} that, if we define effects as differences, then formulas for the three mediation effects are given in Table ????
+(\textcolor{red}{Add a table at end of Section \ref{sec:MA_defns}}).
+Since these mediation effects are all linear combinations of the expected counterfactuals, their asymptotic variances are easily obtained.
+See Table \ref{tab:med_diffs}.
+
+\begin{table}
+    \centering
+    \begin{tabular}{|c|c|c|}
+        \hline
+        Effect & Definition & Asymptotic Variance\\
+        \hline
+        Total & $\kappa(x, x) - \kappa(x', x')$ &  $\lambda^2(x, x) + \lambda^2(x', x') - 2 \lambda(x, x)  \lambda(x', x')  \rho(x, x; x', x')$\\
+        Direct & $\kappa(x, x') - \kappa(x', x')$ & $\lambda^2(x, x') + \lambda^2(x', x') - 2 \lambda(x, x')  \lambda(x', x')  \rho(x, x'; x', x')$\\
+        Indirect & $\kappa(x, x) - \kappa(x, x')$ & $\lambda^2(x, x) + \lambda^2(x, x') - 2 \lambda(x, x)  \lambda(x, x')  \rho(x, x; x, x')$\\
+        \hline
+    \end{tabular}
+    \caption{Mediation effects defined on difference scale.}
+    \label{tab:med_diffs}
+\end{table}
+
 \subsection{Alternative Effect Definitions}

### 2024-07-16 papers/Manuscript - Exact Asymptotics.tex
Finish draft of UQ for mediation effects

@@
 \subsection{Estimation and UQ}
+\label{sec:CF_UQ}
 
@@
 See Table \ref{tab:med_diffs}.
+Alternatively, these asymptotic variances can be derived by applying the $\delta$-method to the map from various $\hat{\kappa}$s to the mediation effect.
+For example, the total effect can be obtained by first computing the asymptotic covariance matrix of $\hat{\kappa}(x,x)$ and $\hat{\kappa}(x',x')$, then pre- and post-multiplying by the gradient of the map $(\kappa_1, \kappa_2) \mapsto \kappa_1 - \kappa_2$.
 
@@
 
-\subsection{Alternative Effect Definitions}
-
-\textcolor{red}{This section needs to be completely re-written. The following might contain useable parts, but no guarantees.}
-
-When the outcome, $Y$, is continuous, it is natural to take the discrepancy function to be $d(y_1, y_2) = y_1 - y_2$ as mentioned above.
-However, when $Y$ is binary, it is less obvious which scale to use.
-In this setting, expected counterfactuals are probabilities, so mediation effects measure the discrepancy between two probabilities.
-One option is to use the same difference preferred for continuous outcome; we refer to this as the risk-difference scale.
-Alternatives include $d(y_1, y_2) = y_1 / y_2$, the risk-ratio scale, and $d(y_1, y_2) = [y_1 / (1 - y_1)] / [y_2 / (1 - y_2)]$, the odds-ratio scale.
-We focus only on the binary outcome setting, although our results are easily extended to handle continuous outcomes.
-While our method can handle any choice of discrepancy function $d$ \hl{(differentiable?
-Whatever is necessary to apply the $\delta$-method later; might need to be continuously differentiable)}, in the interest of accessibility we specialize our results to the three discrepancies just mentioned for binary outcomes.
+Suppose instead that mediation effects are desired on a different scale, such as risk-ratios or odds-ratios for binary outcomes.
+In this case, we adjust the definitions of the various effects accordingly, and estimate each expected counterfactual $\kappa$ with the corresponding plug-in estimator, $\hat{\kappa}$.
+Uncertainty quantification proceeds by applying the $\delta$-method to the function which maps various $\hat{\kappa}$s to the desired mediation effect.
+For example, the limiting variance of the total effect on risk-ratio scale can be obtained by first computing the asymptotic covariance matrix of $\hat{\kappa}(x,x)$ and $\hat{\kappa}(x',x')$, then pre- and post-multiplying by the gradient of the map $(\kappa_1, \kappa_2) \mapsto \kappa_1 / \kappa_2$.
+Formulas for these asymptotic variances tend to be quite long, but are easily implemented computationally (see our package ????).
 

### 2024-07-16 papers/Manuscript - Exact Asymptotics.tex
Add short discussion of simultaneous estimation of mediation effects

@@
 
+While the above discussion focuses on estimation of each mediation effect individually, it is straightforward to extend our analysis to simultaneous estimation of, for example, the total, direct and indirect effects on a single scale.
+This has the advantage of estimating covariances between mediation effects, rather than just the effects' marginal variances.
+
 \subsection{Old}

### 2024-07-17 papers/Manuscript - Exact Asymptotics.tex
Delete old discussion of mediation effects identification

@@
 \newcommand{\zdens}{\frac{e^{-z^2/2}}{\sqrt{2\pi}}}
+\newcommand{\indep}{\perp \!\!\! \perp}
 
@@
 One extension which is of particular interest is to dependent data via multilevel, or mixed-effects, models.
-Mixed-effects regression models involve the introduction of random, unobserved coefficients to an existing regression \hl{(awk)}.
+Mixed-effects regression models involve the introduction of random, unobserved coefficients to an existing regression.
 A common setting in which such a model arises is clustered data, where the random coefficients differ across clusters, but are constant within each cluster.
@@
 In particular, our method can be applied to estimate mediation effects on whatever scale is of interest (e.g., risk difference, risk ration, odds ratio).
+We focus our presentation on a binary outcome and binary mediator, although the extension to one or both being continuous is straightforward.
 
@@
 \begin{outline}
-    \1 Define counterfactuals (ctfs) and nested counterfactuals 
-    \1 Define mediation effects
-        \2 Continuous outcome
-        \2 Binary outcome on risk difference, risk ratio and odds ratio scales
-    \1 Identification of mediation effects
-        \2 See \citet{Ima10I} for non-parametric identification of nested counterfactuals
-    \1 Regression modelling
-        \2 Restrict to binary outcome?
-        \2 Fixed effects regression (brief)
-        \2 Mixed effects regression
-        \2 Prediction of random effects
+    \1 \st{Define counterfactuals (ctfs) and nested counterfactuals }
+    \1 \st{Define mediation effects}
+        \2 \st{Continuous outcome}
+        \2 \st{Binary outcome on risk difference, risk ratio and odds ratio scales}
+    \1 \st{Identification of mediation effects}
+        \2 \st{See} \citet{Ima10I} \st{for non-parametric identification of nested counterfactuals}
+    \1 \st{Regression modelling}
+        \2 \st{Restrict to binary outcome?}
+        \2 \st{Fixed effects regression (brief)}
+        \2 \st{Mixed effects regression}
+        \2 \st{Prediction of random effects}
 \end{outline}
@@
 
-Our approach to causal mediation analysis is based on the counterfactual framework of ????.
+Our approach to causal mediation analysis is based on the well-known counterfactual framework \citep[see, e.g.,][]{Van09,Pea12}.
 Briefly, let $Y$ be an outcome of interest, $X$ be an exposure which is a causal driver of $Y$, and $M$ be a mediator, which influences $Y$ and is influenced by $X$.
-Figure \ref{fig:simp_med} shows a causal diagram representing this relationship.
+Figure \ref{fig:simp_med} shows a minimal causal diagram representing this relationship.
 We call the top arrow the direct effect of $X$ on $Y$, and the bottom path through $M$ the indirect effect of $Y$.
@@
 There are three main types of mediation effect: total, direct and indirect, although the scale on which these are measured can vary.
-For concreteness, we focus here only on discrepancies defined as differences; i.e., those of the form $Y_{x_1, M_{x_1'}} - Y_{x_2, M_{x_2'}}$.
+For concreteness, we focus here only on effects defined as differences; i.e., those of the form $Y_{x_1, M_{x_1'}} - Y_{x_2, M_{x_2'}}$.
 Extending our results to effects defined on different scales (e.g.
-ratios) is straightforward.
-We will return briefly to this in Section ????.
+ratios) is straightforward; see Section \ref{sec:med_effs}.
 
-Proceeding now to the actual definitions, we define the total effect of $X$ on $Y$ to be $TE(x, x') = EY_x - EY_{x'}$.
-The direct effect is defined as $DE(x, x') = Y_{x, M_{x'}} - Y_{x', M_{x'}}$, and the indirect effect is $IE(x, x') = Y_{x, M_{x}} - Y_{x, M_{x'}}$.
-See, e.g., ????
-for motivation and discussion of these definitions.
+Proceeding now to the definitions, we set the total effect of $X$ on $Y$ to be $TE(x, x') = EY_x - EY_{x'}$.
+The direct effect is $DE(x, x') = Y_{x, M_{x'}} - Y_{x', M_{x'}}$, and the indirect effect is $IE(x, x') = Y_{x, M_{x}} - Y_{x, M_{x'}}$.
+See, e.g., \citet{Pea12} for motivation and discussion of these definitions.
 
@@
 In fact, their Theorem 1 is somewhat more general, giving an expression for the density of the nested counterfactual rather than its expected value.
+We henceforth assume that Sequential Ignorability is satisfied by the variables in our model.
 
@@
 %
-Following the usual approach, we model the random effects $U$ and $V$ as normally distributed, allowing for correlation between effects from the same model, but assuming independence between models.
+We model the random effects $U$ and $V$ as normally distributed, allowing for correlation between effects from the same model, but assuming independence between models (i.e.
+$U \indep V$).
 Let $U \sim N(0, \Gamma_Y)$ and $V \sim N(0, \Gamma_M)$.
@@
 From traditional M-estimator theory, the asymptotic covariance of our regression estimators is approximately the inverse Hessian of their respective objective functions at the maximizers.
-These objective functions differ from likelihoods only in that Laplace's method is used to approximate the marginal likelihood of the data from the joint likelihood between the observed data and random effects.
+These objective functions differ from log-likelihoods only in that Laplace's method is used to approximate the marginal likelihood of the data from its joint likelihood with the random effects.
 
-We omit here an explicit formula for the asymptotic covariance of $\hat{\kappa}$, \textcolor{red}{but one can be found in our \texttt{R} package (\texttt{MultMedUQ}?).
+We omit here an explicit formula for the asymptotic covariance of $\hat{\kappa}$, \textcolor{red}{but one can be found in our \texttt{R} package\footnote{Name idea: \texttt{MultMedUQ}?}.
 See also the supplemental material for an accompanying derivation.} For concreteness however, write $\sqrt{n} (\hat{\kappa}(x, x') - \kappa(x, x')) \rightsquigarrow N(0, \lambda^2(x, x'))$.
@@
 \subsection{Mediation Effects}
+\label{sec:med_effs}
 
@@
 
-\subsection{Old}
-
-In this section, we define the total, direct and indirect mediation effects in terms of nested counterfactuals.
-We begin by applying the Mediation Formula of \citet{Pea12} to identify nested counterfactuals with simple functions of conditional expectations.
-
-We begin by observing that mediation effects are often defined in terms of nested counterfactuals (regardless of the scale on which these effects are reported).
-\hl{There is some causal inference theory to be done here.
-For now, I'm just going to write what I expect to be true, then later I will go back and add the necessary assumptions.} As such, we begin by identifying a general nested counterfactual with estimable quantities.
-To this end, write $Y(x, m)$ for the counterfactual value of $Y$ when $X$ and $M$ are set to $x$ and $m$ respectively.
-Similarly, write $M(x)$ for the counterfactual value of $M$ when $X=x$.
-We write $Y(x, M(x'))$ for the value of $Y$ when $X$ is set to $x$ and $M$ is set to what it would have been if $X$ were $x'$.
-We refer to $Y(x, M(x'))$ as a nested counterfactual.
-Note that, when $x \neq x'$, the nested counterfactual is necessarily unobservable.
-Nevertheless, given regularity conditions, we can write
-%
-\begin{equation}
-    \bE Y(x, M(x')) = \int \bE(Y | M=m, X=x) \bP(M = dm | X = x') \label{eq:id_nest}
-\end{equation}
-%
-A similar expression holds conditional on pre-treatment covariates,
-\begin{equation}
-    \bE Y_c(x, M(x')) = \int \bE(Y | M=m, X=x, C=c) \bP(M = dm | X = x', C=c) \label{eq:id_nest_cov}
-\end{equation}
-%
-Having identified expected nested counterfactuals with integrals of conditional quantities, we can use regression to estimate the RHS of Equations \ref{eq:id_nest} and \ref{eq:id_nest_cov}.
-This identification with regression-based quantities is especially simple if $M$ is binary.
-Here, we get $\bP(M=m | \cdot) = \bE(M=m | \cdot)$, where the latter quantity is popularly modelled with logistic regression.
-For simplicity, we hereafter assume that the mediator, $M$, is binary.
-\hl{If not, we need to do something different.
-One option is to extract the conditional density of $(M | \cdot)$, then perform the integral.
-Another is to do regression with response $\bE(Y | M=m, X=x, C=c)$, viewed as a function of $m$.
-There is likely an equivalence between these two methods; I can come back to this later.}
-
-The above development allows only fixed-effects in the regression models.
-If we want to also incorporate random effects, we simply view Equations \ref{eq:id_nest} and \ref{eq:id_nest_cov} as having had these random effects marginalized out.
-That is, if we write $U$ and $V$ for the random effects in our models for $Y$ and $M$ respectively, and $G_V$, $G_V$ for the random effects' distributions, then we can re-write Equations \ref{eq:id_nest} and \ref{eq:id_nest_cov} as
-%
-\begin{align}
-    \bE Y(x, M(x')) &= \int \bE(Y | U=u, M=m, X=x) \bP(M = dm |V=v, X = x') G_U(du) G_V(dv) \label{eq:ran_id_nest} \\
-    \bE Y_c(x, M(x')) &= \int \bE(Y | U=u, M=m, X=x, C=c) \bP(M = dm | V=v, X = x', C=c) G_U(du) G_V(dv) \label{eq:ran_id_nest_cov}
-\end{align}
-
-\subsection{Ideas for CI Theory}
-Applying the Counterfactual Unnesting Theorem of \citet{Cor21}, and provided that the conditions specified therein are satisfied, we have
-
-Alternatively, see Theorem 1 of \citet{Ima10I}.
-This is specifically the result I want, not the corresponding identification results of \citet{Ima10II}, which focuses more narrowly on identification of mediation effects.
-Note that, in \citet{Ima10I}, the function $f$ is a density (with respect to some unspecified measure, possibly counting measure).
-In order to get the above formulas for expected values of $Y$, we do need to assume that $\bE Y$ is finite so we can apply Fubini's Theorem (the one for finite integrals).
-
 \section{Estimation and Inference}
@@
 \begin{outline}
-    \1 Estimate models using \texttt{lme4}
-    \1 UQ for model parameters (mention \texttt{merDeriv} package)
-    \1 UQ for nested counterfactuals via $\delta$-method
-    \1 Simultaneous UQ for triples of mediation effects (i.e. total, direct and indirect)
-        \2 Can also do all 9 simultaneously
-    \1 UQ for predicted group-level effects?
+    \1 \st{Estimate models using \texttt{lme4}}
+    \1 \st{UQ for model parameters (mention \texttt{merDeriv} package)}
+    \1 \st{UQ for nested counterfactuals via $\delta$-method}
+    \1 \st{Simultaneous UQ for triples of mediation effects (i.e. total, direct and indirect)}
+        \2 \st{Can also do all 9 simultaneously}
+    \1 UQ for predicted group-level effects? (not in this paper)
 \end{outline}
@@
     \1 Real data
-        \2 Trust study dataset?
+        \2 Trust study dataset? (this will be the focus of an applied paper)
 \end{outline}

### 2024-07-17 papers/Manuscript - Exact Asymptotics.tex
Delete completed outlines

@@
 
-\begin{outline}
-    \1 Literature review
-        \2 \citeauthor{Sam23}
-        \2 Imai et al.
-        \2 SEM world
-    \1 Overview of our contribution
-        \2 Analytical UQ for mediation effects
-        \2 Existing work is Monte Carlo-based. E.g. Imai's method, bootstrap
-\end{outline}
-
 Mediation analysis is a central problem in modern causal inference.
@@
 \section{Multilevel Mediation Analysis}
-
-\begin{outline}
-    \1 \st{Define counterfactuals (ctfs) and nested counterfactuals }
-    \1 \st{Define mediation effects}
-        \2 \st{Continuous outcome}
-        \2 \st{Binary outcome on risk difference, risk ratio and odds ratio scales}
-    \1 \st{Identification of mediation effects}
-        \2 \st{See} \citet{Ima10I} \st{for non-parametric identification of nested counterfactuals}
-    \1 \st{Regression modelling}
-        \2 \st{Restrict to binary outcome?}
-        \2 \st{Fixed effects regression (brief)}
-        \2 \st{Mixed effects regression}
-        \2 \st{Prediction of random effects}
-\end{outline}
 
@@
 
-\section{Estimation and Inference}
-
-\begin{outline}
-    \1 \st{Estimate models using \texttt{lme4}}
-    \1 \st{UQ for model parameters (mention \texttt{merDeriv} package)}
-    \1 \st{UQ for nested counterfactuals via $\delta$-method}
-    \1 \st{Simultaneous UQ for triples of mediation effects (i.e. total, direct and indirect)}
-        \2 \st{Can also do all 9 simultaneously}
-    \1 UQ for predicted group-level effects? (not in this paper)
-\end{outline}
-
 \section{Empirical Investigation}

### 2024-07-17 papers/Manuscript - Exact Asymptotics.tex
One pass at editing the manuscript

@@
 Mediation analysis is a central problem in modern causal inference.
-Many scientific and public health questions have the form of separating the direct effect of some exposure on an outcome from the indirect effect of that exposure via a mediator.
-Many authors have developed methods to address this problem; gradually increasing in complexity.
-Early work by \citet{Bar86} laid the foundation for future developments, but did not use the machinery of counterfactual outcomes.
+Many scientific and public health questions require separating the direct effect of some exposure on an outcome from the indirect effect of that exposure via a mediator.
+Many authors have developed methods to address this problem.
+Notable early work includes \citet{Bar86}, who laid the foundation for future developments, but did not use the machinery of counterfactual outcomes.
 
-There have been several approaches to the analysis of causal mediation.
-One group established a non-parametric identification result \citep{Ima10II}, and used this result to estimate mediation effects in various contexts \citep{Ima10I, Ima11}.
+There have been several approaches to embedding mediation analysis within the framework of causal inference.
+One group establishes a non-parametric identification result \citep{Ima10II}, and uses this result to estimate mediation effects in various contexts \citep{Ima10I, Ima11}.
 Their methodology is implemented in the \texttt{R} package \texttt{mediation} \citep{Tin14}.
@@
 
-Another approach began with the work of \citet{Van09} on continuous outcomes, and was later extended to handle various modifications to the basic model \citep{Van10, Van13, Van14}.
-Of particular interest to us is the modification to handle binary outcomes \citep{Van10}.
-Effects are defined on the odds-ratio scale, and the outcome is assumed to be rare.
+Another approach begins with the work of \citet{Van09} on the identification and estimation of mediation analysis with continuous outcomes.
+This work is later extended in various ways \citep{Van10, Van13, Van14}.
+Of particular interest to us is the work by \citet{Van10}, which allows for binary outcomes.
+These authors define effects on the odds-ratio scale, and the outcome is assumed to be rare.
 Later work by \citet{Sam21} removes the rare-outcome assumption and extends the work of \citet{Van09} to handle effects on the risk-difference, risk-ratio and odds-ratio scales.
@@
 
-Briefly, causal mediation analysis is based on the counterfactual, or potential outcome framework.
-Let $Y$ be an outcome of interest and $X$ be an exposure.
-We write $Y(x)$ for the value $Y$ would have attained if, possibly counter to fact, $X$ had been set to the value $x$.
-Introducing a mediator, $M$, we write $M(x)$ for the value $M$ would have attained if $X=x$, and $Y(x, m)$ for the value of $Y$ when $X=x$ and $M=m$.
-Note that every individual in the population has values for each of the above quantities, $Y(x)$, $M(x)$, $Y(x,m)$ at every possible value of $x$ and $m$.
+Briefly, causal mediation analysis is based on the counterfactual, or potential outcome framework \citep[see, e.g.,][]{Van09,Pea12}.
+Let $Y$ be an outcome of interest which depends on an exposure variable, $X$.
+We write $Y_x$ for the value $Y$ would have attained if, possibly counter to fact, $X$ had been set to the value $x$.
+Consider a third variable, $M$, known as a mediator, which influences $Y$ and is influenced by $X$.
+We write $M_x$ for the value $M$ would have attained if $X=x$, and $Y_{x,m}$ for the value of $Y$ when $X=x$ and $M=m$.
+Note that every individual in the population has values for each of the above quantities, $Y_x$, $M_x$, $Y_{x,m}$ at every possible value of $x$ and $m$.
 Unfortunately, in practice we only observe $Y$ and $M$ for the values of $x$ and $m$ which actually occurred.
-This is known as the ``fundamental problem of causal inference'' \citep{Din18, Hol86}.
+This limit on what can possibly be measured is known as the ``fundamental problem of causal inference'' \citep{Din18, Hol86}.
 
@@
 Under standard assumptions, such as consistency and no unmeasured confounders \citep[see, e.g.,][]{Pea09}, we can estimate expected conterfactuals as functions of conditional expectations.
-From this point, mediation analysis reduces to a problem of classical statistics; one which can be solved using traditional regression methodology.
+This estimability is referred to as ``identification'' of the expected counterfactuals.
+Assuming that identification holds, mediation analysis reduces to a problem of estimating conditional expectations, which is extremely well studied in the linear and non-linear regression literature.
 The difference between continuous and binary outcomes (or mediators) is essentially addressed by choosing between linear and logistic regression.
@@
 One extension which is of particular interest is to dependent data via multilevel, or mixed-effects, models.
-Mixed-effects regression models involve the introduction of random, unobserved coefficients to an existing regression.
+Mixed-effects regression involve the introduction of random, unobserved coefficients to an existing regression.
 A common setting in which such a model arises is clustered data, where the random coefficients differ across clusters, but are constant within each cluster.
 See, e.g., \citet{Dem04} for an overview of mixed-effects methodology.
-When applied to mediation analysis, mixed-effects methods allow for modelling group-specific mediation effects, and the effect of this heterogeneity on the estimation of global effects.
+When applied to mediation analysis, mixed-effects methods allow for modelling group-specific mediation effects, as well as the effect of this heterogeneity on the estimation of global effects.
 
@@
 In particular, our method can be applied to estimate mediation effects on whatever scale is of interest (e.g., risk difference, risk ration, odds ratio).
-We focus our presentation on a binary outcome and binary mediator, although the extension to one or both being continuous is straightforward.
+Our model extends the work of \citet{Sam23} both by allowing for arbitrarily distributed outcome and mediator (instead of requiring a specific configuration of discrete or continuous variables), and by the introduction of random-effects.
+For concreteness, we focus our presentation on a binary outcome and binary mediator, although the extension to one or both being continuous is straightforward.
 
@@
 
-We define the counterfactual $Y_x$ as the value $Y$ would assume when $X=x$.
-Similarly, write $M_x$ for the value $M$ would assume when $X=x$.
-Next, write $Y_{x,m}$ for the value $Y$ would assume if $X$ and $M$ were set to $x$ and $m$ respectively.
-Combining these ideas, we get the ``nested counterfactual'' $Y_{x, M_{x'}}$, which is the value of $Y$ when $X=x$ and $M$ is set to whatever it would have been when $X=x'$.
+Recall that we define $Y_x$, $M_x$ and $Y_{x,m}$ as the values assumed by $Y$ or $M$ when we either set $X=x$ or simultaneously set $X=x$ and $M=m$.
+Combining these ideas, we define the ``nested counterfactual'' $Y_{x, M_{x'}}$ as the value of $Y$ when $X=x$ and $M$ is set to whatever it would have been when $X=x'$.
 It is common to link ordinary and nested conterfactuals by making the consistency assumption, which states $Y_x = Y_{x, M_x}$.
@@
 For concreteness, we focus here only on effects defined as differences; i.e., those of the form $Y_{x_1, M_{x_1'}} - Y_{x_2, M_{x_2'}}$.
-Extending our results to effects defined on different scales (e.g.
+Extending our results to effects defined on other scales (e.g.
 ratios) is straightforward; see Section \ref{sec:med_effs}.
 
-Proceeding now to the definitions, we set the total effect of $X$ on $Y$ to be $TE(x, x') = EY_x - EY_{x'}$.
-The direct effect is $DE(x, x') = Y_{x, M_{x'}} - Y_{x', M_{x'}}$, and the indirect effect is $IE(x, x') = Y_{x, M_{x}} - Y_{x, M_{x'}}$.
+Proceeding now to the effects, we define the total effect of $X$ on $Y$ to be $TE(x, x') = EY_x - EY_{x'}$.
+We define the direct effect as $DE(x, x') = Y_{x, M_{x'}} - Y_{x', M_{x'}}$, and the indirect effect as $IE(x, x') = Y_{x, M_{x}} - Y_{x, M_{x'}}$.
 See, e.g., \citet{Pea12} for motivation and discussion of these definitions.
@@
 
-A fundamental problem of causal inference is that we can only ever observe one counterfactual outcome on a particular individual.
-In mediation analysis, this problem is even worse, since many of our definitions involve the nested counterfactual, $Y_{x, M_{x'}}$, which when $x \neq x'$ cannot be observed on any individual.
+The so-called fundamental problem of causal inference is that we can only ever observe one counterfactual outcome on a particular individual.
+In mediation analysis, this problem is even worse, since many of our definitions involve the nested counterfactual, $Y_{x, M_{x'}}$, which when $x \neq x'$ is unobservable by construction.
 Nevertheless, \citet{Ima10I} give conditions under which the population average of a nested counterfactual can be expressed in terms of conditional expectations, possibly conditional on one or more additional covariates, $W$.
@@
 For concreteness, we take $Y$ and $M$ to both be binary.
-The extension of our method to continuous outcome and/or mediator is straightforward \textcolor{red}{(comment on sums $\rightarrow$ integrals and quadrature?)}.
+The extension of our method to continuous outcome and/or mediator is straightforward.
 In this case, Equation (\ref{eq:ID}) has a particularly simple form:
@@
 
-We extend this regression modelling with the introduction of random effects \citep[see, e.g.][]{Dem04}.
-Specifically, we include random effects for the intercept and $X$ in our model for $M$, and for the intercept, $X$ and $M$ in our model for $Y$.
-Let $U$ and $V$ be the random effects for our models of $Y$ and $M$ respectively.
+We extend the above regression models by introducing random-effects \citep[see, e.g.][]{Dem04}.
+Specifically, we include random-effects for the intercept and $X$ in our model for $M$, and for the intercept, $X$ and $M$ in our model for $Y$.
+Let $U$ and $V$ be the random-effects for our models of $Y$ and $M$ respectively.
 We can re-write Equation (\ref{eq:ID_bin}) as
@@
 %
-We model the random effects $U$ and $V$ as normally distributed, allowing for correlation between effects from the same model, but assuming independence between models (i.e.
+We model the random-effects $U$ and $V$ as normally distributed, allowing for correlation between effects from the same model, but assuming independence between models (i.e.
 $U \indep V$).
@@
 
-We derive an expression for the first term in Equation \eqref{eq:ID_bin_ran}, then report results for the other three terms \textcolor{red}{(awk)}.
+In order to obtain a more useful expression for the expected counterfactual $\bE (Y_{x, M_{x'}} | W=w)$, we must expand the terms on the right-hand side of \eqref{eq:ID_bin_ran}.
+We derive such an expression for the first term, then report similar expansions for the other three terms.
 
@@
 Let $\mu_Y = \alpha_0 + \alpha_X x + A_W^T w$, so that the fixed-effects component of $\eta_Y$ is $\mu_Y + \alpha_M m$, and let $\xi_Y = U_0 + U_X x + U_M m$ be the random-effects component of $\eta_Y$.
-For convenience, we will also write $\gamma^2_Y(c_1, c_2, c_3) = (c_1, c_2, c_3) \Gamma_Y (c_1, c_2, c_3)^T$, so that $\bV \xi_Y = \gamma^2_Y(1, x, m)$ (recall that $\Gamma_Y$ is the covariance matrix of $U$).
+For convenience, we will also write $\gamma^2_Y(c_1, c_2, c_3) = (c_1, c_2, c_3) \Gamma_Y (c_1, c_2, c_3)^T$, so that $\bV \eta_Y = \bV \xi_Y = \gamma^2_Y(1, x, m)$ (recall that $\Gamma_Y$ is the covariance matrix of $U$).
 
@@
 %
-Importantly, the integral in Equation (\ref{eq:pY1}) is univariate (and thus amenable to numerical evaluation using quadrature). This integral arises often enough that we give it a name. Let
+Importantly, the integral in Equation (\ref{eq:pY1}) is univariate (and thus amenable to numerical evaluation using quadrature).
+The form of this integral arises frequently enough that we give it a name.
+Let
 %
@@
 %
-Returning now to the other terms in Equation \eqref{eq:ID_bin_ran}, similar expressions hold.
+Returning to the other terms in Equation \eqref{eq:ID_bin_ran}, similar expressions hold.
 Write $\eta_M = (\beta_0 + \beta_X x + B_W^T w) + (V_0 + V_X x)$ for the linear predictor of $M$ based on $X$ and $W$.
@@
 
-Although not obvious from our notation, Equation \eqref{eq:Psi_bin_ran} depends only on the parameters of our two regression models (as well as the levels of our covariates, $x$, $x'$ and $w$).
+Although not explicit in our notation, Equation \eqref{eq:Psi_bin_ran} depends only on the parameters of our two regression models (as well as the levels of our covariates, $x$, $x'$ and $w$).
 Estimation of the expected counterfactual thus proceeds by first fitting these two regression models to our data, then plugging-in our estimates of the two sets of model parameters to Equation \eqref{eq:Psi_bin_ran}.
@@
 %
-\begin{equation}
-    \bE (Y_{x, M_{x'}} | W=w) = \kappa(x, x', w; \alpha, \Gamma_Y, \beta, \Gamma_Y) \label{eq:kappa},
-\end{equation}
+\begin{align}
+    \bE (Y_{x, M_{x'}} | W=w) = \kappa(x, x', w; \alpha, \Gamma_Y, \beta, \Gamma_Y) \label{eq:kappa}\\
+    &= \kappa(x, x', w), 
+\end{align}
 %
@@
 %
-We often omit the parameters from $\kappa$, and $w$ from both $\kappa$ and $\hat{\kappa}$.
+We often suppress $w$ in both $\kappa$ and $\hat{\kappa}$.
 
@@
 Next, we stack these two covariances into a single $2\times 2$ block-diagonal matrix (\hl{assuming asymptotic independence between models, see} \citealp{Bau06}), then pre- and post-multiply by the Jacobian of $\kappa$.
-From traditional M-estimator theory, the asymptotic covariance of our regression estimators is approximately the inverse Hessian of their respective objective functions at the maximizers.
-These objective functions differ from log-likelihoods only in that Laplace's method is used to approximate the marginal likelihood of the data from its joint likelihood with the random effects.
+From traditional M-estimator theory, the asymptotic covariance of our regression estimators is approximately the inverse Hessian of their respective objective functions at the maximizers (given appropriate regularity conditions).
+These objective functions differ from log-likelihoods only in that Laplace's method is used to approximate the marginal likelihood of the data from its joint likelihood with the random-effects.
 
@@
 See also the supplemental material for an accompanying derivation.} For concreteness however, write $\sqrt{n} (\hat{\kappa}(x, x') - \kappa(x, x')) \rightsquigarrow N(0, \lambda^2(x, x'))$.
-We will often also need the limiting covariance of $\kappa$ evaluated at two sets of values for $x$ and $x'$.
+We will often also need the limiting joint dependence of $\kappa$ evaluated at two sets of values for $x$ and $x'$.
 In this case we write $\rho(x_1, x'_1; x_2, x'_2)$ for the asymptotic correlation between $\hat{\kappa}(x_1, x'_1)$ and $\hat{\kappa}(x_2, x'_2)$.
@@
 
-Recall that we define the three mediation effects, total, direct and indirect, in terms of expected counterfactuals of the form given in Equation \eqref{eq:kappa}.
+Recall from Section \ref{sec:MA_defns} that we define the three mediation effects, total, direct and indirect, in terms of expected counterfactuals of the form given in Equation \eqref{eq:kappa}.
 Estimation and uncertainty quantification for mediation effects thus follows immediately from the corresponding analysis of expected counterfactuals.
@@
 See Table \ref{tab:med_diffs}.
-Alternatively, these asymptotic variances can be derived by applying the $\delta$-method to the map from various $\hat{\kappa}$s to the mediation effect.
+Alternatively, these asymptotic variances can be derived by applying the $\delta$-method to the map from the relevant $\hat{\kappa}$'s to a particular mediation effect.
 For example, the total effect can be obtained by first computing the asymptotic covariance matrix of $\hat{\kappa}(x,x)$ and $\hat{\kappa}(x',x')$, then pre- and post-multiplying by the gradient of the map $(\kappa_1, \kappa_2) \mapsto \kappa_1 - \kappa_2$.
+Since this map is linear, the $\delta$-method is exact.
 
@@
 In this case, we adjust the definitions of the various effects accordingly, and estimate each expected counterfactual $\kappa$ with the corresponding plug-in estimator, $\hat{\kappa}$.
-Uncertainty quantification proceeds by applying the $\delta$-method to the function which maps various $\hat{\kappa}$s to the desired mediation effect.
+Uncertainty quantification proceeds by applying the $\delta$-method to the function which maps the relevant $\hat{\kappa}$'s to the desired mediation effect.
 For example, the limiting variance of the total effect on risk-ratio scale can be obtained by first computing the asymptotic covariance matrix of $\hat{\kappa}(x,x)$ and $\hat{\kappa}(x',x')$, then pre- and post-multiplying by the gradient of the map $(\kappa_1, \kappa_2) \mapsto \kappa_1 / \kappa_2$.
-Formulas for these asymptotic variances tend to be quite long, but are easily implemented computationally (see our package ????).
+Formulas for these asymptotic variances tend to be quite long, but are easily implemented in software (see our package ????).
 
 While the above discussion focuses on estimation of each mediation effect individually, it is straightforward to extend our analysis to simultaneous estimation of, for example, the total, direct and indirect effects on a single scale.
-This has the advantage of estimating covariances between mediation effects, rather than just the effects' marginal variances.
+This has the advantage of also yielding estimated covariances between mediation effects, rather than just the effects' marginal variances.
 

### 2025-02-03 papers/Manuscript - Exact Asymptotics (William).tex
Copy manuscript for uploading to Overleaf
MacEwan Presentation

--- 2024-07-17 papers/Manuscript - Exact Asymptotics.tex
@@
 Once we have estimates and UQ for expected counterfactuals, the extension to total, direct and indirect effects is straightforward.
+    \1 I have written ????
+everywhere that I need to track-down a citation.
+This has the advantages of both showing up when I search for ???
+(indicating a broken reference), and being different from ???
+(so I can tell that it's deliberate, not a broken reference).
 \end{outline}
@@
 \begin{align}
-    \bE (Y_{x, M_{x'}} | W=w) = & \left[  \Psi(\mu_Y + \alpha_M, \gamma_Y(1, x, 1)) \cdot \Psi(\mu_M , \gamma_Y(1, x')) \right] + \label{eq:Psi_bin_ran}\\
-    & \left[ \Psi(\mu_Y, \gamma_Y(1, x, 0)) \cdot \Psi(- \mu_M , \gamma_Y(1, x')) \right] \nonumber
+    \bE (Y_{x, M_{x'}} | W=w) = & \left[  \Psi(\mu_Y + \alpha_M, \gamma_Y(1, x, 1)) \cdot \Psi(\mu_M , \gamma_M(1, x')) \right] + \label{eq:Psi_bin_ran}\\
+    & \left[ \Psi(\mu_Y, \gamma_Y(1, x, 0)) \cdot \Psi(- \mu_M , \gamma_M(1, x')) \right] \nonumber
 \end{align}
@@
 \begin{align}
-    \bE (Y_{x, M_{x'}} | W=w) = \kappa(x, x', w; \alpha, \Gamma_Y, \beta, \Gamma_Y) \label{eq:kappa}\\
+    \bE (Y_{x, M_{x'}} | W=w) &= \kappa(x, x', w; \alpha, \Gamma_Y, \beta, \Gamma_Y) \label{eq:kappa}\\
     &= \kappa(x, x', w), 
@@
 Next, we stack these two covariances into a single $2\times 2$ block-diagonal matrix (\hl{assuming asymptotic independence between models, see} \citealp{Bau06}), then pre- and post-multiply by the Jacobian of $\kappa$.
-From traditional M-estimator theory, the asymptotic covariance of our regression estimators is approximately the inverse Hessian of their respective objective functions at the maximizers (given appropriate regularity conditions).
+From classical M-estimator theory, the asymptotic covariance of our regression estimators is approximately the inverse Hessian of their respective objective functions at the maximizers (given appropriate regularity conditions).
 These objective functions differ from log-likelihoods only in that Laplace's method is used to approximate the marginal likelihood of the data from its joint likelihood with the random-effects.
@@
 We omit here an explicit formula for the asymptotic covariance of $\hat{\kappa}$, \textcolor{red}{but one can be found in our \texttt{R} package\footnote{Name idea: \texttt{MultMedUQ}?}.
-See also the supplemental material for an accompanying derivation.} For concreteness however, write $\sqrt{n} (\hat{\kappa}(x, x') - \kappa(x, x')) \rightsquigarrow N(0, \lambda^2(x, x'))$.
+See also the supplemental material for an accompanying derivation.} For concreteness however, write $\sqrt{n} (\hat{\kappa}(x, x') - \kappa(x, x')) \rightsquigarrow N(0, \lambda^2(x, x'))$, where $\rightsquigarrow$ denotes convergence in distribution.
 We will often also need the limiting joint dependence of $\kappa$ evaluated at two sets of values for $x$ and $x'$.

### 2025-02-03 presentations/MacEwan 2025-01/SSC 2024.tex
MacEwan Presentation

--- 2024-05-30 presentations/SSC 2024/SSC 2024.tex
@@
     \begin{itemize}
-        \item I.e. Add random effects which vary across clusters \newline
+        \item I.e. Add random effects that vary across clusters \newline
     \end{itemize}
@@
     \begin{itemize}
-        \item Coefficients of these new covariates are random variables which vary across groups/clusters \newline
+        \item Coefficients of these new covariates are random variables that vary across groups/clusters \newline
     \end{itemize}
@@
     \begin{itemize}
-        \item Old model: $Y = \alpha_0 + \alpha_1 X_1 + \ldots + \alpha_p X_p + \varepsilon$
-        \item New model: $Y = \alpha_0 + \alpha_1 X_1 + \ldots + \alpha_p X_p + \mathbf{u_1 Z_1 + \ldots + u_q Z_q} + \varepsilon$ \newline
+        \item Old model: \newline
+        $Y = \alpha_0 + \alpha_1 X_1 + \ldots + \alpha_p X_p + \varepsilon$
+        \item New model:  \newline
+        $Y = \alpha_0 + \alpha_1 X_1 + \ldots + \alpha_p X_p$ \newline
+        \hspace{1cm} $\mathbf{+ u_1 Z_1 + \ldots + u_q Z_q} + \varepsilon$ 
     \end{itemize}
@@
     \begin{itemize}
-        \item \textbf{Bootstrap}
+        \item \textbf{Bootstrap} (Rado Ramasy, yesterday)
         \item Quasi-Bayesian Monte Carlo

### 2025-02-03 scripts/Old_vs_New_SEs.R

@@ -1,0 +1,256 @@
+
+
+j = 1
+
+n = N_K_combs[j, "N"]
+K = N_K_combs[j, "K"]
+
+# Reference values for X and W, the confounders
+## Note: We compute the mediation effect at X_pred vs X_pred + 1, so X_pred=0 is a good choice
+X_pred = 0
+W_pred = c(1,1,1)
+
+
+
+# Use tryCatch to omit any datasets leading to convergence issues
+tryCatch({
+  ############## Un-quote
+  all_Xs = list()
+  all_Ws = list()
+
+  for (k in 1:K) {
+    X = rnorm(n, mean = 0, sd = 1)
+    W = matrix(rnorm(n * p_conf, mean = 0, sd = 1),
+               nrow = n,
+               ncol = p_conf)
+
+    all_Xs[[k]] = X
+    all_Ws[[k]] = W
+  }
+
+  # Generate M
+  all_Ms = list()
+  for (k in 1:K) {
+    eta_vec_fixed = a_0 + a_1 * all_Xs[[k]] + all_Ws[[k]] %*% A_2
+
+    ## Add random effects
+    a_ran = mvrnorm(1, mu = rep(0, 2), Sigma = Sigma_a)
+    eta_vec = eta_vec_fixed + a_ran[1] + a_ran[2] * all_Xs[[k]]
+
+    ## Generate M
+    p_M_vec = expit(eta_vec)
+    M = rbinom(n, size = 1, prob = p_M_vec)
+    all_Ms[[k]] = M
+  }
+
+  # Generate Y
+  all_Ys = list()
+  for (k in 1:K) {
+    zeta_vec_fixed = b_0 + b_1 * all_Ms[[k]] + b_2 * all_Xs[[k]] + all_Ws[[k]] %*%
+      B_3
+
+    ## Add random effects
+    b_ran = mvrnorm(1, mu = rep(0, 2), Sigma = Sigma_b)
+    zeta_vec = zeta_vec_fixed + b_ran[1] + b_ran[2] * all_Xs[[k]]
+
+    ## Generate Y
+    p_Y_vec = expit(zeta_vec)
+    Y = rbinom(n, size = 1, prob = p_Y_vec)
+    all_Ys[[k]] = Y
+  }
+
+
+  # Consolidate groups
+  X = do.call(c, all_Xs)
+  W = do.call(rbind, all_Ws)
+  M = do.call(c, all_Ms)
[…]

### 2025-02-06 presentations/Nasri Lab 2025/Nasri Lab 2025.tex
Presentation for Bouchra's group

--- 2024-05-30 presentations/SSC 2024/SSC 2024.tex
@@
 \begin{frame}{Mediation Analysis}
-    Separate \textbf{Total Effect} of $X$ on $Y$ into
-    \begin{itemize}
-        \item \textbf{Direct Effect}
-        \item \textbf{Indirect Effect}\newline
-    \end{itemize}
+    \begin{outline}
+        \1 Separate \textbf{Total Effect} of $X$ on $Y$ into
+            \2 \textbf{Direct Effect}
+            \2 \textbf{Indirect Effect}\newline
 
-    Traditionally, use regression
+        \1 Define mediation effects using counterfactuals
+        \1 Differences, ratios, odds-ratios
+    \end{outline}    
 \end{frame}
@@
 \begin{frame}{Mediation Analysis}
-    Continuous outcome and mediator:
-    \begin{itemize}
-        \item $Y = \alpha_0 + \alpha_1 M + \alpha_2 X + \varepsilon_Y$
-        \item $M = \beta_0 + \beta_1 X + \varepsilon_M$ \newline
-    \end{itemize}
+    \begin{outline}
+        \1 We only observe one outcome per individual\newline
 
-    \textbf{Direct Effect}: $\alpha_2$
-    \begin{itemize}
-        \item ``$X$ in $Y$''
-    \end{itemize}
-    \textbf{Indirect Effect}: $\alpha_1 \cdot \beta_1$
-    \begin{itemize}
-        \item ``$M$ in $Y$'' $\cdot$ ``$X$ in $M$''
-    \end{itemize}
-    \textbf{Total Effect}: $\alpha_2 + \alpha_1 \cdot \beta_1$
+        \1 Explore population-level effects by averaging\newline
 
+        \1 Identify expected counterfactuals with conditional expectations
+            \2 Now it's a regression problem
+    \end{outline}
 \end{frame}
 
-\begin{frame}{Mediation Analysis}
-    Popular approach
-    \begin{itemize}
-        \item A bit outdated\ldots \newline
-    \end{itemize}
+\begin{frame}
+    \frametitle{Mediation Analysis}
+    \begin{outline}
+        \1 Fit two regression models
+            \2 Mediator given exposure
+            \2 Outcome given mediator and exposure
+            \2 Both may include confounders \newline
 
-    More popular: Causal mediation analysis
-
+        \1 Linear vs Logistic
+        \1 Fixed- vs Mixed-Effects
+            \2 Single- vs Multi-Level
+    \end{outline}
 \end{frame}
 
-\begin{frame}{Causal Inference}
-    Assume that $X$ \textit{causes} $Y$\newline
+\begin{frame}
+    \frametitle{Mediation Analysis}
+    \begin{outline}
+        \1 Fit two regression models
+            \2 Mediator given exposure
+            \2 Outcome given mediator and exposure
+            \2 Both may include confounders \newline
 
-    Counterfactuals:
-    \begin{itemize}
-        \item What value would $Y$ take if $X$ were set to a particular level?
-        \item Write $Y_x$ for the value of $Y$ when $X=x$
-        \item If $X\neq x$ then $Y_x$ is literally a ``counterfactual'' 
-    \end{itemize}
+        \1 Linear vs \textbf{Logistic}
+        \1 Fixed- vs \textbf{Mixed-Effects}
+            \2 Single- vs Multi-Level
+    \end{outline}
 \end{frame}
 
-\begin{frame}{Causal Inference}
-    Example:
-    \begin{itemize}
-        \item Alice only reads scientific publications and will follow all lockdown mandates
-        \item What if she instead only read Facebook?\newline
-        \item $Y_{Science}(Alice) = \mathrm{follow}$
-        \item $Y_{Facebook}(Alice) = \mathrm{follow}$
-    \end{itemize}
+\begin{frame}
+    \frametitle{Mediation Analysis}
+    \begin{outline}
+        \1 Model fitting involves intractable integrals over unobserved random effects
+            \2 Evaluate using quadrature 
+        \1 Estimated mediation effects depend on coefficients and RE covariances\newline
+
+        \1 Uncertainty quantification:
+            \2 Quasi-Bayesian Monte Carlo
+            \2 $\delta$-method
+    \end{outline}
 \end{frame}
 
-\begin{frame}{Causal Inference}
-    Example:
-    \begin{itemize}
-        \item Bob also only reads scientific publications and will follow all lockdown mandates, but is more susceptible to being influenced \newline
-        \item $Y_{Science}(Bob) = \mathrm{follow}$
-        \item $Y_{Facebook}(Bob) = \mathrm{not\ follow}$
-    \end{itemize}
+\begin{frame}
+    \frametitle{Uncertainty Quantification}
+    \begin{outline}
+        \1 Quasi-Bayesian Monte Carlo
+            \2 Monte Carlo $\delta$-Method \newline
+        \end{outline}
+
+        \begin{enumerate}
+            \item Estimate sampling distribution of regression parameters
+            \item Simulate from estimated sampling distribution
+            \item Compute mediation effects for simulated parameters
+        \end{enumerate}
 \end{frame}
 
-\begin{frame}{Causal Inference}
-    \begin{itemize}
-        \item We only observe one outcome per individual\newline
+\begin{frame}
+    \frametitle{Uncertainty Quantification}
+    \begin{outline}
+        \1 Quasi-Bayesian Monte Carlo
+            \2 Monte Carlo $\delta$-Method \newline
 
-        \item Explore population-level effects by averaging\newline
+        \1 Advantage: Flexible        
+        \1 Disadvantage: Computational \newline
 
-        \item Define mediation effects in terms of expected counterfactuals
-    \end{itemize}
+        \1 Existing implementation in \texttt{R}: \textit{mediation}
+            \2 Limited
+        \end{outline}
 \end{frame}
 
-\begin{frame}{Causal Inference}
-    \textbf{Total Effect}: $\mathbb{E}(Y_{x'} - Y_{x})$
-    \begin{itemize}
-        \item Effect on outcome when we change exposure from $X=x$ to $X=x'$ \newline
-    \end{itemize}
+\begin{frame}
+    \frametitle{Uncertainty Quantification}
+    \begin{outline}
+        \1 $\delta$-Method \newline
+        \end{outline}
 
-    Other effects involve dependence on a mediator:
-    \begin{itemize}
-        \item $Y_{xm}$: Value of outcome when
-        \begin{itemize}
-            \item Exposure ($X$) is set to $x$
-            \item Mediator ($M$) is set to $m$
-        \end{itemize}
-        \item $M_x$: Value of mediator when
-        \begin{itemize}
-            \item Exposure ($X$) is set to $x$
-        \end{itemize}
-        \item ``Nested Counterfactuals'': $Y_{x M_x}$ or $Y_{x M_{x'}}$
-    \end{itemize}
-
+        \begin{enumerate}
+            \item Estimate (asymptotic) sampling distribution of regression parameters
+            \item Compute Jacobian of map from reg pars to mediation effects
+            \item Multiply asymp. covariance by Jacobian
+        \end{enumerate}
 \end{frame}
 
-\begin{frame}{Causal Mediation Analysis}
-    \textbf{Controlled Direct Effect}: $\mathbb{E}(Y_{x'm} - Y_{xm})$
-    \begin{itemize}
-        \item Effect of changing exposure with mediator held fixed \newline
-    \end{itemize}
+\begin{frame}
+    \frametitle{Uncertainty Quantification}
+    \begin{outline}
+        \1 $\delta$-Method \newline
 
-    \textbf{Natural Direct Effect}: $\mathbb{E}(Y_{x'M_{x}} - Y_{xM_{x}})$
-    \begin{itemize}
-        \item Effect of changing exposure when we don't interfere with the mediator  \newline
-    \end{itemize}
+        \1 Advantage: Analytical
+        \1 Disadvantage: Asymptotic \newline
 
-    \textbf{Natural Indirect Effect}: $\mathbb{E}(Y_{xM_{x'}} - Y_{xM_{x}})$
-    \begin{itemize}
-        \item Effect of changing which exposure value is seen by the mediator while holding fixed which exposure value is seen by the outcome
-    \end{itemize}
+        \1 No existing \texttt{R} implementation
+            \2 Until now!
+            \2 Use \texttt{glmmTMB}, not \texttt{lme4}
+        \end{outline}
 \end{frame}
 
-\begin{frame}{\CMA}
-    In our example
-    \begin{itemize}
-        \item Controlled Direct Effect: Effect of increasing news trustworthiness if the whole population followed guidelines in the past
-        \item Natural Direct Effect: Effect of increasing news trustworthiness independent of any induced change in past compliance
-        \item Natural Indirect Effect: Effect of changing past compliance if everyone only got news from Facebook
-    \end{itemize}
+\begin{frame}
+    \frametitle{Comparison}
+    \begin{outline}
+        \1 Simulate 200 datasets
+            \2 \textbf{100} groups, \textbf{500} individuals per group \newline
+
+        \1 Build confidence intervals using $\delta$- and MC $\delta$-methods \newline
+
+        \1 Also build Wald interval using Monte Carlo (empirical) covariance matrix
+        \end{outline}
 \end{frame}
 
-\begin{frame}{\CMA}
-    We can't measure all required counterfactuals
-    \begin{itemize}
-        \item E.g., $Y_x$ or $Y_{x'}$, not both \newline
-    \end{itemize}
-
-    Expected counterfactuals related to conditional expectations
-
-    \begin{itemize}
-        \item     Under strong assumptions, $\bE Y_x = \bE(Y | X=x)$ \newline
-    \end{itemize}
-
-    Nested counterfactuals more complicated
-    \begin{itemize}
-        \item More on this later
-    \end{itemize}
+\begin{frame}
+    \frametitle{Comparison}
+    \begin{table}[h!]
+        \centering
+        \begin{tabular}{|cc|ccc|}
+            \hline
+            Effect & Scale & $\delta$ & MC $\delta$ & Empirical\\
+            \hline
+            Total & Diff & 0.940 & 0.940 & 0.940 \\
+            & Ratio & 0.935 & 0.930& 0.955 \\
+            & OR & 0.955 & 0.960 & 0.950 \\ 
+            \hline
+            Direct & Diff & 0.960 & 0.960 & 0.960 \\
+            & Ratio & 0.935 & 0.930& 0.945 \\
+            & OR & 0.970 & 0.970 & 0.950 \\
+            \hline
+            Indirect & Diff & 0.950 & 0.955 & 0.965 \\
+            & Ratio & 0.955 & 0.955& 0.960 \\
+            & OR & 0.955 & 0.955 & 0.960 \\
+            \hline
+        \end{tabular}
+    \end{table}
 \end{frame}
 
-\begin{frame}{\CMA}
-    How does causality change our analysis? \newline
+\begin{frame}
+    \frametitle{Comparison}
+    \begin{outline}
+        \1 Simulate 200 datasets
+            \2 \textbf{10} groups, \textbf{1000} individuals per group \newline
 
-    Still fit regression models, but include interaction terms between exposure and mediator
-    \begin{itemize}
-        \item $Y = \alpha_0 + \alpha_1 M + \alpha_2 X + \alpha_3 M \cdot X + \varepsilon_Y$
-        \item $M = \beta_0 + \beta_1 X + \varepsilon_M$ \newline
-    \end{itemize}
+        \1 Build confidence intervals using $\delta$- and MC $\delta$-methods \newline
 
-    Direct and indirect effects now depend on the levels of the exposure
+        \1 Also build Wald interval using Monte Carlo (empirical) covariance matrix
 
+        \end{outline}
 \end{frame}
 
-\begin{frame}{\CMA\ -- Extensions}
-    Discussion so far has involved continuous mediator and outcome
-    \begin{itemize}
-        \item What about binary or categorical? \newline
-    \end{itemize}
-
-    Individuals might also be clustered
-    \begin{itemize}
-        \item E.g. Within countries
-    \end{itemize}
-
+\begin{frame}
+    \frametitle{Comparison}
+    \begin{table}[h!]
+        \centering
+        \begin{tabular}{|cc|ccc|}
+            \hline
+            Effect & Scale & $\delta$ & MC $\delta$ & Empirical\\
+            \hline
+            Total & Diff & 0.935 & 0.928 & 0.954 \\
+            & Ratio & 0.961 & 0.974 & 0.967 \\
+            & OR & 0.935 & 0.928 & 0.967 \\ 
+            \hline
+            Direct & Diff & 0.948 & 0.928 & 0.941 \\
+            & Ratio & 0.928 & 0.935 & 0.967 \\
+            & OR & 0.954 & 0.935 & 0.948 \\
+            \hline
+            Indirect & Diff & 0.974 & 0.967 & 0.961 \\
+            & Ratio & 0.980 & 0.974 & 0.961 \\
+            & OR & 0.948 & 0.928 & 0.948 \\
+            \hline
+        \end{tabular}
+    \end{table}
 \end{frame}
 
-\begin{frame}{\CMA\ -- Extensions}
-    Handling binary variables is pretty straightforward
+\begin{frame}
+    \frametitle{Trust Study}
     \begin{outline}
-        \1 Instead of linear regression, use logistic regression
-        \1 Re-define mediation effects based on expected counterfactuals
-            \2 I.e. Counterfactual probabilities
-        \1 New formulas for relating mediation effects to regression coefficients \newline
+        \1 Applying our method to the trust study dataset
+            \2 Compare low-high trustworthiness \newline
+
+        \1 CIs for all effect-scale pairs
+            \2 9 Intervals \newline
+
+        \1 $\delta$ and MC $\delta$
     \end{outline}
-
-    Extend to more than 2 categories using binary indicators
 \end{frame}
 
-\begin{frame}{\CMA\ -- Extensions}
-    Clustered data more complicated \newline
-
-    Standard approach is multi-level modelling
-    \begin{itemize}
-        \item I.e. Add random effects which vary across clusters \newline
-    \end{itemize}
-
-    Combined with categorical variables: 
-    \begin{itemize}
-        \item Generalized linear mixed models (GLMMs)
-    \end{itemize}
+\begin{frame}
+    \frametitle{Trust Study}
+    \centering
+    \includegraphics[height=0.9\textheight, width=0.9\textwidth, keepaspectratio]{Plots/CI_Plot-diff.pdf}
 \end{frame}
 
-\begin{frame}{\GLMMs}
-    The core idea is to augment our set of covariates
-    \begin{itemize}
-        \item Coefficients of these new covariates are random variables which vary across groups/clusters \newline
-    \end{itemize}
-
-    In the linear setting:
-    \begin{itemize}
-        \item Old model: $Y = \alpha_0 + \alpha_1 X_1 + \ldots + \alpha_p X_p + \varepsilon$
-        \item New model: $Y = \alpha_0 + \alpha_1 X_1 + \ldots + \alpha_p X_p + \mathbf{u_1 Z_1 + \ldots + u_q Z_q} + \varepsilon$ \newline
-    \end{itemize}
-
+\begin{frame}
+    \frametitle{Trust Study}
+    \centering
+    \includegraphics[height=0.9\textheight, width=0.9\textwidth, keepaspectratio]{Plots/CI_Plot-rat.pdf}
 \end{frame}
 
-\begin{frame}{\GLMMs}
-    The $Z$'s are fixed, known covariates
-    The $u$'s are random variables
-    \begin{itemize}
-        \item I.e. Random effects \newline
-    \end{itemize}
-
-    It's possible for the $X$'s and $Z$'s to overlap
-    \begin{itemize}
-        \item The coefficient on such a covariate has the form $\alpha_j + u_{k}$
-        \item I.e. Mixed effect
-    \end{itemize}
-
+\begin{frame}
+    \frametitle{Trust Study}
+    \centering
+    \includegraphics[height=0.9\textheight, width=0.9\textwidth, keepaspectratio]{Plots/CI_Plot-OR.pdf}
 \end{frame}
 
-\begin{frame}{\GLMMs}
-    Extend to generalized linear models in the usual way \newline
+\begin{frame}
+    \frametitle{Next Steps}
+    \begin{outline}
+        \1 Compare with Imai et al. (\textit{mediation} package)
 
-    Choose response distribution and link function as for ordinary GLMs \newline
+        \1 Sensitivity to values of confounders
 
-    Linear predictor now has a random effects component \newline
-\end{frame}
+        \1 Parametric bootstrap
 
-\begin{frame}{\GLMMs}
-    Why bother?
-    \begin{itemize}
-        \item E.g. Measured some but not all levels of a categorical variable
-        \item Estimate covariance matrix of random effects
-        \item Test for non-zero variance of each random effect \newline
-    \end{itemize}
-
-    ``Predict'' level of random effects for each group
-    \begin{itemize}
-        \item Conditional mean or conditional mode of random effects given response
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{\GLMMs}
-    In our example:
-    \begin{itemize}
-        \item Data collected from 11 different countries
-        \item Explicitly model inter-country variability
-        \item Predict country-specific random effects
-        \item Use country-specific coefficients in formulas for mediation effects
-        \item Test for significant mediation effects within each country \newline
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{Multilevel Mediation Analysis}
-    Uncertainty quantification for mixed-effects models can be challenging  \newline
-
-    Strategies include:
-    \begin{itemize}
-        \item Bootstrap
-        \item Quasi-Bayesian Monte Carlo
-        \item $\delta$-method
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{Multilevel Mediation Analysis}
-    Uncertainty quantification for mixed-effects models can be challenging  \newline
-
-    Strategies include:
-    \begin{itemize}
-        \item \textbf{Bootstrap}
-        \item Quasi-Bayesian Monte Carlo
-        \item $\delta$-method
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{Multilevel Mediation Analysis}
-    Uncertainty quantification for mixed-effects models can be challenging  \newline
-
-    Strategies include:
-    \begin{itemize}
-        \item Bootstrap
-        \item Quasi-Bayesian Monte Carlo
-        \item $\bm{\delta}$\textbf{-method}
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{Multilevel Mediation Analysis}
-    Recall: Mediation effects defined using nested counterfactuals -- $ Y_{xM_{x'}}$
-    \begin{itemize}
-        \item Value of $Y$ when $X$ is set to $x$, and
-        \item $M$ is set to whatever value it would have if $X$ were set to $x'$ \newline
-    \end{itemize}
-
-    Under strong assumptions, use the ``Mediation Formula''
-
-\end{frame}
-
-\begin{frame}{Multilevel Mediation Analysis}
-    Mediation Formula:
-    \begin{align*}
-        \action<+->{\bE Y_{xM_{x'}} &= \bE_{M | X=x'} \bE (Y | X=x, M) \\}
-    &\action<+->{= \sum_{m=0}^1 \mathbb{P}(Y = 1 | X=x, M=m) \mathbb{P}(M = m | X = x')}
-    \end{align*}   
-    \action<+->{
-    Logistic regression model makes this (relatively) simple:
-    \begin{align*}
-        & \mathbb{P}(Y = 1 | X=x, M=m) = \mathrm{logit}^{-1}(\mathrm{linear\ predictor})\\
-        & \mathbb{P}(M = m | X = x') = \mathrm{logit}^{-1}(\mathrm{different\ linear\ predictor})
-    \end{align*}}
-
-\end{frame}
-
-\begin{frame}{Multilevel Mediation Analysis}
-    Estimating $\bE Y_{xM_{x'}}$ messy, but not \textit{hard}\newline
-
-    Uncertainty quantification based on asymptotic covariance of regression parameters
-    \begin{itemize}
-        \item Use $\delta$-method
-        \item Need derivatives of $\bE Y_{xM_{x'}}$ wrt regression parameters
-        \item Very messy, not particularly hard
-    \end{itemize}
-
-\end{frame}
-
-\begin{frame}{Multilevel Mediation Analysis}
-    Uncertainty quantification for mediation effects similar
-    \begin{itemize}
-        \item More $\delta$-method \newline
-    \end{itemize}
-
-    Very similar for GLMs and GLMMs
-    \begin{itemize}
-        \item Latter has more regression parameters
-        \item Use \texttt{merDeriv} package to supplement \texttt{lme4}
-    \end{itemize}
-
-\end{frame}
-
-\begin{frame}{Putting it All Together}
-    \begin{itemize}
-        \item Define direct, indirect and total effects using counterfactuals \newline
-
-        \item Estimate these effects across countries using generalized linear mixed models \newline
-
-        \item Compute standard errors for estimated effects using the $\delta$-method
-
-    \end{itemize}
-
+        \1 Group-specific effects 
+    \end{outline}
 \end{frame}
